### 2021-06-27 scripts/osa.Rmd
reorg

---
title: "Tropical wet secondary forest regeneration patterns"
author: "Nicholas L Medina"
date: "29/12/2018"
output:
  html_document: default
  pdf_document: default
---

# Introduction (ideas...)

Primary tropical forests are idolized for their - role/importance of secondary forests

## Community composition

-   factors affecting their aboveground biomass accumulation Aboveground biomass accumulation is likely affected by interspecific interactions, namely among N-fixers and non-N-fixers. While N-fixers are thought to facilitate the growth of other plants (CITATIONS FROM YOUNG AG), recent data suggest that competition, such as for P, may be more important for interspecific interactions than expected facilitation by N supply (Taylor et al 2017). This makes sense, given that P availability, rather than that of N, is more limiting in tropical ecosystems compared to temperate systems, especially given that they are estimated to have sufficient N to support decadal aboveground biomass accumulation (Brookshire et al 2019). Furthermore, with regard to nutrient dynamics, N-fixing trees in tropical forests are also expected to be major sources of N2O (Kou-Giesbrecht & Menge 2019), further offsetting biomass balances.

    -   limiting nutrients likely fluctuating/oscillatory (Sullivan et al 2014)
    -   interactions and effects sizes are likely variable at species-level (Keller et al 2013; Gei & Powers 2013)

-   factors affecting their species coexistence

    -   relevance of stage structuring by specific gravity? (Nature EE 2019 2ndFOR)
    -   predicted distributions of species/wood densities

## This study

-   census of \~20yo secondary forest derived from an abandoned [common name for Bombacopsis] plantation. The base species is known to do well in dry forests, but not in the wet lowland forests like on the Osa Peninsula.

## Hypotheses

-   biomass varies with:

    -   distance

        -   species dispersal modes/kernels

    -   light (Max)

-   plantation presence promotes/affects:

    -   growth (biomass)

    -   succession (diversity)

# Methods

## *Study site*

This study was set in the lowland wet tropical rainforest of the Osa Peninsula in SW Puntarenas, Costa Rica. The site is property of Osa Conservation (osaconservation.org) located at the Greg Gund Conservation Center. The censused parcel was 20 ha and undergoing secondary succession for \~20 years after abandonment as a wood plantation using *Bombacopsis quinata*, which unfortunately is better suited to the dry forests of Guanacaste and the Nicoya Peninsula (CITATION). Trees ≥10 cm in diameter were...The study was done during the rainy season months June - August 2013.

## *Census design*

Geographical information systems software (ArcGIS, WEBSITE, and QGIS, [WEBSITE]) were used to create a polygon of the 20-ha parcel, which was divided into six 50-m discrete bands (0-300m) that represented a gradient of distance from the known adjacent primary forest (Fig S1). This distance was made discrete as a simpler alternative to finding the shortest linear distance between the plot and primary forest edge. The number of plots set up in each band was coarsely scaled to the of area each distance bance, such that the progression of plots contained in each band from closest to furthest from primary forest was 10, 8, 5, 3, and 1 respectively. Each plot was 21 *x* 21m and set up June - August 2013.

In each plot, every stem ≥10cm in [diameter/circumference?] was measured for diameter at breast height (DBH, breast height = \~2.5 - 3.1m) and total height. Tree height was measured by using a rangefinder (COMPANY/BRAND Bushnell?) to measure from an observer to the crown (where it was visible), and then to the trunk at breast height; these 2 distances were then used to triangulate and calculate the tree's height using Pythagoras's Theorem. Following typical forestry guidelines, in cases where a tree split into ≥2 stems at breast height, each stem was measured separately; in cases where it only initially split above breast height, it was measured as 1 stem.

## *Aboveground biomass*

Aboveground biomass was estimated for each stem using the allometric equation developed by Chave *et al* (2014), which performed similar to those in Chave *et al* (2005) developed specifically for wet tropical forests:

$$AGB = 0.0673 (p D^2 H) ^.976$$

with *D* in cm, *H* in m, and *p* in g cm\^-3. Measured DBHs and heights were used, along with wood densities ( *p* ) exracted from the literature. In most cases, these values were only available at the genus level, and unrepresented taxa were assumed to be 0.58 as stated by the World Agroforestry Database (CITE). References used for this step are given in *Table S1*.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F)
#loadpacks
library(tidyverse)
library(BIOMASS)
library(rstatix)
library(ggpubr)
library(ggbeeswarm) #jitter/quasi-random
library(ggpmisc)
library(car)
library(vegan)
#theme
styleOsa <- theme_bw() +
  theme(text = element_text(size = 12))
theme_set(styleOsa)
#import
setwd("~/Documents/u/0/art/res/GitHub/dat/osa")
dat <- read.csv("osa.csv")
```

```{r newTaxa2017}
taxasp <- correctTaxo(genus = dat$gen, 
                      species = dat$sp)  #notneeded

taxa <- getTaxonomy(dat$gen)  #useful
dat$fam <- taxa$family
getWD <- getWoodDensity(genus = dat$gen, 
                        species = dat$sp,
                        family = dat$fam, 
                        stand = dat$plot)
dat$spg <- getWD$meanWD
dat <- subset(dat, dist != "NA")#&gen!="-")#&sp!="sp.")
dat$kg17 <- computeAGB(D = dat$dap, WD = dat$spg, 
                       H = dat$h) * 1000

# both methods do seem equal
```

```{r Chave2014fam, include=F}
attach(dat)
dat$kg14 <- 0.0673 * (spg * dap^2 * h)^0.976

#updatefams
for (name in dat$fam) {
  #if (name == "Bombacaceae" | 
        #name == "Sterculiaceae") {
    #name <- "Malvaceae"
  #}
  #if (name == "Cecropiaceae") {
    #name <- "Urticaceae"
  #}
}
detach(dat)
```

## *Abiotic plot characteristics*

### *Canopy cover*

Light reaching the forest floor was measured at the center of each plot, at chest height, using a densiometer (BRAND). According to manufacturer instructions, the final value used for the plot was an average of 4 different readings taken facing each of the 4 cardinal directions: North, South, East, and West.

### *Topography*

The slope of the forest floor in each plot was measured using [a rangefinder by measuring the distance from one (specifc?) corner of the plot to another (specific?) corner.]

## *Statistical analysis*

```{r QC, include=F}
#dat <- subset(dat, dist != "NA")#&gen!="-")#&sp!="sp.")
# most w/ no Genus ID are Tiliaceae

# dat <- subset(dat, kg < 3000)
#subsets:
datB <- subset(dat, gen == "Bombacopsis")
datBno <- subset(dat, gen != "Bombacopsis")
datV <- subset(dat, gen == "Vochysia")
```

Analysis ideas from lit: - ...shade species may do better here vs. clear-cutting?

```{r groupSubplots, include=F}
datp <- group_by(.data = dat,
                 dist, plot) %>%
  #summarize_all(list(~mean(.), ~sqrt(length(.))))
  summarise(dap = mean(dap),
            h = mean(h),
            kg14 = mean(kg14),
            kg17 = mean(kg17),
            kg = mean(kg),
            r = mean(length(unique(gen))),
            rf = mean(length(unique(fam))),
            h = mean(h),
            n = n(),
            spg = mean(spg),
            W = mean(W),
            N = mean(N),
            luz = mean(luz),
            incl = mean(incl))
datpB <- group_by(.data = datB,
                  dist, plot) %>%
  summarise(dap = mean(dap),
            h = mean(h),
            kg14 = mean(kg14),
            kg17 = mean(kg17),
            kg = mean(kg),
            r = mean(length(unique(gen))),
            rf = mean(length(unique(fam))),
            h = mean(h),
            n = n(),
            spg = mean(spg),
            W = mean(W),
            N = mean(N),
            luz = mean(luz),
            incl = mean(incl))
datpBno <- group_by(.data = datBno,
                    dist, plot) %>%
  summarise(dap = mean(dap),
            h = mean(h),
            kg14 = mean(kg14),
            kg17 = mean(kg17),
            kg = mean(kg),
            r = mean(length(unique(gen))),
            rf = mean(length(unique(fam))),
            h = mean(h),
            n = n(),
            spg = mean(spg),
            W = mean(W),
            N = mean(N),
            luz = mean(luz),
            incl = mean(incl))

datpt <- group_by(.data = dat,
                  dist, fam, gen, plot) %>%
  summarise(kg17 = mean(kg17),
            kg14 = mean(kg14),
            kg = mean(kg),
            freq = n(),
            spg = mean(spg),
            h = mean(h))
```

```{r QChypoth}
dat1 <- dat %>% 
  mutate(Bq = if_else(gen == "Bombacopsis", 
                      true = "Planted", 
                      false = "New"))
datg <- dat1 %>% 
  group_by(Bq, dist, plot) %>% #groupbyallindepvars
  get_summary_stats(type = "median_mad")
datgBno <- datg %>% filter(Bq == "New")
```

# Results

## Tree biomass

-   also grouped by *Bombacopsis* ID, re-making old polished figures (again) new trendlines and shorter code

-   Bq doesn't seem to help b/c biomass is lower with it

    -   makes sense, since Bq is not adapted...

```{r H1}
datgBnokg17 <- datgB %>% filter(variable == "kg17")
datg %>% filter(variable == "kg17") %>%
  ggplot(aes(x = dist, y = median)) + 
  geom_quasirandom(color = "gray") +
  stat_summary(fun = "median", color = "blue") +
  stat_summary(fun.data = "median_mad", 
               geom = "errorbar", color = "blue") +

  stat_summary(data = datgBnokg17, 
               fun = "median",
               position = position_dodge()) + #?
  stat_summary(data = datgBnokg17,
               fun.data = "median_mad",
               geom = "errorbar",
               position = position_dodge()) + #?
  scale_y_continuous(trans = "log1p")
```

-   explain the loss w/ wood density panel? (vs. diameter or height)

## Functional/dispersal groups

-   But did Bq provide shade?

    -   or favor other [dispersal] groups of taxa?

```{r H2}

```

# Attic

### *Census structure*

#### Abiotic: Light and slope

```{r envDistStat}
# measured only @ plot level

#checkmanova
luzinclmaovcheck <- lm(luz ~ incl, data = datp)
luzinclmaovcheck %>% residuals() %>% shapiro_test() #pass
luzinclmaovcheck %>% summary() #no-keepseparate
# summary(lm(luz ~ n + plot, data = datp))  # noMANOVA

#luz
plot(datp$dist, datp$luz)
leveneTest(luz ~ as.factor(dist), data = datp)  #pass
luzstat <- lm(luz ~ dist, data = datp)
luzstat %>% residuals() %>% shapiro_test() #pass
luzstat %>% summary()  #no

#incl
plot(datp$dist, datp$incl)
leveneTest(incl ~ as.factor(dist), data = datp)  #pass
inclstat <- aov(incl ~ dist, data = datp)
inclstat %>% residuals() %>% shapiro_test() #justpass
inclstat %>% summary() #no
summary(lm(luz ~ poly(dist,2), data = datp))  #no

plot(datp$dist, datp$incl)
summary(aov(incl ~ dist, data = datp))  #no
summary(lm(incl ~ poly(dist,2), data = datp))  #no
#plot(datp$dist, datp$incl)
```

*Table S\#.* ANOVA results for environmental variation in (A) light and (B) terrain slope by distance from primary forest edge.

#### Timber planting

```{r nBdistStat}
shapiro.test(log1p(datpB$n))  #pass
# summary(lm(h ~ n, data = datpB))  # noMANOVA
summary(aov(log1p(n) ~ dist + plot,
            data = datpB))  # plot
emmeans()
```

*Table S\#.* ANOVA results for *B. quinata* plot stem density by distance from primary forest edge.

```{r nBnoDistStat}
shapiro.test(datpBno$n)  #pass
summary(aov(n ~ dist + plot,
            data = datpBno))  # plot
```

```{r nDistStat}
shapiro.test(datp$n)  #pass
summary(aov(n ~ dist + plot,
            data = datp))  #plot
```

But does this plot effect on stem density scale up to affect biomass? No, so plot ID doesn't affect function.

```{r kg14stat}
shapiro.test(log1p(datp$kg14))  #pass
summary(aov(kg14 ~ n + plot,
            data = datp))  #no!
```

```{r kg17stat}
shapiro.test(log10(datp$kg17))  #pass
leveneTest(log10(kg17) ~ as.factor(dist), data = datp)  #pass
summary(aov(log10(kg17) ~ dist + r,
            data = datp))  #no
summary(aov(log10(kg17) ~ poly(dist,2) + r,
            data = datp))  #no
```

### *Canopy structure*

-   height and [mean] SPG (Díaz et al 2016) by dist, or PCoA

```{r HspgDistStat}
# check for MANOVA need
summary(lm(h ~ spg, data = datp))  # yes

# summary(manova(cbind(h, spg) ~ dist + plot,
#               data = dat))  # dist, plot
summary(manova(cbind(h, spg) ~ dist,
               data = datp))  # dist

shapiro.test(datp$spg)  #pass
summary(lm(spg ~ dist, data = datp))

shapiro.test(datp$h)  #pass
summary(aov(h ~ dist, data = datp))

# check for MANOVA need
summary(lm(h ~ spg, data = datpBno))  # yes
summary(manova(cbind(h, spg) ~ dist + plot,
               data = datpBno))  # dist,plot

# these make sense given h & spg factor into biomass
```

*Table S\#*. MANOVA results showing effects of distance from primary forest edge and plot ID on mean plot height and wood density.

```{r hBdistStat}
summary(aov(h ~ dist * plot,
            data = datB))  # dist, plot
summary(aov(h ~ dist + plot,
            data = datpB))  # dist --> why? competition?
```

*Table S\#.* ANOVA results at (A) tree and (B) plot levels for *B. quinata* height by distance from primary forest edge.

### *Biomass structure*

```{r kgStat}
shapiro.test(log10(datp$kg17))  #pass
summary(lm(log10(kg17) ~ dap + h + spg + n, 
           data = datp))  #dap,spg
summary(manova(cbind(log10(kg17), dap, spg) ~ dist, 
               data = datp))  #yes

shapiro.test(log10(datp$kg17))  #pass
leveneTest(kg17 ~ as.factor(dist), data = datp)  #pass
summary(aov(log10(kg17) ~ poly(dist,2),
            data = datp))  #close...
summary(aov(log10(kg17) ~ dist + r,
            data = datp))  #no--final? 1/6/21

shapiro.test(log1p(datp$kg14))  #pass
# check MANOVA need
# summary(lm(kg14 ~ n, data = datpBno))
shapiro.test(log1p(datpBno$kg14))  #pass
shapiro.test(log1p(datpB$kg14))  #pass

summary(lm(poly(log1p(kg14),2) ~ dist + rf +
             n, data = datp))  #yes but wrong
summary(lm(log1p(kg14) ~ dist + plot + n + rf + N + W + luz + incl,
           data = datp))  # n
summary(lm(log1p(kg14) ~ dist + plot + rf + N + W + luz + incl,
           data = datp))  #none
summary(lm(log1p(kg14) ~ dist + plot + n + rf + N + W + luz + incl,
           data = datpBno))  # n,N,W,luz
summary(lm(log1p(kg14) ~ dist + plot + n + N + W + luz + incl,
           data = datpB))  #none

summary(lm(poly(kg14, degree = 2) ~ dist + plot,
            data = dat))  # dist, plot (both weak)
# which of these 2, above and below?
summary(lm(poly(kg14, degree = 2) ~ dist + plot,
            data = datp))  # dist, plot
summary(aov(kg14 ~ dist + plot,
            data = datB))  # dist
# above prob driven by height along dist
summary(aov(kg14 ~ dist + plot,
            data = datBno))
```

*Table S\#*. (A-D) Linear model and (E, F) ANOVA results for (A, D) total mean, (B, E) *B. quinata*, and (C, F) other taxon biomass with respect to (A-C) stem density and (D-F) distance from primary forest edge and plot ID.

```{r nStat}
shapiro.test(datp$n)  #pass

summary(aov(n ~ poly(dist,2),
            data = datp))  #yes

summary(lm(n ~ luz + incl + N + W,
            data = datp))
summary(aov(n ~ dist + plot,
            data = datp))  # plot
summary(aov(n ~ poly(dist, degree = 2) + plot,
            data = datp))  # dist, plot
summary(aov(n ~ dist + plot,
            data = datpB))  # plot
summary(aov(n ~ dist + plot,
            data = datpBno))  # plot
```

*Table S\#*. Linear model and ANOVA results for stem density with respect to distance from primary forest edge and plot ID.

```{r dapstat}
shapiro.test(datp$dap)  #pass
leveneTest(dap ~ as.factor(dist), data = datp)  #pass
summary(lm(dap ~ dist, data = datp))  #marginal

summary(lm(dap ~ h, data = datp))
summary(manova(cbind(dap, h) ~ dist, data = datp))  #yes

summary(manova(cbind(dap, h, spg) ~ dist, data = datp))  #yes
```

### *Community structure*

-   diversity predicted by height and/or biomass (Pan et al 2018)

```{r ShStat}
shapiro.test(datp$rf)  #pass
summary(lm(rf ~ dist, #+ n + kg14,
        data = datp))  # why significant w/ some var combos but not others?
summary(lm(rf ~ poly(dist,2),
        data = datp))

shapiro.test(datp$rf^2)  #pass
summary(lm(poly(rf,2) ~ dist,
        data = datp))  #no
summary(lm(poly(rf,2) ~ dist,
        data = datsumd))  #no
summary(lm(r ~ h + kg14 + n,
           data = datp))  # h
#summary(lm(r ~ h + kg14 + n,
#           data = datpBno))  # h
```

*Table S\#*. Linear model results showing plot genus richness explained by mean plot height.

-   Other (non-richness) diversity measures, by distance to edge?

```{r SdistStat}
shapiro.test(datp$r)  #pass
summary(aov(r ~ poly(dist,2), 
            data = subset(datp, gen != NA)))  #soclose...
summary(lm(r ~ poly(dist,2), data = datp))  #soclose...

shapiro.test(datp$rf)  #pass
summary(lm(rf ~ dist, data = datp))  #marginal
summary(lm(rf ~ poly(dist,2), data = datp))  #yes
summary(aov(rf ~ poly(dist,2), data = datp))  #yes
```

```{r kgSstat_no}
shapiro.test(log10(dat$kg17))  #no
kruskal.test(dat$kg17 ~ dat$gen)
```

# Results

"Large trees increase biomass by reducing richness [density] in a regenerating tropical timber plantation".

```{r groupDist}
datsumd <- datp %>% group_by(dist) %>%
  summarise(dap.se = sd(dap)/sqrt(length(dap)),
            dap = mean(dap),
            h.se = sd(h)/sqrt(length(h)),
            h = mean(h),
            kg14.se = sd(kg14)/sqrt(length(kg14)),
            kg14 = mean(kg14),
            kg17.se = sd(kg17)/sqrt(length(kg17)),
            kg17 = mean(kg17),
            kg.se = sd(kg)/sqrt(length(kg)),
            kg = mean(kg),
            r.se = sd(r)/sqrt(length(r)),
            r = mean(r),
            rf.se = sd(rf)/sqrt(length(rf)),
            rf = mean(rf),
            n.se = sd(n)/sqrt(length(n)),
            n = mean(n),
            spg.se = sd(spg)/sqrt(length(spg)),
            spg = mean(spg),
            luz.se = sd(luz)/sqrt(length(luz)),
            luz = mean(luz),
            incl.se = sd(incl)/sqrt(length(incl)),
            incl = mean(incl), na.rm = T)
datpsum <- group_by(.data = dat, dist) %>%
  summarize(kg14.se = sd(kg14) / sqrt(length(kg14)),
            kg14 = mean(kg14))
```

## Abiotic conditions

```{r luz_incl}
datp1 <- dat %>% group_by(dist, plot) %>% get_summary_stats(type = "median")

datp1 %>% filter(variable == "luz") %>% 
  ggplot(aes(x = dist, y = median)) + geom_quasirandom(color = "grey") + 
  stat_summary(fun = "median") + 
  stat_summary(fun.data = "median_mad", geom = "errorbar")
```

## Census structure

### Stem density

Mean total plot stem density was explained by distance from primary forest edge and plot. The relationship with distance from primary forest edge was positively quadratic.

```{r NdistPlot}
map <- datp %>% ggplot(aes(x = W, y = N, color = n, cex = kg14)) +  # add dist bands?
  scale_color_gradient(low = "gray", high = "black") + geom_point()

datp %>% ggplot(aes(x = n, y = kg14)) + geom_point() + scale_y_log10()

shapiro.test(datsumd$n)  #pass

ndistplot <- ggplot(data = datsumd, aes(x = dist, y = n)) +
  geom_point(data = datp, color = "gray", position = position_jitter()) +
  geom_point(cex = 3) +
  geom_errorbar(aes(ymax = n + n.se, ymin = n - n.se), width = 0.5) +
  geom_smooth(method = "lm", color = "black",
              formula = y ~ poly(x, degree = 2)) +
  stat_poly_eq(formula = y ~ poly(x, degree = 2), parse = T,
               aes(label = paste(stat(..p.value.label..),
                                 stat(..adj.rr.label..), 
                                 stat(..eq.label..), sep = "~~~~~"))) +
  labs(x = "Distance from primary forest edge",
       y = expression(paste("Trees (stems 400 m"^-2,")")))

summary(aov(n ~ poly(dist,2), data = datsumd))

nplot <- ggplot(datp, aes(x = dist, y = n / 400 * 10000)) +
  geom_quasirandom() +
  #geom_point(position = position_jitter()) +
  geom_smooth(method = "lm", formula = y ~ poly(x,2),
              color = "black", lty = "dashed") +
  stat_poly_eq(formula = y ~ poly(x,2), parse = T, size = 3,
               aes(label = paste(stat(..p.value.label..),
                                 stat(..adj.rr.label..),
                                 stat(..eq.label..), sep = "~~~~~"))) +
  labs(x = "Distance from primary forest edge (m)",
       y = expression(paste("Trees (ha"^-1,")")))
```

*Fig \#\#*. (A) Map showing total stem density and GPS coordinates of census plots (21 x 21 m), and (B) plot showing stem density by distance from primary forest edge.

Mean plot height was explained by mean plot specific wood density and distance from primary forest edge. These were negative relationships, which were further explained by a positive relationship between distance from primary forest edge and mean plot specific wood density.

## Canopy structure

```{r HspgPlot}
dat %>% group_by(plot) %>% ggplot(aes(x = spg, y = h)) +
  geom_quasirandom(color = "gray") + stat_summary(fun = "median") +
  #scale_y_continuous(trans = "log10") +
  geom_smooth(method = "lm") +
  stat_poly_eq(formula = y ~ x, parse = T,
               aes(label = paste(stat(..p.value.label..), #bug
                                 stat(..eq.label..),
                                 stat(..adj.rr.label..), sep = "~~~~~")))
```

*Fig \#\#*. Plot showing relationship betwen mean plot (black, individual trees in gray) tree heights and specific wood density.

```{r HspgDistPlot}
shapiro.test(datsumd$h)  #pass

hplot <- ggplot(data = datp, aes(x = dist, y = h)) +
  geom_point(data = dat, color = "gray", position = position_jitter()) +
  geom_point(size = 3, position = position_jitter()) +
  #geom_smooth(method = "lm", formula = y ~ x, color = "black") +
  stat_poly_eq(formula = y ~ x, parse = T, size = 3,
               aes(label = paste(stat(..p.value.label..),  #no
                                 stat(..rr.label..),
                                 stat(..eq.label..), sep = "~~~~~"))) +
  labs(x = "Distance from primary forest edge (m)", y = "Height (m)")

summary(aov(h ~ poly(dist,2), data = datsumd))  #no

hBplot <- ggplot(data = datpB,
       aes(x = dist, y = h)) +
  geom_point(data = datB,
             color = "gray",
             position = position_jitter()) +
  geom_point() +
  geom_smooth(method = "lm") +
  stat_regline_equation(aes(label = paste(..rr.label..,
                                          ..eq.label..,
                                          sep = "~~~~~"))) +
  #scale_y_continuous(trans = "log1p") +
  theme_bw() +
  labs(x = "Distance from edge of primary forest",
       y = "Height (m)")

hBnoplot <- ggplot(data = datpBno,
       aes(x = dist, y = h)) +
  geom_point(data = datBno,
             color = "gray",
             position = position_jitter()) +
  geom_point() +
  geom_smooth(method = "lm") +
  stat_regline_equation(aes(label = paste(..rr.label..,
                                          ..eq.label..,
                                          sep = "~~~~~"))) +
  #scale_y_continuous(trans = "log1p") +
  theme_bw() +
  labs(x = "Distance from edge of primary forest (m)",
       y = "Height (m)")

shapiro.test(datsumd$spg)  #pass

spgplot <- ggplot(data = datp,
       aes(x = dist, y = spg)) +
  geom_point(data = dat,
             color = "gray",
             position = position_jitter()) +
  geom_point(size = 3, position = position_jitter()) +
  geom_smooth(method = "lm", color = "black") +
  stat_poly_eq(formula = y ~ x, parse = T, size = 3,
               aes(label = paste(stat(..p.value.label..),  #yes
                                 stat(..adj.rr.label..),
                                 stat(..eq.label..), sep = "~~~~~"))) +
  labs(x = "Distance from primary forest edge (m)",
       y = expression(paste("Wood density (g cm"^"-3",")")))
summary(aov(spg ~ dist, data = datsumd))  #no

ggarrange(spgplot, hplot, 
          labels = c("a", "b"),
          nrow = 2)
```

*Fig \#\#*. Plots showing the relationship between mean plot (black, individual trees shown in gray) (A) heights and (B) specific wood densities.

## Biomass structure

Mean plot biomass was explained by distance from primary forest edge and mean total plot stem density. The trend with edge distance was negatively quadratic, while that with mean total plot stem density was linearly negative. These relationships correspond with that between edge distance and mean plot stem density, which was instead positively quadratic (previous figure).

*Taken together, these data show that mean total plot biomass was lower closer to the primary forest, due to increases in stem density, primarily with trees with low wood density and presumably faster growth and reproduction rates (CITE?).*

Finally, there was an effect of East-West location on mean plot biomass, but despite observing a downward slope on the southwest corner of the parcel, this general East-West co-variation with slope did not statistically explain mean plot biomass variation.

```{r kgnPlot}
kgnplot <- ggplot(data = datp,
       aes(x = n, y = kg14)) +
  scale_y_log10() +
  geom_point() +
  geom_smooth(method = "lm") +
  stat_regline_equation(formula = log10(y) ~ x,
                        aes(label = paste(..rr.label..,
                                          ..eq.label..,
                                          sep = "~~~~~"))) +
  theme_bw() +
  labs(x = "Tree stems (per 400 sq m)",
       y = "Aboveground plot biomass (kg)")
```

*Fig \#\#*. Plot showing relationship between mean plot biomass and mean plot stem density.

```{r kgDistPlotFig1}
summary(lm(log10(kg17) ~ dist + rf,
           data = datp))  #no

shapiro.test(datsumd$kg17)  #pass

agbplot <- ggplot(data = datp,
       aes(x = dist, y = kg17 / 400 * 10000 / 1000)) +
  geom_quasirandom(color = "gray") +
  #geom_point(data = subset(datpt, gen == "Bombacopsis"),
  #           color = "black", cex = 3, shape = "square") +
  #geom_smooth(data = subset(datpt, gen == "Bombacopsis"),
  #            method = "lm", color = "black", se = F,
  #            lty = "dashed") +
  #stat_poly_eq(formula = y ~ x,
  #             data = subset(datpt, gen == "Bombacopsis"),
  #             aes(label = paste(stat(..p.value.label..),
  #                               stat(..rr.label..),
  #                               stat(..eq.label..),
  #                               sep = "~~~"))) +
                        #label.y.npc = "bottom",
                        #color = "red") +
  geom_point(cex = 3, position = position_jitter()) +
  #geom_errorbar(aes(ymax = kg14 + kg14.se,
                    #ymin = kg14 - kg14.se)) +
  scale_y_continuous(trans = "log10") +
  #geom_smooth(method = "lm", color = "black", lty = "dashed") +
  stat_poly_eq(formula = y ~ x, parse = T, size = 3,
               aes(label = paste(stat(..p.value.label..),
                                 stat(..rr.label..),
                                 stat(..eq.label..),
                                 sep = "~~~~~"))) +
  labs(x = "Distance from primary forest edge (m)",
       y = expression(paste("Biomass (Mg ha"^-1,")")))
  #coord_fixed(ratio = 100)

summary(aov(kg17 ~ poly(dist,2), data = datsumd))  #0.24

ggarrange(agbplot, comp,
          labels = c("A", "B"),
          nrow = 2)
```

```{r trynewfig1}
group_by(datpt, dist, fam) %>%
  summarize(kg14se = sd(kg14) / sqrt(length(kg14)),
            kg14 = mean(kg14)) %>%
ggplot(aes(x = reorder(fam, -kg14),
           y = kg14)) +
  geom_point() +
  scale_y_continuous(trans = "log10") +
  geom_errorbar(aes(ymax = kg14 + kg14se,
                    ymin = kg14 - kg14se)) +
  facet_wrap( ~ dist) +
  theme(axis.text.x = element_text(angle = 90,
                                   size = 6))
```

```{r newstat}
hist(log1p(dat$kg14))
shapiro.test(log1p(dat$kg14))
stat <- summary(lm(log10(kg14) ~ poly(dist, 2) * fam,
                   data = dat))
stat2 <- subset(stat$coefficients,
                stat$coefficients[,4] <= 0.1)
```

*Fig \#\#*. Biomass (plot mean in black, individual trees in gray, *B. quinata* in orange) changes at various distance from primary forest edge.

```{r dapPlot}
dapplot <- ggplot(datp, aes(x = dist, y = dap)) +
  geom_point(data = dat, color = "gray",
             position = position_jitter()) +
  geom_point(size = 3, position = position_jitter()) +
  geom_smooth(method = "lm", lty = "dashed", color = "black") +
  stat_poly_eq(formula = y ~ x, parse = T, size = 3,
               aes(label = paste(stat(..p.value.label..),
                                 stat(..adj.rr.label..),
                                 stat(..eq.label..),
                                 sep = "~~~~~"))) +
  labs(x = "Distance from primary forest edge (m)",
       y = "Diameter (cm)") +
  scale_y_continuous(trans = "log10")
```

### *Community structure*

-   decide which plot represents anticipated stats...(prob PCoA)

```{r distFamPlot}
group_by(datpt, dist, fam) %>%
  summarize(kg17 = mean(kg17)) %>%
ggplot(aes(x = dist, y = kg17, fill = fam)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(trans = "log10")
```

```{r famDistPlot}
famdistplot <- ggplot(data = datp,
       aes(x = dist, y = rf)) +
  geom_point(position = position_jitter()) +
  labs(x = "Distance from primary forest edge (m)",
       y = "Diversity (α, family)") +
  geom_smooth(method = "lm", color = "black",
              formula = y ~ poly(x,2)) +
  stat_poly_eq(formula = y ~ poly(x,2), parse = T, size = 3,
               aes(label = paste(stat(..p.value.label..),  #yes
                                 stat(..adj.rr.label..),
                                 stat(..eq.label..),
                                 sep = "~~~~~")))

ggplot(data = datpB,
       aes(x = dist, y = rf)) + geom_point()
ggplot(data = datpBno,
       aes(x = dist, y = rf)) + geom_point()
```

```{r genDistPlot}
# make PCoA

ggplot(data = datpt,
       aes(x = dist, y = kg14, fill = fam)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(trans = "log10") +
  # show by plot?
  theme_bw()
```

*Fig \#\#*. Plot showing biomass partitioned by genus as a function of distance from primary forest edge.

Plot richness was predicted by mean plot height, which supports previous findings (Pan et al 2018).

```{r ShDistPlot}
ggplot(data = datp,
       aes(x = h, y = r)) +
  geom_point(cex = 3) +
  geom_smooth(method = "lm") +
  stat_regline_equation(aes(label = paste(..rr.label..,
                                          ..eq.label..,
                                          sep = "~~~~~"))) +
  theme_bw()
ggplot(data = datsumd,
       aes(x = dist, y = r)) +
  geom_point(data = datp,
             color = "gray") +
  geom_point(cex = 3) +
  geom_errorbar(aes(ymax = r + r.se,
                    ymin = r - r.se)) +
  geom_smooth(method = "lm",
              formula = y ~ poly(x, degree = 2)) +
  stat_regline_equation(formula = y ~ poly(x, degree = 2),
                        aes(label = paste(..rr.label..,
                                          ..eq.label..,
                                          sep = "~~~~~"))) +
  theme_bw()
```

*Fig \#\#*. Plot showing relationship between richness and (A) mean plot tree height and (B) distance from primary forest edge.

```{r fig1mix}
ggarrange(spgplot, famdistplot,
          hplot, kgnplot,
          labels = c("a", "b", "c", "d"),
          nrow = 2, ncol = 2)
```

```{r fig1newest}
fig1 <- ggarrange(agbplot, spgplot, 
                  dapplot, hplot,
                  nplot, famdistplot,
          labels = c("a", "b", "c", "d", "e", "f"),
          nrow = 3, ncol = 2)
```

Testing factors affecting divergence of successional trajectories (Norden, Vandermeer et al 2015):

```{r plotTraj}
datps <- as.data.frame(datp)
library(plot3D)
points3D(x = datps$n, y = log10(datps$kg14), z = datps$r,
         colvar = datps$dist)
plot(datps$n, datps$r, col = datps$dist)

plot(log10(datps$kg14), datps$r, col = datps$dist)
summary(lm(r ~ log10(kg14), data = datps))  #no

plot(datps$n, log10(datps$kg14))
summary(lm(log10(kg14) ~ n, data = datps))  #no
```

This could be on a downward divergence trajectory, compared to positive [expected] pattern from other [referenced] Costa Rican forests? Actually, statistically, probably no divergence...

### Distributions

```{r SAD}
ggplot(dat, aes(x = kg17)) +
                  #color = fam)) +  #toomanycolors
  stat_bin(geom = "point") +
  scale_x_continuous(trans = "log10") +
  facet_wrap(~dist)
```

```{r SAD2}
ggplot(datpt, aes(x = reorder(fam,-kg17), y = kg17)) +
  geom_point() + facet_wrap(~dist) +
  scale_y_continuous(trans = "log10") +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
```

```{r mesa}
datpt$Mg17 <- datpt$kg17 / 1000 / 400 * 10000

group_by(datpt, dist, fam) %>%
  summarize(Mg17.se = sd(Mg17) / sqrt(length(Mg17)),
            Mg17 = mean(Mg17)) %>%
  ggplot(aes(x = reorder(fam,-Mg17), y = Mg17)) +
  geom_point() + facet_wrap(~dist) +
  scale_y_continuous(trans = "log10") +
  labs(x = "Taxa (family)", 
       y = expression(paste("Biomass (Mg ha"^-1,")"))) +
  theme(axis.text.x = element_text(angle = -90, size = 7,
                                   vjust = 0.5, hjust = 0),
        panel.grid.minor.y = element_line(color = "gray")) +
  geom_errorbar(aes(ymax = Mg17 + Mg17.se,
                    ymin = Mg17 - Mg17.se))
```

```{r famhist}
ggplot(datpt, aes(x = fam, y = freq)) +
  geom_point() + facet_wrap(~dist) +
  theme(axis.text.x = element_text(angle = -90))
```

```{r table1mesa}
datptt <- group_by(datpt, dist, fam) %>%
  summarize(Mg17.se = sd(kg17 / 1000 / 400 * 10000) / 
              sqrt(length(kg17 / 1000 / 400 * 10000)),
            Mg17 = mean(kg17 / 1000 / 400 * 10000))
```

```{r fig2famsubset}
datptsub <- subset(datpt, 
                   fam == "Vochysiaceae" | fam == "Annonaceae" |
                     fam == "Malvaceae" | fam == "Moraceae" |
                     fam == "Fabaceae" | fam == "Rubiaceae")

shapiro.test(log10(datptsub$Mg17+1))  #no
leveneTest(log10(Mg17+1) ~ as.factor(dist), data = datptsub)  #pass
summary(lm(log10(Mg17+1) ~ dist + fam, data = datptsub))  #?

ggplot(datpt, aes(x = dist, y = kg17)) +
  geom_point(position = position_jitter()) + 
  scale_y_continuous(trans = "log10") + facet_wrap(~fam) +
  geom_smooth(method = "lm", color = "black", se = F) +
  stat_poly_eq(formula = y ~ x, parse = T,
               aes(label = paste(stat(..p.value.label..),
                                 stat(..adj.rr.label..),
                                 stat(..eq.label..),
                                 sep = "~~~~~")))

fig2 <- ggplot(datptsub, aes(x = dist, y = (Mg17 + 1))) +
  geom_point(position = position_jitter()) + 
  scale_y_continuous(trans = "log10") + facet_wrap(~fam) +
  geom_smooth(method = "lm", color = "black", 
              se = F, lty = "dashed") +
  stat_poly_eq(formula = y ~ x, parse = T, size = 2.25,
               aes(label = paste(stat(..p.value.label..),
                                 stat(..rr.label..),
                                 stat(..eq.label..),
                                 sep = "~~~~~"))) +
  labs(x = "Distance from primary forest edge (m)",
       y = expression(paste("Biomass (Mg ha"^-1,")")))
```

```{r fig3responsNo}
datptsubsub <- subset(datptsub, 
                      fam == "Vochysiaceae" | fam == "Malvaceae")
ggplot(datptsubsub, aes(x = spg, y = h, shape = fam)) +
  geom_point() + #facet_wrap(~dist) +
  geom_smooth(method = "lm", color = "black", lty = "dashed") +
  stat_poly_eq(formula = y ~ x, parse = T, size = 2,
               aes(label = paste(stat(..p.value.label..),
                                 stat(..rr.label..),
                                 stat(..eq.label..),
                                 sep = "~~~~~")))
```

### Composition

```{r tallyMatrix}
plots <- unique(dat$plot)
fams <- unique(dat$fam)
rich <- length(fams)
tallytable <- matrix(0, nrow = length(plots), ncol = rich)
colnames(tallytable) <- fams
rownames(tallytable) <- plots
for (p in 1:length(rownames(tallytable))) {
  comm <- subset(dat, plot == plots[p])
  tally <- count(comm, fam)
  for (col in 1:length(colnames(tallytable))) {
    taxon <- fams[col]
    if (taxon %in% tally$fam) {
      place <- match(taxon, tally$fam)
      tallytable[p, col] <- tally$n[place]
    } else {}
  }
}
```

```{r DAPmatrix}
DAPtable <- matrix(0, nrow = length(plots), ncol = rich)
colnames(DAPtable) <- fams
rownames(DAPtable) <- plots
for (p in 1:length(rownames(DAPtable))) {
  comm <- subset(dat, plot == plots[p]) %>%
    group_by(fam) %>% summarize(dap = mean(dap))
  for (col in 1:length(colnames(DAPtable))) {
    taxon <- fams[col]
    if (taxon %in% comm$fam) {
      place <- match(taxon, comm$fam)
      DAPtable[p, col] <- comm$dap[place]
    } else {}
  }
}
```

```{r MgMatrix}
MgMatrix <- matrix(0, nrow = length(plots), ncol = rich)
colnames(MgMatrix) <- fams
rownames(MgMatrix) <- plots
for (p in 1:length(rownames(MgMatrix))) {
  comm <- subset(dat, plot == plots[p]) %>%
    group_by(fam) %>% summarize(Mg17 = mean(kg17/400/1000*10000))
  for (col in 1:length(colnames(MgMatrix))) {
    taxon <- fams[col]
    if (taxon %in% comm$fam) {
      place <- match(taxon, comm$fam)
      MgMatrix[p, col] <- log10(comm$Mg17[place] + 1)  #log?
    } else {}
  }
}
```

```{r PCA}
#normalitycheck?
adonis2(tallytable ~ dist + luz, data = datp)  #yes
adonis2(DAPtable ~ dist + luz, data = datp)  #yes
adonis2(MgMatrix ~ dist + luz, data = datp)  #yes4log10matrix

#dists <- vegdist(tallytable)
pca <- rda(tallytable)
biplot(pca)
#ordiplot(pca)
ordiellipse(pca, datp$dist, conf = 0.95)
#plot(rda(tallytable ~ dist, data = datp))
#plot(rda(tallytable ~ 1,))

ggtable <- as.data.frame(cbind(pca$CA$u[,1], pca$CA$u[,2],
                               datp$dist))
ggplot(ggtable, aes(x = ggtable[,1], y = ggtable[,2],
                    shape = as.factor(ggtable[,3]))) +
  geom_point() + stat_conf_ellipse() ##check
```

```{r PCA+}

```

#### try Chao?

```{r div}
datp$div <- diversity(tallytable)

shapiro.test(datp$div)  #pass
summary(lm(div ~ poly(dist,2), data = datp))  #yes
plot(datp$dist, datp$div)
```

### Interactions

```{r index}
dat$Mg17 <- dat$kg17 /400*1000/10000
indexes <- c()
for (p in plots) {
  subcomm <- subset(dat, plot == p)
  trees <- length(subcomm$Mg17)
  for (tree in 1:trees) {
    focus <- subcomm[tree,]$Mg17
    index <- 0
    for (othertree in 1:trees) {
      neighbor <- subcomm$Mg17[othertree]
      subindex <- (focus + 1) / (neighbor + 1)
      index <- index + subindex
    }
    indexes <- c(indexes, index)
  }
}
dat$index <- indexes
```

```{r indexstat}
datpti <- group_by(dat, dist, fam, plot) %>%
  summarize(ii.se = sd(index) / sqrt(length(index)),
            ii = mean(index))
datptisub <- subset(datpti, 
                    fam == "Vochysiaceae" | fam == "Malvaceae" | 
                      fam == "Rubiaceae" | fam == "Fabaceae")

shapiro.test(log10(datpti$ii))  #soclose...
hist(log10(datpti$ii))
shapiro.test(log10(datptisub$ii))  #pass
summary(lm(ii ~ dist, data = datptisub))  #no

ggplot(datptisub, aes(x = dist, y = ii)) +
  geom_point(position = position_jitter()) + facet_wrap(~fam) +
  geom_smooth(method = "lm", formula = y ~ poly(x,2),
              lty = "dashed", color = "black") +
  stat_poly_eq(formula = y ~ poly(x,2), parse = T,  #yes
               aes(label = paste(stat(..p.value.label..),
                                 stat(..adj.rr.label..),
                                 stat(..eq.label..),
                                 sep = "~~~~~"))) +

  labs(x = "Distance from primary forest",
       y = "Interaction index")
```

#### Dispersal

```{r dispTable}
dispTable <- data.frame(fam = dat$fam,
                        gen = dat$gen,
                        sp = dat$sp,
                        plot = dat$plot)
splist <- dispTable %>%
  group_by(fam, gen, sp) %>% summarize(count = n()) %>%
  #group_by(fam, gen, sp) %>% 
  #summarize(count.se = sd(count+1)/sqrt(length(count+1)),
  #          count.m = mean(count)) %>% 
  arrange(desc(count))
write.csv(splist, file = "splist.csv")
```

# NEW

```{r new4.20}
datqc <- subset(dat, dist != "NA" & fam != "-")
group_by(datqc, dist, fam) %>%
  summarize(n = n(),
            kg14 = mean(kg14)) %>%
  ggplot(aes(x = kg14,
             fill = fam)) +
  stat_bin(na.rm = T,
           binwidth = 0.5,  # min = log(3.7 kg)
           geom = "bar") +
  facet_wrap(~dist) +
  scale_x_continuous(trans = "log10") +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title.align = 0.5) +
  labs(x = "Aboveground biomass (kg)",
       y = "Frequency (stems)",
       fill = "Family") +
  coord_fixed(0.5)

comp <- ggplot(data = datpt,
       aes(x = dist, y = kg14)) +
  geom_bar(stat = "identity",
           aes(fill = fam),
           position = "fill") +
  theme_classic() +
  labs(x = "Distance away from primary forest edge (m)",
       y = "Proportion of aboveground biomass (kg)",
       fill = "Family") +
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 12),
        legend.position = "right",
        legend.title.align = 0.5,
        legend.direction = "vertical",
        legend.text = element_text(size = 8)) +
  coord_fixed(ratio = 500)
```

```{r ecdf_kg14}
# not yet
```

```{r community}
community <- data.frame()
for (name in 1:length(unique(dat$fam))) {
  community$name <- c(NA)  # DEBUG HERE
}
for (i in unique(dat$plot)) {
  datplot <- subset(dat)
}
tally <- add_count(as.tbl(dat), fam)
```

```{r NMDS}
datNMDS <- dat[,c(2,12)]
```

## *Other ongoing analyses...*

```{r kgDistHist}
ggplot(data = dat, aes(x = kg14,
                        color = as.factor(rep))) +
  stat_bin(geom = "point") +
  scale_x_continuous(trans = "log10") 
  facet_wrap(~dist)  # overall is more skewed
ggplot(data = datB, aes(x = kg14,
                        color = as.factor(rep))) +
  stat_bin(geom = "point") +
  scale_x_continuous(trans = "log10") +
  facet_wrap(~dist)
ggplot(data = datBno, aes(x = kg14,
                        color = as.factor(rep))) +
  stat_bin(geom = "point") +
  scale_x_continuous(trans = "log10") +
  facet_wrap(~dist)
```

```{r hDistHist}
ggplot(data = dat,
       aes(x = h,
           color = as.factor(rep))) +
  stat_bin(geom = "point") +
  #scale_x_continuous(trans = "log10") 
  facet_wrap(~dist)  # overall is also skewed
ggplot(data = datB,
       aes(x = h,
           color = as.factor(rep))) +
  stat_bin(geom = "point") +
  facet_wrap(~dist)  # why are pochote also skewed in height??
ggplot(data = datBno,
       aes(x = h,
           color = as.factor(rep))) +
  stat_bin(geom = "point") +
  facet_wrap(~dist)
```

*Figs \#*. Histograms showing plot (A) biomass and (B) height distributions by distance from primary forest edge.

-   forest biomass is complex (Pan et al 2013)
-   but biomass distributions here are log-normal?!
-   check for *power-law fits* of height and biomass histograms among edge plots

# CODE/PLOT ATTIC, DONT READ

```{r supplemental, include=F}
ggplot(datsumd, aes(x = dist, y = luz)) +
  geom_point() +
  geom_errorbar(aes(ymax = luz + luz.se,
                    ymin = luz - luz.se))
ggplot(datsumd, aes(x = dist, y = n)) +
  geom_point() +
  geom_errorbar(aes(ymax = n + n.se,
                    ymin = n - n.se))
ggplot(datsumd, aes(x = dist, y = spg)) +
  geom_point() +
  geom_errorbar(aes(ymax = spg + spg.se,
                    ymin = spg - spg.se))
```

\#\#Fig1.Biomass,distance,genus

```{r kgdist, include=F}
##1a
ggplot(datsumd, aes(x = dist, y = kg)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymax = kg + kg.se,
                    ymin = kg - kg.se)) +
  theme_bw() +
  xlab(label = "Distance from primary forest edge (m)")+
  ylab(label = "Plot aboveground biomass (kg)")

summary(lm(data = datp, kg ~ poly(dist, degree = 2)))

##1b
datgt <- group_by(.data = datpt, dist,gen)
datsumt <- summarise(datgt,
                     kg.se = sd(kg)/sqrt(length(kg)),
                     kg = mean(kg),
                     freq.se = sd(freq)/sqrt(length(freq)),
                     freq = mean(freq))
ggplot(datsumt, aes(x = dist, y = kg, color = gen)) +
  geom_point(stat="identity") +
  geom_errorbar(aes(ymax = kg + kg.se,
                    ymin = kg - kg.se)) +
  #scale_y_continuous(trans = "log10") +
  theme_bw() +
  xlab(label = "Distance from primary forest edge (m)") +
  ylab(label = "Plot aboveground biomass (kg)") +
  labs(color = "Genus")

subset(datsumt, kg > 3000)  #Lonchocarpus outlier
```

Fig 1a. Intermediate distances (150m from edge) tended (p=0.26) to show higher biomass.

(Outliers seem to be relevant for variance at intermediate distances...)

```{r fig1bfollowup, include=F}
##maxkg?
max(datsumt[datsumt$dist==150,]$kg)
```

Lonchocarpus individual is huge. Remove?

\#\#FigA1.Vstem,distance

```{r vstem}

```

\#\#Bombacopsis biomass by distance

```{r kgdistBomba, include=F}
datBgp <- group_by(datB, dist,plot)
datBp <- summarise(datBgp, kg = mean(kg), freq = n())

datBg <- group_by(datBp, dist)
datBsumd <- summarise(datBg,
                      kg.se = sd(kg)/sqrt(length(kg)),
                      kg = mean(kg),
                      freq.se = sd(freq)/sqrt(length(freq)),
                      freq = mean(freq))

ggplot(datBsumd, aes(x = dist, y = kg)) +
  geom_point() +
  geom_errorbar(aes(ymax = kg + kg.se,
                    ymin = kg - kg.se))

summary(lm(kg~dist,data=datBsumd))

ggplot(datBsumd, aes(x = dist, y = freq)) +
  geom_point() +
  geom_errorbar(aes(ymax = freq + freq.se,
                    ymin = freq - freq.se))

summary(lm(freq~dist,data=datBsumd))
```

Figs 1b and c. (b) Biomass of plantation species significantly increased with distance from undisturbed forest edge (p\<0.05, R2=0.70), with (b) no change in the number of *Bombacopsis* stems (p\>0.05).

Together, Figs 1a and 1b show that biomass was higher in areas where the plantation species *Bombacopsis* stems grew smaller on average.

(This may suggest either competition or facilitation--check biomass ditribution of major contributing genera...)

```{r kgsp, include=F}
ggplot(datsumt[datsumt$dist==150,],
       aes(x=reorder(gen,-kg),y=log10(kg)))+
  geom_point()+
  theme(axis.text.x=element_text(hjust=0,angle=-45))
  #facet_wrap(~dist)

summary(lm(data=datsumt, log10(kg)~c(1:length(gen))*dist))
```

Fig 2a. Rank-biomass curves (by genus) significantly (but weakly) fit a negative exponential (p=0.01, R2=0.05).

From Fig 1, highest biomass was in 150m, which this figure shows mostly comes from 1 *Lonchocarpus*, 1 *Tabebuia*, 1 [/3] *Ficus*, 2 [/10] *Hyeronima*, 1 [/2] *Apeiba*, 4 [/23] *Vochysia*, and 8 [/16] *Bombacopsis* trees (ranked sequentially). Excluding individual outlier tree sizes, it could be

\#\#Vochysia biomass by distance

```{r kgdistVochysia, include=F}
datVgp <- group_by(datV, dist,plot)
datVp <- summarise(datVgp, kg = mean(kg), freq = n())

datVg <- group_by(datVp, dist)
datVsumd <- summarise(datVg,
                      kg.se = sd(kg)/sqrt(length(kg)),
                      kg = mean(kg),
                      freq.se = sd(freq)/sqrt(length(freq)),
                      freq = mean(freq))

ggplot(datVsumd,aes(x = dist, y = kg)) +
  geom_point() +
  geom_errorbar(aes(ymax = kg + kg.se,
                    ymin = kg - kg.se))

ggplot(datVsumd,aes(x = dist, y = freq)) +
  geom_point() +
  geom_errorbar(aes(ymax = freq + freq.se,
                    ymin = freq - freq.se))
summary(lm(freq~dist,data=datVsumd))
```

Fig 2b and c. [...] Actually, it seems like loss of biomass may not due to competition between *Bombacopsis* and any single taxon, given that the trends of lower richness with higher biomass observed here seem to be due to singletons.

\*\*Does this suggest a kind of portfolio effect, but at the level of productive individuals (vs. species), and in the opposite direction due to competition?

This could be due to a lower chance of sampling species in an area with a few big trees. -- What does the stem number distribution look like by distance and/or genus?

```{r stemdist, include=F}
ggplot(datsumt,aes(x=dist,y=freq,color=gen))+
  geom_point(stat="identity")+
  geom_errorbar(aes(ymax=freq+freq.se,
                    ymin=freq-freq.se))
  #scale_y_continuous(trans="log10")
```

Seems like it is true, that fewer individuals of rare species occurred in plots that also had the most massive (and the least massive) individual trees.

```{r rabund, include=F}
ggplot(datpt,aes(x=reorder(gen,-kg)))+
  geom_histogram(stat="count")+
  theme(axis.text.x=element_text(angle=-45,hjust=0))
  #facet_wrap(~dist)
```

```{r kghist, include=F}
ggplot(datp,aes(x=log10(kg)))+
  geom_histogram(bins=10)
#testforlog-normal?
```

Fig \#a. Frequency distribution of biomass tends to appear log-normal.

```{r kghistBomba, include=F}
ggplot(datBp,aes(x=kg))+
  geom_histogram(binwidth=log10(15))
#testforlog-normal?
```

DISCUSSION - explain hump b/c proximity on road side + -- lower richness b/c earlier-successional (20y) - Vochysia is wind-dispersed; is avg 200m?

### Community composition

#### Shannon diversity

```{r rdist, include=F}
plot(datp$dist, datp$r)

datsumr <- summarise(.data = datp,
                     r.se = sd(r)/sqrt(length(r)),
                     r = mean(r))
ggplot(data = datsumr, aes(x = dist, y = r)) +
  geom_point() +
  geom_errorbar(aes(ymax = r + r.se,
                    ymin = r - r.se))

summary(lm(r~poly(dist,degree=2),data=datp))
```

Fig 3. Richness-by-distance relationship significantly fits a negative quadratic model (p\<0.01, R2=0.22).

(Which [rare] species were lost?)

Together, Figs 1 and 3 show that higher biomass tended to occur in plots with lower taxonomic richness.

```{r Skg, include=F}
Skgsum<-summary(lm(log10(kg)~log10(r),data=datsumd[datsumd$dist!=300,]))
ggplot(datsumd[datsumd$dist!=300,],aes(x=r,y=kg))+
  geom_point()+
  geom_errorbarh(aes(xmax=r+r.se,xmin=r-r.se))+
  geom_errorbar(aes(ymax=kg+kg.se,ymin=kg-kg.se))+
  scale_x_continuous(trans="log10")+
  scale_y_continuous(trans="log10")+
  geom_abline(intercept=Skgsum$coefficients[1],
              slope=Skgsum$coefficients[2])
```

Fig 4. Biomass-richness relationship marginally significantly fits a straight line on a log-log plot (p=0.066, R2=0.63).

\#Conclusions Higher biomass tended to occur in areas that also showed lower taxonomic richness at \>10 years after abandonment of the *Bombacopsis* tree plantation.

\#\#Attic \#\#\#Vochysia

```{r groupV, include=F}
#datV <- read.csv("~/Documents/u/0/dat/osa/osa.csv")
#datV <- subset(datV, subset=gen=="Vochysia")

#datV <- group_by(datV, dist,n)
#datV <- summarise(datV, kg=mean(kg))

#datV <- group_by(datV, dist)
#datV <- summarise(datV, kg.se=sd(kg)/sqrt(length(kg)),
#                 kg=mean(kg))
```

```{r dist_kgV}
#ggplot(datV, aes(x=dist,y=kg)) +
#  geom_point(stat='identity', size=3) +
#  geom_errorbar(aes(ymax=kg+kg.se, ymin=kg-kg.se)) +
#  theme_classic()
```

### 2021-06-28 article/article.Rmd
made outline of sections, looking to make planned fig1 next time

---
title: "Osa"
output: html_notebook
---

# Introduction

1.  importance of reforestation

2.  patterns of reforestation

    1.  general

    2.  tropical

3.  things that affect reforestation

    1.  edge effects

    2.  plantation structure?

4.  [for NGO--start biomass census]

    1.  Does leaving the plantation structure help reforestation?

        1.  shade

        2.  edge effects (i.e. buffer)

# Results

## Shade / Canopy

1.  shade x distance + timber sp

## Edge effects

1.  biomass x distance + timber sp

    -   can break down biomass by component

2.  diversity x distance [+ 1 timber sp]

    -   can break down by diversity index

# Discussion

1.  assess hypotheses
2.  infer causes of patterns - plantation structure, old legacy trees, or species traits?
3.  compare inferences to other studies
4.  recommend future conservation approaches, ecological studies

# Methods

## Study site

-   osa peninsula (GPS)
-   wet lowland rainforest
-   greg gund conservation center, osa conservation

## Plot design

-   QGIS/ArcGIS 50m bands 0-300m from forest
-   30 plots, random within each band, allocated by area: 11, 8, 5, 3, 2, 1
-   20x20m

## Data collection

### Canopy light and slope

-   densiometer in 4 directions at center of plot
-   slope?

### Tree biomass

-   DBH, H using rangefinder (BUSHNELL)

-   chave 2014, updated to chave 2017, little difference in result

-   unique individual tag/plate \#s

-   species IDs

    -   local knowledge of common names

    -   taxonomy matches using local field guides

-   wood density using Agroforestry Wood Database in 2013, updated using BIOMASS package

## Statistical analysis

-   code available (somewhere..., DRYAD, GitHub)

-   R v\#\#\#

    -   BIOMASS package v\#\#\#

        -   taxa (family, maybe genus) updated from raw IDs using field guides

        -   matched wood densities from package database

    -   tidyverse v\#\#\# = summary stats, plotting

    -   vegan v\#\#\# = community analysis

# Acknowledgements

-   osa conservation - max villalobos, [dan student harvard alum], agustín (& annia), manuel director
-   lehigh group - don, emma, ?, ?
-   brandeis WOW funding

### 2021-06-28 analysis/clean.R

@@ -1,6 +1,6 @@
 library(vroom)
 
-dat <- vroom("osa.csv")
+dat <- vroom("data/raw/osa.csv")
 
 
 #updatetaxa2017

### 2021-06-28 analysis/stats.R

@@ -1,0 +1,1 @@
+library(rstatix)

### 2021-06-29 article/article.Rmd
outline, re-start clean base stats

@@
 title: "Osa"
-output: html_notebook
+output: 
+  html_notebook: 
+    toc: yes
+    code_folding: hide
+    theme: spacelab
+    fig_caption: yes
 ---
@@
     -   can break down biomass by component
+    -   maybe system-level curve (e.g. biomas hist) + timber sp
 
@@
     -   can break down by diversity index
+    -   compare to NMDS?
+
+    1.  system-level curve (e.g. rank-abund) x distance [+ 1 timber sp]
 

### 2021-06-29 analysis/clean.R

@@ -1,6 +1,6 @@
 library(vroom)
 
-dat <- vroom("data/raw/osa.csv")
+dat <- vroom("../data/raw/osa.csv") #rel2analysisfiles
 
 
 #updatetaxa2017
@@ -26,7 +26,7 @@ dat$kg17 <- computeAGB(D = dat$dap, WD = dat$spg,
 attach(dat)
 dat$kg14 <- 0.0673 * (spg * dap^2 * h)^0.976
 
-#updatefams
+#updatefams--old
 for (name in dat$fam) {
   #if (name == "Bombacaceae" |
   #name == "Sterculiaceae") {
@@ -39,6 +39,3 @@ for (name in dat$fam) {
 detach(dat)
 
 # both methods do seem equal
-
-
-library(tidyverse)

### 2021-06-29 analysis/stats.R

@@ -1,1 +1,8 @@
+source("clean.R")
+
+library(tidyverse)
 library(rstatix)
+
+dats <- dat %>%
+  group_by(dist, plot) %>%
+  get_summary_stats(type = "median_mad")

### 2021-06-29 analysis/clean.R
got nested table w sum and other stats, now trying to add final col for binary eval of pvals

@@ -1,13 +1,15 @@
+library(here) #fromprojdir
 library(vroom)
 
-dat <- vroom("../data/raw/osa.csv") #rel2analysisfiles
+dat <- vroom(here("data/raw/osa.csv"))
 
 
 #updatetaxa2017
 library(BIOMASS)
 
 taxasp <- correctTaxo(genus = dat$gen,
-                      species = dat$sp)  #notneeded
+                      species = dat$sp, #notneeded
+                      useCache = T)
 
 taxa <- getTaxonomy(dat$gen)  #useful
 dat$fam <- taxa$family

### 2021-07-01 analysis/clean.R
adding col to eval stat test assumptions

@@ -1,5 +1,5 @@
 library(here) #fromprojdir
-library(vroom)
+library(vroom) #loadastbl
 
 dat <- vroom(here("data/raw/osa.csv"))
 

### 2021-07-01 analysis/stats.R

@@ -10,7 +10,10 @@ plotVarsTbl <- dat %>%
   nest(vardata = -variable)
 
 
-dattaxasum <- dat %>%
+treeVarsTbl <- dat %>%
   group_by(dist, plot, fam, gen, sp) %>%
   select(dap, h, spg, kg17) %>%
-  get_summary_stats(type = "median_mad") #-n
+  #mutate(richness?)
+  select(-n) %>% #deprecated
+  get_summary_stats(type = "median_mad") %>%
+  nest(vardata = -variable)

### 2021-07-02 analysis/stats.R
added a stat eval col, trying to add biotic plot-level vars

@@ -4,16 +4,16 @@ library(rstatix)
 
 plotVarsTbl <- dat %>%
   group_by(dist, plot) %>%
-  select(N, W, luz, incl) %>%
+  select(luz, incl, #abiotic
+         dap, h, spg, kg17) %>%
   get_summary_stats(type = "median") %>%
-  select(-n) %>% #deprecated
   nest(vardata = -variable)
 
 
 treeVarsTbl <- dat %>%
   group_by(dist, plot, fam, gen, sp) %>%
   select(dap, h, spg, kg17) %>%
-  #mutate(richness?)
-  select(-n) %>% #deprecated
+  #mutate(richness?) #maybevegan
   get_summary_stats(type = "median_mad") %>%
   nest(vardata = -variable)
+

### 2021-07-03 analysis/clean.R
cleaned clean-data file, glanced at tree-level data, need to choose level of analysis for some variables

@@ -1,35 +1,23 @@
 library(here) #fromprojdir
 library(vroom) #loadastbl
-
-dat <- vroom(here("data/raw/osa.csv"))
-
-
-#updatetaxa2017
+library(tidyverse)
 library(BIOMASS)
 
-taxasp <- correctTaxo(genus = dat$gen,
-                      species = dat$sp, #notneeded
-                      useCache = T)
+rawData <- vroom(here("data/raw/osa.csv")) %>%
+  filter(!is.na(dist)) %>% #gen!=na,sp!=sp.?
+  mutate(fixedNames = correctTaxo(genus = gen,
+                                  species = sp,
+                                  useCache = T)) %>%
+  mutate(fam = getTaxonomy(gen)$family) %>% #useful
+  mutate(spg = getWoodDensity(genus = gen,
+                              species = sp,
+                              family = fam,
+                              stand = plot)$meanWD) %>%
+  mutate(kg14 = 0.0673 * (spg * dap^2 * h)^0.976) %>%
+  mutate(kg17 = computeAGB(D = dap, WD = spg, H = h) * 1000)
 
-taxa <- getTaxonomy(dat$gen)  #useful
-dat$fam <- taxa$family
-getWD <- getWoodDensity(genus = dat$gen,
-                        species = dat$sp,
-                        family = dat$fam,
-                        stand = dat$plot)
-dat$spg <- getWD$meanWD
-dat <- subset(dat, dist != "NA")#&gen!="-")#&sp!="sp.")
-dat$kg17 <- computeAGB(D = dat$dap, WD = dat$spg,
-                       H = dat$h) * 1000
-
-
-#chave2014
-
-attach(dat)
-dat$kg14 <- 0.0673 * (spg * dap^2 * h)^0.976
-
-#updatefams--old
-for (name in dat$fam) {
+#attic--newfams
+for (name in rawData$fam) {
   #if (name == "Bombacaceae" |
   #name == "Sterculiaceae") {
   #name <- "Malvaceae"
@@ -38,6 +26,4 @@ for (name in dat$fam) {
   #name <- "Urticaceae"
   #}
 }
-detach(dat)
-
-# both methods do seem equal
+# both fam methods do seem equal

### 2021-07-03 analysis/stats.R

@@ -1,19 +1,19 @@
-source(here("analysis/clean.R"))
-library(tidyverse)
-library(rstatix)
+library(here); source(here("analysis/clean.R"))
+library(tidyverse); library(rstatix)
 
-plotVarsTbl <- dat %>%
+plotVarsTbl <- rawData %>%
   group_by(dist, plot) %>%
+  mutate(richness = n_distinct(sp)) %>%
   select(luz, incl, #abiotic
-         dap, h, spg, kg17) %>%
+         dap, h, spg, kg17, #biotic
+         richness) %>% #taxonomic
   get_summary_stats(type = "median") %>%
-  nest(vardata = -variable)
+  nest(varData = -variable)
 
 
-treeVarsTbl <- dat %>%
+treeVarsTbl <- rawData %>%
   group_by(dist, plot, fam, gen, sp) %>%
   select(dap, h, spg, kg17) %>%
-  #mutate(richness?) #maybevegan
   get_summary_stats(type = "median_mad") %>%
-  nest(vardata = -variable)
-
+  mutate(median1 = log10(median + 1)) %>% #4statTests
+  nest(varData = -variable)

### 2021-07-05 analysis/clean.R
made progress on embedding non-normal glms to exploration pipeline/tibble, now choosing distr family by AIC

@@ -1,20 +1,19 @@
 library(here) #fromprojdir
 library(vroom) #loadastbl
-library(tidyverse)
-library(BIOMASS)
+library(tidyverse); library(BIOMASS)
 
 rawData <- vroom(here("data/raw/osa.csv")) %>%
   filter(!is.na(dist)) %>% #gen!=na,sp!=sp.?
   mutate(fixedNames = correctTaxo(genus = gen,
                                   species = sp,
-                                  useCache = T)) %>%
-  mutate(fam = getTaxonomy(gen)$family) %>% #useful
-  mutate(spg = getWoodDensity(genus = gen,
+                                  useCache = T),
+         fam = getTaxonomy(gen)$family, #useful
+         spg = getWoodDensity(genus = gen,
                               species = sp,
                               family = fam,
-                              stand = plot)$meanWD) %>%
-  mutate(kg14 = 0.0673 * (spg * dap^2 * h)^0.976) %>%
-  mutate(kg17 = computeAGB(D = dap, WD = spg, H = h) * 1000)
+                              stand = plot)$meanWD,
+         kg14 = 0.0673 * (spg * dap^2 * h)^0.976,
+         kg17 = computeAGB(D = dap, WD = spg, H = h) * 1000)
 
 #attic--newfams
 for (name in rawData$fam) {

### 2021-07-05 analysis/stats.R

@@ -10,7 +10,6 @@ plotVarsTbl <- rawData %>%
   get_summary_stats(type = "median") %>%
   nest(varData = -variable)
 
-
 treeVarsTbl <- rawData %>%
   group_by(dist, plot, fam, gen, sp) %>%
   select(dap, h, spg, kg17) %>%

### 2021-07-07 analysis/stats.R
included non-normal stats, attempting entropy by size class, not species

@@ -16,3 +16,5 @@ treeVarsTbl <- rawData %>%
   get_summary_stats(type = "median_mad") %>%
   mutate(median1 = log10(median + 1)) %>% #4statTests
   nest(varData = -variable)
+
+#funcw/formulahard

### 2021-07-11 analysis/disfit.R
start vegan stats part

@@ -1,0 +1,56 @@
+library(poweRlaw)
+
+
+disfit <- function(RealFreq) {  #freqvector
+  #Clauset2009,Gillespie2015
+
+  Pl <- displ$new(round(RealFreq))
+  Pl$setPars(estimate_pars(Pl))
+  PlVar <- var(bootstrap(Pl)$bootstraps$pars) #variance
+  #plot(Pl)
+
+  PlP <- bootstrap_p(Pl)$p  #>0.1passes
+
+  Exp <- disexp$new(round(RealFreq))
+  Exp$setPars(estimate_pars(Exp))
+  ExpVar <- var(bootstrap(Exp)$bootstraps$pars)
+
+  Pois <- dispois$new(round(RealFreq))
+  Pois$setPars(estimate_pars(Pois))
+  PoisVar <- var(bootstrap(Pois)$bootstraps$pars)
+
+  Lognorm <- dislnorm$new(round(RealFreq))
+  Lognorm$setPars(estimate_pars(Lognorm))
+  LognormVar <- bootstrap(Lognorm)$bootstraps
+  LognormVar1 <- var(LognormVar$pars1)
+  LognormVar2 <- var(LognormVar$pars2, na.rm = T)
+
+
+  #null=bothOK #1sided=arg1==arg2
+  PlExpP <- compare_distributions(Pl, Exp)$p_one_sided  #<0.05=arg1better
+  PlLognormP <- compare_distributions(Pl, Lognorm)$p_one_sided
+  PlPoisP <- compare_distributions(Pl, Pois)$p_one_sided
+
+
+  results <- data.frame("PlP" = PlP,
+                        "PlExpP" = PlExpP, #pvals
+                        "PlLognormP" = PlLognormP,
+                        "PlPoisP" = PlPoisP,
+
+                        #slopes/pars
+                        "PlPar" = Pl$pars,
+                        "ExpPar" = Exp$pars,
+                        "PoisPar" = Pois$pars,
+                        "LognormPar1" = Lognorm$pars[1],
+                        "LognormPar2" = Lognorm$pars[2],
+
+                        #variances
+                        "PlVar" = PlVar,
+                        "ExpVar" = ExpVar,
+                        "PoisVar" = PoisVar,
+                        "LognormVar1" = LognormVar1,
+                        "LognormVar2" = LognormVar2)
+
+
+  return(results)
+}

### 2021-07-13 analysis/stats.R
added entropy calc, need to mutate it before nesting and stat pipeline

@@ -4,6 +4,12 @@ library(tidyverse); library(rstatix)
 plotVarsTbl <- rawData %>%
   group_by(dist, plot) %>%
   mutate(richness = n_distinct(sp)) %>%
+  add_count(gen, name = "n_gen") %>%
+  pivot_wider(names_from = gen, values_from = n_gen,
+              values_fill = 0, values_fn = sum,
+              names_repair = "universal") %>%
+  select(-(ID : richness)) %>%
+  diversity() #mutate
   select(luz, incl, #abiotic
          dap, h, spg, kg17, #biotic
          richness) %>% #taxonomic
@@ -15,7 +21,6 @@ treeVarsTbl <- rawData %>%
   select(dap, h, spg, kg17) %>%
   get_summary_stats(type = "median_mad") %>%
   mutate(median1 = log10(median + 1)) %>% #4statTests
-  #mutate(spTbl = rawData %>% pivot_wider(gen, median)) %>%
   nest(varData = -variable)
 
 #funcw/formulahard

### 2021-07-15 analysis/clean.R
finally found line causing tibl length issues after pivot_wider, line 98. but there should also be 1 more.

@@ -15,6 +15,8 @@ rawData <- vroom(here("data/raw/osa.csv")) %>%
          kg14 = 0.0673 * (spg * dap^2 * h)^0.976,
          kg17 = computeAGB(D = dap, WD = spg, H = h) * 1000)
 
+rawData$fixedNames
+
 #attic--newfams
 for (name in rawData$fam) {
   #if (name == "Bombacaceae" |

### 2021-07-15 analysis/stats.R

@@ -1,20 +1,27 @@
 library(here); source(here("analysis/clean.R"))
-library(tidyverse); library(rstatix);
+library(tidyverse); library(rstatix); library(skimr)
 library(vegan) #diversity()
+
+rawData$ID %>% sort(na.last=T)
+spTbl$ID %>% sort(na.last=T)
+check1 <- rawData %>% filter(is.na(ID)) %>% arrange(plot,fam)
+check2 <- spTbl %>% filter(is.na(ID)) %>% arrange(plot,fam)
+check1$fam == check2$fam #line98
+check1$fam[2:length(check1$fam)] == check2
+view(check1); view(check2)
 
 spTbl <- rawData %>% #4vegan
   add_count(gen, plot, name = "n_gen", sort = T) %>%
   pivot_wider(names_from = gen, values_from = n_gen,
               values_fn = length, #rmwarn
               values_fill = 0) %>% #noid_col,leavedups
-  select(-c(ID, dist, rep:kg17)) #keepplot
+  select(-c(dist, rep:side, sp:kg17)) #keepplot
 #bugeats2rows...
 
 plotVarsTbl <- rawData %>%
-  mutate(entropy = spTbl %>% diversity())
+  mutate(spTbl = spTbl) #bug
   group_by(dist, plot) %>%
-  mutate(richness = n_distinct(sp))
-length(plotVarsTbl$richness)
+  mutate(richness = n_distinct(sp)) %>%
 
   select(luz, incl, #abiotic
          dap, h, spg, kg17, #biotic
@@ -25,7 +32,7 @@ length(plotVarsTbl$richness)
 treeVarsTbl <- rawData %>%
   group_by(dist, plot, fam, gen, sp) %>%
   select(dap, h, spg, kg17) %>%
-  get_summary_stats(type = "median_mad") %>%
+  get_summary_stats(type = "median_mad")
   mutate(median1 = log10(median + 1)) %>% #4statTests
   nest(varData = -variable)
 

### 2021-07-16 analysis/clean.R
pivoted into sp tbl, but diversity = 0 since left at tree vs. plot level. next is fixing 0s

@@ -1,21 +1,25 @@
 library(here) #fromprojdir
 library(vroom) #loadastbl
-library(tidyverse); library(BIOMASS)
+#library(skimr) #glance
+library(tidyverse);
+library(BIOMASS)
 
-rawData <- vroom(here("data/raw/osa.csv")) %>%
+rawData <- vroom(
+  here("data/raw/osa.csv")) %>%
   filter(!is.na(dist)) %>% #gen!=na,sp!=sp.?
-  mutate(fixedNames = correctTaxo(genus = gen,
-                                  species = sp,
-                                  useCache = T),
-         fam = getTaxonomy(gen)$family, #useful
-         spg = getWoodDensity(genus = gen,
-                              species = sp,
-                              family = fam,
-                              stand = plot)$meanWD,
-         kg14 = 0.0673 * (spg * dap^2 * h)^0.976,
-         kg17 = computeAGB(D = dap, WD = spg, H = h) * 1000)
-
-rawData$fixedNames
+  mutate(
+    fixedNames = correctTaxo(genus = gen, #df
+                             species = sp,
+                             useCache = T),
+    fam = getTaxonomy(gen)$family, #useful
+    spg = getWoodDensity(genus = gen,
+                         species = sp,
+                         family = fam,
+                         stand = plot)$meanWD,
+    kg14 = 0.0673 * (spg * dap^2 * h)^0.976,
+    kg17 = computeAGB(D = dap,
+                      WD = spg,
+                      H = h) * 1000)
 
 #attic--newfams
 for (name in rawData$fam) {

### 2021-07-16 analysis/disfit.R

@@ -1,55 +1,74 @@
 library(poweRlaw)
 
 
-disfit <- function(RealFreq) {  #freqvector
+disfit <- function(freqData) {
+  #input=freqvector
   #Clauset2009,Gillespie2015
 
-  Pl <- displ$new(round(RealFreq))
-  Pl$setPars(estimate_pars(Pl))
-  PlVar <- var(bootstrap(Pl)$bootstraps$pars) #variance
+  Pl <- displ$new(round(freqData))
+  Pl$setPars(
+    estimate_pars(Pl))
+  #variance
+  PlVar <- var(
+    bootstrap(Pl)$bootstraps$pars)
   #plot(Pl)
 
   PlP <- bootstrap_p(Pl)$p  #>0.1passes
 
-  Exp <- disexp$new(round(RealFreq))
-  Exp$setPars(estimate_pars(Exp))
-  ExpVar <- var(bootstrap(Exp)$bootstraps$pars)
+  Exp <- disexp$new(round(freqData))
+  Exp$setPars(
+    estimate_pars(Exp))
+  ExpVar <- var(
+    bootstrap(Exp)$bootstraps$pars)
 
-  Pois <- dispois$new(round(RealFreq))
-  Pois$setPars(estimate_pars(Pois))
-  PoisVar <- var(bootstrap(Pois)$bootstraps$pars)
+  Pois <- dispois$new(round(freqData))
+  Pois$setPars(
+    estimate_pars(Pois))
+  PoisVar <- var(
+    bootstrap(Pois)$bootstraps$pars)
 
-  Lognorm <- dislnorm$new(round(RealFreq))
-  Lognorm$setPars(estimate_pars(Lognorm))
-  LognormVar <- bootstrap(Lognorm)$bootstraps
+  Lognorm <- dislnorm$new(round(freqData))
+  Lognorm$setPars(
+    estimate_pars(Lognorm))
+  LognormVar <-
+    bootstrap(Lognorm)$bootstraps
   LognormVar1 <- var(LognormVar$pars1)
-  LognormVar2 <- var(LognormVar$pars2, na.rm = T)
+  LognormVar2 <- var(LognormVar$pars2,
+                     na.rm = T)
 
 
   #null=bothOK #1sided=arg1==arg2
-  PlExpP <- compare_distributions(Pl, Exp)$p_one_sided  #<0.05=arg1better
-  PlLognormP <- compare_distributions(Pl, Lognorm)$p_one_sided
[…]

### 2021-07-16 analysis/stats.R

@@ -1,39 +1,54 @@
-library(here); source(here("analysis/clean.R"))
-library(tidyverse); library(rstatix); library(skimr)
+library(here);
+source(
+  here("analysis/clean.R"))
+library(rstatix);
 library(vegan) #diversity()
 
-rawData$ID %>% sort(na.last=T)
-spTbl$ID %>% sort(na.last=T)
-check1 <- rawData %>% filter(is.na(ID)) %>% arrange(plot,fam)
-check2 <- spTbl %>% filter(is.na(ID)) %>% arrange(plot,fam)
-check1$fam == check2$fam #line98
-check1$fam[2:length(check1$fam)] == check2
-view(check1); view(check2)
 
-spTbl <- rawData %>% #4vegan
-  add_count(gen, plot, name = "n_gen", sort = T) %>%
-  pivot_wider(names_from = gen, values_from = n_gen,
-              values_fn = length, #rmwarn
-              values_fill = 0) %>% #noid_col,leavedups
-  select(-c(dist, rep:side, sp:kg17)) #keepplot
-#bugeats2rows...
+#4vegan
+view(spTbl) <- rawData %>%
+  #groups
+  add_count(gen, plot,
+            name = "n_gen",
+            sort = T) %>%
+  #leavedups
+  mutate(
+    newID = row_number()) %>%
+  pivot_wider(
+    id_cols = c(newID),
+    names_from = gen,
+    values_from = n_gen,
+    #rmwarn
+    #values_fn = length,
+    values_fill = 0) %>%
+  select(-newID)
 
 plotVarsTbl <- rawData %>%
-  mutate(spTbl = spTbl) #bug
+  mutate(
+    #alreadygrouped
+    "entropy" = spTbl %>%
+      diversity()) %>%
+  #filter(entropy > 0) #debug
   group_by(dist, plot) %>%
-  mutate(richness = n_distinct(sp)) %>%
+  mutate(
+    "richness" = n_distinct(sp)) %>%
 
   select(luz, incl, #abiotic
-         dap, h, spg, kg17, #biotic
-         richness) %>% #taxonomic
-  get_summary_stats(type = "median_mad") %>%
+         dap, h, spg, kg17,
+         richness, entropy) %>%
+  get_summary_stats(
+    type = "median_mad") %>%
   nest(varData = -variable)
 
 treeVarsTbl <- rawData %>%
-  group_by(dist, plot, fam, gen, sp) %>%
[…]

### 2021-07-20 analysis/clean.R
only used glance for parametric and shapiro tests, but didn't seem to work for glms.

@@ -16,9 +16,9 @@ rawData <- vroom(
     #useful
     "fam" = getTaxonomy(gen) %>%
       pull(family),
-    "spg" = getWoodDensity(genus = "gen",
-                           species = "sp",
-                           family = "fam",
+    "spg" = getWoodDensity(genus = gen,
+                           species = sp,
+                           family = fam,
                            stand = plot) %>%
       pull(meanWD),
     "kg17" = computeAGB(D = dap,

### 2021-07-20 analysis/checkVars.R
fimally learned to  subset and eval() formula components for use in general function. made  one for exploratory stats mutate-modify-tbl flow.

@@ -1,0 +1,156 @@
+library(here);
+source(
+  here("analysis/stats.R"))
+library(tidymodels)
+library(ggbeeswarm) #quasirandom()
+library(ggpmisc) #stat_poly_eq()
+library(ggpubr) #ggarrange()
+source(
+  here("analysis/disfit.R")) #disfit()
+
+
+checkVars <- function(
+  .nestedVarDataTbl,
+  .formula
+  ) {
+
+  .nestedVarDataTbl %>%
+    mutate(
+      "statTest" = varData %>%
+        modify(
+          ~ .x %>%
+            lm(formula = .formula)
+          ),
+      statPrint = statTest %>%
+        modify(
+          ~ .x %>%
+            tidy() %>%
+            full_join(.,
+                      .x %>%
+                        glance()
+                      ) %>%
+            filter(
+              !is.na(term)
+              )
+          ),
+
+      normalTest = statTest %>%
+        modify(
+          ~ .x %>%
+            augment() %>%
+            pull(.resid) %>%
+            shapiro_test()
+          ),
+      #OGdata
+      varianceTest = varData %>%
+        modify(
+          ~ .x %>%
+            levene_test(
+              formula =
+                .formula[[2]] %>%
+                eval() ~
+                .formula[[3]] %>%
+                eval() %>%
+                as_factor()
+              )
+          ),
+      "isNormal" = normalTest %>%
+        modify(
+          ~ if_else(
+            .x$p.value > 0.055,
+            T,F)
+          ),
+      "isHomosced" = varianceTest %>%
+        modify(
+          ~ if_else(
+            .x$p > 0.055,
+            T, F)
+          )
+    ) %>%
+    #list2vec
+    unnest(
+      c(isNormal, isHomosced)
+      ) %>%
+    mutate(
+      "isModelOK" = if_else(
+        isNormal & isHomosced,
+        T, F),
+      "isSignif" = statPrint %>%
+        modify_if(
+          isModelOK,
+          ~ if_else(
+            #mainterm
+            .x$p.value[2] < 0.105,
+            T,F),
+          .else = ~ NA)
+      ) %>%
+    unnest(isSignif) %>%
+    #nonnormalstats
+    #nestedmodifyishard
+    mutate(
+      "statTestPois" = varData %>%
+        modify_if(
+          !isModelOK,
+          ~ .x %>%
+            glm(
+              formula = mainModel,
+              family = poisson()
+              ) %>%
+            summary(),
[…]

### 2021-07-20 analysis/checkVars.R
generalized some exploratory stat functions/workflows and filtered significant variables.

@@ -1,156 +1,346 @@
 library(here);
 source(
-  here("analysis/stats.R"))
-library(tidymodels)
+  here("analysis/stats.R")
+)
+library(tidymodels) #glance()
 library(ggbeeswarm) #quasirandom()
 library(ggpmisc) #stat_poly_eq()
 library(ggpubr) #ggarrange()
 source(
-  here("analysis/disfit.R")) #disfit()
+  here("analysis/disfit.R")
+) #disfit()
 
 
-checkVars <- function(
-  .nestedVarDataTbl,
-  .formula
+getStats <- function(
+  ...nestedVarDataTbl,
+  ...formula
+) {
+
+  ...nestedVarDataTbl %>%
+    statFit(
+      ...formula
+    ) %>%
+    statEval(
+      ...formula
+    ) %>%
+
+    #nonnormaldistr
+    statFitNon(
+      ...formula
+    ) %>%
+    statEvalNon() %>%
+
+    statGraph(
+      ...formula
+    ) %>%
+
+  return()
+}
+
+
+
+statFit <- function(
+  ...nestedVarDataTbl,
+  ....formula
   ) {
 
-  .nestedVarDataTbl %>%
+  ...nestedVarDataTbl %>%
+
     mutate(
-      "statTest" = varData %>%
+      "statTest" =
+        varData %>%
         modify(
           ~ .x %>%
-            lm(formula = .formula)
+            lm(
+              formula = ....formula
+              )
           ),
-      statPrint = statTest %>%
+      statPrint =
+        statTest %>%
         modify(
           ~ .x %>%
             tidy() %>%
-            full_join(.,
-                      .x %>%
-                        glance()
-                      ) %>%
+            full_join(
+              .,
+              .x %>%
+                glance()
+              ) %>%
             filter(
-              !is.na(term)
+              !is.na(
+                term
+                )
               )
-          ),
+          )
+    ) %>%
 
-      normalTest = statTest %>%
+    return()
+}
+
+
+statEval <- function(
+  ...statFitTbl,
+  ....formula
+  ) {
+
+  ...statFitTbl %>%
+
+    mutate(
+      normalTest =
+        statTest %>%
         modify(
           ~ .x %>%
             augment() %>%
-            pull(.resid) %>%
+            pull(
+              .resid
+              ) %>%
             shapiro_test()
           ),
       #OGdata
-      varianceTest = varData %>%
+      varianceTest =
+        varData %>%
         modify(
           ~ .x %>%
[…]

### 2021-07-20 analysis/clean.R

@@ -4,23 +4,59 @@ library(vroom) #loadastbl
 library(tidyverse);
 library(BIOMASS)
 
+
 rawData <- vroom(
-  here("data/raw/osa.csv")) %>%
-  #gen!=na,sp!=sp.?
-  filter(
-    !is.na(dist)) %>%
-  mutate(
-    "newName" = correctTaxo(genus = gen,
-                            species = sp,
-                            useCache = T),
-    #useful
-    "fam" = getTaxonomy(gen) %>%
-      pull(family),
-    "spg" = getWoodDensity(genus = gen,
-                           species = sp,
-                           family = fam,
-                           stand = plot) %>%
-      pull(meanWD),
-    "kg17" = computeAGB(D = dap,
-                        WD = spg,
-                        H = h) * 1000)
+  here("data/raw/osa.csv")
+)
+
+cleanData <- function(
+  ...rawData
+) {
+
+  ...rawData %>%
+
+    #gen!=na,sp!=sp.?
+    filter(
+      !is.na(
+        dist
+      )
+    ) %>%
+
+    mutate(
+      "newName" = correctTaxo(
+        genus = gen,
+        species = sp,
+        useCache = T
+      ),
+
+      #useful
+      "fam" = getTaxonomy(
+        gen
+      ) %>%
+        pull(
+          family
+        ),
+
+      "spg" = getWoodDensity(
+        genus = gen,
+        species = sp,
+        family = fam,
+        stand = plot
+      ) %>%
+        pull(
+          meanWD
+        ),
+
+      "kg17" = computeAGB(
+        D = dap,
+        WD = spg,
+        H = h
+      ) * 1000
+    ) %>%
+
+    return()
+}
+
+cleanData <- rawData %>%
+  cleanData()
+

### 2021-07-20 analysis/stats.R

@@ -1,55 +1,110 @@
 library(here);
 source(
-  here("analysis/clean.R"))
+  here("analysis/clean.R")
+)
 library(rstatix);
 library(vegan) #diversity()
 
 
 #4vegan
-spTbl <- rawData %>%
-  add_count(gen,
-            plot, #group
-            name = "n_gen",
-            sort = T) %>%
+spTbl <- cleanData %>%
+  add_count(
+    gen,
+    plot, #group
+    name = "n_gen",
+    sort = T
+    ) %>%
   pivot_wider(
     id_cols = plot,
     names_from = gen,
     values_from = n_gen,
     #rmwarn
     values_fn = length,
-    values_fill = 0) %>%
+    values_fill = 0
+    ) %>%
   mutate(
-    "entropy" = diversity(.))
+    "entropy" =
+      diversity(
+        .
+        )
+    )
 
 
-plotVarsTbl <- rawData %>%
-  left_join(.,
-            spTbl %>%
-              select(plot,
-                     entropy),
-            by = "plot",
-            copy = T) %>%
-  group_by(dist, plot) %>%
+plotVarsTbl <- cleanData %>%
+  left_join(
+    .,
+    spTbl %>%
+      select(
+        plot,
+        entropy
+        ),
+    by = "plot",
+    copy = T
+    ) %>%
+
+  group_by(
+    dist,
+    plot
+    ) %>%
   mutate(
-    "richness" = n_distinct(sp)) %>%
-  select(luz, incl, #abiotic
-         dap, h, spg, kg17,
-         richness, entropy) %>%
+    "richness" =
+      n_distinct(
+        sp
+        )
+    ) %>%
+  select(
+    #abiotic
+    luz,
+    incl,
+
+    #biotic
+    dap,
+    h,
+    spg,
+    kg17,
+
+    #community
+    richness,
+    entropy
+    ) %>%
+
   get_summary_stats(
-    type = "median_mad") %>%
-  nest("varData" = -variable)
+    type = "median_mad"
+    ) %>%
+  nest(
+    "varData" = -variable
+    )
 
 
 #rid?
-treeVarsTbl <- rawData %>%
-  group_by(dist, plot,
-           fam, gen, sp) %>%
-  select(dap, h, spg, kg17) %>%
+treeVarsTbl <- cleanData %>%
+  group_by(
+    dist,
+    plot,
+
+    fam,
+    gen,
+    sp
+    ) %>%
+  select(
[…]

### 2021-07-21 analysis/fullstats.R
almost there, quick viz buggy left.

@@ -11,26 +11,27 @@ source(
 ) #disfit()
 
 
-getStats <- function(
+getStatsTbl <- function(
   ...nestedVarDataTbl,
+  #user's
   ...formula
 ) {
 
   ...nestedVarDataTbl %>%
-    statFit(
+    statFitTbl(
       ...formula
     ) %>%
-    statEval(
+    addStatEval(
       ...formula
     ) %>%
 
     #nonnormaldistr
-    statFitNon(
+    addStatFitNon(
       ...formula
     ) %>%
-    statEvalNon() %>%
+    addStatEvalNon() %>%
 
-    statGraph(
+    addStatGraph(
       ...formula
     ) %>%
 
@@ -39,7 +40,7 @@ getStats <- function(
 
 
 
-statFit <- function(
+statFitTbl <- function(
   ...nestedVarDataTbl,
   ....formula
   ) {
@@ -76,8 +77,18 @@ statFit <- function(
     return()
 }
 
+#wrapper4consistency"-2"
+statFitTbl2 <- function(
+  ...nestedVarDataTbl,
+  ....formula2
+) {
+  statFitTbl(
+    ...nestedVarDataTbl,
+    ....formula2)
+}
 
-statEval <- function(
+
+addStatEval <- function(
   ...statFitTbl,
   ....formula
   ) {
@@ -89,10 +100,7 @@ statEval <- function(
         statTest %>%
         modify(
           ~ .x %>%
-            augment() %>%
-            pull(
-              .resid
-              ) %>%
+            residuals() %>%
             shapiro_test()
           ),
       #OGdata
@@ -167,10 +175,57 @@ statEval <- function(
 }
 
 
+addStatEval2 <- function(
+  ...statFitTbl2,
+  ....formula2
+) {
+
+  ...statFitTbl2 %>%
+
+    mutate(
+      curve =
+        statPrint %>%
+        modify(
+          ~ .x %>%
+            filter(
+              term !=
+                "(Intercept)"
+            ) %>%
+            pull(
+              term
+            )
+        ),
+      "pvals" =
+        statPrint %>%
+        modify(
+          ~ .x %>%
+            filter(
+              term !=
+                "(Intercept)"
+            ) %>%
+            pull(
+              p.value
+            )
+        ),
+      "isSignif2" =
+        pvals %>%
+        modify(
+          ~ if_else(
+            .x <
+              0.055,
+            T, F
[…]

### 2021-07-21 analysis/fullstats.R
still buggy label.

@@ -39,6 +39,24 @@ getStatsTbl <- function(
 }
 
 
+getStatsTbl2 <- function(
+  ...nestedVarDataTbl,
+  #user's
+  ...formula2
+) {
+
+  ...nestedVarDataTbl %>%
+    statFitTbl2(
+      ...formula2
+    ) %>%
+    addStatEval2(
+      ...formula2
+    ) %>%
+
+    return()
+}
+
+
 
 statFitTbl <- function(
   ...nestedVarDataTbl,
@@ -377,9 +395,13 @@ addStatGraph <- function(
               formula = y ~ x,
               parse = T,
               aes(
-                label = stat(
-                  ..p.value.label..,
-                  ..adj.rr.label..
+                label = paste(
+                  after_stat(
+                    p.value.label
+                    ),
+                  after_stat(
+                    adj.rr.label
+                    )
                   )
                 )
               ) +
@@ -427,18 +449,26 @@ addStatGraph2 <- function(
             ) +
             geom_quasirandom() +
             geom_smooth(
-              method = "lm"
+              method = "lm",
+              formula = y ~ poly(
+                x,
+                2
+              )
             ) +
             stat_poly_eq(
               formula = y ~ poly(
                 x,
                 2
               ),
-              parse = T,
+              parse = F,
               aes(
-                label = stat(
-                  ..p.value.label..,
-                  ..adj.rr.label..
+                label = paste(
+                  after_stat(
+                    p.value.label
+                    ),
+                  after_stat(
+                    adj.rr.label
+                  )
                 )
               )
             ) +
@@ -451,7 +481,7 @@ addStatGraph2 <- function(
               ),
               #fixtitle
               title = deparse(
-                curve
+                variable
               )
             )
         )

### 2021-08-03 analysis/clean.R
simple ord, matrix format issue

@@ -9,6 +9,7 @@ rawData <- vroom(
   here("data/raw/osa.csv")
 )
 
+
 cleanData <- function(
   ...rawData
 ) {
@@ -56,6 +57,7 @@ cleanData <- function(
 
     return()
 }
+
 
 cleanData <- rawData %>%
   cleanData()

### 2021-08-03 analysis/stats.R

@@ -108,3 +108,21 @@ treeVarsTbl <- cleanData %>%
     )
 
 #funcw/formulahard
+
+
+distMat <- vegdist(
+  {
+    spTbl %>%
+      select(
+        -c(
+          entropy,
+          plot
+        )
+      )
+  }
+)
+
+ordStat <- adonis(
+  distMat ~ plot,
+  cleanData #bug--as.matrix...
+)

### 2021-08-03 analysis/statsTbl.R

@@ -11,13 +11,17 @@ source(
 ) #disfit()
 
 
+#main
 getStatsTbl <- function(
+  #has"varData"col
   ...nestedVarDataTbl,
   #user's
-  ...formula
+  ...formula,
+  ...spTbl = NULL
 ) {
 
   ...nestedVarDataTbl %>%
+
     statFitTbl(
       ...formula
     ) %>%
@@ -32,6 +36,10 @@ getStatsTbl <- function(
     addStatEvalNon() %>%
 
     addStatGraph(
+      ...formula
+    ) %>%
+
+    addOrdStat(
       ...formula
     ) %>%
 
@@ -378,11 +386,11 @@ addStatGraph <- function(
             ggplot(
               aes(
                 x = {
-                  ....formula[[2]] %>%
+                  ....formula[[3]] %>%
                     eval()
                 },
                 y = {
-                  ....formula[[3]] %>%
+                  ....formula[[2]] %>%
                     eval()
                 }
               )
@@ -393,7 +401,7 @@ addStatGraph <- function(
               ) +
             stat_poly_eq(
               formula = y ~ x,
-              parse = T,
+              parse = F,
               aes(
                 label = paste(
                   after_stat(
@@ -438,11 +446,11 @@ addStatGraph2 <- function(
             ggplot(
               aes(
                 x = {
-                  ....formula[[2]] %>%
+                  ....formula[[3]] %>%
                     eval()
                 },
                 y = {
-                  ....formula[[3]] %>%
+                  ....formula[[2]] %>%
                     eval()
                 }
               )
@@ -490,3 +498,28 @@ addStatGraph2 <- function(
     return()
 }
 
+
+#loading...
+addOrdStat <- function(
+  ...addStatGraph,
+  ...spTbl,
+  ....formula
+) {
+
+  ...addStatGraph %>%
+
+    mutate(
+      distmat = vegdist(
+        ...spTbl
+      ),
+      ordStat = adonis2(
+        ....formula,
+        {
+          ...addStatGraph %>%
+            pull(
+              varData
+            )
+        }
+      )
+    )
+}

### 2021-08-04 analysis/stats.R
fixed ord

@@ -110,25 +110,49 @@ treeVarsTbl <- cleanData %>%
 #funcw/formulahard
 
 
+#ordinate
+
+commTbl <- spTbl %>%
+  select(
+    -c(
+      entropy,
+      plot
+    )
+  )
+
 distMat <- vegdist(
-  {
-    spTbl %>%
-      select(
-        -c(
-          entropy
-        )
-      )
-  }
+  commTbl
 )
 
+metaTbl <- cleanData %>%
+  distinct(
+    plot,
+    dist
+  )
+
+
+#stat
+
+# set.seed(
+#   079523
+# )
 ordStat <- adonis(
   distMat ~ dist,
-  #bug--as.matrix...
-  {
-    cleanData %>%
-      distinct(
-        plot,
-        dist
-      )
-  }
+  metaTbl,
+  99999
+)
+
+
+ord <- metaMDS(
+  commTbl
+)
+
+ordTbl <- ord %>%
+  scores() %>%
+  as_tibble()
+
+
+ordFullTbl <- cbind(
+  metaTbl,
+  ordTbl
 )

### 2021-08-04 analysis/stats.R
wrote ordination function

@@ -120,7 +120,7 @@ commTbl <- spTbl %>%
     )
   )
 
-distMat <- vegdist(
+distMat = vegdist(
   commTbl
 )
 
@@ -129,32 +129,3 @@ metaTbl <- cleanData %>%
     plot,
     dist
   )
-
-
-#stat
-
-# set.seed(
-#   079523
-# )
-ordStat <- adonis(
-  distMat ~ dist,
-  metaTbl,
-  99999
-)
-
-
-ord <- rda(
-  commTbl
-)
-
-
-ordTbl <- scores(
-  ord
-)$sites %>%
-  as_tibble()
-
-
-ordFullTbl <- cbind(
-  metaTbl,
-  ordTbl
-)

### 2021-08-05 analysis/stats.R
made higher level curves

@@ -72,8 +72,27 @@ plotVarsTbl <- cleanData %>%
     type = "median_mad"
     ) %>%
   nest(
-    "varData" = -variable
-    )
+    "varData0" = -variable
+    ) %>%
+
+  #nextlevel
+  mutate(
+    "varData" = varData0 %>%
+      modify(
+        ~ .x %>%
+          group_by(
+            dist
+          ) %>%
+          summarize(
+            median = median(
+              median
+            ),
+            mad = mad(
+              median
+            )
+          )
+      )
+  )
 
 
 #rid?
@@ -129,3 +148,13 @@ metaTbl <- cleanData %>%
     plot,
     dist
   )
+
+
+ordModel <- distMat ~
+  dist
+
+ordStat = adonis(
+  ordModel,
+  metaTbl,
+  99999
+)

### 2021-08-05 analysis/stats.R
need to reduce reg funcs

@@ -72,12 +72,12 @@ plotVarsTbl <- cleanData %>%
     type = "median_mad"
     ) %>%
   nest(
-    "varData0" = -variable
+    "varData" = -variable
     ) %>%
 
   #nextlevel
   mutate(
-    "varData" = varData0 %>%
+    "varData1" = varData %>%
       modify(
         ~ .x %>%
           group_by(
@@ -129,6 +129,9 @@ treeVarsTbl <- cleanData %>%
 #funcw/formulahard
 
 
+
+
+
 #ordinate
 
 commTbl <- spTbl %>%
@@ -150,8 +153,7 @@ metaTbl <- cleanData %>%
   )
 
 
-ordModel <- distMat ~
-  dist
+ordModel <- distMat ~ dist
 
 ordStat = adonis(
   ordModel,

### 2021-08-05 analysis/statsTbl.R

@@ -35,7 +35,7 @@ getStatsTbl <- function(
     ) %>%
     addStatEvalNon() %>%
 
-    addStatGraph(
+    addGraph(
       ...formula
     )
 }
@@ -49,7 +49,7 @@ getStatsTbl2 <- function(
 ) {
 
   ...nestedVarDataTbl %>%
-    statFitTbl2(
+    statFitTbl(
       ...formula2
     ) %>%
     addStatEval2(
@@ -68,7 +68,7 @@ getStatsTbl1 <- function(
 
   ...nestedVarDataTbl %>%
 
-    statFitTbl(
+    statFitTbl1(
       ...formula
     ) %>%
     addStatEval1(
@@ -81,7 +81,36 @@ getStatsTbl1 <- function(
     ) %>%
     addStatEvalNon() %>%
 
-    addStatGraph(
+    addGraph(
+      ...formula
+    )
+}
+
+
+
+getStatsTbl1g2 <- function(
+  #has"varData"col
+  ...nestedVarDataTbl,
+  #user's
+  ...formula
+) {
+
+  ...nestedVarDataTbl %>%
+
+    statFitTbl1(
+      ...formula
+    ) %>%
+    addStatEval1(
+      ...formula
+    ) %>%
+
+    #nonnormaldistr
+    addStatFitNon(
+      ...formula
+    ) %>%
+    addStatEvalNon() %>%
+
+    addGraph2(
       ...formula
     )
 }
@@ -103,10 +132,11 @@ statFitTbl <- function(
         modify(
           ~ .x %>%
             lm(
-              formula = ....formula
+              formula =
+                ....formula
               )
           ),
-      statPrint =
+      "statPrint" =
         statTest %>%
         modify(
           ~ .x %>%
@@ -126,14 +156,41 @@ statFitTbl <- function(
 }
 
 
-#wrapper4consistency"-2"
-statFitTbl2 <- function(
+
+statFitTbl1 <- function(
   ...nestedVarDataTbl,
-  ....formula2
+  ....formula
 ) {
-  statFitTbl(
-    ...nestedVarDataTbl,
-    ....formula2)
+
+  ...nestedVarDataTbl %>%
+
+    mutate(
+      "statTest" =
+        varData1 %>%
+        modify(
+          ~ .x %>%
+            lm(
+              formula =
+                ....formula
+            )
+        ),
+      "statPrint" =
+        statTest %>%
+        modify(
+          ~ .x %>%
+            tidy() %>%
+            full_join(
[…]

### 2021-08-10 analysis/statsTbl.R
used poly fits, zooming in to sp. from here

@@ -71,24 +71,18 @@ getStatsTbl1 <- function(
     statFitTbl1(
       ...formula
     ) %>%
-    addStatEval1(
+    addStatEval(
       ...formula
     ) %>%
 
-    #nonnormaldistr
-    addStatFitNon(
-      ...formula
-    ) %>%
-    addStatEvalNon() %>%
-
-    addGraph(
+    addGraph1(
       ...formula
     )
 }
 
 
 
-getStatsTbl1g2 <- function(
+getStatsTbl12 <- function(
   #has"varData"col
   ...nestedVarDataTbl,
   #user's
@@ -100,17 +94,7 @@ getStatsTbl1g2 <- function(
     statFitTbl1(
       ...formula
     ) %>%
-    addStatEval1(
-      ...formula
-    ) %>%
-
-    #nonnormaldistr
-    addStatFitNon(
-      ...formula
-    ) %>%
-    addStatEvalNon() %>%
-
-    addGraph2(
+    addStatEval2(
       ...formula
     )
 }
@@ -134,8 +118,8 @@ statFitTbl <- function(
             lm(
               formula =
                 ....formula
-              )
-          ),
+            )
+        ),
       "statPrint" =
         statTest %>%
         modify(
@@ -166,7 +150,7 @@ statFitTbl1 <- function(
 
     mutate(
       "statTest" =
-        varData1 %>%
+        varData1 %>%  #diffhere
         modify(
           ~ .x %>%
             lm(
@@ -279,7 +263,7 @@ addStatEval <- function(
       ) %>%
 
     mutate(
-      R2 =
+      "R2adj" =
         statPrint %>%
         modify(
           ~ .x %>%
@@ -293,7 +277,7 @@ addStatEval <- function(
         )
     ) %>%
     unnest(
-      R2
+      R2adj
     )
 }
 
@@ -307,7 +291,8 @@ addStatEval2 <- function(
   ...statFitTbl2 %>%
 
     mutate(
-      curveName =
+      #diffhere
+      "curveName" =
         statPrint %>%
         modify(
           ~ .x %>%
@@ -339,8 +324,11 @@ addStatEval2 <- function(
               0.055,
             T, F
           )
-        ),
-      "R2" =
+        )
+    ) %>%
+
+    mutate(
+      "R2adj" =
         statPrint %>%
         modify(
           ~ .x %>%
@@ -354,85 +342,7 @@ addStatEval2 <- function(
         )
[…]

### 2021-08-12 analysis/stats.R
QC'd taxa tbl for shapiro

@@ -6,6 +6,7 @@ library(rstatix);
 library(vegan) #diversity()
 
 
+
 #4vegan
 spTbl <- cleanData %>%
   add_count(
@@ -28,6 +29,7 @@ spTbl <- cleanData %>%
         .
         )
     )
+
 
 
 plotVarsTbl <- cleanData %>%
@@ -95,6 +97,7 @@ plotVarsTbl <- cleanData %>%
   )
 
 
+
 #rid?
 treeVarsTbl <- cleanData %>%
   group_by(
@@ -130,8 +133,6 @@ treeVarsTbl <- cleanData %>%
 
 
 
-
-
 #ordinate
 
 commTbl <- spTbl %>%
@@ -160,3 +161,59 @@ ordStat = adonis(
   metaTbl,
   99999
 )
+
+
+
+#taxa
+
+# which sp only have 2 dists?
+spStatTbl <- cleanData %>%
+  select(
+    dist,
+    plot,
+    gen
+  ) %>%
+  add_count(
+    gen,
+    plot,
+  ) %>%
+  distinct(
+    gen,
+    plot,
+    .keep_all = T
+  ) %>%
+  nest(
+    "varData" = -gen
+  ) %>%
+
+  #QCshapiro
+  mutate(,
+    nDist = varData %>%
+      modify(
+        ~ .x %>%
+          pull(
+            dist
+          ) %>%
+          n_distinct()
+      ),
+    varN = varData %>%
+      modify(
+        ~ .x %>%
+          pull(
+            n
+          ) %>%
+          var()
+      )
+  ) %>%
+  unnest(
+    c(
+      nDist,
+      varN
+    )
+  ) %>%
+  filter(
+    nDist >=
+      3,
+    varN >
+      0
+  )

### 2021-08-13 analysis/stats.R
isSignif unnest bug

@@ -187,7 +187,7 @@ spStatTbl <- cleanData %>%
   ) %>%
 
   #QCshapiro
-  mutate(,
+  mutate(
     nDist = varData %>%
       modify(
         ~ .x %>%

### 2021-08-13 analysis/statsTbl.R

@@ -243,7 +243,11 @@ addStatEval <- function(
           isNormal &
             isHomosced,
           T, F
-        ),
+        )) %>%
+    unnest(
+      isModelOK
+    ) %>%
+    mutate(
       "isSignif" =
         statPrint %>%
         modify_if(

### 2021-09-01 article/article.Rmd
adding col to eval stat test assumptions
included non-normal stats, attempting entropy by size class, not species
added entropy calc, need to mutate it before nesting and stat pipeline
updated article, directions...

@@
 
+Forested areas in the tropics offer many [economic and regulatory] services like timber, fuel, biomass/carbon storage, and biodiversity maintenance.
+Today, most forested areas in the tropics are secondary (FAO 2020) and show legacy effects of human activity like logging and intensive agriculture.
+
+In more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e.
+near 1:1 replacement) the targeted tree cutting that is done using information collected about the ecological profiles and ages (when possible) of trees (Duah-Gyamfi et al 2014, Edwards et al 2014).
+However, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015), but also usually show negative trade-offs between services, like that with biodiversity, due to monoculture planting designs.
+In some cases where management decisions are made primarily based on regional economic but not local ecological information, often lacking access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.
+After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g.
+based on surrounding forest proximity.
+
+Previous studies show some consistent patterns of secondary forest succession, while others show inconsistent trajectories.
+For example, syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years) with respect to tree height, above-ground biomass (Poorter et al 2016), and also species richness in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).
+Higher richness may be enough to maintain carbon storage (Liu et al 2018), but lagging changes in composition may contribute to reduced effects of diversity on other forest functions (Griess et al 2012).
+Unlike richness, changes to species composition likely last up to centuries, showing a strong legacy of early-successional communities that are mostly fast-growing, low-wood-density, shade-intolerant pioneering species (Tabarelli et al 2014, Tabarelli & Lopes 2008).
+Slower changes in composition can be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, which may also be managed, possibly explaining why different tropical forest trajectories diverge over time (Nordon et al 2015, Christensen 2014, Munroe et al 2013, Chazdon et al 2007).
+
+Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009), with underlying arguments based on metapopulation theory in ecology (Levins 1969, MacArthur & Wilson 1967), but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.
+Especially for abandoned land that regenerates via various processes (Stanturf et al 2014), the proximity to less disturbed habitat can affect the composition of the regrowing community by changing effective seed dispersal kernels and probability of establishment (Wandrag et al 2017, Guariguata & Ostertag 2001), along with animal disperser communities themselves (Gilroy et al 2014).
+While dispersal is complicated (Muller-Landau & Hardesty 2005), it can broadly correlate with simpler plant traits such as height and seed mass (Tamme et al 2014, Thomson et al 2011).
+Edge effects can also operate via the style of management, including logging decisions (Edwards et al 2014, Clark et al 2012) and land-use history generally (Omeja et al 2012), can also affect stand diversity maintenance, with examples of high diversity (compared to older adjacent stands) in agroforests (Oliveira-Neto et al 2017) and native timber plantations (Pryde et al 2015), possibly by changing early tree survival probabilities, growth rates, and/or competition, but surprisingly less likely facilitation by nitrogen-fixing trees (Lai et al 2018).
+It may also be that due to both high recruitment and mortality (turnover) of pioneering species' stems, transitions from early- to late-successional communities may be more complex than gradual (van Greugel et al 2007).
+The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density (César et al 2021), and could therefore require a switch in focal timber species (Karsten et al 2014).
+
+Here we study the regeneration of an abandoned timber plantation as a function of distance from surrounding native forest, as means of understanding how these local ecological principles can support forest stewardship (Hulshoff & Powers 2019).
+We hypothesized the planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity.
+We also considered how the legacy plantation canopy structure mediated forest stand effects via changes in abiotic conditions like light availability and microtopography (Powers & Peréz-Aviles 2012).
+
 # Results
@@
     -   can break down biomass by component
-    -   maybe system-level curve (e.g. biomas hist) + timber sp
+
+    -   maybe system-level curve (e.g. biomass hist) + timber sp
+
+        -   metric hist, vs. sp-abund/diversity hist below
+
+    -   max DAP as trait?
 
@@
 
-## Study site
-
--   osa peninsula (GPS)
--   wet lowland rainforest
--   greg gund conservation center, osa conservation
-
-## Plot design
-
--   QGIS/ArcGIS 50m bands 0-300m from forest
--   30 plots, random within each band, allocated by area: 11, 8, 5, 3, 2, 1
--   20x20m
-
-## Data collection
-
-### Canopy light and slope
-
--   densiometer in 4 directions at center of plot
--   slope?
-
-### Tree biomass
-
--   DBH, H using rangefinder (BUSHNELL)
-
--   chave 2014, updated to chave 2017, little difference in result
-
--   unique individual tag/plate \#s
-
--   species IDs
-
-    -   local knowledge of common names
-
-    -   taxonomy matches using local field guides
-
--   wood density using Agroforestry Wood Database in 2013, updated using BIOMASS package
-
 ## Statistical analysis
-
--   code available (somewhere..., DRYAD, GitHub)
 
@@
 
+        -   justify metrics, i.e. entropy vs. hists
+
+## Study site
+
+This study was done in the lowland wet tropical rainforest of the Osa Peninsula in SW Puntarenas, Costa Rica. The forest fragment used was acquired by the Natural
+Resources Defense Council (nrdc.org) and managed by the local non-profit
+organization Osa Conservation (osaconservation.org), located at their Greg Gund Conservation Center, near the more widely-used Piro research station.
+The focal parcel was 20 ha in size and was undergoing secondary succession after having been abandoned \~20 years ago as a wood plantation for *Bombacopsis quinata*, which has usually been better suited to the dry forests of Guanacaste and the Nicoya Peninsula (Pérez Cordero & Kanninen 2002).
+The secondary forest stand was roughly triangular and surrounded by primary forest on the S and NW sides.
+This census was done in 2013 during the rainy season months between June and August.
+
+## Census design
+
+Geographic information systems software (QGIS 2.0, qgis.org; ArcGIS 10, arcgis.com) was used to divide the parcel area into six discrete strata representing 50-m intervals from the observed primary forest edge to the center of the opposing NE edge, \~300 m away.
+This design meant that strata were smaller further from the primary forest edge, so accordingly, fewer plots were randomly positioned inside each stratum: the 0-50, 50-100, 100-150, 150-200, 200-250, and 250-300 m strata had eleven, eight, five, four, two, and one plot(s), respectively.
+The edge distance variable was descretized into strata this way as a more feasible alternative to measuring plot-edge distances after plot placement, which would have taken much longer compared to the improved spatial resolution.
+Each plot was 21 x 21 m and oriented North, and the total area represented by all the surveyed plots was \~1 ha or 5% of the total stand area, comparable to similar studies (Onyekwelu & Olabiwannu 2016).
+
+## Plot measurements / Data collection
+
+### Canopy cover & Topography
+
+Light reaching the forest floor was measured at the center of each plot, at chest
+height, using a densiometer (Foresty Suppliers, Inc.).
+Following manufacturer instructions, the final plot value was the average of 4 different readings taken facing each of the 4 cardinal directions: North, South, East, and West.
+
+### [Topography]
+
+The slope of the forest floor in each plot was measured using [a rangefinder by measuring the distance from one (specific?) corner of the plot to another (specific?) corner.]
+
+### Tree measurements
+
+In each plot, the diameter at breast height (DBH = 2.5 -- 3 m), crown height, and
+taxonomic ID was recorded for every stem ≥10 cm in [diameter/circumference?].
+Following common forestry guidelines, in cases where a tree split into 2 or more
+stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem. Tree
+height was measured by using a rangefinder (Bushnell, Forestry Suppliers, Inc.)
+to measure from an observer to the crown (from where it was visible), and then
+to the trunk at breast height; these 2 distances were then used to triangulate
+and calculate the tree's height using Pythagoras's Theorem.
+
+Trees were distinguished based on local ecological knowledge of common names, followed by matching to Latin binomial taxonomic names using printed local field guides available.
+
+### Aboveground biomass
+
+Aboveground biomass (AGB) was estimated for each stem using the allometric equation developed by Chave et al. (2014) specific for wet tropical forests,
+which has performed slightly better compared to both pantropical and previous other models by Chave et al. (2005) (Ngomanda et al 2014, Alvarez et al 2012),
+with improved local accuracy adding height (Khan et al 2020, Domke et al 2012)
+and wood density (specific gravity) (van Breugel et al 2011):
+
+*AGB = 0.0673 x (ρ D^2^ H)^0.976^*
+
+with *D* in cm, *H* in m, and *p* in g cm^-3^.
+Measured DBHs and heights were used, along with wood densities (*ρ*) extracted from available literature.
+In most cases, these values were only available at the genus level, and unrepresented taxa were assumed to be 0.58, an assumption recommended by the World Agroforestry Database (worldagroforestry.org).
+[References used for this step are given in *Table S1*.] Original values from 2013 were then updated using the pipeline developed in the BIOMASS package for R by Réjou-Méchain et al.
+(2017), which replaced unknown wood density values with plot-level averages instead.
+However, these updates did not appear to change the main results of this study.
+
+### Statistical analyses
+
+All variables were averaged at the plot level, including biomass converted to a per hectare basis, thereby setting *n* = 30 for all analyses reported.
+
+All response variables passed Shapiro-Wilk normality and Levene's homogeneity tests before being subject to parametric tests---either multi-way ANOVA or one-way ANOVA and linear regression, or generalized (non-linear) regression.
+
+Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.
+These variables are also shown in figures with their own separate analyses, exclusively one-way ANOVAs with additional relevant parameters from their linear regression counterparts (*i.e.* for *R^2^*and *slope*).
+Standard one-way ANOVAs were also run for hill slope and canopy
+light availability separately. Non-linear regressions were run as linear
+regressions with slight modifications using the function "poly()" in the
+"stats" package for R software, which was only used for stem density and
+RICHNESS variables. All individual trees censused were included in all
+analyses, including unidentified taxa, but these averaged to only 1 group for
+taxa-related analyses, namely RICHNESS with distance, having a minor effect. All
+tests included only distance to primary forest edge as the dependent variable,
+and all data was analyzed in R v3.6.1 software (R Core Team 2019).
+
+-   Allen and Kim (2020) as some justification for family-level analysis (their MI forest site shares \~6 / their 20 families with this site)
+
+```{=html}
+<!-- -->
+```
+-   Neighborhood influence metric (Lorimer 1983, applied in Looney et al 2016)
+-   Data and analytical code available on GitHub.../DRYAD
+
 # Acknowledgements

### 2021-09-01 analysis/stats.R

@@ -166,7 +166,6 @@ ordStat = adonis(
 
 #taxa
 
-# which sp only have 2 dists?
 spStatTbl <- cleanData %>%
   select(
     dist,
@@ -216,4 +215,20 @@ spStatTbl <- cleanData %>%
       3,
     varN >
       0
+  ) %>%
+
+  mutate(
+    "varData1" =
+      varData %>%
+      modify(
+        ~ .x %>%
+          group_by(
+            dist
+          ) %>%
+          summarize(
+            n = median(
+              n
+            )
+          )
+      )
   )

### 2021-09-02 article/article.Rmd
fixed bugs, revisiting framework

@@
 Forested areas in the tropics offer many [economic and regulatory] services like timber, fuel, biomass/carbon storage, and biodiversity maintenance.
-Today, most forested areas in the tropics are secondary (FAO 2020) and show legacy effects of human activity like logging and intensive agriculture.
 
-In more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e.
-near 1:1 replacement) the targeted tree cutting that is done using information collected about the ecological profiles and ages (when possible) of trees (Duah-Gyamfi et al 2014, Edwards et al 2014).
-However, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015), but also usually show negative trade-offs between services, like that with biodiversity, due to monoculture planting designs.
-In some cases where management decisions are made primarily based on regional economic but not local ecological information, often lacking access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.
-After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g.
-based on surrounding forest proximity.
+At the same time, most forested areas in the tropics are secondary (FAO 2020), showing legacy effects of human activity like logging and intensive agriculture.
 
-Previous studies show some consistent patterns of secondary forest succession, while others show inconsistent trajectories.
-For example, syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years) with respect to tree height, above-ground biomass (Poorter et al 2016), and also species richness in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).
-Higher richness may be enough to maintain carbon storage (Liu et al 2018), but lagging changes in composition may contribute to reduced effects of diversity on other forest functions (Griess et al 2012).
-Unlike richness, changes to species composition likely last up to centuries, showing a strong legacy of early-successional communities that are mostly fast-growing, low-wood-density, shade-intolerant pioneering species (Tabarelli et al 2014, Tabarelli & Lopes 2008).
-Slower changes in composition can be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, which may also be managed, possibly explaining why different tropical forest trajectories diverge over time (Nordon et al 2015, Christensen 2014, Munroe et al 2013, Chazdon et al 2007).
+As a result, tropical reforestation is a widespread ecological process that must be better understood to optimize [local] tradeoffs in ecosystem services [CITE?].
 
-Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009), with underlying arguments based on metapopulation theory in ecology (Levins 1969, MacArthur & Wilson 1967), but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.
+Previous studies have shown some general/consistent patterns of secondary forest succession, although other studies show inconsistent trajectories.
+
+For example, syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years) with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).
+
+However, while restoring species richness may be enough to maintain carbon storage (Liu et al 2018), it does not guarantee the restoration of species composition.
+
+Instead, changes to species composition likely last up to centuries, showing a strong legacy of priority effects from early-successional communities that are mostly fast-growing, low-wood-density, shade-intolerant pioneering species (Tabarelli et al 2014, Tabarelli & Lopes 2008).
+
+Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests (Nordon et al 2015, Christensen 2014, Munroe et al 2013, Chazdon et al 2007).
+
+but lagging changes in composition may contribute to reduced effects of diversity on other forest functions (Griess et al 2012).
+
+Restoring species composition more quickly is likely a place where active forest managers can have the most unique impacts.
+
+For land managers / foresters with interest in conservation, inheriting land affected by past decisions is a reality.
+One of the legacies is where the land was placed.
+To best manage it in the future, an initial assessment of the ecological status of the land is important to understand it's likely trajectory, and what it will need in the future for tailed services/conservation/protection.
+
+Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009),
+
+with underlying arguments based on metapopulation theory in ecology (Levins 1969, MacArthur & Wilson 1967), but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.
+
 Especially for abandoned land that regenerates via various processes (Stanturf et al 2014), the proximity to less disturbed habitat can affect the composition of the regrowing community by changing effective seed dispersal kernels and probability of establishment (Wandrag et al 2017, Guariguata & Ostertag 2001), along with animal disperser communities themselves (Gilroy et al 2014).
+
 While dispersal is complicated (Muller-Landau & Hardesty 2005), it can broadly correlate with simpler plant traits such as height and seed mass (Tamme et al 2014, Thomson et al 2011).
-Edge effects can also operate via the style of management, including logging decisions (Edwards et al 2014, Clark et al 2012) and land-use history generally (Omeja et al 2012), can also affect stand diversity maintenance, with examples of high diversity (compared to older adjacent stands) in agroforests (Oliveira-Neto et al 2017) and native timber plantations (Pryde et al 2015), possibly by changing early tree survival probabilities, growth rates, and/or competition, but surprisingly less likely facilitation by nitrogen-fixing trees (Lai et al 2018).
+
+Edge effects can also operate via the style of management, including logging decisions (Edwards et al 2014, Clark et al 2012) and land-use history generally (Omeja et al 2012), can also affect stand diversity maintenance,
+
+with examples of high diversity (compared to older adjacent stands) in agroforests (Oliveira-Neto et al 2017) and native timber plantations (Pryde et al 2015),
+
+possibly by changing early tree survival probabilities, growth rates, and/or competition,
+
+but surprisingly less likely facilitation by nitrogen-fixing trees (Lai et al 2018).
+
 It may also be that due to both high recruitment and mortality (turnover) of pioneering species' stems, transitions from early- to late-successional communities may be more complex than gradual (van Greugel et al 2007).
+
 The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density (César et al 2021), and could therefore require a switch in focal timber species (Karsten et al 2014).
 
-Here we study the regeneration of an abandoned timber plantation as a function of distance from surrounding native forest, as means of understanding how these local ecological principles can support forest stewardship (Hulshoff & Powers 2019).
-We hypothesized the planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity.
+To understand the ecological principles behind secondary forest regeneration (Hulshoff & Powers 2019), here we study the regeneration of an abandoned timber plantation as a function of distance from surrounding native forest.
+
+We hypothesized that planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity.
+
 We also considered how the legacy plantation canopy structure mediated forest stand effects via changes in abiotic conditions like light availability and microtopography (Powers & Peréz-Aviles 2012).
@@
 
+In more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e.
+near 1:1 replacement) the targeted tree cutting that is done using information collected about the ecological profiles and ages (when possible) of trees (Duah-Gyamfi et al 2014, Edwards et al 2014).
+However, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015), but also usually show negative trade-offs between services, like that with biodiversity, due to monoculture planting designs.
+In some cases where management decisions are made primarily based on regional economic but not local ecological information, often lacking access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.
+After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g.
+based on surrounding forest proximity.
+
 # Methods
@@
 
-This study was done in the lowland wet tropical rainforest of the Osa Peninsula in SW Puntarenas, Costa Rica. The forest fragment used was acquired by the Natural
-Resources Defense Council (nrdc.org) and managed by the local non-profit
-organization Osa Conservation (osaconservation.org), located at their Greg Gund Conservation Center, near the more widely-used Piro research station.
+This study was done in the lowland wet tropical rainforest of the Osa Peninsula in SW Puntarenas, Costa Rica.
+The forest fragment used was acquired by the Natural Resources Defense Council (nrdc.org) and managed by the local non-profit organization Osa Conservation (osaconservation.org), located at their Greg Gund Conservation Center, near the more widely-used Piro research station.
 The focal parcel was 20 ha in size and was undergoing secondary succession after having been abandoned \~20 years ago as a wood plantation for *Bombacopsis quinata*, which has usually been better suited to the dry forests of Guanacaste and the Nicoya Peninsula (Pérez Cordero & Kanninen 2002).
@@
 
-Light reaching the forest floor was measured at the center of each plot, at chest
-height, using a densiometer (Foresty Suppliers, Inc.).
+Light reaching the forest floor was measured at the center of each plot, at chest height, using a densiometer (Foresty Suppliers, Inc.).
 Following manufacturer instructions, the final plot value was the average of 4 different readings taken facing each of the 4 cardinal directions: North, South, East, and West.
@@
 
-In each plot, the diameter at breast height (DBH = 2.5 -- 3 m), crown height, and
-taxonomic ID was recorded for every stem ≥10 cm in [diameter/circumference?].
-Following common forestry guidelines, in cases where a tree split into 2 or more
-stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem. Tree
-height was measured by using a rangefinder (Bushnell, Forestry Suppliers, Inc.)
-to measure from an observer to the crown (from where it was visible), and then
-to the trunk at breast height; these 2 distances were then used to triangulate
-and calculate the tree's height using Pythagoras's Theorem.
+In each plot, the diameter at breast height (DBH = 2.5 -- 3 m), crown height, and taxonomic ID was recorded for every stem ≥10 cm in [diameter/circumference?].
+Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
+Tree height was measured by using a rangefinder (Bushnell, Forestry Suppliers, Inc.) to measure from an observer to the crown (from where it was visible), and then to the trunk at breast height; these 2 distances were then used to triangulate and calculate the tree's height using Pythagoras's Theorem.
 
@@
 
-Aboveground biomass (AGB) was estimated for each stem using the allometric equation developed by Chave et al. (2014) specific for wet tropical forests,
-which has performed slightly better compared to both pantropical and previous other models by Chave et al. (2005) (Ngomanda et al 2014, Alvarez et al 2012),
-with improved local accuracy adding height (Khan et al 2020, Domke et al 2012)
-and wood density (specific gravity) (van Breugel et al 2011):
+Aboveground biomass (AGB) was estimated for each stem using the allometric equation developed by Chave et al.
+(2014) specific for wet tropical forests, which has performed slightly better compared to both pantropical and previous other models by Chave et al.
+(2005) (Ngomanda et al 2014, Alvarez et al 2012), with improved local accuracy adding height (Khan et al 2020, Domke et al 2012) and wood density (specific gravity) (van Breugel et al 2011):
 
@@
 These variables are also shown in figures with their own separate analyses, exclusively one-way ANOVAs with additional relevant parameters from their linear regression counterparts (*i.e.* for *R^2^*and *slope*).
-Standard one-way ANOVAs were also run for hill slope and canopy
-light availability separately. Non-linear regressions were run as linear
-regressions with slight modifications using the function "poly()" in the
-"stats" package for R software, which was only used for stem density and
-RICHNESS variables. All individual trees censused were included in all
-analyses, including unidentified taxa, but these averaged to only 1 group for
-taxa-related analyses, namely RICHNESS with distance, having a minor effect. All
-tests included only distance to primary forest edge as the dependent variable,
-and all data was analyzed in R v3.6.1 software (R Core Team 2019).
+Standard one-way ANOVAs were also run for hill slope and canopy light availability separately.
+Non-linear regressions were run as linear regressions with slight modifications using the function "poly()" in the "stats" package for R software, which was only used for stem density and RICHNESS variables.
+All individual trees censused were included in all analyses, including unidentified taxa, but these averaged to only 1 group for taxa-related analyses, namely RICHNESS with distance, having a minor effect.
+All tests included only distance to primary forest edge as the dependent variable, and all data was analyzed in R v3.6.1 software (R Core Team 2019).
 
@@
 
-```{=html}
-<!-- -->
-```
 -   Neighborhood influence metric (Lorimer 1983, applied in Looney et al 2016)

### 2021-09-03 article/article.Rmd
workshopped intro

@@
 
-Forested areas in the tropics offer many [economic and regulatory] services like timber, fuel, biomass/carbon storage, and biodiversity maintenance.
+Forest regeneration has the potential to store \#\#C and preserve \#\# diversity globally in the coming decades (UN, Bonn challenge).
 
-At the same time, most forested areas in the tropics are secondary (FAO 2020), showing legacy effects of human activity like logging and intensive agriculture.
+To realize this potential, it is fundamental that we better understand how general ecological processes manifest under widespread conservation contexts.
 
-As a result, tropical reforestation is a widespread ecological process that must be better understood to optimize [local] tradeoffs in ecosystem services [CITE?].
+At local scales, regeneration is strongly affected by landscape [ecological] heterogeneity as well as management history.
 
-Previous studies have shown some general/consistent patterns of secondary forest succession, although other studies show inconsistent trajectories.
+Most tropical forests are secondary (FAO 2020), meaning that forest regeneration is widespread.
 
-For example, syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years) with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).
+However, patterns of forest regeneration remain unclear, especially at more local scales (\< \~50 ha).
 
-However, while restoring species richness may be enough to maintain carbon storage (Liu et al 2018), it does not guarantee the restoration of species composition.
+Still, forests provide key functions like wood, carbon storage, and biodiversity maintenance.
 
-Instead, changes to species composition likely last up to centuries, showing a strong legacy of priority effects from early-successional communities that are mostly fast-growing, low-wood-density, shade-intolerant pioneering species (Tabarelli et al 2014, Tabarelli & Lopes 2008).
+Therefore, better understanding forest regeneration is an integral part of achieving more sustainable development.
+
+Some general patterns of forest regeneration have emerged, but this also includes the lack of pattern in some cases.
+
+Syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years), mainly with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).
+
+However, while restoring species richness may be enough to maintain some ecosystem functions like carbon storage (Liu et al 2018), it does not guarantee the restoration of species composition.
+
+Instead, changes to species composition likely last up to centuries, showing a strong legacy of priority effects from pioneering, early-successional communities that are mostly fast-growing, low-wood-density, and shade-intolerant (Tabarelli et al 2014, Tabarelli & Lopes 2008).
+
+forests show legacy effects of common human activity, like logging and intensive agriculture/agroforestry/silviculture (FAO 2020).
+
+[High] Regional biodiversity [can] affects/increases the uncertainty about the local occurrences and outcomes of specific species interactions,
+
+such as after dispersal and/or niche overlap [CITE-niche, neutral theory].
+
+[At the same time,] Diverse local management activity modifies demographic states/profiles of various taxa's population dynamics.
+
+For example, active timber plantations strongly promote the population growth of focal taxa at the expense of non-focal taxa.
+
+After abandonment, however, the ecosystem likely reflects how continuous regional ecological processes look different given past management.
+
+such as seed dispersal from nearby primary forest edges.
+
+[These] Interactions between background ecology and human management ultimately have both local and general implications for the potential conservation/stability of ecosystem services/functions
+
+like biomass/carbon storage and biodiversity maintenance,
+
+as well as manager livelihood.
 
@@
 
-In more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e.
+Under more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e.
 near 1:1 replacement) the targeted tree cutting that is done using information collected about the ecological profiles and ages (when possible) of trees (Duah-Gyamfi et al 2014, Edwards et al 2014).
-However, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015), but also usually show negative trade-offs between services, like that with biodiversity, due to monoculture planting designs.
-In some cases where management decisions are made primarily based on regional economic but not local ecological information, often lacking access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.
+
+In contrast, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015) while also usually showing negative trade-offs with other services, like that with biodiversity given reliance on monoculture planting designs.
+
+In some cases where management decisions are made primarily based on regional economic but not local ecological information, due to limited access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.
+
 After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g.

### 2021-09-08 article/article.Rmd
polished intro for now

@@
     fig_caption: yes
+bibliography: references.bib
 ---
@@
 
-1.  importance of reforestation
+1.  importance of reforestation/regeneration/regrowth
 
@@
 
+    2.  How does the forest regenerate by distance to primary forest?
+
+        1.  how much [effort] do we need to re-seed? / how much can we leave to nature?
+
+Global forest and landscape regeneration is a key conservation and climate change mitigation strategy, adopted internationally (\@UN14 Paris, Bonn Challenge).
+
+-   Tropical forests store most terrestrial biomass @pan2013 and biodiversity (CITE, but are losing functions, especially in the neotropics @hubau2020, as most tropical forests become secondary (\@FAO20.
+
+Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
+
+-   Neotropical secondary forests accumulate biomass \>10x faster than old-growth forests @poorter2016, and offset \~10% of carbon emissions from Amazonian deforestation @smith2020 .
+-   They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y @rozendaal2019 .
+-   However, secondary neotropical forests diverge widely in taxonomic composition @norden2015 .
+
+Divergent communities among secondary forests may warrant more localized conservation efforts, studying relative effects of generalized community processes.
+
+Half of all forests lie half of a kilometer from an ecotone @haddad2015, suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.
+
+-   Tropical forests lose 22% carbon along edges \>100m, in part due to shorter trees and less leaf mass @ordway2020 .
+
+Here we report the edge effects of an abandoned wet neotropical timber plantation.
+
+-   We hypothesized that remnant timber trees would mediate forest regrowth by maintaining uniform shade that would suppress shade-intolerant / pioneer trees / taxa and allow for more diversity.
+
 Forest regeneration has the potential to store \#\#C and preserve \#\# diversity globally in the coming decades (UN, Bonn challenge).
-
-To realize this potential, it is fundamental that we better understand how general ecological processes manifest under widespread conservation contexts.
 
@@
 
-Most tropical forests are secondary (FAO 2020), meaning that forest regeneration is widespread.
-
 However, patterns of forest regeneration remain unclear, especially at more local scales (\< \~50 ha).
-
-Still, forests provide key functions like wood, carbon storage, and biodiversity maintenance.
-
-Therefore, better understanding forest regeneration is an integral part of achieving more sustainable development.
-
-Some general patterns of forest regeneration have emerged, but this also includes the lack of pattern in some cases.
 
@@
 
+Edge effects are known for degrading primary forest, but they can also be used to assist reforestation.
+
+Traditional edge effects act by mortality of large trees that makes room for pioneer species to enter, but existing interior pioneer species also have dispersal radii that overlap with the adjacent exterior, potentially facilitating regeneration.
+
+Encroachment vs. facilitation relies on a balance between mortality (per area) and dispersal + establishment + stem growth.
+
+So pioneer species can be helpful for reforestation though they are typically seen as bad for old forest conservation.
+
 Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009),
@@
 
+-   how can we expect it to regrow?
+-   how far/much [effort] do we need to re-seed?
+
 We hypothesized that planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity.
@@
 
-1.  shade x distance + timber sp
+1.  fig1
 
-## Edge effects
+    1.  shade x distance @ordway2020 + timber sp?
 
-1.  biomass x distance + timber sp
+    2.  height @ordway2020
 
-    -   can break down biomass by component
+    3.  wood density @poorter2019
+
+    4.  total biomass x dist @ordway2020 + timber sp?
 
@@
         -   max DAP as trait?
+
+## Edge effects
+
+1.  fig2
+
+    1.  richness
 
@@
 -   Neighborhood influence metric (Lorimer 1983, applied in Looney et al 2016)
+
 -   Data and analytical code available on GitHub.../DRYAD

### 2021-09-08 analysis/stats.R

@@ -7,18 +7,31 @@ library(vegan) #diversity()
 
 
 
+#userInput
+taxRank <- quote(
+  gen
+)
+
 #4vegan
 spTbl <- cleanData %>%
   add_count(
-    gen,
+    {
+      {
+        taxRank
+      }
+    },
     plot, #group
-    name = "n_gen",
+    name = "n_tax",
     sort = T
-    ) %>%
+    ) #%>%
   pivot_wider(
     id_cols = plot,
-    names_from = gen,
-    values_from = n_gen,
+    names_from = {
+      {
+        taxRank
+      }
+    },
+    values_from = n_tax,
     #rmwarn
     values_fn = length,
     values_fill = 0
@@ -92,6 +105,21 @@ plotVarsTbl <- cleanData %>%
             mad = mad(
               median
             )
+          )
+      )
+  ) %>%
+
+  # hist parms
+  mutate(
+    "sys" = varData %>%
+      modify(
+        ~ .x %>%
+
+          pull(
+            median
+          ) %>%
+          hist(
+            .
           )
       )
   )
@@ -170,19 +198,35 @@ spStatTbl <- cleanData %>%
   select(
     dist,
     plot,
-    gen
+    {
+      {
+        taxRank
+      }
+    }
   ) %>%
   add_count(
-    gen,
+    {
+      {
+        taxRank
+      }
+    },
     plot,
   ) %>%
   distinct(
-    gen,
+    {
+      {
+        taxRank
+      }
+    },
     plot,
     .keep_all = T
   ) %>%
   nest(
-    "varData" = -gen
+    "varData" = -{
+      {
+        taxRank
+      }
+    }
   ) %>%
 
   #QCshapiro

### 2021-09-10 article/article.Rmd
fixed underlying data toggle in fig function

@@
 output:
+  html_document:
+    toc: yes
+    df_print: paged
   html_notebook:
@@
 
--   Tropical forests store most terrestrial biomass @pan2013 and biodiversity (CITE, but are losing functions, especially in the neotropics @hubau2020, as most tropical forests become secondary (\@FAO20.
+-   Tropical forests store most terrestrial biomass ( @pan2013 ) and biodiversity (CITE, but are losing functions, especially in the neotropics ( @hubau2020 ), as most tropical forests become secondary (\@FAO20.
 
@@
 
--   Neotropical secondary forests accumulate biomass \>10x faster than old-growth forests @poorter2016, and offset \~10% of carbon emissions from Amazonian deforestation @smith2020 .
--   They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y @rozendaal2019 .
--   However, secondary neotropical forests diverge widely in taxonomic composition @norden2015 .
+-   Neotropical secondary forests accumulate biomass \>10x faster than old-growth forests ( @poorter2016 ), and offset \~10% of carbon emissions from Amazonian deforestation ( @smith2020 ).
+-   They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y ( @rozendaal2019 ).
+-   However, secondary neotropical forests diverge widely in taxonomic composition ( @norden2015 ).
 
@@
 
-Half of all forests lie half of a kilometer from an ecotone @haddad2015, suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.
+Half of all forests lie half of a kilometer from an ecotone ( @haddad2015 ), suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.
 
--   Tropical forests lose 22% carbon along edges \>100m, in part due to shorter trees and less leaf mass @ordway2020 .
+-   Tropical forests lose 22% carbon along edges \>100m, in part due to shorter trees and less leaf mass ( @ordway2020 ).
 
@@
 -   We hypothesized that remnant timber trees would mediate forest regrowth by maintaining uniform shade that would suppress shade-intolerant / pioneer trees / taxa and allow for more diversity.
+
+    -   supported by inference?!
 
@@
 
-## Shade / Canopy
+```{r setup, echo = F, message = F}
+knitr::opts_chunk$set(
+  echo = F,
+  message = F,
+  warning = F
+)
+library(here)
+```
+
+## Biomass
 
@@
 
-    1.  shade x distance @ordway2020 + timber sp?
+    1.  shade x distance ( @ordway2020 ) + timber sp?
 
-    2.  height @ordway2020
+    2.  height ( @ordway2020 )
 
-    3.  wood density @poorter2019
+    3.  wood density ( @poorter2019 )
 
-    4.  total biomass x dist @ordway2020 + timber sp?
+    4.  total biomass x dist ( @ordway2020 ) + timber sp?
 
@@
 
-## Edge effects
+[RESULTS TEXT]
+
+```{r fig1}
+source(
+  here(
+    "figs/fig1.R"
+  )
+)
+fig1
+```
+
+## Composition
 
@@
         1.  system-level curve (e.g. rank-abund) x distance [+ 1 timber sp]
+
+```{r fig2}
+
+```
 

### 2021-09-10 analysis/stats.R

@@ -2,6 +2,9 @@ library(here);
 source(
   here("analysis/clean.R")
 )
+source(
+  here("analysis/statsTbl.R")
+)
 library(rstatix);
 library(vegan) #diversity()
 
@@ -23,7 +26,7 @@ spTbl <- cleanData %>%
     plot, #group
     name = "n_tax",
     sort = T
-    ) #%>%
+    ) %>%
   pivot_wider(
     id_cols = plot,
     names_from = {
@@ -40,8 +43,8 @@ spTbl <- cleanData %>%
     "entropy" =
       diversity(
         .
-        )
-    )
+      )
+  )
 
 
 
@@ -52,21 +55,21 @@ plotVarsTbl <- cleanData %>%
       select(
         plot,
         entropy
-        ),
+      ),
     by = "plot",
     copy = T
-    ) %>%
+  ) %>%
 
   group_by(
     dist,
     plot
-    ) %>%
+  ) %>%
   mutate(
     "richness" =
       n_distinct(
         sp
-        )
-    ) %>%
+      )
+  ) %>%
   select(
     #abiotic
     luz,
@@ -81,14 +84,14 @@ plotVarsTbl <- cleanData %>%
     #community
     richness,
     entropy
-    ) %>%
+  ) %>%
 
   get_summary_stats(
     type = "median_mad"
-    ) %>%
+  ) %>%
   nest(
     "varData" = -variable
-    ) %>%
+  ) %>%
 
   #nextlevel
   mutate(
@@ -119,45 +122,11 @@ plotVarsTbl <- cleanData %>%
             median
           ) %>%
           hist(
-            .
+            .,
+            plot = F
           )
       )
   )
-
-
-
-#rid?
-treeVarsTbl <- cleanData %>%
-  group_by(
-    dist,
-    plot,
-
-    fam,
-    gen,
-    sp
-    ) %>%
-  select(
-    dap,
-    h,
-    spg,
-    kg17
-    ) %>%
-  get_summary_stats(
-    type = "median_mad"
-    ) %>%
-  #4statTests
-  mutate(
-    median1 =
-      log10(
-        median + 1
-        )
-    ) %>%
-
-  nest(
-    varData = -variable
[…]

### 2021-09-10 figs/fig1.R

@@ -46,6 +46,8 @@ yAxisLabel4 <- quote(
   "Aboveground biomass (kg)"
 )
 
+
+
 varForm <- median ~ dist
 xAxisLabel <- quote(
   "Distance to secondary forest edge (m)"
@@ -67,7 +69,6 @@ graph1 <-  plotResultsTbl1 %>%
     yAxisLabel1,
     ..addGroups = T
   )
-
 graph1
 
 
@@ -81,10 +82,9 @@ graph2 <- plotResultsTbl1 %>%
     varForm[[2]],
     xAxisLabel,
     yAxisLabel2,
-    underData,
-    ..addGroups = T
+    ..addGroups = T,
+    underData
   )
-
 graph2
 
 
@@ -93,43 +93,14 @@ graph2
 
 graph3 <- plotResultsTbl1 %>%
   dotGraph(
-    measure2,
+    measure3,
     varForm[[3]],
     varForm[[2]],
     xAxisLabel,
-    yAxisLabel2
-  ) +
-
-  geom_point(
-    data = {
-      plotResultsTbl1 %>%
-        filter(
-          variable == measure3
-        ) %>%
-        select(
-          "varData1"
-        ) %>%
-        unnest()
-    },
-    color = graphPointColor,
-    size = graphPointSize
-  ) %>%
-
-
-  #move
-  append_layers(
-    geom_quasirandom(
-      data = cleanData,
-      aes(
-        y = measure3
-      ),
-      color = "gray",
-      size = 1
-    ),
-
-    position = "bottom"
+    yAxisLabel3,
+    underData,
+    ..addGroups = T
   )
-
 graph3
 
 
@@ -138,47 +109,17 @@ graph3
 
 graph4 <- plotResultsTbl1 %>%
   dotGraph(
-    measure2,
+    measure4,
     varForm[[3]],
     varForm[[2]],
     xAxisLabel,
-    yAxisLabel2
+    yAxisLabel4,
+    underData,
+    ..addGroups = T
   ) +
-
-  geom_point(
-    data = {
-      plotResultsTbl1 %>%
-        filter(
-          variable == measure4
-        ) %>%
-        select(
-          "varData1"
-        ) %>%
-        unnest()
-    },
-    color = graphPointColor,
-    size = graphPointSize
-  ) %>%
-
-
-  #move
-  append_layers(
-    geom_quasirandom(
-      data = cleanData,
-      aes(
-        y = measure4
-      ),
[…]

### 2021-09-14 article/article.Rmd
fig2 graph3 worked/fixed

@@
 
-Divergent communities among secondary forests may warrant more localized conservation efforts, studying relative effects of generalized community processes.
+Divergent communities among secondary forests may ultimately warrant more localized conservation efforts, studying relative effects of generalized community processes.
 

### 2021-09-14 analysis/clean.R

@@ -27,7 +27,7 @@ cleanData <- function(
       "newName" = correctTaxo(
         genus = gen,
         species = sp,
-        useCache = T
+        useCache = F
       ),
 
       #useful

### 2021-09-14 analysis/stats.R

@@ -163,6 +163,12 @@ ordStat = adonis(
 
 #taxa
 
+#user
+taxMetric <- quote(
+  kg17
+)
+
+
 spStatTbl <- cleanData %>%
   select(
     dist,
@@ -170,6 +176,11 @@ spStatTbl <- cleanData %>%
     {
       {
         taxRank
+      }
+    },
+    {
+      {
+        taxMetric
       }
     }
   ) %>%
@@ -181,6 +192,7 @@ spStatTbl <- cleanData %>%
     },
     plot,
   ) %>%
+  #collapse
   distinct(
     {
       {
@@ -239,8 +251,11 @@ spStatTbl <- cleanData %>%
             dist
           ) %>%
           summarize(
-            n = median(
-              n
+            "taxMetric" = median(
+              {
+                taxMetric %>%
+                  eval()
+              }
             )
           )
       )
@@ -250,9 +265,19 @@ spStatTbl <- cleanData %>%
 #ORGANIZE
 
 
+#user
+mainMetric <- quote(
+  median
+)
+
+
 #get
 
-mainModel <- median ~ dist
+mainModel <- {
+  mainMetric %>%
+    eval()
+} ~ dist
+
 
 plotResultsTbl <- plotVarsTbl %>%
   getStatsTbl(
@@ -262,7 +287,10 @@ plotResultsTbl <- plotVarsTbl %>%
 
 #get2
 
-mainModel2 <- median ~ poly(
+mainModel2 <- {
+  mainMetric %>%
+    eval()
+} ~ poly(
   dist,
   2
 )
@@ -284,12 +312,6 @@ plotResultsTbl1 <- plotVarsTbl %>%
 
 
 #get12
-
-mainModel2 <- median ~ poly(
-  dist,
-  2
-)
-
 plotResultsTbl12 <- plotVarsTbl %>%
   getStatsTbl12(
     mainModel2
@@ -305,7 +327,11 @@ plotResultsTbl12 <- plotVarsTbl %>%
 
 #getTaxa
 
-taxModel <- n ~ dist
+taxModel <- {
+  taxMetric %>%
+    eval()
+} ~ dist
+
 
 taxaResultsTbl <- spStatTbl %>%
   getStatsTbl(
@@ -315,7 +341,10 @@ taxaResultsTbl <- spStatTbl %>%
 
 #getTaxa2
 
[…]

### 2021-09-14 analysis/statsTbl.R

@@ -351,7 +351,7 @@ addStatEval2 <- function(
             )
         ),
       "isSignif2" =
-        pvals %>%
+        pval %>%
         modify(
           ~ if_else(
             .x <

### 2021-09-14 figs/fig2.R

@@ -4,6 +4,7 @@ source(
     "figs/style.R"
   )
 )
+#taxMetric
 source(
   here(
     "analysis/stats.R"
@@ -36,20 +37,33 @@ yAxisLabel2 <- quote(
 )
 
 measure3 <- quote(
-
+  Vochysia
 )
 yAxisLabel3 <- quote(
-
+  glue(
+    measure3,
+    " prevalence"
+  )
 )
 
 
-varForm <- n ~ dist
-varForm1 <- median ~ dist
+varForm <- {
+  mainMetric %>%
+    eval()
+} ~ dist
+#check.env...
+taxForm <- {
+  taxMetric %>%
+    eval()
+} ~ dist
 xAxisLabel <- quote(
   "Distance to secondary forest edge (m)"
 )
 statData12 <- plotResultsTbl1
-statData34 <- taxaResultsTbl1
+statData34 <- taxaResultsTbl1 %>%
+  rename(
+    "variable" = "gen"
+  )
 underData <- cleanData
 
 
@@ -59,8 +73,8 @@ underData <- cleanData
 graph1 <- statData12 %>%
   dotGraph(
     measure1,
-    varForm1[[3]],
-    varForm1[[2]],
+    varForm[[3]],
+    varForm[[2]],
     xAxisLabel,
     yAxisLabel1,
     ..addGroups = T
@@ -73,29 +87,29 @@ graph1
 graph2 <- statData12 %>%
   dotGraph(
     measure2,
-    varForm1[[3]],
-    varForm1[[2]],
+    varForm[[3]],
+    varForm[[2]],
     xAxisLabel,
     yAxisLabel2,
     ..addGroups = T
   )
+
+#addLine
 graph2
 
 
 #panel
 
-#HERE
-
 graph3 <- statData34 %>%
   dotGraph(
     measure3,
-    varForm[[3]],
-    varForm[[2]],
+    taxForm[[3]],
+    taxForm[[2]],
     xAxisLabel,
     yAxisLabel3,
-    ..addGroups = T,
-    ..y1 = varForm1[[2]]
-  )
+    ..addGroups = T
+  ) +
+  scale_y_log10()
 graph3
 
 
@@ -104,8 +118,8 @@ graph3
 graph4 <- statData34 %>%
   dotGraph(
     measure4,
-    varForm[[3]],
-    varForm[[2]],
+    taxForm[[3]],
+    taxForm[[2]],
     xAxisLabel,
     yAxisLabel4,
     ..addGroups = T

### 2021-09-14 figs/figStyle.R

@@ -1,71 +1,0 @@
-library(tidyverse)
-#geom_quasirandom()
-library(ggbeeswarm)
-library(here)
-source(
-  here(
-    "analysis/statsTbl.R"
-  )
-)
-
-
-style <- theme_bw() +
-  theme()
-
-theme_set(
-  style
-)
-
-graphPointColor <- "black"
-graphPointSize <- 3
-
-
-lineGraph <- function(
-  #graphData
-  #varData1
-  ..varData,
-  ..x,
-  ..y,
-  ..xlab,
-  ..ylab
-) {
-
-  ..varData %>%
-
-    ggplot(
-      aes(
-        x = {
-          ..x %>%
-            eval()
-        },
-        y = {
-          ..y %>%
-            eval()
-        }
-      )
-    ) +
-
-    geom_quasirandom(
-      color = "black",
-      shape = 21,
-      fill = "white",
-      size = 2
-    ) +
-
-    labs(
-      x = xlab,
-      y = ylab
-    ) +
-
-    geom_point(
-      data = {
-        graphData %>%
-          select(
-            "varData1"
-          ) %>%
-          unnest()
-      },
-      color = graphPointColor,
-      size = graphPointSize
-    )
-}

### 2021-09-14 figs/style.R

@@ -32,7 +32,6 @@ dotGraph <- function(
   ..xlab,
   ..ylab,
   ..addGroups = F,
-  ..y1 = ..y,
   ..cleanData = NULL
 ) {
 
@@ -119,6 +118,7 @@ dotGraph <- function(
             showP >
               0.001,
             round(
+              showP,
               3
             ),
             "< 0.001"
@@ -184,13 +184,7 @@ dotGraph <- function(
             unnest()
         },
         color = "black",
-        size = 3,
-        aes(
-          y = {
-            ..y1 %>%
-              eval()
-          }
-        )
+        size = 3
       )
 
   } else {

### 2021-10-21 analysis/statsTbl.R
dispersal modes, varify plus other props

@@ -201,17 +201,31 @@ addStatEval <- function(
         ),
       #OGdata
       varyTest =
-        varData %>%
-        modify(
-          ~ .x %>%
-            levene_test(
-              formula =
-                ....formula[[2]] %>%
-                eval() ~
-                ....formula[[3]] %>%
-                eval() %>%
-                as_factor()
+        ifelse(
+          #noLmerYet
+          {
+            ....formula[[3]] %>%
+              length()
+          } == 1,
+          {
+            varData %>%
+              modify(
+                ~ .x %>%
+                  levene_test(
+                    formula =
+                      ....formula[[2]] %>%
+                      eval() ~
+                      ....formula[[3]] %>%
+                      eval() %>%
+                      as_factor()
+                  )
+              )
+          },
+          list(
+            tibble(
+              "p" = NA
             )
+          )
         ),
       "isNormal" =
         normalTest %>%
@@ -223,15 +237,21 @@ addStatEval <- function(
             T, F
           )
         ),
+#     )
+# }
       "isHomosced" =
         varyTest %>%
-        modify(
+        modify_if(
+          !is.na(
+              varyTest
+          ),
           ~ if_else(
-            .x$
-              p >
-              0.055,
-            T, F
-          )
+              .x$
+                p >
+                0.055,
+              T, F
+          ),
+          ~ NA
         )
     ) %>%
     #list2vec
@@ -298,16 +318,30 @@ addStatEval <- function(
 
     mutate(
       "R2adj" =
-        statPrint %>%
-        modify(
-          ~ .x %>%
-            filter(
-              term !=
-                "(Intercept)"
-            ) %>%
-            pull(
-              adj.r.squared
+        #noLmerYet
+        ifelse(
+          {
+            ....formula[[3]] %>%
+              length()
+          } == 1,
+          {
+            statPrint %>%
+            modify(
+              ~ .x %>%
+                filter(
+                  term !=
+                    "(Intercept)"
+                ) %>%
+                pull(
+                  adj.r.squared
+                )
             )
+          },
+          list(
+            tibble(
+              "adj.r.squared" = NA
+            )
+          )
         )
     ) %>%
     unnest(
@@ -502,7 +536,7 @@ addStatEvalNon <- function(
     mutate(
       #morefamilies?
       "pickPval" =
-        if_else(
+        ifelse(
           pickAIC == poisAIC,
           poisPval,
           if_else(

### 2021-10-21 figs/fig2.R

@@ -46,6 +46,16 @@ yAxisLabel3 <- quote(
   )
 )
 
+measure4 <- quote(
+  Ficus
+)
+yAxisLabel4 <- quote(
+  glue(
+    measure4,
+    " prevalence"
+  )
+)
+
 
 varForm <- {
   mainMetric %>%
@@ -125,6 +135,9 @@ graph4 <- statData34 %>%
     ..addGroups = T
   )
 graph4
+#Ficus2few...
+
+#HERE
 
 
 #panel

### 2021-10-21 figs/style.R

@@ -13,6 +13,7 @@ library(gginnards)
 
 style <- theme_bw() +
   theme()
+#dark?
 
 theme_set(
   style

### 2021-11-05 analysis/attic/osa.Rmd
filtered NAs in joint cleanData

--- 2021-06-27 scripts/osa.Rmd
@@
 #loadpacks
-library(tidyverse)
-library(BIOMASS)
+
 library(rstatix)
@@
 theme_set(styleOsa)
-#import
-setwd("~/Documents/u/0/art/res/GitHub/dat/osa")
-dat <- read.csv("osa.csv")
-```
-
-```{r newTaxa2017}
-taxasp <- correctTaxo(genus = dat$gen, 
-                      species = dat$sp)  #notneeded
-
-taxa <- getTaxonomy(dat$gen)  #useful
-dat$fam <- taxa$family
-getWD <- getWoodDensity(genus = dat$gen, 
-                        species = dat$sp,
-                        family = dat$fam, 
-                        stand = dat$plot)
-dat$spg <- getWD$meanWD
-dat <- subset(dat, dist != "NA")#&gen!="-")#&sp!="sp.")
-dat$kg17 <- computeAGB(D = dat$dap, WD = dat$spg, 
-                       H = dat$h) * 1000
-
-# both methods do seem equal
-```
-
-```{r Chave2014fam, include=F}
-attach(dat)
-dat$kg14 <- 0.0673 * (spg * dap^2 * h)^0.976
-
-#updatefams
-for (name in dat$fam) {
-  #if (name == "Bombacaceae" | 
-        #name == "Sterculiaceae") {
-    #name <- "Malvaceae"
-  #}
-  #if (name == "Cecropiaceae") {
-    #name <- "Urticaceae"
-  #}
-}
-detach(dat)
+
 ```

### 2021-11-05 osa.tex
filtered NAs in joint cleanData

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Tropical wet secondary forest regeneration patterns},
            pdfauthor={Nicholas L Medina},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Tropical wet secondary forest regeneration patterns}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Nicholas L Medina}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{29/12/2018}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Primary tropical forests are idolized for their - role/importance of
secondary forests

\hypertarget{community-composition}{%
\subsection{\texorpdfstring{\emph{Community
composition}}{Community composition}}\label{community-composition}}

\begin{itemize}
\item
  factors affecting their aboveground biomass accumulation Aboveground
  biomass accumulation is likely affected by interspecific interactions,
  namely among N-fixers and non-N-fixers. While N-fixers are thought to
  facilitate the growth of other plants (CITATIONS FROM YOUNG AG),
  recent data suggest that competition, such as for P, may be more
  important for interspecific interactions than expected facilitation by
  N supply (Taylor et al 2017). This makes sense, given that P
  availability, rather than that of N, is more limiting in tropical
  ecosystems compared to temperate systems, especially given that they
  are estimated to have sufficient N to support decadal aboveground
  biomass accumulation (Brookshire et al 2019). Furthermore, with regard
  to nutrient dynamics, N-fixing trees in tropical forests are also
  expected to be major sources of N2O (Kou-Giesbrecht \& Menge 2019),
  further offsetting biomass balances.
\item
  \begin{itemize}
  \tightlist
  \item
    limiting nutrients likely fluctuating/oscillatory (Sullivan et al
    2014)
  \end{itemize}
\item
  \begin{itemize}
  \tightlist
  \item
    interactions and effects sizes are likely variable at species-level
    (Keller et al 2013; Gei \& Powers 2013)
  \end{itemize}
\item
  factors affecting their species coexistence
\item
  \begin{itemize}
  \tightlist
  \item
    relevance of stage structuring by specific gravity? (Nature EE 2019
    2ndFOR)
  \end{itemize}
\item
  \begin{itemize}
  \tightlist
  \item
    predicted distributions of species/wood densities
  \end{itemize}
\end{itemize}

\hypertarget{this-study}{%
\subsection{\texorpdfstring{\emph{This
study}}{This study}}\label{this-study}}

\begin{itemize}
\tightlist
\item
  census of \textasciitilde{}20yo secondary forest derived from an
  abandoned {[}common name for Bombacopsis{]} plantation. The base
  species is known to do well in dry forests, but not in the wet lowland
  forests like on the Osa Peninsula.
\end{itemize}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{study-site}{%
\subsection{\texorpdfstring{\emph{Study
site}}{Study site}}\label{study-site}}

This study was set in the lowland wet tropical rainforest of the Osa
Peninsula in SW Puntarenas, Costa Rica. The site is property of Osa
Conservation (osaconservation.org) located at the Greg Gund Conservation
Center. The censused parcel was 20 ha and undergoing secondary
succession for \textasciitilde{}20 years after abandonment as a wood
plantation using \emph{Bombacopsis quinata}, which unfortunately is
better suited to the dry forests of Guanacaste and the Nicoya Peninsula
(CITATION). Trees ≥10 cm in diameter were\ldots{}The study was done
during the rainy season months June - August 2013.

\hypertarget{census-design}{%
\subsection{\texorpdfstring{\emph{Census
design}}{Census design}}\label{census-design}}

Geographical information systems software (ArcGIS, WEBSITE, and QGIS,
{[}WEBSITE{]}) were used to create a polygon of the 20-ha parcel, which
was divided into six 50-m discrete bands (0-300m) that represented a
gradient of distance from the known adjacent primary forest (Fig S1).
This distance was made discrete as a simpler alternative to finding the
shortest linear distance between the plot and primary forest edge. The
number of plots set up in each band was coarsely scaled to the of area
each distance bance, such that the progression of plots contained in
each band from closest to furthest from primary forest was 10, 8, 5, 3,
and 1 respectively. Each plot was 21 \emph{x} 21m and set up June -
August 2013.

In each plot, every stem ≥10cm in {[}diameter/circumference?{]} was
measured for diameter at breast height (DBH, breast height =
\textasciitilde{}2.5 - 3.1m) and total height. Tree height was measured
by using a rangefinder (COMPANY/BRAND Bushnell?) to measure from an
observer to the crown (where it was visible), and then to the trunk at
breast height; these 2 distances were then used to triangulate and
calculate the tree's height using Pythagoras's Theorem. Following
typical forestry guidelines, in cases where a tree split into ≥2 stems
at breast height, each stem was measured separately; in cases where it
only initially split above breast height, it was measured as 1 stem.

\hypertarget{aboveground-biomass}{%
\subsection{\texorpdfstring{\emph{Aboveground
biomass}}{Aboveground biomass}}\label{aboveground-biomass}}

Aboveground biomass was estimated for each stem using the allometric
equation developed by Chave \emph{et al} (2014), which performed similar
to those in Chave \emph{et al} (2005) developed specifically for wet
tropical forests:

\emph{AGB = 0.0673 x (p D\^{}2 H)\^{}0.976}

with \emph{D} in cm, \emph{H} in m, and \emph{p} in g cm\^{}-3. Measured
DBHs and heights were used, along with wood densities ( \emph{p} )
exracted from the literature. In most cases, these values were only
available at the genus level, and unrepresented taxa were assumed to be
0.58 as stated by the World Agroforestry Database (CITE). References
used for this step are given in \emph{Table S1}.

\hypertarget{abiotic-plot-characteristics}{%
\subsection{\texorpdfstring{\emph{Abiotic plot
characteristics}}{Abiotic plot characteristics}}\label{abiotic-plot-characteristics}}

\hypertarget{canopy-cover}{%
\subsubsection{\texorpdfstring{\emph{Canopy
cover}}{Canopy cover}}\label{canopy-cover}}

Light reaching the forest floor was measured at the center of each plot,
at chest height, using a densiometer (BRAND). According to manufacturer
instructions, the final value used for the plot was an average of 4
different readings taken facing each of the 4 cardinal directions:
North, South, East, and West.

\hypertarget{topography}{%
\subsubsection{\texorpdfstring{\emph{Topography}}{Topography}}\label{topography}}

The slope of the forest floor in each plot was measured using {[}a
rangefinder by measuring the distance from one (specifc?) corner of the
plot to another (specific?) corner.{]}

\hypertarget{statistical-analysis}{%
\subsection{\texorpdfstring{\emph{Statistical
analysis}}{Statistical analysis}}\label{statistical-analysis}}

Analysis ideas from lit: - \ldots{}shade species may do better here
vs.~clear-cutting?

\hypertarget{census-structure}{%
\subsubsection{\texorpdfstring{\emph{Census
structure}}{Census structure}}\label{census-structure}}

\begin{verbatim}
##             Df Sum Sq Mean Sq F value Pr(>F)  
## dist         1    0.1    0.12   0.004 0.9500  
## plot         1  142.5  142.47   4.753 0.0428 *
## dist:plot    1    6.8    6.81   0.227 0.6393  
## Residuals   18  539.6   29.98                 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\emph{Table S\#.} ANOVA results for \emph{B. quinata} plot stem density
by distance from primary forest edge.

\begin{verbatim}
##             Df Sum Sq Mean Sq F value Pr(>F)  
## dist         1   22.6    22.6   0.329  0.571  
## plot         1  331.4   331.4   4.817  0.037 *
## Residuals   27 1857.4    68.8                 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\begin{verbatim}
##             Df Sum Sq Mean Sq F value Pr(>F)
## dist         1    0.0    0.02   0.001  0.978
## plot         1    0.0    0.00   0.000  0.994
## Residuals   27  859.9   31.85
\end{verbatim}

\begin{verbatim}
##             Df Sum Sq Mean Sq F value Pr(>F)
## dist         1    365   365.5   1.280  0.268
## plot         1    439   439.1   1.537  0.226
## Residuals   27   7712   285.6
\end{verbatim}

\emph{Table S\#.} ANOVA results for environmental variation in (A) light
and (B) terrain slope by distance from primary forest edge.

\hypertarget{canopy-structure}{%
\subsubsection{\texorpdfstring{\emph{Canopy
structure}}{Canopy structure}}\label{canopy-structure}}

\begin{itemize}
\tightlist
\item
  height and {[}mean{]} SPG (Díaz et al 2016) by dist, or PCoA
\end{itemize}

\begin{verbatim}
##           Df  Pillai approx F num Df den Df   Pr(>F)   
## dist       1 0.40709   8.5824      2     25 0.001453 **
## plot       1 0.22378   3.6036      2     25 0.042153 * 
## n          1 0.00828   0.1044      2     25 0.901292   
## Residuals 26                                           
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\emph{Table S\#}. MANOVA results showing effects of distance from
primary forest edge and plot ID on mean plot height and wood density.

\begin{verbatim}
##              Df Sum Sq Mean Sq F value   Pr(>F)    
## dist          1   85.4   85.42  14.549 0.000171 ***
## plot          1   19.7   19.66   3.349 0.068395 .  
## dist:plot     1   23.8   23.76   4.047 0.045278 *  
## Residuals   261 1532.5    5.87                     
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\begin{verbatim}
##             Df Sum Sq Mean Sq F value Pr(>F)  
## dist         1   9.34   9.338   4.375 0.0509 .
## plot         1   0.31   0.307   0.144 0.7090  
## n            1   2.73   2.725   1.277 0.2733  
## Residuals   18  38.42   2.134                 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\emph{Table S\#.} ANOVA results at (A) tree and (B) plot levels for
\emph{B. quinata} height by distance from primary forest edge.

\begin{verbatim}
## 
## Call:
## lm(formula = kg14 ~ n + N + W + luz + incl, data = datp)
## 
## Residuals:
##    Min     1Q Median     3Q    Max 
## -82.00 -32.71  -1.70  16.90 128.52 
## 
## Coefficients:
##               Estimate Std. Error t value Pr(>|t|)  
## (Intercept) -2.934e+05  7.794e+05  -0.376   0.7099  
## n           -1.803e+00  9.241e-01  -1.951   0.0629 .
## N            2.144e+04  1.306e+04   1.643   0.1135  
## W            1.359e+03  9.081e+03   0.150   0.8823  
## luz          2.692e+00  2.467e+00   1.091   0.2861  
## incl         7.389e-01  6.476e-01   1.141   0.2651  
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 55.98 on 24 degrees of freedom
## Multiple R-squared:  0.2004, Adjusted R-squared:  0.03387 
## F-statistic: 1.203 on 5 and 24 DF,  p-value: 0.3375
\end{verbatim}

\begin{verbatim}
## 
## Call:
## lm(formula = kg14 ~ n + N + W + luz + incl, data = datpBno)
## 
## Residuals:
##    Min     1Q Median     3Q    Max 
## -91.13 -44.18 -11.50  23.32 156.35 
## 
## Coefficients:
##               Estimate Std. Error t value Pr(>|t|)  
## (Intercept) -1.149e+06  9.813e+05  -1.171   0.2532  
## n           -3.177e+00  1.639e+00  -1.939   0.0644 .
## N            2.986e+04  1.591e+04   1.877   0.0727 .
## W            1.078e+04  1.143e+04   0.943   0.3552  
## luz          5.138e+00  3.113e+00   1.651   0.1118  
## incl         1.101e+00  8.582e-01   1.283   0.2119  
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 70.37 on 24 degrees of freedom
## Multiple R-squared:  0.2459, Adjusted R-squared:  0.08883 
## F-statistic: 1.565 on 5 and 24 DF,  p-value: 0.2077
\end{verbatim}

\begin{verbatim}
## 
## Call:
## lm(formula = kg14 ~ n + N + W + luz + incl, data = datpB)
## 
## Residuals:
##     Min      1Q  Median      3Q     Max 
## -68.473 -22.095  -0.378  18.756 121.415 
## 
## Coefficients:
##               Estimate Std. Error t value Pr(>|t|)   
## (Intercept)  2.490e+06  7.232e+05   3.443  0.00334 **
## n           -2.528e-01  1.867e+00  -0.135  0.89397   
## N           -1.247e+04  1.405e+04  -0.888  0.38788   
## W           -2.863e+04  8.329e+03  -3.437  0.00338 **
## luz         -4.005e+00  2.642e+00  -1.516  0.14906   
## incl         4.903e-01  6.922e-01   0.708  0.48894   
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 48.08 on 16 degrees of freedom
## Multiple R-squared:  0.4372, Adjusted R-squared:  0.2613 
## F-statistic: 2.486 on 5 and 16 DF,  p-value: 0.07555
\end{verbatim}

\begin{verbatim}
## Response 1 :
## 
## Call:
## lm(formula = `1` ~ dist + plot, data = dat)
## 
## Residuals:
##      Min       1Q   Median       3Q      Max 
## -0.00973 -0.00752 -0.00576 -0.00015  0.56404 
## 
## Coefficients:
##              Estimate Std. Error t value Pr(>|t|)
## (Intercept) -0.001544   0.002501  -0.617    0.537
## dist        -0.001298   0.003857  -0.337    0.737
## plot         0.001307   0.003862   0.338    0.735
## 
## Residual standard error: 0.0284 on 1239 degrees of freedom
## Multiple R-squared:  0.0004468,  Adjusted R-squared:  -0.001167 
## F-statistic: 0.2769 on 2 and 1239 DF,  p-value: 0.7582
## 
## 
## Response 2 :
## 
## Call:
## lm(formula = `2` ~ dist + plot, data = dat)
## 
## Residuals:
##      Min       1Q   Median       3Q      Max 
## -0.21901 -0.00303  0.00765  0.01152  0.40945 
## 
## Coefficients:
##              Estimate Std. Error t value Pr(>|t|)  
## (Intercept) -0.001857   0.002496  -0.744   0.4571  
## dist        -0.007380   0.003849  -1.917   0.0554 .
## plot         0.007375   0.003854   1.913   0.0559 .
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 0.02834 on 1239 degrees of freedom
## Multiple R-squared:  0.004565,   Adjusted R-squared:  0.002958 
## F-statistic: 2.841 on 2 and 1239 DF,  p-value: 0.05874
\end{verbatim}

\begin{verbatim}
## Response 1 :
## 
## Call:
## lm(formula = `1` ~ dist + plot, data = datp)
## 
## Residuals:
##      Min       1Q   Median       3Q      Max 
## -0.22486 -0.13069 -0.03585  0.03691  0.54840 
## 
## Coefficients:
##             Estimate Std. Error t value Pr(>|t|)
## (Intercept) -0.08527    0.11517  -0.740    0.465
## dist        -0.07817    0.16403  -0.477    0.638
## plot         0.07863    0.16429   0.479    0.636
## 
## Residual standard error: 0.19 on 27 degrees of freedom
## Multiple R-squared:  0.02533,    Adjusted R-squared:  -0.04687 
## F-statistic: 0.3508 on 2 and 27 DF,  p-value: 0.7073
## 
## 
## Response 2 :
## 
## Call:
## lm(formula = `2` ~ dist + plot, data = datp)
## 
## Residuals:
##      Min       1Q   Median       3Q      Max 
## -0.29484 -0.14106 -0.01064  0.08475  0.39195 
## 
## Coefficients:
##             Estimate Std. Error t value Pr(>|t|)  
## (Intercept) -0.09838    0.10500  -0.937   0.3571  
## dist        -0.31294    0.14954  -2.093   0.0459 *
## plot         0.31280    0.14977   2.088   0.0463 *
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 0.1732 on 27 degrees of freedom
## Multiple R-squared:  0.1899, Adjusted R-squared:  0.1299 
## F-statistic: 3.165 on 2 and 27 DF,  p-value: 0.05823
\end{verbatim}

\begin{verbatim}
##              Df  Sum Sq Mean Sq F value   Pr(>F)    
## dist          1  354161  354161  42.242 4.05e-10 ***
## plot          1    1990    1990   0.237    0.627    
## Residuals   262 2196631    8384                     
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\begin{verbatim}
##              Df    Sum Sq Mean Sq F value Pr(>F)
## dist          1      1012    1012   0.006  0.938
## plot          1     32528   32528   0.196  0.658
## Residuals   974 161488210  165799
\end{verbatim}

\emph{Table S\#}. (A-D) Linear model and (E, F) ANOVA results for (A, D)
total mean, (B, E) \emph{B. quinata}, and (C, F) other taxon biomass
with respect to (A-C) stem density and (D-F) distance from primary
forest edge and plot ID.

\begin{verbatim}
## 
## Call:
## lm(formula = n ~ luz + incl + N + W, data = datp)
## 
## Residuals:
##     Min      1Q  Median      3Q     Max 
## -14.809  -8.642  -2.051   5.632  27.496 
## 
## Coefficients:
##               Estimate Std. Error t value Pr(>|t|)
## (Intercept) -3.341e+04  1.686e+05  -0.198    0.844
## luz          2.693e-02  5.339e-01   0.050    0.960
## incl         1.302e-01  1.377e-01   0.945    0.354
## N            3.941e+03  2.713e+03   1.453    0.159
## W            3.618e+00  1.965e+03   0.002    0.999
## 
## Residual standard error: 12.12 on 25 degrees of freedom
## Multiple R-squared:  0.1531, Adjusted R-squared:  0.01756 
## F-statistic:  1.13 on 4 and 25 DF,  p-value: 0.3652
\end{verbatim}

\begin{verbatim}
##             Df Sum Sq Mean Sq F value Pr(>F)  
## dist         1     54    54.3   0.387 0.5393  
## plot         1    483   483.5   3.440 0.0746 .
## Residuals   27   3795   140.6                 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\begin{verbatim}
##                        Df Sum Sq Mean Sq F value Pr(>F)  
## poly(dist, degree = 2)  2    722   360.8   2.954 0.0698 .
## plot                    1    437   436.5   3.574 0.0699 .
## Residuals              26   3175   122.1                 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\begin{verbatim}
##             Df Sum Sq Mean Sq F value Pr(>F)  
## dist         1    0.1    0.12   0.004 0.9489  
## plot         1  142.5  142.47   4.954 0.0383 *
## Residuals   19  546.4   28.76                 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\begin{verbatim}
##             Df Sum Sq Mean Sq F value Pr(>F)  
## dist         1   22.6    22.6   0.329  0.571  
## plot         1  331.4   331.4   4.817  0.037 *
## Residuals   27 1857.4    68.8                 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\emph{Table S\#}. Linear model and ANOVA results for stem density with
respect to distance from primary forest edge and plot ID.

\hypertarget{community-structure}{%
\subsubsection{\texorpdfstring{\emph{Community
structure}}{Community structure}}\label{community-structure}}

\begin{itemize}
\tightlist
\item
  diversity predicted by height and/or biomass (Pan et al 2018)
\end{itemize}

\begin{verbatim}
## 
## Call:
## lm(formula = r ~ h + kg14 + n, data = datp)
## 
## Residuals:
##    Min     1Q Median     3Q    Max 
## -3.363 -1.378 -0.082  1.567  4.169 
## 
## Coefficients:
##              Estimate Std. Error t value Pr(>|t|)    
## (Intercept) 17.546249   3.042766   5.767 4.51e-06 ***
## h           -0.749553   0.265174  -2.827  0.00893 ** 
## kg14         0.008226   0.007414   1.109  0.27741    
## n           -0.037015   0.033893  -1.092  0.28480    
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 2.139 on 26 degrees of freedom
## Multiple R-squared:  0.2952, Adjusted R-squared:  0.2139 
## F-statistic:  3.63 on 3 and 26 DF,  p-value: 0.026
\end{verbatim}

\emph{Table S\#}. Linear model results showing plot genus richness
explained by mean plot height.

\begin{itemize}
\tightlist
\item
  Other (non-richness) diversity measures, by distance to edge?
\end{itemize}

\hypertarget{results}{%
\section{Results}\label{results}}

``Large trees increase biomass by reducing richness {[}density{]} in a
regenerating tropical timber plantation''.

\hypertarget{census-structure-1}{%
\subsection{Census structure}\label{census-structure-1}}

Mean total plot stem density was explained by distance from primary
forest edge and plot. The relationship with distance from primary forest
edge was positively quadratic.

\includegraphics{osa_files/figure-latex/mapNdistPlot-1.pdf}

\begin{verbatim}
## Warning: Removed 1 rows containing missing values (geom_errorbar).
\end{verbatim}

\includegraphics{osa_files/figure-latex/mapNdistPlot-2.pdf}

\emph{Fig \#\#}. (A) Map showing total stem density and GPS coordinates
of census plots (21 x 21 m), and (B) plot showing stem density by
distance from primary forest edge.

Mean plot height was explained by mean plot specific wood density and
distance from primary forest edge. These were negative relationships,
which were further explained by a positive relationship between distance
from primary forest edge and mean plot specific wood density.

\hypertarget{canopy-structure-1}{%
\subsection{Canopy structure}\label{canopy-structure-1}}

\includegraphics{osa_files/figure-latex/HspgPlot-1.pdf}

\emph{Fig \#\#}. Plot showing relationship betwen mean plot (black,
individual trees in gray) tree heights and specific wood density.

\begin{verbatim}
## Warning: Removed 1 rows containing missing values (geom_errorbar).
\end{verbatim}

\includegraphics{osa_files/figure-latex/HspgDistPlot-1.pdf}

\begin{verbatim}
## Warning: Removed 1 rows containing missing values (geom_errorbar).
\end{verbatim}

\includegraphics{osa_files/figure-latex/HspgDistPlot-2.pdf}

\emph{Fig \#\#}. Plots showing the relationship between mean plot
(black, individual trees shown in gray) (A) heights and (B) specific
wood densities.

\hypertarget{biomass-structure}{%
\subsection{Biomass structure}\label{biomass-structure}}

Mean plot biomass was explained by distance from primary forest edge and
mean total plot stem density. The trend with edge distance was
negatively quadratic, while that with mean total plot stem density was
linearly negative. These relationships correspond with that between edge
distance and mean plot stem density, which was instead positively
quadratic (previous figure).

\emph{Taken together, these data show that mean total plot biomass was
lower closer to the primary forest, due to increases in stem density,
primarily with trees with low wood density and presumably faster growth
and reproduction rates (CITE?).}

Finally, there was an effect of East-West location on mean plot biomass,
but despite observing a downward slope on the southwest corner of the
parcel, this general East-West co-variation with slope did not
statistically explain mean plot biomass variation.

\includegraphics{osa_files/figure-latex/kgnPlot-1.pdf}

\emph{Fig \#\#}. Plot showing relationship between mean plot biomass and
mean plot stem density.

\begin{verbatim}
## Warning: Removed 1 rows containing missing values (geom_errorbar).
\end{verbatim}

\includegraphics{osa_files/figure-latex/kgDistPlot-1.pdf}

\emph{Fig \#\#}. Biomass (plot mean in black, individual trees in gray,
\emph{B. quinata} in orange) changes at various distance from primary
forest edge.

\hypertarget{community-structure-1}{%
\subsubsection{\texorpdfstring{\emph{Community
structure}}{Community structure}}\label{community-structure-1}}

\begin{itemize}
\tightlist
\item
  decide which plot represents anticipated stats\ldots{}(prob PCoA)
\end{itemize}

\includegraphics{osa_files/figure-latex/genDistPlot-1.pdf}

\emph{Fig \#\#}. Plot showing biomass partitioned by genus as a function
of distance from primary forest edge.

Plot richness was predicted by mean plot height, which supports previous
findings (Pan et al 2018).

\includegraphics{osa_files/figure-latex/ShDistPlot-1.pdf}

\begin{verbatim}
## Warning: Removed 1 rows containing missing values (geom_errorbar).
\end{verbatim}

\includegraphics{osa_files/figure-latex/ShDistPlot-2.pdf}

\emph{Fig \#\#}. Plot showing relationship between richness and (A) mean
plot tree height and (B) distance from primary forest edge.

\hypertarget{other-ongoing-analyses}{%
\subsection{\texorpdfstring{\emph{Other ongoing
analyses\ldots{}}}{Other ongoing analyses\ldots{}}}\label{other-ongoing-analyses}}

\begin{verbatim}
## `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
\end{verbatim}

\includegraphics{osa_files/figure-latex/kgDistHist-1.pdf}

\begin{verbatim}
## <ggproto object: Class FacetWrap, Facet, gg>
##     compute_layout: function
##     draw_back: function
##     draw_front: function
##     draw_labels: function
##     draw_panels: function
##     finish_data: function
##     init_scales: function
##     map_data: function
##     params: list
##     setup_data: function
##     setup_params: function
##     shrink: TRUE
##     train_scales: function
##     vars: function
##     super:  <ggproto object: Class FacetWrap, Facet, gg>
\end{verbatim}

\begin{verbatim}
## `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
\end{verbatim}

\includegraphics{osa_files/figure-latex/kgDistHist-2.pdf}

\begin{verbatim}
## `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
\end{verbatim}

\includegraphics{osa_files/figure-latex/kgDistHist-3.pdf}

\begin{verbatim}
## `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
\end{verbatim}

\includegraphics{osa_files/figure-latex/hDistHist-1.pdf}

\begin{verbatim}
## `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
\end{verbatim}

\includegraphics{osa_files/figure-latex/hDistHist-2.pdf}

\begin{verbatim}
## `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
\end{verbatim}

\includegraphics{osa_files/figure-latex/hDistHist-3.pdf}

\emph{Figs \#}. Histograms showing plot (A) biomass and (B) height
distributions by distance from primary forest edge.

\begin{itemize}
\tightlist
\item
  forest biomass is complex (Pan et al 2013)
\item
  but biomass distributions here are log-normal?!
\item
  check for \emph{power-law fits} of height and biomass histograms among
  edge plots
\end{itemize}

\hypertarget{codeplot-attic-dont-read}{%
\section{CODE/PLOT ATTIC, DONT READ}\label{codeplot-attic-dont-read}}

\#\#Fig1.Biomass,distance,genus

Fig 1a. Intermediate distances (150m from edge) tended (p=0.26) to show
higher biomass.

(Outliers seem to be relevant for variance at intermediate
distances\ldots{})

Lonchocarpus individual is huge. Remove?

\#\#FigA1.Vstem,distance

\#\#Bombacopsis biomass by distance

Figs 1b and c. (b) Biomass of plantation species significantly increased
with distance from undisturbed forest edge (p\textless{}0.05, R2=0.70),
with (b) no change in the number of \emph{Bombacopsis} stems
(p\textgreater{}0.05).

Together, Figs 1a and 1b show that biomass was higher in areas where the
plantation species \emph{Bombacopsis} stems grew smaller on average.

(This may suggest either competition or facilitation--check biomass
ditribution of major contributing genera\ldots{})

Fig 2a. Rank-biomass curves (by genus) significantly (but weakly) fit a
negative exponential (p=0.01, R2=0.05).

From Fig 1, highest biomass was in 150m, which this figure shows mostly
comes from 1 \emph{Lonchocarpus}, 1 \emph{Tabebuia}, 1 {[}/3{]}
\emph{Ficus}, 2 {[}/10{]} \emph{Hyeronima}, 1 {[}/2{]} \emph{Apeiba}, 4
{[}/23{]} \emph{Vochysia}, and 8 {[}/16{]} \emph{Bombacopsis} trees
(ranked sequentially). Excluding individual outlier tree sizes, it could
be

\#\#Vochysia biomass by distance

Fig 2b and c. {[}\ldots{}{]} Actually, it seems like loss of biomass may
not due to competition between \emph{Bombacopsis} and any single taxon,
given that the trends of lower richness with higher biomass observed
here seem to be due to singletons.

**Does this suggest a kind of portfolio effect, but at the level of
productive individuals (vs.~species), and in the opposite direction due
to competition?

This could be due to a lower chance of sampling species in an area with
a few big trees. -- What does the stem number distribution look like by
distance and/or genus?

Seems like it is true, that fewer individuals of rare species occurred
in plots that also had the most massive (and the least massive)
individual trees.

Fig \#a. Frequency distribution of biomass tends to appear log-normal.

DISCUSSION - explain hump b/c proximity on road side + -- lower richness
b/c earlier-successional (20y) - Vochysia is wind-dispersed; is avg
200m?

\hypertarget{community-composition-1}{%
\subsubsection{Community composition}\label{community-composition-1}}

\hypertarget{shannon-diversity}{%
\paragraph{Shannon diversity}\label{shannon-diversity}}

Fig 3. Richness-by-distance relationship significantly fits a negative
quadratic model (p\textless{}0.01, R2=0.22).

(Which {[}rare{]} species were lost?)

Together, Figs 1 and 3 show that higher biomass tended to occur in plots
with lower taxonomic richness.

Fig 4. Biomass-richness relationship marginally significantly fits a
straight line on a log-log plot (p=0.066, R2=0.63).

\#Conclusions Higher biomass tended to occur in areas that also showed
lower taxonomic richness at \textgreater{}10 years after abandonment of
the \emph{Bombacopsis} tree plantation.

\#\#Attic \#\#\#Vochysia

\end{document}

### 2021-11-05 analysis/disfit.R

@@ -1,79 +1,0 @@
-library(tidyverse)
-library(poweRlaw)
-
-
-disfit <- function(freqData) {
-  #input=freqvector
-  #Clauset2009,Gillespie2015
-
-  Pl <- displ$new(round(freqData))
-  Pl$setPars(
-    estimate_pars(Pl))
-  #variance
-  PlVar <-
-    bootstrap(Pl)$bootstraps$pars %>%
-    var()
-  #plot(Pl)
-
-  PlP <- bootstrap_p(Pl)$p  #>0.1passes
-
-  Exp <- disexp$new(round(freqData))
-  Exp$setPars(
-    estimate_pars(Exp))
-  ExpVar <-
-    bootstrap(Exp)$bootstraps$pars %>%
-    var()
-
-  Pois <- dispois$new(round(freqData))
-  Pois$setPars(
-    estimate_pars(Pois))
-  PoisVar <-
-    bootstrap(Pois)$bootstraps$pars %>%
-    var()
-
-  Lognorm <- dislnorm$new(round(freqData))
-  Lognorm$setPars(
-    estimate_pars(Lognorm))
-  LognormVar <-
-    bootstrap(Lognorm)$bootstraps
-  LognormVar1 <- var(LognormVar$pars1)
-  LognormVar2 <- var(LognormVar$pars2,
-                     na.rm = T)
-
-
-  #null=bothOK #1sided=arg1==arg2
-  PlExpP <-
-    compare_distributions(Pl,
-                          Exp)$p_one_sided
-  #<0.05=arg1better
-  PlLognormP <-
-    compare_distributions(Pl,
-                          Lognorm)$p_one_sided
-  PlPoisP <-
-    compare_distributions(Pl,
-                          Pois)$p_one_sided
-
-
-  results <- tibble(
-    "PlP" = PlP,
-    "PlExpP" = PlExpP, #pvals
-    "PlLognormP" = PlLognormP,
-    "PlPoisP" = PlPoisP,
-
-    #slopes/pars
-    "PlPar" = Pl$pars,
-    "ExpPar" = Exp$pars,
-    "PoisPar" = Pois$pars,
-    "LognormPar1" = Lognorm$pars[1],
-    "LognormPar2" = Lognorm$pars[2],
-
-    #variances
-    "PlVar" = PlVar,
-    "ExpVar" = ExpVar,
-    "PoisVar" = PoisVar,
-    "LognormVar1" = LognormVar1,
[…]

### 2021-11-05 analysis/stats.R

@@ -12,6 +12,8 @@ library(tidyverse)
 library(rstatix);
 library(vegan) #diversity()
 library(oir)
+library(ggbeeswarm)
+library(ggpmisc)
 
 
 

### 2021-11-05 analysis/statsTbl.R

@@ -1,859 +1,0 @@
-library(here);
-# source(
-#   here("analysis/stats.R")
-# ) #...$varData
-library(tidymodels) #glance()
-library(ggbeeswarm) #quasirandom()
-library(ggpmisc) #stat_poly_eq()
-library(ggpubr) #ggarrange()
-# source(
-#   here("analysis/disfit.R")
-# ) #disfit()
-
-
-#main
-
-getStatsTbl <- function(
-  #has"varData"col
-  ...nestedVarDataTbl,
-  #user's
-  ...formula
-) {
-
-  ...nestedVarDataTbl %>%
-
-    statFitTbl(
-      ...formula
-    ) %>%
-    addStatEval(
-      ...formula
-    ) %>%
-
-    #nonnormaldistr
-    addStatFitNon(
-      ...formula
-    ) %>%
-    addStatEvalNon() %>%
-
-    addGraph(
-      ...formula
-    )
-}
-
-
-
-getStatsTbl2 <- function(
-  ...nestedVarDataTbl,
-  #user's
-  ...formula2
-) {
-
-  ...nestedVarDataTbl %>%
-    statFitTbl(
-      ...formula2
-    ) %>%
-    addStatEval2(
-      ...formula2
-    )
-}
-
-
-
-getStatsTbl1 <- function(
-  #has"varData1"col
-  ...nestedVarDataTbl,
-  #user's
-  ...formula
-) {
-
-  ...nestedVarDataTbl %>%
-
-    statFitTbl1(
-      ...formula
-    ) %>%
-    addStatEval(
-      ...formula
-    ) %>%
-
-    addStatFitNon(
-      ...formula
-    ) %>%
-    addStatEvalNon() %>%
-
-    addGraph1(
-      ...formula
-    )
-}
-
-
-
-getStatsTbl12 <- function(
-  #has"varData1"col
-  ...nestedVarDataTbl,
-  #user's
-  ...formula
-) {
-
-  ...nestedVarDataTbl %>%
-
-    statFitTbl1(
-      ...formula
-    ) %>%
-    addStatEval2(
-      ...formula
-    )
-}
-
-
-
-#sub-funcs
-
-statFitTbl <- function(
-  ...nestedVarDataTbl,
-  ....formula
-) {
-
-  ...nestedVarDataTbl %>%
-
-    mutate(
-      "statTest" =
-        varData %>%
-        modify(
-          ~ .x %>%
-            lm(
-              formula =
-                ....formula
[…]

### 2021-11-05 data/clean.R

@@ -22,6 +22,11 @@ dispersal <- vroom(
     gen,
     sp,
     mainDispersal
+  ) %>%
+  filter(
+    !is.na(
+      gen
+    )
   )
 
 
@@ -78,5 +83,10 @@ cleanData <- rawData %>%
   cleanData() %>%
   full_join(
     dispersal
+  ) %>%
+  filter(
+    !is.na(
+      dist
+    )
   )
 

### 2021-11-05 figs/fig1.R

@@ -1,8 +1,6 @@
 library(here)
-source(
-  here(
-  "figs/style.R"
-  )
+i_am(
+  "figs/fig1.R"
 )
 source(
   here(
@@ -12,6 +10,8 @@ source(
 #plotResultsTbl
 library(glue)
 #annotate()
+
+library(oir)
 
 
 #userInput

### 2021-11-05 figs/fig2.R

@@ -1,8 +1,6 @@
 library(here)
-source(
-  here(
-    "figs/style.R"
-  )
+i_am(
+  "figs/fig2.R"
 )
 #taxMetric
 source(
@@ -13,6 +11,8 @@ source(
 #plotResultsTbl
 library(glue)
 #annotate()
+
+library(oir)
 
 
 #plan:

### 2021-11-05 figs/style.R

@@ -1,200 +1,0 @@
-library(tidyverse)
-#geom_quasirandom()
-library(ggbeeswarm)
-library(here)
-# source(
-#   here(
-#     "analysis/statsTbl.R"
-#   )
-# )
-library(gginnards)
-#append_layers()
-library(glue)
-
-
-style <- theme_bw() +
-  theme()
-#dark?
-
-theme_set(
-  style
-)
-
-
-dotGraph <- function(
-  ..varData,
-  # $variable
-  # $varData
-  # $varData1
-  # $pval
-  # $pickPval
-  ..var,
-  ..x,
-  ..y,
-  ..xlab,
-  ..ylab,
-  ..addGroups = F,
-  ..cleanData = NULL
-) {
-
-  graph <- ..varData %>%
-
-    filter(
-      variable == ..var
-    ) %>%
-    select(
-      "varData"
-    ) %>%
-    unnest(
-      everything()
-    ) %>%
-
-
-    ggplot(
-      aes(
-        x = {
-          ..x %>%
-            eval()
-        },
-        y = {
-          ..y %>%
-            eval()
-        }
-      )
-    ) +
-
-    geom_quasirandom(
-      color = "black",
-      shape = 21,
-      fill = "white",
-      size = 2
-    ) +
-
-    labs(
-      x = {
-        ..xlab %>%
-          eval()
-        },
-      y = {
-        ..ylab %>%
-          eval()
-      }
-    ) +
-
-    annotate(
-      "text",
-      label = glue(
-        "P = ",
-        {
-          varResult <- ..varData %>%
-            filter(
-              variable == ..var
-            )
-
-          checkResult <- varResult %>%
-            pull(
-              isModelOK
-            )
-
-          #if_else()2strict
-          showP <- ifelse(
-            checkResult &
-              !is.na(
-                checkResult
-              ),
-            {
-              varResult %>%
-              pull(
-                "pval"
-              )
-            },
-            {
-              varResult %>%
-              pull(
-                "pickPval"
-              )
-            }
-          ) %>%
-            last() %>%
-            as.double()
-
-          ifelse(
-            showP >
-              0.001,
-            round(
-              showP,
-              3
-            ),
-            "< 0.001"
-          )
-        }
-      ),
-      x = 1,
-      y = 1,
-      hjust = 0
-    )
-
-
-  #toggles
-
-
-  graph <- if(
-    is.null(
-      ..cleanData
-    )
[…]

### 2022-01-18 talk/talk.Rmd
save fig1

---
title: "R Notebook"
output: html_notebook
---

```{r setup, echo=F}
knitr::opts_chunk$set(
  echo=F, quiet=T)
library(
  here
)
i_am(
  "talk/talk.Rpres"
)
#fixRender
library(
  httpuv
)
```

Outline
===

- Introduction
- Methods
- Results
- Discussion

Introduction: Tropical reforestation
===

- Over half are secondary, globally

- Managed recovery can help support local communities

- To make the most of ongoing natural processes, ecological dynamics should be better understood

Introduction: Community patterns
===

- Species richness recovers quickly

- But community composition may take centuries

- So balance of ecosystem service recovery unclear

Introduction: This study
===

- Question:

- Hypothesis:

- Predictions:

Methods: Study site
===

- Wet rainforest, Osa Peninsula, Costa Rica

- 20 ha. timber plantation, became NRDC preserve post-abandonment ~2003

- Surrounded by unmanaged/primary forest on 3 sides, road on other

```{r site}
# nature_img.png
```

Methods: Study design
===

- 300 m transect from unmanaged forest edge inward

- Divided into 50 m bands using Q-, ArcGIS

- 30 plots randomly assigned, but weighted by band area

![](map.png)

Methods: Census
===

- DBH, height stems ≥10 cm using rangefinder

- ID by local field guide personnel and books

- plot canopy closure w/ densiometer, and ground slope

# Results: Forest traits along edge

```{r fig1}
source(
  here(
    "figs/fig1.R"
  )
)
fig1
```

Results: Composition along edge
===

```{r fig2}
source(
  here(
    "figs/fig2.R"
  )
)
fig2
```

Results: Dispersal modes along edge
===

```{r fig3}
source(
  here(
    "figs/fig3.R"
  )
)
fig3
```

Discussion
===

<!-- For more details on authoring R presentations please visit <https://support.rstudio.com/hc/en-us/articles/200486468>. -->

### 2022-01-18 analysis/stats.R

@@ -17,7 +17,7 @@ library(ggpmisc)
 
 
 
-#userInput--parentFile
+# #userInput--parentFile
 # taxRank <- quote(
 #   gen
 # )
@@ -393,7 +393,7 @@ mainModel0 <- {
 
 
 dispResultsTbl0 <- spStatTbl %>%
-  oir::statFitTbl(
+  oir::getStatsTbl(
     mainModel0
   )
 

### 2022-01-18 figs/fig1.R

@@ -57,7 +57,7 @@ yAxisLabel1d <- quote(
 
 varForm <- median ~ dist
 xAxisLabel <- quote(
-  "Distance to secondary forest edge (m)"
+  "Distance to edge (m)"
 )
 statData <- plotResultsTbl1
 underData <- cleanData
@@ -69,13 +69,14 @@ underData <- cleanData
 
 graph1a <-  statData %>%
   dotGraph(
-    measure1,
+    measure1a,
     varForm[[3]],
     varForm[[2]],
     xAxisLabel,
-    yAxisLabel1,
+    yAxisLabel1a,
     ..addBins = T,
-    ..cleanData = underData
+    ..cleanData = underData,
+    ..addCenters = T
   ) +
   scale_y_continuous(
     trans = "log10"
@@ -88,13 +89,15 @@ graph1a
 
 graph1b <- statData %>%
   dotGraph(
-    measure2,
+    measure1b,
     varForm[[3]],
     varForm[[2]],
     xAxisLabel,
-    yAxisLabel2,
+    yAxisLabel1b,
     ..addBins = T,
-    ..cleanData = underData
+    ..cleanData = underData,
+    ..addCenters = T,
+    ..addLines = T
   )
 graph1b
 
@@ -104,13 +107,14 @@ graph1b
 
 graph1c <- statData %>%
   dotGraph(
-    measure3,
+    measure1c,
     varForm[[3]],
     varForm[[2]],
     xAxisLabel,
-    yAxisLabel3,
+    yAxisLabel1c,
     ..addBins = T,
-    ..cleanData = underData
+    ..cleanData = underData,
+    ..addCenters = T
   )
 graph1c
 
@@ -120,12 +124,13 @@ graph1c
 
 graph1d <- statData %>%
   dotGraph(
-    measure4,
+    measure1d,
     varForm[[3]],
     varForm[[2]],
     xAxisLabel,
-    yAxisLabel4,
-    ..addBins = T
+    yAxisLabel1d,
+    ..addBins = T,
+    ..addCenters = T
   )
 graph1d
 
@@ -137,7 +142,24 @@ fig1 <- ggarrange(
   graph1c,
   graph1d,
   nrow = 2,
-  ncol = 2
+  ncol = 2,
+  labels = c(
+    "a",
+    "b",
+    "c",
+    "d"
+  )
 )
 
+
 fig1
+
+
+ggsave(
+  "fig1.pdf",
+  fig1,
[…]

### 2022-01-18 figs/fig2.R

@@ -79,7 +79,7 @@ taxForm <- {
     eval()
 } ~ dist
 xAxisLabel <- quote(
-  "Distance to secondary forest edge (m)"
+  "Distance to edge (m)"
 )
 statData2ab <- plotResultsTbl1
 statData2cd <- taxaResultsTbl1 %>%

### 2022-01-18 figs/fig3.R

@@ -2,65 +2,66 @@ library(
   here
 )
 i_am(
-  "figs/fig3"
+  "figs/fig3.R"
 )
 taxRank <- quote(
   mainDispersal
 )
+#re-run
+#taxaResultsTbl1
 source(
   here(
-    "figs/stats.R"
+    "analysis/stats.R"
   )
 )
-# source(
-#   here(
-#     "figs/fig1.R"
-#   )
-# )
-library(oir)
 
 
-#plan:
-#dist-x-Xkernels4dispersalModes
 
-measure31 <- quote(
-  wind
-)
-yAxisLabel1 <- quote(
-  "Aboveground biomass (kg)"
-)
-varForm1 <- {
-  taxMetric %>%
-    eval()
-} ~ dist * mainDispersal
-xAxisLabel <- quote(
+xAxisLabel3 <- quote(
   "Distance to secondary forest edge (m)"
 )
-statData1 <- taxaResultsTbl %>%
-  rename(
-    "variable" = "mainDispersal"
-  ) %>%
+yAxisLabel3 <- quote(
+  "Aboveground biomass (kg)"
+)
+statData3 <- taxaResultsTbl1 %>%
+  unnest(
+    "varData"
+  )
 
-  filter(
-    !is.na(
-      variable
+
+
+theme_set(
+  style
+)
+
+fig3 <- statData3 %>%
+  # dotGraph()
+  ggplot(
+    aes(
+      x = dist,
+      y = {{
+        taxMetric
+      }},
+      color = mainDispersal
     )
-  ) #waterSignif
-# underData <-
-
-
-
-#panel
-
-graph1 <- statData1 %>%
-  dotGraph(
-    measure31,
-    varForm1[[3]][[2]],
-    varForm1[[2]],
-    xAxisLabel,
-    yAxisLabel1
-    # ..useGroups = varForm1[[3]][[3]]
-) +
-  scale_y_log10()
-
-graph1
+  ) +
+  geom_quasirandom() +
+  labs(
+    x = xAxisLabel3,
+    y = yAxisLabel3
+  ) +
+  scale_y_log10() +
+  theme(
+    legend.position = "top"
+  ) +
+  # geom_smooth(
+  #   method = "lm",
+  #   se = F
+  # ) +
+  stat_summary(
+    fun.data = "median_mad",
+    position = position_dodge(
+      width = 25
+    )
+  )
+fig3

### 2022-03-04 article/disc.Rmd
draft talk,fig3 refs

--- 2021-09-08 article/article.Rmd
@@
 ---
-title: "Osa"
-output: 
-  html_notebook: 
-    toc: yes
-    code_folding: hide
-    theme: spacelab
-    fig_caption: yes
-bibliography: references.bib
+title: "Discussion"
+output: html_notebook
 ---
-
-# Introduction
-
-1.  importance of reforestation/regeneration/regrowth
-
-2.  patterns of reforestation
-
-    1.  general
-
-    2.  tropical
-
-3.  things that affect reforestation
-
-    1.  edge effects
-
-    2.  plantation structure?
-
-4.  [for NGO--start biomass census]
-
-    1.  Does leaving the plantation structure help reforestation?
-
-        1.  shade
-
-        2.  edge effects (i.e. buffer)
-
-    2.  How does the forest regenerate by distance to primary forest?
-
-        1.  how much [effort] do we need to re-seed? / how much can we leave to nature?
-
-Global forest and landscape regeneration is a key conservation and climate change mitigation strategy, adopted internationally (\@UN14 Paris, Bonn Challenge).
-
--   Tropical forests store most terrestrial biomass @pan2013 and biodiversity (CITE, but are losing functions, especially in the neotropics @hubau2020, as most tropical forests become secondary (\@FAO20.
-
-Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
-
--   Neotropical secondary forests accumulate biomass \>10x faster than old-growth forests @poorter2016, and offset \~10% of carbon emissions from Amazonian deforestation @smith2020 .
--   They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y @rozendaal2019 .
--   However, secondary neotropical forests diverge widely in taxonomic composition @norden2015 .
-
-Divergent communities among secondary forests may warrant more localized conservation efforts, studying relative effects of generalized community processes.
-
-Half of all forests lie half of a kilometer from an ecotone @haddad2015, suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.
-
--   Tropical forests lose 22% carbon along edges \>100m, in part due to shorter trees and less leaf mass @ordway2020 .
-
-Here we report the edge effects of an abandoned wet neotropical timber plantation.
-
--   We hypothesized that remnant timber trees would mediate forest regrowth by maintaining uniform shade that would suppress shade-intolerant / pioneer trees / taxa and allow for more diversity.
-
-Forest regeneration has the potential to store \#\#C and preserve \#\# diversity globally in the coming decades (UN, Bonn challenge).
-
-At local scales, regeneration is strongly affected by landscape [ecological] heterogeneity as well as management history.
-
-However, patterns of forest regeneration remain unclear, especially at more local scales (\< \~50 ha).
-
-Syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years), mainly with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).
-
-However, while restoring species richness may be enough to maintain some ecosystem functions like carbon storage (Liu et al 2018), it does not guarantee the restoration of species composition.
-
-Instead, changes to species composition likely last up to centuries, showing a strong legacy of priority effects from pioneering, early-successional communities that are mostly fast-growing, low-wood-density, and shade-intolerant (Tabarelli et al 2014, Tabarelli & Lopes 2008).
-
-forests show legacy effects of common human activity, like logging and intensive agriculture/agroforestry/silviculture (FAO 2020).
-
-[High] Regional biodiversity [can] affects/increases the uncertainty about the local occurrences and outcomes of specific species interactions,
-
-such as after dispersal and/or niche overlap [CITE-niche, neutral theory].
-
-[At the same time,] Diverse local management activity modifies demographic states/profiles of various taxa's population dynamics.
-
-For example, active timber plantations strongly promote the population growth of focal taxa at the expense of non-focal taxa.
-
-After abandonment, however, the ecosystem likely reflects how continuous regional ecological processes look different given past management.
-
-such as seed dispersal from nearby primary forest edges.
-
-[These] Interactions between background ecology and human management ultimately have both local and general implications for the potential conservation/stability of ecosystem services/functions
-
-like biomass/carbon storage and biodiversity maintenance,
-
-as well as manager livelihood.
-
-Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests (Nordon et al 2015, Christensen 2014, Munroe et al 2013, Chazdon et al 2007).
-
-but lagging changes in composition may contribute to reduced effects of diversity on other forest functions (Griess et al 2012).
-
-Restoring species composition more quickly is likely a place where active forest managers can have the most unique impacts.
-
-For land managers / foresters with interest in conservation, inheriting land affected by past decisions is a reality.
-One of the legacies is where the land was placed.
-To best manage it in the future, an initial assessment of the ecological status of the land is important to understand it's likely trajectory, and what it will need in the future for tailed services/conservation/protection.
-
-Edge effects are known for degrading primary forest, but they can also be used to assist reforestation.
-
-Traditional edge effects act by mortality of large trees that makes room for pioneer species to enter, but existing interior pioneer species also have dispersal radii that overlap with the adjacent exterior, potentially facilitating regeneration.
-
-Encroachment vs. facilitation relies on a balance between mortality (per area) and dispersal + establishment + stem growth.
-
-So pioneer species can be helpful for reforestation though they are typically seen as bad for old forest conservation.
-
-Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009),
-
-with underlying arguments based on metapopulation theory in ecology (Levins 1969, MacArthur & Wilson 1967), but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.
-
-Especially for abandoned land that regenerates via various processes (Stanturf et al 2014), the proximity to less disturbed habitat can affect the composition of the regrowing community by changing effective seed dispersal kernels and probability of establishment (Wandrag et al 2017, Guariguata & Ostertag 2001), along with animal disperser communities themselves (Gilroy et al 2014).
-
-While dispersal is complicated (Muller-Landau & Hardesty 2005), it can broadly correlate with simpler plant traits such as height and seed mass (Tamme et al 2014, Thomson et al 2011).
-
-Edge effects can also operate via the style of management, including logging decisions (Edwards et al 2014, Clark et al 2012) and land-use history generally (Omeja et al 2012), can also affect stand diversity maintenance,
-
-with examples of high diversity (compared to older adjacent stands) in agroforests (Oliveira-Neto et al 2017) and native timber plantations (Pryde et al 2015),
-
-possibly by changing early tree survival probabilities, growth rates, and/or competition,
-
-but surprisingly less likely facilitation by nitrogen-fixing trees (Lai et al 2018).
-
-It may also be that due to both high recruitment and mortality (turnover) of pioneering species' stems, transitions from early- to late-successional communities may be more complex than gradual (van Greugel et al 2007).
-
-The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density (César et al 2021), and could therefore require a switch in focal timber species (Karsten et al 2014).
-
-To understand the ecological principles behind secondary forest regeneration (Hulshoff & Powers 2019), here we study the regeneration of an abandoned timber plantation as a function of distance from surrounding native forest.
-
--   how can we expect it to regrow?
--   how far/much [effort] do we need to re-seed?
-
-We hypothesized that planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity.
-
-We also considered how the legacy plantation canopy structure mediated forest stand effects via changes in abiotic conditions like light availability and microtopography (Powers & Peréz-Aviles 2012).
-
-# Results
-
-## Shade / Canopy
-
-1.  fig1
-
-    1.  shade x distance @ordway2020 + timber sp?
-
-    2.  height @ordway2020
-
-    3.  wood density @poorter2019
-
-    4.  total biomass x dist @ordway2020 + timber sp?
-
-        -   maybe system-level curve (e.g. biomass hist) + timber sp
-
-            -   metric hist, vs. sp-abund/diversity hist below
-
-        -   max DAP as trait?
-
-## Edge effects
-
-1.  fig2
-
-    1.  richness
-
-    2.  diversity x distance [+ 1 timber sp]
-
-        -   can break down by diversity index
-        -   compare to NMDS?
-
-        1.  system-level curve (e.g. rank-abund) x distance [+ 1 timber sp]
-
-# Discussion
 
@@
 based on surrounding forest proximity.
-
-# Methods
-
-## Statistical analysis
-
--   R v\#\#\#
-
-    -   BIOMASS package v\#\#\#
-
-        -   taxa (family, maybe genus) updated from raw IDs using field guides
-
-        -   matched wood densities from package database
-
-    -   tidyverse v\#\#\# = summary stats, plotting
-
-    -   vegan v\#\#\# = community analysis
-
-        -   justify metrics, i.e. entropy vs. hists
-
-## Study site
-
-This study was done in the lowland wet tropical rainforest of the Osa Peninsula in SW Puntarenas, Costa Rica.
-The forest fragment used was acquired by the Natural Resources Defense Council (nrdc.org) and managed by the local non-profit organization Osa Conservation (osaconservation.org), located at their Greg Gund Conservation Center, near the more widely-used Piro research station.
-The focal parcel was 20 ha in size and was undergoing secondary succession after having been abandoned \~20 years ago as a wood plantation for *Bombacopsis quinata*, which has usually been better suited to the dry forests of Guanacaste and the Nicoya Peninsula (Pérez Cordero & Kanninen 2002).
-The secondary forest stand was roughly triangular and surrounded by primary forest on the S and NW sides.
-This census was done in 2013 during the rainy season months between June and August.
-
-## Census design
-
-Geographic information systems software (QGIS 2.0, qgis.org; ArcGIS 10, arcgis.com) was used to divide the parcel area into six discrete strata representing 50-m intervals from the observed primary forest edge to the center of the opposing NE edge, \~300 m away.
-This design meant that strata were smaller further from the primary forest edge, so accordingly, fewer plots were randomly positioned inside each stratum: the 0-50, 50-100, 100-150, 150-200, 200-250, and 250-300 m strata had eleven, eight, five, four, two, and one plot(s), respectively.
-The edge distance variable was descretized into strata this way as a more feasible alternative to measuring plot-edge distances after plot placement, which would have taken much longer compared to the improved spatial resolution.
-Each plot was 21 x 21 m and oriented North, and the total area represented by all the surveyed plots was \~1 ha or 5% of the total stand area, comparable to similar studies (Onyekwelu & Olabiwannu 2016).
-
-## Plot measurements / Data collection
-
-### Canopy cover & Topography
-
-Light reaching the forest floor was measured at the center of each plot, at chest height, using a densiometer (Foresty Suppliers, Inc.).
-Following manufacturer instructions, the final plot value was the average of 4 different readings taken facing each of the 4 cardinal directions: North, South, East, and West.
-
-### [Topography]
-
-The slope of the forest floor in each plot was measured using [a rangefinder by measuring the distance from one (specific?) corner of the plot to another (specific?) corner.]
-
-### Tree measurements
-
-In each plot, the diameter at breast height (DBH = 2.5 -- 3 m), crown height, and taxonomic ID was recorded for every stem ≥10 cm in [diameter/circumference?].
-Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
-Tree height was measured by using a rangefinder (Bushnell, Forestry Suppliers, Inc.) to measure from an observer to the crown (from where it was visible), and then to the trunk at breast height; these 2 distances were then used to triangulate and calculate the tree's height using Pythagoras's Theorem.
-
-Trees were distinguished based on local ecological knowledge of common names, followed by matching to Latin binomial taxonomic names using printed local field guides available.
-
-### Aboveground biomass
-
-Aboveground biomass (AGB) was estimated for each stem using the allometric equation developed by Chave et al.
-(2014) specific for wet tropical forests, which has performed slightly better compared to both pantropical and previous other models by Chave et al.
-(2005) (Ngomanda et al 2014, Alvarez et al 2012), with improved local accuracy adding height (Khan et al 2020, Domke et al 2012) and wood density (specific gravity) (van Breugel et al 2011):
-
-*AGB = 0.0673 x (ρ D^2^ H)^0.976^*
-
-with *D* in cm, *H* in m, and *p* in g cm^-3^.
-Measured DBHs and heights were used, along with wood densities (*ρ*) extracted from available literature.
-In most cases, these values were only available at the genus level, and unrepresented taxa were assumed to be 0.58, an assumption recommended by the World Agroforestry Database (worldagroforestry.org).
-[References used for this step are given in *Table S1*.] Original values from 2013 were then updated using the pipeline developed in the BIOMASS package for R by Réjou-Méchain et al.
-(2017), which replaced unknown wood density values with plot-level averages instead.
-However, these updates did not appear to change the main results of this study.
-
-### Statistical analyses
-
-All variables were averaged at the plot level, including biomass converted to a per hectare basis, thereby setting *n* = 30 for all analyses reported.
-
-All response variables passed Shapiro-Wilk normality and Levene's homogeneity tests before being subject to parametric tests---either multi-way ANOVA or one-way ANOVA and linear regression, or generalized (non-linear) regression.
-
-Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.
-These variables are also shown in figures with their own separate analyses, exclusively one-way ANOVAs with additional relevant parameters from their linear regression counterparts (*i.e.* for *R^2^*and *slope*).
-Standard one-way ANOVAs were also run for hill slope and canopy light availability separately.
-Non-linear regressions were run as linear regressions with slight modifications using the function "poly()" in the "stats" package for R software, which was only used for stem density and RICHNESS variables.
-All individual trees censused were included in all analyses, including unidentified taxa, but these averaged to only 1 group for taxa-related analyses, namely RICHNESS with distance, having a minor effect.
-All tests included only distance to primary forest edge as the dependent variable, and all data was analyzed in R v3.6.1 software (R Core Team 2019).
-
--   Allen and Kim (2020) as some justification for family-level analysis (their MI forest site shares \~6 / their 20 families with this site)
-
--   Neighborhood influence metric (Lorimer 1983, applied in Looney et al 2016)
-
--   Data and analytical code available on GitHub.../DRYAD
-
-# Acknowledgements
-
--   osa conservation - max villalobos, [dan student harvard alum], agustín (& annia), manuel director
--   lehigh group - don, emma, ?, ?
--   brandeis WOW funding

### 2022-03-04 article/intro.Rmd
draft talk,fig3 refs

--- 2021-09-14 article/article.Rmd
@@
 ---
-title: "Osa"
-output:
-  html_document:
-    toc: yes
-    df_print: paged
-  html_notebook:
-    toc: yes
-    code_folding: hide
-    theme: spacelab
-    fig_caption: yes
-bibliography: references.bib
+title: "Introduction"
+output: html_notebook
 ---
-
-# Introduction
 
@@
 We also considered how the legacy plantation canopy structure mediated forest stand effects via changes in abiotic conditions like light availability and microtopography (Powers & Peréz-Aviles 2012).
-
-# Results
-
-```{r setup, echo = F, message = F}
-knitr::opts_chunk$set(
-  echo = F,
-  message = F,
-  warning = F
-)
-library(here)
-```
-
-## Biomass
-
-1.  fig1
-
-    1.  shade x distance ( @ordway2020 ) + timber sp?
-
-    2.  height ( @ordway2020 )
-
-    3.  wood density ( @poorter2019 )
-
-    4.  total biomass x dist ( @ordway2020 ) + timber sp?
-
-        -   maybe system-level curve (e.g. biomass hist) + timber sp
-
-            -   metric hist, vs. sp-abund/diversity hist below
-
-        -   max DAP as trait?
-
-[RESULTS TEXT]
-
-```{r fig1}
-source(
-  here(
-    "figs/fig1.R"
-  )
-)
-fig1
-```
-
-## Composition
-
-1.  fig2
-
-    1.  richness
-
-    2.  diversity x distance [+ 1 timber sp]
-
-        -   can break down by diversity index
-        -   compare to NMDS?
-
-        1.  system-level curve (e.g. rank-abund) x distance [+ 1 timber sp]
-
-```{r fig2}
-
-```
-
-# Discussion
-
-1.  assess hypotheses
-2.  infer causes of patterns - plantation structure, old legacy trees, or species traits?
-3.  compare inferences to other studies
-4.  recommend future conservation approaches, ecological studies
-
-Under more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e.
-near 1:1 replacement) the targeted tree cutting that is done using information collected about the ecological profiles and ages (when possible) of trees (Duah-Gyamfi et al 2014, Edwards et al 2014).
-
-In contrast, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015) while also usually showing negative trade-offs with other services, like that with biodiversity given reliance on monoculture planting designs.
-
-In some cases where management decisions are made primarily based on regional economic but not local ecological information, due to limited access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.
-
-After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g.
-based on surrounding forest proximity.
-
-# Methods
-
-## Statistical analysis
-
--   R v\#\#\#
-
-    -   BIOMASS package v\#\#\#
-
-        -   taxa (family, maybe genus) updated from raw IDs using field guides
-
-        -   matched wood densities from package database
-
-    -   tidyverse v\#\#\# = summary stats, plotting
-
-    -   vegan v\#\#\# = community analysis
-
-        -   justify metrics, i.e. entropy vs. hists
-
-## Study site
-
-This study was done in the lowland wet tropical rainforest of the Osa Peninsula in SW Puntarenas, Costa Rica.
-The forest fragment used was acquired by the Natural Resources Defense Council (nrdc.org) and managed by the local non-profit organization Osa Conservation (osaconservation.org), located at their Greg Gund Conservation Center, near the more widely-used Piro research station.
-The focal parcel was 20 ha in size and was undergoing secondary succession after having been abandoned \~20 years ago as a wood plantation for *Bombacopsis quinata*, which has usually been better suited to the dry forests of Guanacaste and the Nicoya Peninsula (Pérez Cordero & Kanninen 2002).
-The secondary forest stand was roughly triangular and surrounded by primary forest on the S and NW sides.
-This census was done in 2013 during the rainy season months between June and August.
-
-## Census design
-
-Geographic information systems software (QGIS 2.0, qgis.org; ArcGIS 10, arcgis.com) was used to divide the parcel area into six discrete strata representing 50-m intervals from the observed primary forest edge to the center of the opposing NE edge, \~300 m away.
-This design meant that strata were smaller further from the primary forest edge, so accordingly, fewer plots were randomly positioned inside each stratum: the 0-50, 50-100, 100-150, 150-200, 200-250, and 250-300 m strata had eleven, eight, five, four, two, and one plot(s), respectively.
-The edge distance variable was descretized into strata this way as a more feasible alternative to measuring plot-edge distances after plot placement, which would have taken much longer compared to the improved spatial resolution.
-Each plot was 21 x 21 m and oriented North, and the total area represented by all the surveyed plots was \~1 ha or 5% of the total stand area, comparable to similar studies (Onyekwelu & Olabiwannu 2016).
-
-## Plot measurements / Data collection
-
-### Canopy cover & Topography
-
-Light reaching the forest floor was measured at the center of each plot, at chest height, using a densiometer (Foresty Suppliers, Inc.).
-Following manufacturer instructions, the final plot value was the average of 4 different readings taken facing each of the 4 cardinal directions: North, South, East, and West.
-
-### [Topography]
-
-The slope of the forest floor in each plot was measured using [a rangefinder by measuring the distance from one (specific?) corner of the plot to another (specific?) corner.]
-
-### Tree measurements
-
-In each plot, the diameter at breast height (DBH = 2.5 -- 3 m), crown height, and taxonomic ID was recorded for every stem ≥10 cm in [diameter/circumference?].
-Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
-Tree height was measured by using a rangefinder (Bushnell, Forestry Suppliers, Inc.) to measure from an observer to the crown (from where it was visible), and then to the trunk at breast height; these 2 distances were then used to triangulate and calculate the tree's height using Pythagoras's Theorem.
-
-Trees were distinguished based on local ecological knowledge of common names, followed by matching to Latin binomial taxonomic names using printed local field guides available.
-
-### Aboveground biomass
-
-Aboveground biomass (AGB) was estimated for each stem using the allometric equation developed by Chave et al.
-(2014) specific for wet tropical forests, which has performed slightly better compared to both pantropical and previous other models by Chave et al.
-(2005) (Ngomanda et al 2014, Alvarez et al 2012), with improved local accuracy adding height (Khan et al 2020, Domke et al 2012) and wood density (specific gravity) (van Breugel et al 2011):
-
-*AGB = 0.0673 x (ρ D^2^ H)^0.976^*
-
-with *D* in cm, *H* in m, and *p* in g cm^-3^.
-Measured DBHs and heights were used, along with wood densities (*ρ*) extracted from available literature.
-In most cases, these values were only available at the genus level, and unrepresented taxa were assumed to be 0.58, an assumption recommended by the World Agroforestry Database (worldagroforestry.org).
-[References used for this step are given in *Table S1*.] Original values from 2013 were then updated using the pipeline developed in the BIOMASS package for R by Réjou-Méchain et al.
-(2017), which replaced unknown wood density values with plot-level averages instead.
-However, these updates did not appear to change the main results of this study.
-
-### Statistical analyses
-
-All variables were averaged at the plot level, including biomass converted to a per hectare basis, thereby setting *n* = 30 for all analyses reported.
-
-All response variables passed Shapiro-Wilk normality and Levene's homogeneity tests before being subject to parametric tests---either multi-way ANOVA or one-way ANOVA and linear regression, or generalized (non-linear) regression.
-
-Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.
-These variables are also shown in figures with their own separate analyses, exclusively one-way ANOVAs with additional relevant parameters from their linear regression counterparts (*i.e.* for *R^2^*and *slope*).
-Standard one-way ANOVAs were also run for hill slope and canopy light availability separately.
-Non-linear regressions were run as linear regressions with slight modifications using the function "poly()" in the "stats" package for R software, which was only used for stem density and RICHNESS variables.
-All individual trees censused were included in all analyses, including unidentified taxa, but these averaged to only 1 group for taxa-related analyses, namely RICHNESS with distance, having a minor effect.
-All tests included only distance to primary forest edge as the dependent variable, and all data was analyzed in R v3.6.1 software (R Core Team 2019).
-
--   Allen and Kim (2020) as some justification for family-level analysis (their MI forest site shares \~6 / their 20 families with this site)
-
--   Neighborhood influence metric (Lorimer 1983, applied in Looney et al 2016)
-
--   Data and analytical code available on GitHub.../DRYAD
-
-# Acknowledgements
-
--   osa conservation - max villalobos, [dan student harvard alum], agustín (& annia), manuel director
--   lehigh group - don, emma, ?, ?
--   brandeis WOW funding

### 2022-03-04 article/methods.Rmd
draft talk,fig3 refs

---
title: "Methods"
output: html_notebook
---

## Study site

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center **(GPS)** near the Piro Biological Station run by Osa Conservation.

**See () for an ecosystem description.**

The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the dry-adapted species from the Guanacaste region *(Pérez Cordero & Kanninen 2002)* grew poorly in very wet conditions.

This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides *(Fig 1)* and a wide service road on the third NE border, with primary forest beyond it.

This census was done in 2013 during the rainy season months between June and August.

## Census design

Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E *(Fig 1)*, using GIS software *(ArcGIS, QGIS)*.

Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.

The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *(Onyekwelu & Olabiwannu 2016)*.

## Plot measurements / Data collection

### Canopy cover & Topography

Light reaching the forest floor was measured at the center of each plot at chest height using a densiometer *(Forestry Suppliers, Inc.)*, as an average of four readings taken facing each cardinal direction.

<!-- ### [Topography] -->

The slope of the forest floor was measured using a rangefinder *(Bushnell, Forestry Suppliers, Inc)* to measure the distance the diagonal between two plot corners and triangulate the observation angle.

### Tree measurements

The diameter of all stems >10 cm wide were recorded in each census plot.

Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.

Tree height was recorded by measuring distances to both the crown and the stem at chest height (~2.7 m) using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and trangulating the missing side length.

Taxa were identified with localized knowledge and field guides matching common names to Latin binomials, 
and trait information was gathered from the literature.

### Aboveground biomass

Aboveground biomass (AGB) was estimated per stem using the allometric equation by *Chave et al. (2014)*.

The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *Chave et al. (2005) (Ngomanda et al 2014, Alvarez et al 2012)*.

The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *(Khan et al 2020, Domke et al 2012)*.

Wood density (or specific gravity) values when available also improve biomass estimates even more than height *(van Breugel et al 2011)*.

The equation used was:
\
*AGB = 0.0673 x (ρ D^2^ H)^0.976^*
\
with *D* in cm, *H* in m, and *p* in g cm^-3^.

Measured DBHs and heights were used and wood densities (*ρ*) extracted from available literature.

Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be 0.58 following World Agroforestry Database quidelines *(worldagroforestry.org)*.

<!-- *[References used for this step are given in *Table S1*.] * -->

Original values from 2013 were later updated to modern values using the pipeline developed in the *BIOMASS* ( `r `packageVersion("BIOMASS")` ) R software package by *Réjou-Méchain et al. (2017)*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.

<!-- However, these updates did not appear to change the main results of this study. -->

# Statistical analyses

For all response variables, data were analyzed using plot medians and subject to linear regression with the distance to forest edge variable after checks for residual normality and equal variance assumptions using R `getRversion()` *(R Core Team 2019)*.

 <!-- including biomass converted to a per hectare basis,  -->

<!-- Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.  -->

Non-linear regressions were run using the function *poly()* in the *stats* `packageVersion("stats")` R package *()*, 
which was only recorded for stem density and richness variables.

All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.

Community analysis and ordination was done as a PERMANOVA with the *vegan* `packageVersion("vegan")` R package *()*.

Data and code are available at *nmedina17.github.com/osa*.

<!-- -   Allen and Kim (2020) as some justification for family-level analysis (their MI forest site shares \~6 / their 20 families with this site) -->

<!-- -   Neighborhood influence metric (Lorimer 1983, applied in Looney et al 2016) -->

<!-- - justify metrics, i.e. entropy vs. hists -->

### 2022-03-04 article/results.Rmd
draft talk,fig3 refs

---
title: "Results"
output: html_notebook
---

```{r setup, echo = F, message = F}
knitr::opts_chunk$set(
  echo = F,
  message = F,
  warning = F
)
library(here)
```

## Biomass

Distance to primary forest edge was associated with ### more ___, although this did not correlate with ___.

<!-- 1.  shade x distance ( @ordway2020 ) + timber sp? -->
<!-- 2.  height ( @ordway2020 ) -->
<!-- 3.  wood density ( @poorter2019 ) -->
<!-- 4.  total biomass x dist ( @ordway2020 ) + timber sp? -->
<!--   -   maybe system-level curve (e.g. biomass hist) + timber sp -->
<!--   -   metric hist, vs. sp-abund/diversity hist below -->
<!--     -   max DAP as trait? -->

<!-- [RESULTS TEXT] -->

```{r fig1}
source(
  here(
    "figs/fig1.R"
  )
)
fig1
```

## Composition

<!-- 1.  richness -->
<!-- 2.  diversity x distance [+ 1 timber sp] -->
<!--   -   can break down by diversity index -->
<!--     -   compare to NMDS? -->
<!--   1.  system-level curve (e.g. rank-abund) x distance [+ 1 timber sp] -->

```{r fig2}
source(
  here(
    "figs/fig2.R"
  )
)
fig1
```

### 2022-03-04 talk/talk.Rmd
draft talk,fig3 refs

@@
 ---
-title: "R Notebook"
-output: html_notebook
+title: 
+  "Natural regeneration of a tropical timber plantation"
+author: 
+  "Nicholas Medina"
+output: 
+  # slidy_presentation
+  # oislides_presentation
+  # powerpoint_presentation
+  html_notebook
+  # html_document
+  # pdf_document #no
 ---
 
-```{r setup, echo=F}
+```{r setup, echo = F, include = F}
 knitr::opts_chunk$set(
-  echo=F, quiet=T)
+  echo = F,
+  quiet = T
+)
 library(
@@
 i_am(
-  "talk/talk.Rpres"
+  "talk/talk.Rmd"
 )
-#fixRender
-library(
-  httpuv
-)
+# #fixRender
+# library(
+#   httpuv
+# )
+#embed_file()
+# library(
+#   xfun
+# )
+# xfun::pkg_load2(c("htmltools", "mime"))
 ```
@@
 
-Introduction: Tropical reforestation
+I--Global forests are regenerating, but how?
 ===
 
-- Over half are secondary, globally
+- ~Half of forests are tropical, and over half of all forests globally are secondary *(FAO 2020)*
 
-- Managed recovery can help support local communities
+- Services like timber and biodiversity depend on natural and managed regeneration patterns
 
-- To make the most of ongoing natural processes, ecological dynamics should be better understood
+- Regeneration dynamics should be better understood
 
-Introduction: Community patterns
+I--Regeneration is quick and inconsistent
 ===
 
-- Species richness recovers quickly
+- Biomass recovers 90% in 66 yrs *(Poorter et al 2016 PNAS)*
 
-- But community composition may take centuries
+- Taxonomic richness recovers quickly ()
 
-- So balance of ecosystem service recovery unclear
+- Community composition may take centuries to recover, if at all *(Norder et al 2010 PNAS)*
 
-Introduction: This study
+I--Edges shape regeneration
 ===
 
-- Question:
+- 70% forests are <1 km from a habitat edge *(Haddad et al 2015 Science)*
 
-- Hypothesis:
+- Edges effects tend to reduce biodiversity by 13 - 75% via ... *(")*
 
-- Predictions:
+    - stronger in small and isolated fragments
 
-Methods: Study site
+I--Edge effects may differ in wet tropical plantations
 ===
 
-- Wet rainforest, Osa Peninsula, Costa Rica
+- 7% of global forests are planted and planted area has increased over last 30 yrs *(FAO 2020)*
 
-- 20 ha. timber plantation, became NRDC preserve post-abandonment ~2003
+    - Often dense with fast-growing, shade-intolerant taxa
 
-- Surrounded by unmanaged/primary forest on 3 sides, road on other
+- Wet tropical forest regeneration tends to be light-limited, based on tree functional traits *(Poorter et al 2021)*
+
+- Existing plantation demography and shade may skew/buffer edge effects toward net positive instead of negative
+
+I--This study
+===
+
+- Question -- How is wet tropical secondary forest regeneration affected by former plantation status and edges with road and primary forest?
+
+- Hypothesis --
+
+    - Former plantation status facilitates succession by allowing shade-intolerant taxa to grow better due to __denser canopies__
+
+    - Forest edges promotes regeneration by faciltating __dispersal__ of shade-tolerant taxa
+
+- Predictions -- Distance away from primary forest edge is associated with --
+
+    - lower stand biomass
+
+    - lower tree diversity
+
+    - lower shade-tolerant taxa
+
+II--Study site
+===
+
+- Wet tropical rainforest in Osa Peninsula, SW Costa Rica
+
+- 20 ha plantation of dry forest timber, abandoned in ~2003 into NRDC preserve
+
+- Area surrounded by primary forest on two sides and road on third
 
@@
 
-Methods: Study design
+II--Census design
 ===
 
-- 300 m transect from unmanaged forest edge inward
+- Random stratified forest biomass inventory sampling design --
 
-- Divided into 50 m bands using Q-, ArcGIS
+    - divided 300 m range into six 50 m strata, from forest to road edge, using QGIS and ArcGIS
 
-- 30 plots randomly assigned, but weighted by band area
+    - 30, 21 x 21 m plots randomly distributed, but quantity roughly proportional to stratum area
 
-![](map.png)
+    - All plots covered 1 ha total, 5% of entire plantation
 
-Methods: Census
+![](../figs/map.png)
+
+II--Census measurements
 ===
 
-- DBH, height stems ≥10 cm using rangefinder
+- Stems ≥10 cm were measured for DBH and height using rangefinder *(Bushnell, Forestry Suppliers, Inc)*
 
-- ID by local field guide personnel and books
+- Taxa were ID'd with help from local field guides
 
-- plot canopy closure w/ densiometer, and ground slope
+- Plot canopy closure was measured with densiometer
 
-# Results: Forest traits along edge
+II--Statistical analysis
+===
+
+- For all variables--
+
+    - Trend lines follow median values of all plots in distance stratum
+
+- Traits obtained from literature and compared to TRY database *(Kattge et al 2020)*
+
+- Computations done with R `getRversion()` depending on *BIOMASS* `packageVersion("BIOMASS")`, 
+*vegan* `packageVersion("vegan")`, and 
+*tidyverse* `packageVersion("tidyverse")` packages
+
+III--Stand regeneration
+===
+
+- Further from primary forest edge--
+
+    - Biomass and wood density tended to increase
+
+    - Stem density decreased then increased
+
+    - Neither canopy light nor height (not shown) tended to change
+
+- Distance from forest edge tended to explain variation--
+
+    - 10% in wood density
+
+    - ~20% in stem density
 
 ```{r fig1}
-source(
-  here(
-    "figs/fig1.R"
-  )
-)
-fig1
+# source(
+#   here(
+#     "figs/fig1.R"
+#   )
+# )
+# fig1
 ```
 
-Results: Composition along edge
+![](../figs/fig1.png)
+
+III--Composition regeneration
 ===
 
+- Further from forest edge--
+
+    - Richness tended to increase
+
+    - Diversity decreased steeply
+
+    - Composition varied notably, toward PC1
+
+    - Of abundant taxa biomass--
+
+        - Vochysia tended to decrease
+
+        - Ficus decreased
+
 ```{r fig2}
-source(
-  here(
-    "figs/fig2.R"
-  )
-)
-fig2
+# source(
+#   here(
+#     "figs/fig2.R"
+#   )
+# )
 ```
 
-Results: Dispersal modes along edge
+![](../figs/fig2.png)
+
+III--Functional regeneration
 ===
 
+-
+
+-
+
+-
+
 ```{r fig3}
-source(
-  here(
-    "figs/fig3.R"
-  )
-)
-fig3
+# source(
+#   here(
+#     "figs/fig3.R"
+#   )
+# )
+# fig3
 ```
 
-Discussion
+![](../figs/fig3.png)
+
+IV--Hypothesis check
 ===
 
+- Stand biomass is more resilient than biodiversity
+
+    - Further from forest--
+
+        - No trends in biomass
+
+        - Diversity decreased
+
+- Forest edge associated with tree diversity
+
+IV--Processes underlying patterns
+===
+
+- not shade...
+
+- dispersal...
+
+-
+
+IV--Compare to similar studies
+===
+
+- aligns well with major recent papers
+
+-
+
+-
+
+IV--Future studies
+===
+
+- Compare wet tropical plantations at different stages of regeneration
+
+- Compare forests with different initial composition but in similar region (i.e. realized vs. fundamental niches)
+
+- Study ecological dynamics under various management conditions
+
 <!-- For more details on authoring R presentations please visit <https://support.rstudio.com/hc/en-us/articles/200486468>. -->

### 2022-03-04 talk/talk.tex
draft talk,fig3 refs

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\title{Natural regeneration of a tropical timber plantation}
\author{Nicholas Medina}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Natural regeneration of a tropical timber plantation},
  pdfauthor={Nicholas Medina},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\begin{document}
\maketitle

\hypertarget{outline}{%
\section{Outline}\label{outline}}

\begin{itemize}
\tightlist
\item
  Introduction
\item
  Methods
\item
  Results
\item
  Discussion
\end{itemize}

\hypertarget{iglobal-forests-are-regenerating-but-how}{%
\section{I--Global forests are regenerating, but
how?}\label{iglobal-forests-are-regenerating-but-how}}

\begin{itemize}
\item
  \textasciitilde Half of forests are tropical, and over half of all
  forests globally are secondary \emph{(FAO 2020)}
\item
  Services like timber and biodiversity depend on natural and managed
  regeneration patterns
\item
  Regeneration dynamics should be better understood
\end{itemize}

\hypertarget{iregeneration-is-quick-and-inconsistent}{%
\section{I--Regeneration is quick and
inconsistent}\label{iregeneration-is-quick-and-inconsistent}}

\begin{itemize}
\item
  Biomass recovers 90\% in 66 yrs \emph{(Poorter et al 2016 PNAS)}
\item
  Taxonomic richness recovers quickly ()
\item
  Community composition may take centuries to recover, if at all
  \emph{(Norder et al 2010 PNAS)}
\end{itemize}

\hypertarget{iedges-shape-regeneration}{%
\section{I--Edges shape regeneration}\label{iedges-shape-regeneration}}

\begin{itemize}
\item
  70\% forests are \textless1 km from a habitat edge \emph{(Haddad et al
  2015 Science)}
\item
  Edges effects tend to reduce biodiversity by 13 - 75\% via \ldots{}
  \emph{(``)}

  \begin{itemize}
  \tightlist
  \item
    stronger in small and isolated fragments
  \end{itemize}
\end{itemize}

\hypertarget{iedge-effects-may-differ-in-wet-tropical-plantations}{%
\section{I--Edge effects may differ in wet tropical
plantations}\label{iedge-effects-may-differ-in-wet-tropical-plantations}}

\begin{itemize}
\item
  7\% of global forests are planted and planted area has increased over
  last 30 yrs \emph{(FAO 2020)}

  \begin{itemize}
  \tightlist
  \item
    Often dense with fast-growing, shade-intolerant taxa
  \end{itemize}
\item
  Wet tropical forest regeneration tends to be light-limited, based on
  tree functional traits \emph{(Poorter et al 2021)}
\item
  Existing plantation demography and shade may skew/buffer edge effects
  toward net positive instead of negative
\end{itemize}

\hypertarget{ithis-study}{%
\section{I--This study}\label{ithis-study}}

\begin{itemize}
\item
  Question -- How is wet tropical secondary forest regeneration affected
  by former plantation status and edges with road and primary forest?
\item
  Hypothesis --

  \begin{itemize}
  \item
    Former plantation status facilitates succession by allowing
    shade-intolerant taxa to grow better due to \textbf{denser canopies}
  \item
    Forest edges promotes regeneration by faciltating \textbf{dispersal}
    of shade-tolerant taxa
  \end{itemize}
\item
  Predictions -- Distance away from primary forest edge is associated
  with --

  \begin{itemize}
  \item
    lower stand biomass
  \item
    lower tree diversity
  \item
    lower shade-tolerant taxa
  \end{itemize}
\end{itemize}

\hypertarget{iistudy-site}{%
\section{II--Study site}\label{iistudy-site}}

\begin{itemize}
\item
  Wet tropical rainforest in Osa Peninsula, SW Costa Rica
\item
  20 ha plantation of dry forest timber, abandoned in
  \textasciitilde2003 into NRDC preserve
\item
  Area surrounded by primary forest on two sides and road on third
\end{itemize}

\hypertarget{iicensus-design}{%
\section{II--Census design}\label{iicensus-design}}

\begin{itemize}
\item
  Random stratified forest biomass inventory sampling design --

  \begin{itemize}
  \item
    divided 300 m range into six 50 m strata, from forest to road edge,
    using QGIS and ArcGIS
  \item
    30, 21 x 21 m plots randomly distributed, but quantity roughly
    proportional to stratum area
  \item
    All plots covered 1 ha total, 5\% of entire plantation
  \end{itemize}
\end{itemize}

\includegraphics{../figs/map.png}

\hypertarget{iicensus-measurements}{%
\section{II--Census measurements}\label{iicensus-measurements}}

\begin{itemize}
\item
  Stems ≥10 cm were measured for DBH and height using rangefinder
  \emph{(Bushnell, Forestry Suppliers, Inc)}
\item
  Taxa were ID'd with help from local field guides
\item
  Plot canopy closure was measured with densiometer
\end{itemize}

\hypertarget{iistatistical-analysis}{%
\section{II--Statistical analysis}\label{iistatistical-analysis}}

\begin{itemize}
\item
  For all variables--

  \begin{itemize}
  \tightlist
  \item
    Trend lines follow median values of all plots in distance stratum
  \end{itemize}
\item
  Traits obtained from literature and compared to TRY database
  \emph{(Kattge et al 2020)}
\item
  Computations done with R \texttt{getRversion()} depending on
  \emph{BIOMASS} \texttt{packageVersion("BIOMASS")}, \emph{vegan}
  \texttt{packageVersion("vegan")}, and \emph{tidyverse}
  \texttt{packageVersion("tidyverse")} packages
\end{itemize}

\hypertarget{iiistand-regeneration}{%
\section{III--Stand regeneration}\label{iiistand-regeneration}}

\begin{itemize}
\item
  Further from primary forest edge--

  \begin{itemize}
  \item
    Biomass and wood density tended to increase
  \item
    Stem density decreased then increased
  \item
    Neither canopy light nor height (not shown) tended to change
  \end{itemize}
\item
  Distance from forest edge tended to explain variation--

  \begin{itemize}
  \item
    10\% in wood density
  \item
    \textasciitilde20\% in stem density
  \end{itemize}
\end{itemize}

\includegraphics{../figs/fig1.png}

\hypertarget{iiicomposition-regeneration}{%
\section{III--Composition
regeneration}\label{iiicomposition-regeneration}}

\begin{itemize}
\item
  Further from forest edge--

  \begin{itemize}
  \item
    Richness tended to increase
  \item
    Diversity decreased steeply
  \item
    Composition varied notably, toward PC1
  \item
    Of abundant taxa biomass--

    \begin{itemize}
    \item
      Vochysia tended to decrease
    \item
      Ficus decreased
    \end{itemize}
  \end{itemize}
\end{itemize}

\includegraphics{../figs/fig2.png}

\hypertarget{iiifunctional-regeneration}{%
\section{III--Functional
regeneration}\label{iiifunctional-regeneration}}

\begin{itemize}
\tightlist
\item
\item
\item
\end{itemize}

\includegraphics{../figs/fig3.png}

\hypertarget{ivhypothesis-check}{%
\section{IV--Hypothesis check}\label{ivhypothesis-check}}

\begin{itemize}
\item
  Stand biomass is more resilient than biodiversity

  \begin{itemize}
  \item
    Further from forest--

    \begin{itemize}
    \item
      No trends in biomass
    \item
      Diversity decreased
    \end{itemize}
  \end{itemize}
\item
  Forest edge associated with tree diversity
\end{itemize}

\hypertarget{ivprocesses-underlying-patterns}{%
\section{IV--Processes underlying
patterns}\label{ivprocesses-underlying-patterns}}

\begin{itemize}
\item
  not shade\ldots{}
\item
  dispersal\ldots{}
\item
\end{itemize}

\hypertarget{ivcompare-to-similar-studies}{%
\section{IV--Compare to similar
studies}\label{ivcompare-to-similar-studies}}

\begin{itemize}
\item
  aligns well with major recent papers
\item
\item
\end{itemize}

\hypertarget{ivfuture-studies}{%
\section{IV--Future studies}\label{ivfuture-studies}}

\begin{itemize}
\item
  Compare wet tropical plantations at different stages of regeneration
\item
  Compare forests with different initial composition but in similar
  region (i.e.~realized vs.~fundamental niches)
\item
  Study ecological dynamics under various management conditions
\end{itemize}

\end{document}

### 2022-03-04 analysis/stats.R

@@ -1,17 +1,24 @@
+# Refs----
+
+
 library(here);
 i_am(
   "analysis/stats.R"
 )
-source(
-  here("data/clean.R")
-)
+
+#run1----
+
+# source(
+#   here("data/clean.R")
+# )
+
 # source(
 #   here("analysis/statsTbl.R")
 # )
 library(tidyverse)
 library(rstatix);
 library(vegan) #diversity()
-library(oir)
+library(oir) #local
 library(ggbeeswarm)
 library(ggpmisc)
 
@@ -21,6 +28,11 @@ library(ggpmisc)
 # taxRank <- quote(
 #   gen
 # )
+
+
+
+# Species tbl ----
+
 
 #4vegan
 spTbl <- cleanData %>%
@@ -57,6 +69,8 @@ spTbl <- cleanData %>%
 
 
 
+# centers ----
+
 plotVarsTbl <- cleanData %>%
   left_join(
     .,
@@ -69,12 +83,15 @@ plotVarsTbl <- cleanData %>%
     copy = T
   ) %>%
 
+  ## plots ----
+
   group_by(
     dist,
     # mainDispersal,
     plot
   ) %>%
   mutate(
+    "stems" = n(),
     "richness" =
       n_distinct(
         sp
@@ -92,6 +109,7 @@ plotVarsTbl <- cleanData %>%
     kg17,
 
     #community
+    stems,
     richness,
     entropy
   ) %>%
@@ -104,6 +122,9 @@ plotVarsTbl <- cleanData %>%
   ) %>%
 
   #nextlevel
+
+  ## dists ----
+
   mutate(
     "varData1" = varData %>%
       modify(
@@ -141,7 +162,7 @@ plotVarsTbl <- cleanData %>%
 
 
 
-#ordinate
+# ordinate ----
 
 commTbl <- spTbl %>%
   select(
@@ -159,15 +180,24 @@ metaTbl <- cleanData %>%
   distinct(
     plot,
     dist
+  ) %>%
+
+  #qc
+  filter(
+    !is.na(
+      plot |
+        dist
+    )
   )
 
 
 ordModel <- distMat ~ dist
 
-ordStat = adonis(
+ordStat <- adonis(
   ordModel,
   metaTbl,
-  99999
+  # 99999
+  permutations = 99
 )
 
 
@@ -178,7 +208,9 @@ ordTbl <- commTbl %>%
 
 
 
[…]

### 2022-03-04 data/clean.R

@@ -8,6 +8,9 @@ library(tidyverse);
 library(BIOMASS)
 
 
+#import----
+
+##raw----
 rawData <- vroom(
   here("data/raw/osa.csv")
 )
@@ -26,11 +29,76 @@ dispersal <- vroom(
   filter(
     !is.na(
       gen
+    ) &
+      !is.na(
+        mainDispersal
+      )
+  )
+
+##traits----
+traits <- vroom(
+  here(
+    "data/base/try/try19707.csv"
+  )
+) %>% #glimpse()
+  select(
+    AccSpeciesName,
+    TraitName,
+    OrigValueStr,
+    StdValue
+  ) %>%
+
+  ###qc----
+  mutate(
+    StdValue = ifelse(
+      is.na(
+        StdValue
+      ),
+      yes = OrigValueStr,
+      no = StdValue
+    ),
+    "gen" = AccSpeciesName %>%
+      str_split(
+        " "
+      ) %>% #list
+      modify(
+        . %>%
+          first()
+      ) %>%
+      as_vector()
+  ) %>%
+  select(
+    !OrigValueStr
+  ) %>%
+
+  pivot_wider(
+    id_cols = gen,
+    names_from = TraitName,
+    values_from = StdValue
+  )
+
+
+##lifeHist----
+lifeHist <- vroom(
+  here(
+    "data/base/lifeHist.csv"
+  )
+) %>%
+  select(
+    family,
+    genus,
+    dispersalMode,
+    successionalStage
+  ) %>%
+  #qc
+  filter(
+    !is.na(
+      successionalStage
     )
   )
 
 
-cleanData <- function(
+updateRaw <- function(
   ...rawData
 ) {
 
@@ -80,8 +148,11 @@ cleanData <- function(
 }
 
 
+#join----
+
 cleanData <- rawData %>%
-  cleanData() %>%
+  updateRaw() %>%
+
   full_join(
     dispersal
   ) %>%
@@ -89,15 +160,63 @@ cleanData <- rawData %>%
     !is.na(
       dist
     )
-  )
+  ) %>%
+
+  full_join(
+    lifeHist,
+    by = c(
+      "gen" = "genus"
+    )
+  ) %>%
+
+  full_join(
+    traits,
+    by = "gen"
+  ) %>%
+
+  distinct() %>%
+
+  ##qc----
+  rename(
+    "DispersalSyndrome" = "Dispersal syndrome"
+  ) #%>%
+
+#   mutate(
[…]

### 2022-03-04 figs/fig2.R

@@ -26,28 +26,33 @@ library(ggpubr)
 #-taxaKernels--Vochysia,Guatteria,Virola
 
 
+#defs----
+
+
 #userInput
 
 
 measure2a <- quote(
-  richness
-)
-yAxisLabel2a <- "Richness"
-
-measure2b <- quote(
   entropy
 )
-yAxisLabel2b <- quote(
-  "Diversity"
+yAxisLabel2a <- quote(
+  "Diversity (H')"
 )
 
-measure2c <- quote(
+measure2b <- quote(
+  richness
+)
+yAxisLabel2b <- quote(
+  "Richness"
+)
+
+measure2e <- quote(
   Vochysia
 )
 taxMetricUnit <- "kg"
-yAxisLabel2c <- quote(
+yAxisLabel2e <- quote(
   glue(
-    measure2c,
+    measure2e,
     " prevalence",
     " (",
     taxMetricUnit,
@@ -55,14 +60,25 @@ yAxisLabel2c <- quote(
   )
 )
 
-measure2d <- quote(
+measure2f <- quote(
   Ficus
 )
-yAxisLabel2d <- quote(
+yAxisLabel2f <- quote(
   glue(
-    measure2d,
+    measure2f,
     " prevalence",
     " (",
+    taxMetricUnit,
+    ")"
+  )
+)
+
+measure2c <- quote(
+  kg17
+)
+yAxisLabel2c <- quote(
+  glue(
+    "Biomass (",
     taxMetricUnit,
     ")"
   )
@@ -82,18 +98,22 @@ xAxisLabel <- quote(
   "Distance to edge (m)"
 )
 statData2ab <- plotResultsTbl1
-statData2cd <- taxaResultsTbl1 %>%
+statData2ef <- taxaResultsTbl1 %>%
   rename(
-    variable = all_of(
+    "variable" = all_of(
       taxRank
     )
   )
-statData2e <- ordTbl
+statData2c <- ordTbl
 
 
 
-#panel
+#panel----
 
+plotResultsTbl1$pval[[2]] <-
+  #bug=doublePosition
+  plotResultsTbl12$pval[[2 * 2]] %>%
+  last()
 graph2a <- statData2ab %>%
   dotGraph(
     measure2a,
@@ -101,12 +121,13 @@ graph2a <- statData2ab %>%
     varForm[[2]],
     xAxisLabel,
     yAxisLabel2a,
-    ..addCenters = T
[…]

### 2022-03-04 figs/fig3.R

@@ -4,8 +4,10 @@ library(
 i_am(
   "figs/fig3.R"
 )
+#userDiff----
 taxRank <- quote(
-  mainDispersal
+  # mainDispersal
+  successionalStage
 )
 #re-run
 #taxaResultsTbl1
@@ -17,15 +19,36 @@ source(
 
 
 
+#plan----
+
+# - biomass x dist for main disp mode 1
+# - biomass x dist for main disp mode 2
+# - disp kernel (y x dist) for main taxon 1
+# - disp kernel (y x dist) for main taxon 2
+
+
+#defs----
+
+
+measure3a <- quote(
+  "NA"
+)
+
 xAxisLabel3 <- quote(
-  "Distance to secondary forest edge (m)"
+  "Distance to edge (m)"
 )
 yAxisLabel3 <- quote(
   "Aboveground biomass (kg)"
 )
-statData3 <- taxaResultsTbl1 %>%
+
+statData3clean <- taxaResultsTbl1 %>%
   unnest(
     "varData"
+  )
+statData3 <- taxaResultsTbl1 %>%
+  rename(
+    #keepConsistent
+    "variable" = mainDispersal
   )
 
 
@@ -34,7 +57,9 @@ theme_set(
   style
 )
 
-fig3 <- statData3 %>%
+#draft----
+
+fig3 <- statData3clean %>%
   # dotGraph()
   ggplot(
     aes(
@@ -65,3 +90,122 @@ fig3 <- statData3 %>%
     )
   )
 fig3
+
+
+#panel----
+
+
+fig3a <- statData3clean %>%
+  filter(
+    is.na(
+      mainDispersal
+    )
+  ) %>%
+
+  # dotGraph()
+  ggplot(
+    aes(
+      x = dist,
+      y = {{
+        taxMetric
+      }}
+    )
+  ) +
+  geom_quasirandom(
+    data = cleanData,
+    color = "gray"
+  ) +
+  geom_quasirandom(
+    shape = 21,
+    fill = "white"
+  ) +
+  labs(
+    x = xAxisLabel3,
+    y = yAxisLabel3
+  ) +
+  scale_y_log10() +
+  # theme(
+  #   legend.position = "top"
+  # ) +
+  # geom_smooth(
+  #   method = "lm",
+  #   se = F
+  # ) +
+  stat_summary(
+    fun.data = "median_mad",
+    position = position_dodge(
+      width = 25
+    )
+  ) +
+  stat_smooth(
[…]

### 2022-03-25 article/abs.Rmd
made fig3

---
title: "Abstract"
output: html_notebook
---

# ATBC *(400 wds)*

## REVIEW CRITERIA
​
All abstracts will be evaluated and scored by the ATBC2022 Scientific Committee. The Scientific Committee may accept, decline, or offer an opportunity for a revised abstract submission. Proposals will be assessed using the following criteria:

- *Scientific merit*: Likelihood of promoting a significant advancement in our understanding of tropical biology and conservation, innovative or interdisciplinary approaches, or likelihood of novel conceptual synthesis.

- *Broader impacts*: Integration of science with conservation issues, impact on policy, and potential utility for conservation, among others.

- *Appeal to participants of ATBC2022*: Abstract should target topics that will be of interest to the ATBC community.

- Relevance to the meeting's theme: We encourage submission of abstracts that are relevant to the meeting’s theme.

- A clearly written and compelling abstract.

# Draft = Habitat edges affect tree diversity more than biomass regeneration in a wet tropical forest

## Introduction / Background / Justification *(50-80 wds)*

About half of all forests are tropical and secondary, making tropical forest regeneration processes integral to the future of forests globally.

However, syntheses find forest regeneration to be quick yet inconsistent--stand biomass and taxonomic richness recovers almost completely in about half a century, but relative abundances may take centuries to recover or instead may diverge entirely.

Given that three quarters of all forests are within a km of a habitat edge, nearby habitat edges also likely affect biomass and taxonomic regeneration trends, as well as management history legacies.

In wet tropical forests specifically, succession tends to depend on light availability, compared to water availability and hydraulic traits like more open or dry forests, which may also extend to mediate edge effects.

## Objective(s)/Hypothesis(es) *(50-80 wds)*

This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge in a former timber plantation context, and
addresses the hypothesis that old plantations facilitate regeneration by favoring shade tolorant or late successional taxa, and that habitat edges affect regenerating community composition.

## Methods *(50-80 wds)*

A 20 ha timber plantation abandoned ca 20 years ago surrounded by primary forest on two sides and a main service road on a third was censused for over a thousand trees using random stratified sampling plots representing 5% of total parcel area along a 300 m edge distance gradient in the Osa Peninsula, SW Costa Rica, and analysis used relevant databases from the literature.

## Results *(50-80 wds)*

As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same.

Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably.

Tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change.

## Implications/Conclusions *(50-80 wds)*

Overall this study supports main ideas about forest regeneration that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future with timber and diversity trade-offs based on local and regional management activity.

## Keywords (10)

forest
secondary
regeneration
edge
biomass
diversity
community
composition
function
trait

### 2022-03-25 analysis/stats.R

@@ -8,9 +8,9 @@ i_am(
 
 #run1----
 
-# source(
-#   here("data/clean.R")
-# )
+source(
+  here("data/clean.R")
+)
 
 # source(
 #   here("analysis/statsTbl.R")

### 2022-03-25 figs/fig3.R

@@ -16,6 +16,9 @@ source(
     "analysis/stats.R"
   )
 )
+library(
+  glue
+)
 
 
 
@@ -45,10 +48,13 @@ statData3clean <- taxaResultsTbl1 %>%
   unnest(
     "varData"
   )
+
 statData3 <- taxaResultsTbl1 %>%
   rename(
     #keepConsistent
-    "variable" = mainDispersal
+
+    # "variable" = mainDispersal
+    "variable" = successionalStage
   )
 
 
@@ -57,49 +63,17 @@ theme_set(
   style
 )
 
-#draft----
-
-fig3 <- statData3clean %>%
-  # dotGraph()
-  ggplot(
-    aes(
-      x = dist,
-      y = {{
-        taxMetric
-      }},
-      color = mainDispersal
-    )
-  ) +
-  geom_quasirandom() +
-  labs(
-    x = xAxisLabel3,
-    y = yAxisLabel3
-  ) +
-  scale_y_log10() +
-  theme(
-    legend.position = "top"
-  ) +
-  # geom_smooth(
-  #   method = "lm",
-  #   se = F
-  # ) +
-  stat_summary(
-    fun.data = "median_mad",
-    position = position_dodge(
-      width = 25
-    )
-  )
-fig3
-
 
 #panel----
 
 
 fig3a <- statData3clean %>%
   filter(
-    is.na(
-      mainDispersal
-    )
+    !is.na(
+      # mainDispersal
+      successionalStage
+    ) &
+      successionalStage == "Both"
   ) %>%
 
   # dotGraph()
@@ -112,7 +86,14 @@ fig3a <- statData3clean %>%
     )
   ) +
   geom_quasirandom(
-    data = cleanData,
+    data = cleanData %>%
+      filter(
+        !is.na(
+          # mainDispersal
+          successionalStage
+        ) &
+          successionalStage == "Both"
+      ),
     color = "gray"
   ) +
   geom_quasirandom(
@@ -127,15 +108,11 @@ fig3a <- statData3clean %>%
   # theme(
   #   legend.position = "top"
   # ) +
-  # geom_smooth(
-  #   method = "lm",
-  #   se = F
-  # ) +
   stat_summary(
-    fun.data = "median_mad",
-    position = position_dodge(
-      width = 25
-    )
+    fun = "median"
+    # position = position_dodge(
+    #   width = 25
+    # )
   ) +
[…]

### 2022-04-14 reviews/outlineFigsJF.Rmd
update figs1,2, map next

---
title: ""
output: html_notebook
---

<!-- # James Fifer, PhD candidate @ Boston Uni -->

- I dont have any major concerns with your figures, just some more minor aesthetic stuff
- map
    - Like I cant read your map, all the text needs to be way bigger
    - if those numbers are important, put them in white boxes or something so you can see them
    - and put a black frame around  the map
    - I dont know how many taxa you have, but you could add cartoons of the taxa to the map fig in boxes around the map
- plots
    - Sometimes your p values overlap with the dots/lines
    - I think your colors for dist are too similar for such a small figure
    - Sometimes you put P => when the p value is greater than 0.10 and sometimes you write the actual value when its greater that 0.10, so I would just choose one way and keep it consistent

<!-- - Im a fan of doing p <0.10 and then writing the p value when its above 0.10 personally -->
<!-- - Because I am more interested in close non-significance than close significance -->
<!-- - *I should rephrase- Im more interested in how close it was if it was non-sig than how close it was if it was sig -->
    - I dont think you need an R2 for your pca, but I would add variance explained by PC1 and PC2

### 2022-04-14 talk/talk.Rmd
made fig3
update figs1,2, map next

@@
   html_notebook
-  # html_document
   # pdf_document #no
+  # word_document
+bibliography: ../../library.bib
 ---
@@
 
-Outline
-===
+# Outline
 
@@
 
-I--Global forests are regenerating, but how?
-===
+# I--Global forests are regenerating, but how?
 
@@
 
-- Regeneration dynamics should be better understood
+- Regeneration dynamics (e.g. associated biodiversity) should be better understood to optimize sustainable management and service provision
 
-I--Regeneration is quick and inconsistent
-===
+# I--Forest regeneration is overall both quick and inconsistent
 
-- Biomass recovers 90% in 66 yrs *(Poorter et al 2016 PNAS)*
+- Biomass recovers 90% in 66 yrs *[@poorter16]*
 
-- Taxonomic richness recovers quickly ()
+- Taxonomic richness recovers quickly *[@rozendaal19]*
 
-- Community composition may take centuries to recover, if at all *(Norder et al 2010 PNAS)*
+- Community composition may take centuries to recover, if at all *[@norden15]*
 
-I--Edges shape regeneration
-===
+# I--Habitat edges shape regeneration
 
-- 70% forests are <1 km from a habitat edge *(Haddad et al 2015 Science)*
+- 70% forests are <1 km from a habitat edge *[@haddad15]*
 
-- Edges effects tend to reduce biodiversity by 13 - 75% via ... *(")*
+- Edge effects reduce biodiversity 13 - 75%
 
-    - stronger in small and isolated fragments
+    - more so in small and isolated fragments
 
-I--Edge effects may differ in wet tropical plantations
-===
+# I--Edge effects may differ in wet tropical plantations
 
@@
 
-- Wet tropical forest regeneration tends to be light-limited, based on tree functional traits *(Poorter et al 2021)*
+- Wet tropical forest regeneration tends to be light-limited, based on tree functional traits *[@poorter21]*
 
@@
 
-I--This study
-===
+# I--This study
 
@@
 
-II--Study site
-===
+# II--Study site
 
@@
 
-II--Census design
-===
+# II--Census design
 
@@
 
-II--Census measurements
-===
+# II--Census measurements
 
@@
 
-II--Statistical analysis
-===
+# II--Statistical analysis
 
@@
 
-- Traits obtained from literature and compared to TRY database *(Kattge et al 2020)*
+- Traits obtained from literature and compared to TRY database *[@kattge19]*
 
-- Computations done with R `getRversion()` depending on *BIOMASS* `packageVersion("BIOMASS")`, 
-*vegan* `packageVersion("vegan")`, and 
-*tidyverse* `packageVersion("tidyverse")` packages
+- Computations done with R `r getRversion()` depending on *BIOMASS* `r packageVersion("BIOMASS")`, 
+*vegan* `r packageVersion("vegan")`, and 
+*tidyverse* `r packageVersion("tidyverse")` packages
 
-III--Stand regeneration
-===
+# III--Stand regeneration
 
@@
 
-III--Composition regeneration
-===
+# III--Composition regeneration
 
@@
 
-        - Vochysia tended to decrease
+        - *Vochysia* tended to decrease
 
-        - Ficus decreased
+        - *Ficus* decreased
 
@@
 
-III--Functional regeneration
-===
+# III--Functional regeneration
 
--
+- Away from forest edge--
 
--
+    - Biomass of taxa often found in both successional stages decreased
 
--
+    - Biomass based on dispersal mode remained constant
 
@@
 
-IV--Hypothesis check
-===
+# IV--Hypothesis check
 
@@
 
+        - Composition tended to change
+
 - Forest edge associated with tree diversity
 
-IV--Processes underlying patterns
-===
+# IV--Processes underlying patterns
 
-- not shade...
+- Not shade, perhaps given high absolute value and/or evenness of original timber planting
 
-- dispersal...
+- Forest edge may associate with faster succession, given drop in generalist tree taxa with distance
 
--
+- No signal of dispersal mode detected
 
-IV--Compare to similar studies
-===
+# IV--Compare to similar studies
 
-- aligns well with major recent papers
+- Aligns well with several recent studies suggesting that--
 
--
+    - Tree diversity regenerates quicker than taxonomic composition *[@poorter16]*
 
--
+    - Habitat edges lower diversity and generally slow regeneration, like via less seed consumption *[@hohlenwerger22]*
 
-IV--Future studies
-===
+    - Taxa indiscriminate to successional stage are minimally affected by disturbance, like from edges *[@bongers09]*
+
+# IV--Future studies
 
@@
 
+# Acknowledgements
+
+- James Fifer for initial review
+
+- Lehigh group Don Morris and undergrads for field help
+
+- Brandeis University for funding
+
+\pagebreak
+
+# References
+
 <!-- For more details on authoring R presentations please visit <https://support.rstudio.com/hc/en-us/articles/200486468>. -->

### 2022-04-14 talk/talk.tex
update figs1,2, map next

@@
 ]{article}
-\title{Natural regeneration of a tropical timber plantation}
-\author{Nicholas Medina}
-\date{}
-
 \usepackage{amsmath,amssymb}
@@
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 \ifLuaTeX
@@
 \fi
+
+\title{Natural regeneration of a tropical timber plantation}
+\author{Nicholas Medina}
+\date{}
 
@@
 \item
-  Regeneration dynamics should be better understood
+  Regeneration dynamics (e.g.~associated biodiversity) should be better
+  understood to optimize sustainable management and service provision
 \end{itemize}
 
-\hypertarget{iregeneration-is-quick-and-inconsistent}{%
-\section{I--Regeneration is quick and
-inconsistent}\label{iregeneration-is-quick-and-inconsistent}}
+\hypertarget{iforest-regeneration-is-overall-both-quick-and-inconsistent}{%
+\section{I--Forest regeneration is overall both quick and
+inconsistent}\label{iforest-regeneration-is-overall-both-quick-and-inconsistent}}
 
@@
 \item
-  Biomass recovers 90\% in 66 yrs \emph{(Poorter et al 2016 PNAS)}
+  Biomass recovers 90\% in 66 yrs \emph{(Poorter et al. 2016)}
 \item
-  Taxonomic richness recovers quickly ()
+  Taxonomic richness recovers quickly \emph{(Rozendaal et al. 2019)}
 \item
   Community composition may take centuries to recover, if at all
-  \emph{(Norder et al 2010 PNAS)}
+  \emph{(Norden et al. 2015)}
 \end{itemize}
 
-\hypertarget{iedges-shape-regeneration}{%
-\section{I--Edges shape regeneration}\label{iedges-shape-regeneration}}
+\hypertarget{ihabitat-edges-shape-regeneration}{%
+\section{I--Habitat edges shape
+regeneration}\label{ihabitat-edges-shape-regeneration}}
 
@@
 \item
-  70\% forests are \textless1 km from a habitat edge \emph{(Haddad et al
-  2015 Science)}
+  70\% forests are \textless1 km from a habitat edge \emph{(Haddad et
+  al. 2015)}
 \item
-  Edges effects tend to reduce biodiversity by 13 - 75\% via \ldots{}
-  \emph{(``)}
+  Edge effects reduce biodiversity 13 - 75\%
 
@@
   \item
-    stronger in small and isolated fragments
+    more so in small and isolated fragments
   \end{itemize}
@@
   Wet tropical forest regeneration tends to be light-limited, based on
-  tree functional traits \emph{(Poorter et al 2021)}
+  tree functional traits \emph{(Poorter et al. 2021)}
 \item
@@
   Traits obtained from literature and compared to TRY database
-  \emph{(Kattge et al 2020)}
+  \emph{(Kattge and Str 2019)}
 \item
-  Computations done with R \texttt{getRversion()} depending on
-  \emph{BIOMASS} \texttt{packageVersion("BIOMASS")}, \emph{vegan}
-  \texttt{packageVersion("vegan")}, and \emph{tidyverse}
-  \texttt{packageVersion("tidyverse")} packages
+  Computations done with R 4.1.3 depending on \emph{BIOMASS} 2.1.7,
+  \emph{vegan} 2.5.7, and \emph{tidyverse} 1.3.1 packages
 \end{itemize}
@@
 \begin{itemize}
-\tightlist
 \item
+  Away from forest edge--
+
+  \begin{itemize}
   \item
+    Biomass of taxa often found in both successional stages decreased
   \item
+    Biomass based on dispersal mode remained constant
+  \end{itemize}
 \end{itemize}
@@
       Diversity decreased
+    \item
+      Composition tended to change
     \end{itemize}
@@
 \item
-  not shade\ldots{}
+  Not shade, perhaps given high absolute value and/or evenness of
+  original timber planting
 \item
-  dispersal\ldots{}
+  Forest edge may associate with faster succession, given drop in
+  generalist tree taxa with distance
 \item
+  No signal of dispersal mode detected
 \end{itemize}
@@
 \item
-  aligns well with major recent papers
+  Aligns well with several recent studies suggesting that--
+
+  \begin{itemize}
   \item
+    Tree diversity regenerates quicker than taxonomic composition
+    \emph{(Poorter et al. 2016)}
   \item
+    Habitat edges lower diversity and generally slow regeneration, like
+    via less seed consumption \emph{(Hohlenwerger, Tambosi, and Metzger
+    2022)}
+  \item
+    Taxa indiscriminate to successional stage are minimally affected by
+    disturbance, like from edges \emph{(Bongers et al. 2009)}
+  \end{itemize}
 \end{itemize}
@@
 
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-bongers09}{}}%
+Bongers, Frans, Lourens Poorter, William D. Hawthorne, and Douglas
+Sheil. 2009. {``The Intermediate Disturbance Hypothesis Applies to
+Tropical Forests, but Disturbance Contributes Little to Tree
+Diversity.''} \emph{Ecology Letters} 12 (8): 798--805.
+\url{https://doi.org/10.1111/j.1461-0248.2009.01329.x}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-haddad15}{}}%
+Haddad, Nick M., Lars A. Brudvig, Jean Clobert, Kendi F. Davies, Andrew
+Gonzalez, Robert D. Holt, Thomas E. Lovejoy, et al. 2015. {``Habitat
+Fragmentation and Its Lasting Impact on {Earth}'s Ecosystems.''}
+\emph{Science Advances} 1 (2): e1500052.
+\url{https://doi.org/10.1126/sciadv.1500052}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hohlenwerger22}{}}%
+Hohlenwerger, Camila, Leandro Reverberi Tambosi, and Jean Paul Metzger.
+2022. {``Forest Cover and Proximity to Forest Affect Predation by
+Natural Enemies in Pasture and Coffee Plantations Differently.''}
+\emph{Agriculture, Ecosystems \& Environment} 333 (August): 107958.
+\url{https://doi.org/10.1016/j.agee.2022.107958}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kattge19}{}}%
+Kattge, Jens, and Hans Knöll Str. 2019. {``{TRY} Plant Trait Database
+\textendash{} Enhanced Coverage and Open Access,''} 70.
+
+\leavevmode\vadjust pre{\hypertarget{ref-norden15}{}}%
+Norden, Natalia, Héctor A. Angarita, Frans Bongers, Miguel
+Martínez-Ramos, Iñigo Granzow-de la Cerda, Michiel van Breugel, Edwin
+Lebrija-Trejos, et al. 2015. {``Successional Dynamics in {Neotropical}
+Forests Are as Uncertain as They Are Predictable.''} \emph{Proceedings
+of the National Academy of Sciences} 112 (26): 8013--18.
+\url{https://doi.org/10.1073/pnas.1500403112}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-poorter16}{}}%
+Poorter, Lourens, Frans Bongers, T. Mitchell Aide, Angélica M. Almeyda
+Zambrano, Patricia Balvanera, Justin M. Becknell, Vanessa Boukili, et
+al. 2016. {``Biomass Resilience of {Neotropical} Secondary Forests.''}
+\emph{Nature} 530 (7589): 211--14.
+\url{https://doi.org/10.1038/nature16512}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-poorter21}{}}%
+Poorter, Lourens, Dylan Craven, Catarina C. Jakovac, Masha T. van der
+Sande, Lucy Amissah, Frans Bongers, Robin L. Chazdon, et al. 2021.
+{``Multidimensional Tropical Forest Recovery.''} \emph{Science} 374
+(6573): 1370--76. \url{https://doi.org/10.1126/science.abh3629}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rozendaal19}{}}%
+Rozendaal, Danaë M. A., Frans Bongers, T. Mitchell Aide, Esteban
+Alvarez-Dávila, Nataly Ascarrunz, Patricia Balvanera, Justin M.
+Becknell, et al. 2019. {``Biodiversity Recovery of {Neotropical}
+Secondary Forests.''} \emph{Science Advances} 5 (3).
+\url{https://doi.org/10.1126/sciadv.aau3114}.
+
+\end{CSLReferences}
+
 \end{document}

### 2022-04-14 analysis/stats.R

@@ -202,7 +202,7 @@ ordStat <- adonis(
 
 
 ordTbl <- commTbl %>%
-  getOrdVarTbl(
+  oir::getOrdVarTbl(
     metaTbl
   )
 

### 2022-04-14 data/clean.R

@@ -153,6 +153,15 @@ updateRaw <- function(
 cleanData <- rawData %>%
   updateRaw() %>%
 
+  mutate(
+    #perHa
+    kg17 = kg17 / (
+      21 * 21
+    ) * (
+      10000 / 1
+    )
+  ) %>%
+
   full_join(
     dispersal
   ) %>%

### 2022-04-14 figs/fig1.R

@@ -6,6 +6,9 @@ i_am(
 )
 #median,dist
 #plotResultsTbl,cleanData
+taxRank <- quote(
+  gen
+)
 source(
   here(
     "analysis/stats.R"
@@ -31,7 +34,12 @@ measure1a <- quote(
   kg17
 )
 yAxisLabel1a <- quote(
-  "Aboveground biomass (kg)"
+  expression(
+    paste(
+      "Biomass (kg ha"^-1,
+      ")"
+    )
+  )
 )
 measure1b <- quote(
   spg
@@ -39,7 +47,7 @@ measure1b <- quote(
 yAxisLabel1b <- quote(
   expression(
     paste(
-      "Wood density (g"^-1,
+      "Wood density (g",
       " cm"^-3,
       ")"
     )
@@ -51,7 +59,7 @@ measure1c <- quote(
 yAxisLabel1c <- quote(
   expression(
     paste(
-      "Stems (441 m"^-2,
+      "Stems (0.04 ha"^-1,
       ")"
     )
   )
@@ -74,7 +82,7 @@ underData <- cleanData
 
 
 
-#panel
+#panel----
 
 graph1a <-  statData %>%
   dotGraph(
@@ -88,13 +96,17 @@ graph1a <-  statData %>%
     ..addCenters = T
   ) +
   scale_y_continuous(
-    trans = "log10"
+    trans = "log10",
+    labels = trans_format(
+      "log10",
+      math_format()
+    )
   )
 graph1a
 
 
 
-#panel
+#panel----
 
 graph1b <- statData %>%
   dotGraph(
@@ -112,7 +124,7 @@ graph1b
 
 
 
-#panel
+#panel----
 
 statData$pval[[9]] <-
   #varsDoubled
@@ -135,7 +147,7 @@ graph1c
 
 
 
-#panel
+#panel----
 
 graph1d <- statData %>%
   dotGraph(
@@ -148,6 +160,10 @@ graph1d <- statData %>%
     ..addCenters = T
   )
 graph1d
+
+
+
+#arrange----
 
 
 
@@ -170,6 +186,12 @@ fig1 <- ggarrange(
   # widths = 0.25,
   # heights = 0.25
 )
+
+  # annotate_figure(
+  #   bottom = "
[…]

### 2022-04-14 figs/fig2.R

@@ -53,9 +53,10 @@ taxMetricUnit <- "kg"
 yAxisLabel2e <- quote(
   glue(
     measure2e,
-    " prevalence",
+    " ",
     " (",
     taxMetricUnit,
+    " ha^-1",
     ")"
   )
 )
@@ -66,9 +67,10 @@ measure2f <- quote(
 yAxisLabel2f <- quote(
   glue(
     measure2f,
-    " prevalence",
+    " ",
     " (",
     taxMetricUnit,
+    " ha^-1",
     ")"
   )
 )
@@ -80,6 +82,7 @@ yAxisLabel2c <- quote(
   glue(
     "Biomass (",
     taxMetricUnit,
+    " ha^-1",
     ")"
   )
 )
@@ -122,7 +125,10 @@ graph2a <- statData2ab %>%
     xAxisLabel,
     yAxisLabel2a,
     ..addCenters = T,
-    ..addCurve = T
+    ..addCurve = T,
+    ..addPxy = c(
+      1, 0.5
+    )
   )
 graph2a
 
@@ -168,7 +174,10 @@ graph2f <- statData2ef %>%
     xAxisLabel,
     yAxisLabel2f,
     ..addCenters = T,
-    ..addLines = T
+    ..addLines = T,
+    ..addPxy = c(
+      0, 40000
+    )
   )
 graph2f
 #Ficus2few...
@@ -192,10 +201,19 @@ graph2d <- ordTbl %>%
     legend.position = "top"
   ) +
   labs(
-    color = "Dist (m)"
+    color = "Dist (m)",
+    x = glue("PC1 (", ordTbl$PC1_prop[1] * 100, "%)"),
+    y = glue("PC2 (", ordTbl$PC2_prop[1] * 100, "%)")
   ) +
-  scale_color_brewer(
-    direction = -1
+  # scale_color_brewer(
+  #   type = "div", #"qual"
+  #   direction = -1
+  # ) +
+  scale_color_manual(
+    values = c(
+      "brown", "green", "blue",
+      "orange", "red", "yellow"
+    )
   ) +
   annotate(
     "text",
@@ -210,26 +228,28 @@ graph2d <- ordTbl %>%
         )
     ),
     x = -5,
-    y = -5
+    y = -5,
+    size = 2
   ) +
   annotate(
     "text",
     label = glue(
-      "P = ",
+      "R2 = ",
       ordStat$aov.tab$R2 %>%
         first() %>%
         round(
[…]

### 2022-04-14 figs/fig3.R

@@ -41,7 +41,10 @@ xAxisLabel3 <- quote(
   "Distance to edge (m)"
 )
 yAxisLabel3 <- quote(
-  "Aboveground biomass (kg)"
+  paste(
+    "Biomass (kg ha^-1",
+    ")"
+  )
 )
 
 statData3clean <- taxaResultsTbl1 %>%
@@ -82,7 +85,8 @@ fig3a <- statData3clean %>%
       x = dist,
       y = {{
         taxMetric
-      }}
+      }},
+      shape = successionalStage
     )
   ) +
   geom_quasirandom(
@@ -91,8 +95,8 @@ fig3a <- statData3clean %>%
         !is.na(
           # mainDispersal
           successionalStage
-        ) &
-          successionalStage == "Both"
+        ) # &
+    #       successionalStage == "Both"
       ),
     color = "gray"
   ) +
@@ -109,7 +113,8 @@ fig3a <- statData3clean %>%
   #   legend.position = "top"
   # ) +
   stat_summary(
-    fun = "median"
+    fun.data = "median_mad",
+    size = 0.25
     # position = position_dodge(
     #   width = 25
     # )
@@ -150,6 +155,35 @@ fig3a <- statData3clean %>%
     y = 1.1,
     hjust = -0.5,
     size = 2
+  ) +
+  theme(
+    legend.position = "top"
+  ) +
+  labs(
+    shape = "Successional stage"
+  ) +
+
+  theme(
+    #allsmaller
+    legend.text = element_text(
+      size = 6
+    ),
+    legend.key.size = unit(
+      0.2, "cm"
+    ),
+    legend.text.align = 0,
+    legend.key.width = unit(
+      0.2, "cm"
+    ),
+    legend.title = element_text(
+      size = 6
+    ),
+    legend.margin = margin(
+      l = -5,
+      t = -5,
+      b = -5,
+      r = -5
+    )
   )
 fig3a
 
@@ -234,7 +268,7 @@ fig3b <- statData3clean %>%
     legend.position = "top",
     #allsmaller
     legend.text = element_text(
-      size = 8
+      size = 6
     ),
     legend.key.size = unit(
       0.2, "cm"
@@ -244,7 +278,7 @@ fig3b <- statData3clean %>%
       0.2, "cm"
     ),
     legend.title = element_text(
-      size = 8
+      size = 6
     ),
     legend.margin = margin(
       l = -5,
[…]

### 2022-04-25 article/abs.Rmd
draft res,fig3 bug

@@
 This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge in a former timber plantation context, and
-addresses the hypothesis that old plantations facilitate regeneration by favoring shade tolorant or late successional taxa, and that habitat edges affect regenerating community composition.
+addresses the hypothesis that old plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and that habitat edges affect regenerating community composition.
 
@@
 trait
+
+# SACNAS
+
+About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
+Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
+Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, medaited by shade tolerance and dispersal traits, along with management history. 
+This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
+An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature. 
+As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same. 
+Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably. 
+Finally, tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change. 
+Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

### 2022-04-25 article/intro.Rmd
draft res,fig3 bug

@@
 output: html_notebook
+bibliography: ../../library.bib
 ---
 
+# Outline
+
 1.  importance of reforestation/regeneration/regrowth
+2.  patterns of reforestation
+    1.  general
+    2.  tropical
+3.  things that affect reforestation
+    1.  edge effects
+    2.  plantation structure?
+4.  [for NGO--start biomass census]
+    1.  Does leaving the plantation structure help reforestation?
+        1.  shade
+        2.  edge effects (i.e. buffer)
+    2.  How does the forest regenerate by distance to primary forest?
+        1.  how much [effort] do we need to re-seed? / how much can we leave to nature?
 
-2.  patterns of reforestation
-
-    1.  general
-
-    2.  tropical
-
-3.  things that affect reforestation
-
-    1.  edge effects
-
-    2.  plantation structure?
-
-4.  [for NGO--start biomass census]
-
-    1.  Does leaving the plantation structure help reforestation?
-
-        1.  shade
-
-        2.  edge effects (i.e. buffer)
-
-    2.  How does the forest regenerate by distance to primary forest?
-
-        1.  how much [effort] do we need to re-seed? / how much can we leave to nature?
+# Draft
 
@@
 
--   Tropical forests store most terrestrial biomass ( @pan2013 ) and biodiversity (CITE, but are losing functions, especially in the neotropics ( @hubau2020 ), as most tropical forests become secondary (\@FAO20.
+    - Tropical forests store most terrestrial biomass *[@pan13]* and biodiversity (CITE, but are losing functions, especially in the neotropics *[@hubau2020]*, as most tropical forests become secondary *[@FAO20]*.
 
@@
 
--   Neotropical secondary forests accumulate biomass \>10x faster than old-growth forests ( @poorter2016 ), and offset \~10% of carbon emissions from Amazonian deforestation ( @smith2020 ).
--   They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y ( @rozendaal2019 ).
--   However, secondary neotropical forests diverge widely in taxonomic composition ( @norden2015 ).
+    - Neotropical secondary forests accumulate biomass >10x faster than old-growth forests ( @poorter2016 ), and offset ~10% of carbon emissions from Amazonian deforestation *[@smith20]*.
+    - They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y *[@rozendaal19]*.
+    - However, secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*.
 
@@
 
-Half of all forests lie half of a kilometer from an ecotone ( @haddad2015 ), suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.
+Half of all forests lie half of a kilometer from an ecotone *[@haddad15]*, suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.
 
--   Tropical forests lose 22% carbon along edges \>100m, in part due to shorter trees and less leaf mass ( @ordway2020 ).
+    - Tropical forests lose 22% carbon along edges >100m, in part due to shorter trees and less leaf mass *[@ordway20]*.
 
@@
 
+# attic
+
 Forest regeneration has the potential to store \#\#C and preserve \#\# diversity globally in the coming decades (UN, Bonn challenge).
@@
 
-Syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years), mainly with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).
+Syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years), mainly with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions *[@rozendaal19; @onyekwelu16]*.
 
@@
 
-Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests (Nordon et al 2015, Christensen 2014, Munroe et al 2013, Chazdon et al 2007).
+Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests *(@nordon15, Christensen 2014, Munroe et al 2013, Chazdon et al 2007)*.
 
@@
 
-Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009),
+Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management *[@wright09]*,
 
-with underlying arguments based on metapopulation theory in ecology (Levins 1969, MacArthur & Wilson 1967), but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.
+with underlying arguments based on metapopulation theory in ecology *[@levins69; @macarthur67]*, but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.
 
@@
 
-with examples of high diversity (compared to older adjacent stands) in agroforests (Oliveira-Neto et al 2017) and native timber plantations (Pryde et al 2015),
+with examples of high diversity (compared to older adjacent stands) in agroforests *[@oliveira-neto17]* and native timber plantations (Pryde et al 2015),
 

### 2022-04-25 article/merge.Rmd
draft res,fig3 bug

---
title: ""
output: 
  html_notebook
  bookdown::html_document2
    number_sections: false
bibliography: ../../library.bib
---

# Abstract

```{r child='abstract.Rmd'}
```

# Intro

```{r child="intro.Rmd"}
```

# Meth

```{r child="methods.Rmd"}
```

# Res

```{r child="results.Rmd"}
```

# Disc

```{r child="disc.Rmd"}
```

# Refs

### 2022-04-25 article/methods.Rmd
draft res,fig3 bug

@@
 title: "Methods"
-output: html_notebook
+output: 
+  # html_notebook
+  bookdown::html_document2:
+    number_sections: false
+bibliography: ../../library.bib
 ---
+
+```{r echo=F, results='hide', warning=F}
+knitr::opts_chunk$set(echo = F, warning=F)
+library(here)
+```
 
@@
 
-**See () for an ecosystem description.**
+(See *@taylor15* for a broader ecosystem description of the region.)
 
@@
 
-This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides *(Fig 1)* and a wide service road on the third NE border, with primary forest beyond it.
+This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides *(Fig \@ref(fig:map))* and a wide service road on the third NE border, with primary forest beyond it.
 
@@
 
-Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E *(Fig 1)*, using GIS software *(ArcGIS, QGIS)*.
+Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E *(Fig \@ref(fig:map))*, using GIS software *(ArcGIS, QGIS)*.
 
@@
 
-The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *(Onyekwelu & Olabiwannu 2016)*.
+The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.
 
-## Plot measurements / Data collection
+```{r map, fig.cap="Map of conservation area on Osa Peninsula, Costa Rica."}
+knitr::include_graphics(here("figs/map.png"))
+# magick::image_read(path = "../figs/map.png") %>% plot()
+```
+<!-- ![**Fig 1**](../figs/map.png) -->
 
-### Canopy cover & Topography
+## Plot measurements 
+<!-- / Data collection -->
+
+### Canopy cover and topography
 
@@
 
-Tree height was recorded by measuring distances to both the crown and the stem at chest height (~2.7 m) using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and trangulating the missing side length.
+Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.
 
@@
 
-Aboveground biomass (AGB) was estimated per stem using the allometric equation by *Chave et al. (2014)*.
+Aboveground biomass (AGB) was estimated per stem using the allometric equation by ... *[@chave15]*.
 
-The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *Chave et al.
-(2005) (Ngomanda et al 2014, Alvarez et al 2012)*.
+The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by ...
+*[@chave05] (Ngomanda et al 2014, Alvarez et al 2012)*.
 
@@
 The equation used was:
-\
-*AGB = 0.0673 x (ρ D^2^ H)^0.976^*
-\
-with *D* in cm, *H* in m, and *p* in g cm^-3^.
+
+*AGB = 0.0673 x (ρ D^2 H)^0.976*
+
+with *D* in cm, *H* in m, and *p* in g cm^-3.
 
@@
 
-Original values from 2013 were later updated to modern values using the pipeline developed in the *BIOMASS* ( `r `packageVersion("BIOMASS")` ) R software package by *Réjou-Méchain et al.
-(2017)*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.
+Original values from 2013 were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by ...
+*[@réjou-méchain17]*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.
 
@@
 
-For all response variables, data were analyzed using plot medians and subject to linear regression with the distance to forest edge variable after checks for residual normality and equal variance assumptions using R `getRversion()` *(R Core Team 2019)*.
+For all response variables, data were analyzed using plot medians and subject to linear regression with the distance to forest edge variable after checks for residual normality and equal variance assumptions using `r version$version.string`.
 
@@
 
-Non-linear regressions were run using the function *poly()* in the *stats* `packageVersion("stats")` R package *()*, 
+Non-linear regressions were run using the function *poly()* in the *stats* `r packageVersion("stats")` R package *(R Core Team 2022)*, 
 which was only recorded for stem density and richness variables.
@@
 
-Community analysis and ordination was done as a PERMANOVA with the *vegan* `packageVersion("vegan")` R package *()*.
+Community analysis and ordination was done as a PERMANOVA with the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
 
 Data and code are available at *nmedina17.github.com/osa*.
+
+<!-- `r citation()` -->
 

### 2022-04-25 article/results.Rmd
draft res,fig3 bug

@@
 title: "Results"
-output: html_notebook
+output: 
+  # html_notebook
+  bookdown::html_document2:
+    number_sections: false
 ---
 
-```{r setup, echo = F, message = F}
-knitr::opts_chunk$set(
-  echo = F,
-  message = F,
-  warning = F
-)
-library(here)
+```{r setup, echo = F, message = F, warning=F, results=F}
+knitr::opts_chunk$set(echo = F, message = F, warning = F)
+library(here); i_am("article/results.Rmd"); source(here("figs/fig2.R"))
 ```
 
-## Biomass
-
-Distance to primary forest edge was associated with ### more ___, although this did not correlate with ___.
+# Biomass
 
@@
 
-```{r fig1}
-source(
-  here(
-    "figs/fig1.R"
-  )
-)
-fig1
+```{r fig1data}
+wd <- filter(plotResultsTbl1, variable == "spg") #alias
+stems <- filter(plotResultsTbl1, variable == "stems") #alias
+stems12 <- filter(plotResultsTbl12, variable == "stems") %>% distinct() #alias
 ```
 
-## Composition
+Aboveground biomass did not tend to change with distance to primary forest edge 
+*(Fig \@ref(fig:biomass)a)*, despite significant trends with underlying related variables 
+*(Fig \@ref(fig:biomass))*.
+
+Median plot wood density tended to increase marginally significantly 
+(`r glue("p=",round(as.numeric(wd$pval),3))`) 
+with edge distance by 
+`r round(wd$statPrint[[1]]$estimate[[2]],4)` ± 
+`r round(wd$statPrint[[1]]$std.error[[2]],5)` per m 
+*(Fig \@ref(fig:biomass)b)*.
+
+Plot stem density also tended to increase significantly 
+(`r glue("p=",round(as.numeric(stems12$pval[[2]][2]),2))`) 
+with edge distance but more strongly and non-linearly, by 
+~`r round(stems12$statPrint[[1]]$estimate[[2]],2)` ± 
+`r round(stems12$statPrint[[1]]$std.error[[2]],2)` per m 
+*(Fig \@ref(fig:biomass)c)*.
+
+At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge 
+*(Fig \@ref(fig:biomass)d)*.
+
+```{r biomass, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) stem density, and (d) canopy light availability."}
+knitr::include_graphics("../figs/fig1.png")
+```
+
+# Composition
 
@@
 
-```{r fig2}
-source(
-  here(
-    "figs/fig2.R"
-  )
-)
-fig1
+```{r fig2data}
+entropy <- filter(plotResultsTbl1, variable == "entropy") #pReplaced
+entropy12 <- filter(plotResultsTbl12, variable == "entropy") %>% distinct() #slopeR2
+richness <- filter(plotResultsTbl1, variable == "richness")
+ficus <- filter(taxaResultsTbl1, gen == "Ficus")
 ```
+
+Overall diversity changed significantly with distance to primary forest edge 
+*(Fig \@ref(fig:taxa))*.
+
+Alpha diversity significantly 
+(`r glue("p=",round(as.numeric(entropy12$pval[[2]][2]),3))`) 
+decreased non-linearly by 
+~`r round(abs(entropy12$statPrint[[1]]$estimate[[2]]),2)` ± 
+`r round(abs(entropy12$statPrint[[1]]$std.error[[2]]),2)` per m 
+as distance to primary forest edge increased 
+which explained 
+`r "R2x100"`% of variance among distance strata medians 
+*(Fig \@ref(fig:taxa)a)* -- 
+while in contrast taxonomic richness increased slightly 
+(`r glue("p=",round(as.numeric(richness$pval),3))`) 
+and linearly 
+by 
+`r round(richness$statPrint[[1]]$estimate[[2]],2)` ± 
+`r round(richness$statPrint[[1]]$std.error[[2]],3)` per m 
+and had `r "R2x100"`% variance explained among distance strata 
+*(Fig \@ref(fig:taxa)b)*.
+
+Community composition and beta diversity also changed significantly 
+(`r glue("p=",round(ordStat$aov.tab[[6]][1],2))`) with 
+`r round(ordStat$aov.tab$R2[1],2) *100`% 
+variance among distance strata explained by distance to primary forest edge, 
+and the first two principal components explaining 
+`r ordTbl$PC1_prop[1] *100`% and 
+`r ordTbl$PC2_prop[1] *100`% totaling 
+`r (ordTbl$PC1_prop[1] + ordTbl$PC2_prop[1]) *100`% 
+of variance among calculated plot distances 
+*(Fig \@ref(fig:taxa)d)*.
+
+Key abundant taxa `r measure2e` and `r measure2f` showed different responses -- `r measure2e` nearly tended to decrease with distance to primary forest edge 
+*(Fig \@ref(fig:taxa)e)*,
+while `r measure2f` decreased marginally significantly 
+(`r glue("p=",round(as.numeric(ficus$pval),3))`) 
+and linearly by 
+`r round(ficus$statPrint[[1]]$estimate[[2]],2)` ± 
+`r round(ficus$statPrint[[1]]$std.error[[2]],2)` per m 
+with `r "R2x100"`% variance among distance strata explained 
+*(Fig \@ref(fig:taxa)f)*.
+
+```{r taxa, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa."}
+knitr::include_graphics("../figs/fig2.png")
+```
+
+# Functions
+
+```{r fig3data}
+source(here("figs/fig3.R")); succession <- taxaResultsTbl12 #pReplaced
+```
+
+Overall functional regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest 
+*(Fig \@ref(fig:traits))*.
+
+Taxa associated with both early and late successional stages decreased significantly 
+(`r glue("p=",round(as.numeric(succession$pval[[3]]),3))`) 
+and curvi-linearly 
+by 
+~`r round(succession$statPrint[[1]]$estimate[[2]],2)` ± 
+`r round(succession$statPrint[[1]]$std.error[[2]],2)` per m 
+with distance to edge explaining 
+`r "R2x100"`% variance among distance strata medians 
+*(Fig \@ref(fig:traits)a)*.
+
+Dispersal modes did not show consistent trends in biomass with increasing distance to edge 
+*(Fig \@ref(fig:traits)b)*.
+
+```{r traits, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode."}
+knitr::include_graphics("../figs/fig3.png")
+```

### 2022-04-25 talk/talk.Rmd
draft res,fig3 bug

@@
 ```{r setup, echo = F, include = F}
-knitr::opts_chunk$set(
-  echo = F,
-  quiet = T
-)
-library(
-  here
-)
-i_am(
-  "talk/talk.Rmd"
-)
+knitr::opts_chunk$set(echo = F, quiet = T)
+library(here); i_am("talk/talk.Rmd")
 # #fixRender
@@
     - more so in small and isolated fragments
+
+- Edge effects may shape diversity via dispersal overlap, possibly peaking at intermediate values *[@yamaguchi22]*
 

### 2022-04-25 data/clean.R

@@ -1,7 +1,5 @@
 library(here) #fromprojdir
-i_am(
-  "data/clean.R"
-)
+i_am("data/clean.R")
 library(vroom) #loadastbl
 #library(skimr) #glance
 library(tidyverse);
@@ -16,9 +14,7 @@ rawData <- vroom(
 )
 
 dispersal <- vroom(
-  here(
-    "data/dispersal.csv"
-  )
+  here("data/dispersal.csv")
 ) %>%
   select(
     fam,
@@ -37,9 +33,7 @@ dispersal <- vroom(
 
 ##traits----
 traits <- vroom(
-  here(
-    "data/base/try/try19707.csv"
-  )
+  here("data/base/try/try19707.csv")
 ) %>% #glimpse()
   select(
     AccSpeciesName,
@@ -80,9 +74,7 @@ traits <- vroom(
 
 ##lifeHist----
 lifeHist <- vroom(
-  here(
-    "data/base/lifeHist.csv"
-  )
+  here("data/base/lifeHist.csv")
 ) %>%
   select(
     family,

### 2022-04-25 figs/fig1.R

@@ -1,19 +1,12 @@
 # refs ----
 
-library(here)
-i_am(
-  "figs/fig1.R"
-)
+library(here); i_am("figs/fig1.R")
 #median,dist
 #plotResultsTbl,cleanData
 taxRank <- quote(
   gen
 )
-source(
-  here(
-    "analysis/stats.R"
-  )
-)
+source(here("analysis/stats.R"))
 #plotResultsTbl
 library(glue)
 #annotate()
@@ -98,8 +91,7 @@ graph1a <-  statData %>%
   scale_y_continuous(
     trans = "log10",
     labels = trans_format(
-      "log10",
-      math_format()
+      "log10", math_format()
     )
   )
 graph1a
@@ -142,6 +134,7 @@ graph1c <- statData %>%
     # ..cleanData = underData,
     ..addCenters = T,
     ..addCurve = T
+   # ..addLines = T
   )
 graph1c
 

### 2022-04-25 figs/fig2.R

@@ -1,17 +1,10 @@
-library(here)
-i_am(
-  "figs/fig2.R"
-)
+library(here); i_am("figs/fig2.R")
 #taxMetric,mainMetric
 #plotResultsTbl1,taxaResultsTbl1
 taxRank <- quote(
   gen
 )
-source(
-  here(
-    "analysis/stats.R"
-  )
-)
+source(here("analysis/stats.R"))
 #plotResultsTbl
 library(glue)
 #annotate()
@@ -157,7 +150,7 @@ graph2e <- statData2ef %>%
     taxForm[[3]],
     taxForm[[2]],
     xAxisLabel,
-    yAxisLabel2e,
+    eval(yAxisLabel2e),
     ..addCenters = T
   ) +
   scale_y_log10()
@@ -172,7 +165,7 @@ graph2f <- statData2ef %>%
     taxForm[[3]],
     taxForm[[2]],
     xAxisLabel,
-    yAxisLabel2f,
+    eval(yAxisLabel2f),
     ..addCenters = T,
     ..addLines = T,
     ..addPxy = c(
@@ -187,11 +180,9 @@ graph2f
 
 graph2d <- ordTbl %>%
   ggplot(
-    aes(
-      x = PC1,
-      y = PC2,
-      color = as_factor(
-        dist
+    aes(x = PC1, y = PC2,
+        color = as_factor(
+          dist
       )
     )
   ) +
@@ -219,13 +210,8 @@ graph2d <- ordTbl %>%
     "text",
     label = glue(
       "P = ",
-      ordStat$
-        aov.tab$
-        `Pr(>F)` %>%
-        first() %>%
-        round(
-          digits = 2
-        )
+      ordStat$aov.tab$`Pr(>F)` %>%
+        first() %>% round(2)
     ),
     x = -5,
     y = -5,
@@ -234,12 +220,10 @@ graph2d <- ordTbl %>%
   annotate(
     "text",
     label = glue(
-      "R2 = ",
+      "R2",
+      " = ",
       ordStat$aov.tab$R2 %>%
-        first() %>%
-        round(
-          digits = 2
-        )
+        first() %>% round(2)
     ),
     x = 5,
     y = 5,
@@ -393,20 +377,20 @@ fig2
 #save----
 
 
-ggsave(
-  "fig2.pdf",
-  fig2,
-  path = "figs",
-  width = 5,
-  height = 3,
-  units = "in"
[…]

### 2022-04-25 figs/fig3.R

@@ -1,9 +1,4 @@
-library(
-  here
-)
-i_am(
-  "figs/fig3.R"
-)
+library(here); i_am("figs/fig3.R")
 #userDiff----
 taxRank <- quote(
   # mainDispersal
@@ -11,14 +6,9 @@ taxRank <- quote(
 )
 #re-run
 #taxaResultsTbl1
-source(
-  here(
-    "analysis/stats.R"
-  )
-)
-library(
-  glue
-)
+source(here("analysis/stats.R"))
+library(glue)
+library(ggpubr)
 
 
 
@@ -41,9 +31,11 @@ xAxisLabel3 <- quote(
   "Distance to edge (m)"
 )
 yAxisLabel3 <- quote(
-  paste(
-    "Biomass (kg ha^-1",
-    ")"
+  expression(
+    paste(
+      "Biomass (kg ha"^-1,
+      ")"
+    )
   )
 )
 
@@ -106,9 +98,15 @@ fig3a <- statData3clean %>%
   ) +
   labs(
     x = xAxisLabel3,
-    y = yAxisLabel3
+    y = eval(yAxisLabel3)
   ) +
-  scale_y_log10() +
+  # scale_y_log10() +
+  scale_y_continuous(
+    trans = "log10",
+    label = trans_format(
+      "log10", math_format()
+    )
+  ) +
   # theme(
   #   legend.position = "top"
   # ) +
@@ -121,21 +119,13 @@ fig3a <- statData3clean %>%
   ) +
   stat_smooth(
     method = "lm",
-    formula = y ~
-      poly(
-        x,
-        2
-      ),
+    formula = y ~ poly(x, 2),
     color = "black",
     size = 0.5,
     se = F
   ) +
   stat_poly_eq(
-    formula = y ~
-      poly(
-        x,
-        2
-      ),
+    formula = y ~ poly(x, 2),
     size = 2,
     label.x = "right",
     label.y = 1.1
@@ -146,10 +136,7 @@ fig3a <- statData3clean %>%
       "P = ",
       taxaResultsTbl12$
         pval[[2 * 2]] %>%
-        last() %>%
-        round(
-          digits = 3
-        )
+        last() %>% round(3)
     ),
     x = 1,
     y = 1.1,
@@ -209,11 +196,7 @@ taxRank <- quote(
 )
 #re-run
 #taxaResultsTbl1
-source(
-  here(
[…]

### 2022-05-15 article/abs.Rmd
revise intro,bibs

@@
 
-# ATBC *(400 wds)*
+<!-- # ATBC *(400 wds)* -->
 
-## REVIEW CRITERIA
-​
-All abstracts will be evaluated and scored by the ATBC2022 Scientific Committee.
+<!-- ## REVIEW CRITERIA -->
+<!-- ​ -->
+<!-- All abstracts will be evaluated and scored by the ATBC2022 Scientific Committee.
 The Scientific Committee may accept, decline, or offer an opportunity for a revised abstract submission.
-Proposals will be assessed using the following criteria:
+Proposals will be assessed using the following criteria:  -->
 
-- *Scientific merit*: Likelihood of promoting a significant advancement in our understanding of tropical biology and conservation, innovative or interdisciplinary approaches, or likelihood of novel conceptual synthesis.
+<!-- - *Scientific merit*: Likelihood of promoting a significant advancement in our understanding of tropical biology and conservation, innovative or interdisciplinary approaches, or likelihood of novel conceptual synthesis.
+-->
 
-- *Broader impacts*: Integration of science with conservation issues, impact on policy, and potential utility for conservation, among others.
+<!-- - *Broader impacts*: Integration of science with conservation issues, impact on policy, and potential utility for conservation, among others.  -->
 
-- *Appeal to participants of ATBC2022*: Abstract should target topics that will be of interest to the ATBC community.
+<!-- - *Appeal to participants of ATBC2022*: Abstract should target topics that will be of interest to the ATBC community.  -->
 
-- Relevance to the meeting's theme: We encourage submission of abstracts that are relevant to the meeting’s theme.
+<!-- - Relevance to the meeting's theme: We encourage submission of abstracts that are relevant to the meeting’s theme. -->
 
-- A clearly written and compelling abstract.
+<!-- - A clearly written and compelling abstract. -->
 
-# Draft = Habitat edges affect tree diversity more than biomass regeneration in a wet tropical forest
+<!-- # Draft = Habitat edges affect tree diversity more than biomass regeneration in a wet tropical forest -->
 
-## Introduction / Background / Justification *(50-80 wds)*
+<!-- ## Introduction / Background / Justification *(50-80 wds)* -->
 
-About half of all forests are tropical and secondary, making tropical forest regeneration processes integral to the future of forests globally.
+<!-- About half of all forests are tropical and secondary, making tropical forest regeneration processes integral to the future of forests globally. -->
 
-However, syntheses find forest regeneration to be quick yet inconsistent--stand biomass and taxonomic richness recovers almost completely in about half a century, but relative abundances may take centuries to recover or instead may diverge entirely.
+<!-- However, syntheses find forest regeneration to be quick yet inconsistent--stand biomass and taxonomic richness recovers almost completely in about half a century, but relative abundances may take centuries to recover or instead may diverge entirely.
+-->
 
-Given that three quarters of all forests are within a km of a habitat edge, nearby habitat edges also likely affect biomass and taxonomic regeneration trends, as well as management history legacies.
+<!-- Given that three quarters of all forests are within a km of a habitat edge, nearby habitat edges also likely affect biomass and taxonomic regeneration trends, as well as management history legacies.
+-->
 
-In wet tropical forests specifically, succession tends to depend on light availability, compared to water availability and hydraulic traits like more open or dry forests, which may also extend to mediate edge effects.
+<!-- In wet tropical forests specifically, succession tends to depend on light availability, compared to water availability and hydraulic traits like more open or dry forests, which may also extend to mediate edge effects.
+-->
 
-## Objective(s)/Hypothesis(es) *(50-80 wds)*
+<!-- ## Objective(s)/Hypothesis(es) *(50-80 wds)* -->
 
-This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge in a former timber plantation context, and
-addresses the hypothesis that old plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and that habitat edges affect regenerating community composition.
+<!-- This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge in a former timber plantation context, and -->
+<!-- addresses the hypothesis that old plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and that habitat edges affect regenerating community composition. -->
 
-## Methods *(50-80 wds)*
+<!-- ## Methods *(50-80 wds)* -->
 
-A 20 ha timber plantation abandoned ca 20 years ago surrounded by primary forest on two sides and a main service road on a third was censused for over a thousand trees using random stratified sampling plots representing 5% of total parcel area along a 300 m edge distance gradient in the Osa Peninsula, SW Costa Rica, and analysis used relevant databases from the literature.
+<!-- A 20 ha timber plantation abandoned ca 20 years ago surrounded by primary forest on two sides and a main service road on a third was censused for over a thousand trees using random stratified sampling plots representing 5% of total parcel area along a 300 m edge distance gradient in the Osa Peninsula, SW Costa Rica, and analysis used relevant databases from the literature.
+-->
 
-## Results *(50-80 wds)*
+<!-- ## Results *(50-80 wds)* -->
 
-As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same.
+<!-- As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same.
+-->
 
-Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably.
+<!-- Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably. -->
 
-Tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change.
+<!-- Tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change. -->
 
-## Implications/Conclusions *(50-80 wds)*
+<!-- ## Implications/Conclusions *(50-80 wds)* -->
 
-Overall this study supports main ideas about forest regeneration that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future with timber and diversity trade-offs based on local and regional management activity.
+<!-- Overall this study supports main ideas about forest regeneration that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future with timber and diversity trade-offs based on local and regional management activity.
+-->
 
-## Keywords (10)
+<!-- ## Keywords (10) -->
 
-forest
-secondary
-regeneration
-edge
-biomass
-diversity
-community
-composition
-function
-trait
+<!-- forest -->
+<!-- secondary -->
+<!-- regeneration -->
+<!-- edge -->
+<!-- biomass -->
+<!-- diversity -->
+<!-- community -->
+<!-- composition -->
+<!-- function -->
+<!-- trait -->
 
-# SACNAS
+<!-- # SACNAS -- Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest -->
+
+<!-- About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests.  -->
+<!-- Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely.  -->
+<!-- Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
+-->
+<!-- This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
+-->
+<!-- An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.
+-->
+<!-- As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same.
+-->
+<!-- Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably.  -->
+<!-- Finally, tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change. -->
+<!-- Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.
+-->
+
+# ESA -- Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest
+
+<!-- # Background/Question/Methods -->
 
@@
 Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
-Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, medaited by shade tolerance and dispersal traits, along with management history. 
+Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
 This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
 An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.
-As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same. 
-Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably. 
-Finally, tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change. 
+
+<!-- # Results/Conclusions -->
+
+As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ca 10% variation explained, while biomass and canopy light tended to stay the same.
+Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably. 
+Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
 Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

### 2022-05-15 article/disc.Rmd
revise intro,bibs

@@
 output: html_notebook
+bibliography: ../../library.bib
+link-citations: TRUE
 ---
@@
 based on surrounding forest proximity.
+
+The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density *[@césar21]*, and could therefore require a switch in focal timber species *[@karsten14]*.

### 2022-05-15 article/intro.Rmd
revise intro,bibs

--- 2022-03-04 article/intro.Rmd
@@
 output: html_notebook
+bibliography: ../../library.bib
+link-citations: TRUE
 ---
 
-1.  importance of reforestation/regeneration/regrowth
+<!-- # Outline -->
 
-2.  patterns of reforestation
+<!-- 1.  importance of reforestation/regeneration/regrowth -->
+<!-- 2.  patterns of reforestation -->
+<!--     1.  general -->
+<!--     2.  tropical -->
+<!-- 3.  things that affect reforestation -->
+<!--     1.  edge effects -->
+<!--     2.  plantation structure? -->
+<!-- 4.  [for NGO--start biomass census] -->
+<!--     1.  Does leaving the plantation structure help reforestation? -->
+<!--         1.  shade -->
+<!--         2.  edge effects (i.e. buffer) -->
+<!--     2.  How does the forest regenerate by distance to primary forest? -->
+<!--         1.  how much [effort] do we need to re-seed? / how much can we leave to nature? -->
 
-    1.  general
+<!-- # Draft -->
 
-    2.  tropical
+Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*, and while tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and regenerating degraded functions *[@hubau20]* amid insufficient policy support *[@chazdon18]*.
+Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
+Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
+They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80% of old-growth forest levels in just 20 years, especially in wetter regions *[@rozendaal19]*.
+The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
+However, secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]*.
+Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
+As a result, understanding secondary forest regeneration remains important...
 
-3.  things that affect reforestation
+Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, as well as land-use history, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
+About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
+Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.
+The traits of existing taxa also depend on land-use history and previous management *[@omeja12]*, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
+Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
+Overall, while edge effects near degraded habitat edges are often negative, meta-population *[@levins79]* and -community *[@leibold04; @warren15]* theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
+Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.
+<!-- role of N-fixing trees... *[@lai18]* -->
 
-    1.  edge effects
+This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
+We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade that would suppress less shade-tolerant pioneer tree taxa and allow for more tree diversity closer to the primary forest edge.
+    <!-- - supported by inference?! -->
+Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.
 
-    2.  plantation structure?
+<!-- # attic -->
 
-4.  [for NGO--start biomass census]
+<!-- Forest regeneration has the potential to store *## C* and preserve *## diversity* globally in the coming decades (UN, Bonn challenge). -->
 
-    1.  Does leaving the plantation structure help reforestation?
+<!-- At local scales, regeneration is strongly affected by landscape [ecological] heterogeneity as well as management history. -->
 
-        1.  shade
+<!-- However, patterns of forest regeneration remain unclear, especially at more local scales (< ~50 ha). -->
 
-        2.  edge effects (i.e. buffer)
+<!-- Syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (~50-80%, compared to old-growth forests) rather quickly (decades, < 20 years), mainly with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions *[@rozendaal19; @onyekwelu16]*.
+-->
 
-    2.  How does the forest regenerate by distance to primary forest?
+<!-- However, while restoring species richness may be enough to maintain some ecosystem functions like carbon storage (Liu et al 2018), it does not guarantee the restoration of species composition.
+-->
 
-        1.  how much [effort] do we need to re-seed? / how much can we leave to nature?
+<!-- Instead, changes to species composition likely last up to centuries,  -->
 
-Global forest and landscape regeneration is a key conservation and climate change mitigation strategy, adopted internationally (\@UN14 Paris, Bonn Challenge).
+<!-- [High] Regional biodiversity [can] affects/increases the uncertainty about the local occurrences and outcomes of specific species interactions, -->
 
--   Tropical forests store most terrestrial biomass ( @pan2013 ) and biodiversity (CITE, but are losing functions, especially in the neotropics ( @hubau2020 ), as most tropical forests become secondary (\@FAO20.
+<!-- such as after dispersal and/or niche overlap [CITE-niche, neutral theory]. -->
 
-Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
+<!-- [At the same time,] Diverse local management activity modifies demographic states/profiles of various taxa's population dynamics. -->
 
--   Neotropical secondary forests accumulate biomass \>10x faster than old-growth forests ( @poorter2016 ), and offset \~10% of carbon emissions from Amazonian deforestation ( @smith2020 ).
--   They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y ( @rozendaal2019 ).
--   However, secondary neotropical forests diverge widely in taxonomic composition ( @norden2015 ).
+<!-- For example, active timber plantations strongly promote the population growth of focal taxa at the expense of non-focal taxa. -->
 
-Divergent communities among secondary forests may ultimately warrant more localized conservation efforts, studying relative effects of generalized community processes.
+<!-- After abandonment, however, the ecosystem likely reflects how continuous regional ecological processes look different given past management. -->
 
-Half of all forests lie half of a kilometer from an ecotone ( @haddad2015 ), suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.
+<!-- such as seed dispersal from nearby primary forest edges. -->
 
--   Tropical forests lose 22% carbon along edges \>100m, in part due to shorter trees and less leaf mass ( @ordway2020 ).
+<!-- [These] Interactions between background ecology and human management ultimately have both local and general implications for the potential conservation/stability of ecosystem services/functions -->
 
-Here we report the edge effects of an abandoned wet neotropical timber plantation.
+<!-- like biomass/carbon storage and biodiversity maintenance, -->
 
--   We hypothesized that remnant timber trees would mediate forest regrowth by maintaining uniform shade that would suppress shade-intolerant / pioneer trees / taxa and allow for more diversity.
+<!-- as well as manager livelihood. -->
 
-    -   supported by inference?!
+<!-- Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests *(@nordon15, Christensen 2014, Munroe et al 2013, Chazdon et al 2007)*.
+-->
 
-Forest regeneration has the potential to store \#\#C and preserve \#\# diversity globally in the coming decades (UN, Bonn challenge).
+<!-- but lagging changes in composition may contribute to reduced effects of diversity on other forest functions (Griess et al 2012). -->
 
-At local scales, regeneration is strongly affected by landscape [ecological] heterogeneity as well as management history.
+<!-- Restoring species composition more quickly is likely a place where active forest managers can have the most unique impacts. -->
 
-However, patterns of forest regeneration remain unclear, especially at more local scales (\< \~50 ha).
-
-Syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years), mainly with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).
-
-However, while restoring species richness may be enough to maintain some ecosystem functions like carbon storage (Liu et al 2018), it does not guarantee the restoration of species composition.
-
-Instead, changes to species composition likely last up to centuries, showing a strong legacy of priority effects from pioneering, early-successional communities that are mostly fast-growing, low-wood-density, and shade-intolerant (Tabarelli et al 2014, Tabarelli & Lopes 2008).
-
-forests show legacy effects of common human activity, like logging and intensive agriculture/agroforestry/silviculture (FAO 2020).
-
-[High] Regional biodiversity [can] affects/increases the uncertainty about the local occurrences and outcomes of specific species interactions,
-
-such as after dispersal and/or niche overlap [CITE-niche, neutral theory].
-
-[At the same time,] Diverse local management activity modifies demographic states/profiles of various taxa's population dynamics.
-
-For example, active timber plantations strongly promote the population growth of focal taxa at the expense of non-focal taxa.
-
-After abandonment, however, the ecosystem likely reflects how continuous regional ecological processes look different given past management.
-
-such as seed dispersal from nearby primary forest edges.
-
-[These] Interactions between background ecology and human management ultimately have both local and general implications for the potential conservation/stability of ecosystem services/functions
-
-like biomass/carbon storage and biodiversity maintenance,
-
-as well as manager livelihood.
-
-Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests (Nordon et al 2015, Christensen 2014, Munroe et al 2013, Chazdon et al 2007).
-
-but lagging changes in composition may contribute to reduced effects of diversity on other forest functions (Griess et al 2012).
-
-Restoring species composition more quickly is likely a place where active forest managers can have the most unique impacts.
-
-For land managers / foresters with interest in conservation, inheriting land affected by past decisions is a reality.
+<!-- For land managers / foresters with interest in conservation, inheriting land affected by past decisions is a reality.
 One of the legacies is where the land was placed.
 To best manage it in the future, an initial assessment of the ecological status of the land is important to understand it's likely trajectory, and what it will need in the future for tailed services/conservation/protection.
+-->
 
-Edge effects are known for degrading primary forest, but they can also be used to assist reforestation.
+<!-- Edge effects are known for degrading primary forest, but they can also be used to assist reforestation. -->
 
-Traditional edge effects act by mortality of large trees that makes room for pioneer species to enter, but existing interior pioneer species also have dispersal radii that overlap with the adjacent exterior, potentially facilitating regeneration.
+<!-- Traditional edge effects act by mortality of large trees that makes room for pioneer species to enter, but existing interior pioneer species also have dispersal radii that overlap with the adjacent exterior, potentially facilitating regeneration.
+-->
 
-Encroachment vs. facilitation relies on a balance between mortality (per area) and dispersal + establishment + stem growth.
+<!-- Encroachment vs. facilitation relies on a balance between mortality (per area) and dispersal + establishment + stem growth. -->
 
-So pioneer species can be helpful for reforestation though they are typically seen as bad for old forest conservation.
+<!-- So pioneer species can be helpful for reforestation though they are typically seen as bad for old forest conservation. -->
 
-Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009),
+<!-- Especially for abandoned land that regenerates via various processes (Stanturf et al 2014), the proximity to less disturbed habitat can affect the composition of the regrowing community by changing effective seed dispersal kernels and probability of establishment (Wandrag et al 2017, Guariguata & Ostertag 2001), along with animal disperser communities themselves (Gilroy et al 2014).
+-->
 
-with underlying arguments based on metapopulation theory in ecology (Levins 1969, MacArthur & Wilson 1967), but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.
+<!-- Edge effects can also operate via the style of management, including logging decisions (Edwards et al 2014, Clark et al 2012) and land-use history generally (Omeja et al 2012), can also affect stand diversity maintenance, -->
 
-Especially for abandoned land that regenerates via various processes (Stanturf et al 2014), the proximity to less disturbed habitat can affect the composition of the regrowing community by changing effective seed dispersal kernels and probability of establishment (Wandrag et al 2017, Guariguata & Ostertag 2001), along with animal disperser communities themselves (Gilroy et al 2014).
+<!-- possibly by changing early tree survival probabilities, growth rates, and/or competition, -->
 
-While dispersal is complicated (Muller-Landau & Hardesty 2005), it can broadly correlate with simpler plant traits such as height and seed mass (Tamme et al 2014, Thomson et al 2011).
+<!-- but surprisingly less likely facilitation by nitrogen-fixing trees (Lai et al 2018). -->
 
-Edge effects can also operate via the style of management, including logging decisions (Edwards et al 2014, Clark et al 2012) and land-use history generally (Omeja et al 2012), can also affect stand diversity maintenance,
+<!-- It may also be that due to both high recruitment and mortality (turnover) of pioneering species' stems, transitions from early- to late-successional communities may be more complex than gradual (van Greugel et al 2007).
+-->
 
-with examples of high diversity (compared to older adjacent stands) in agroforests (Oliveira-Neto et al 2017) and native timber plantations (Pryde et al 2015),
+<!-- To understand the ecological principles behind secondary forest regeneration (Hulshoff & Powers 2019), here we study the regeneration of an abandoned timber plantation as a function of distance from surrounding native forest.
+-->
 
-possibly by changing early tree survival probabilities, growth rates, and/or competition,
+<!-- -   how can we expect it to regrow? -->
+<!-- -   how far/much [effort] do we need to re-seed? -->
 
-but surprisingly less likely facilitation by nitrogen-fixing trees (Lai et al 2018).
+<!-- We hypothesized that planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity.
+-->
 
-It may also be that due to both high recruitment and mortality (turnover) of pioneering species' stems, transitions from early- to late-successional communities may be more complex than gradual (van Greugel et al 2007).
-
-The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density (César et al 2021), and could therefore require a switch in focal timber species (Karsten et al 2014).
-
-To understand the ecological principles behind secondary forest regeneration (Hulshoff & Powers 2019), here we study the regeneration of an abandoned timber plantation as a function of distance from surrounding native forest.
-
--   how can we expect it to regrow?
--   how far/much [effort] do we need to re-seed?
-
-We hypothesized that planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity.
-
-We also considered how the legacy plantation canopy structure mediated forest stand effects via changes in abiotic conditions like light availability and microtopography (Powers & Peréz-Aviles 2012).
+<!-- We also considered how the legacy plantation canopy structure mediated forest stand effects via changes in abiotic conditions like light availability and microtopography (Powers & Peréz-Aviles 2012).
+-->

### 2022-05-15 article/merge.Rmd
revise intro,bibs

@@
 title: ""
+author: 
+  name: Nicholas Medina
+  affilitation: Brandeis University
+address:
+  code: Brandeis University
+  address: 415 South St, Waltham, MA 02453
 output: 
-  html_notebook
-  bookdown::html_document2
-    number_sections: false
+  # html_notebook
+  bookdown::html_document2:
+    number_sections: FALSE
 bibliography: ../../library.bib
+link-citations: TRUE
 ---
@@
 
-```{r child='abstract.Rmd'}
+```{r child='abs.Rmd'}
 ```
 
-# Intro
+# Introduction
 
@@
 
-# Meth
+# Methods
 
@@
 
-# Res
+# Results
 
-```{r child="results.Rmd"}
-```
+<!-- # ```{r child="results.Rmd"} -->
+<!-- # ``` -->
 
-# Disc
+# Discussion
 
@@
 
-# Refs
+# References

### 2022-05-15 article/methods.Rmd
revise intro,bibs

@@
   bookdown::html_document2:
-    number_sections: false
+    number_sections: FALSE
 bibliography: ../../library.bib
+link-citations: TRUE
 ---
 
-```{r echo=F, results='hide', warning=F}
-knitr::opts_chunk$set(echo = F, warning=F)
-library(here)
+```{r echo=F, results='hide', warning=F, message=F}
+knitr::opts_chunk$set(echo = F, warning = F, message = F); here::i_am("article/methods.Rmd")
 ```
@@
 
-This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center **(GPS)** near the Piro Biological Station run by Osa Conservation.
-
+This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx.
+`r RgoogleMaps::getGeoCode("cabo matapalo")`) near the Piro Biological Station run by Osa Conservation.
 (See *@taylor15* for a broader ecosystem description of the region.)
-
-The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the dry-adapted species from the Guanacaste region *(Pérez Cordero & Kanninen 2002)* grew poorly in very wet conditions.
-
-This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides *(Fig \@ref(fig:map))* and a wide service road on the third NE border, with primary forest beyond it.
-
+The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the dry-adapted species from the Guanacaste region *[@perezcordero02; @hulshof20]* grew poorly in very wet conditions.
+This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \@ref(fig:map)) and a wide service road on the third NE border, with primary forest beyond it.
 This census was done in 2013 during the rainy season months between June and August.
@@
 
-Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E *(Fig \@ref(fig:map))*, using GIS software *(ArcGIS, QGIS)*.
-
+Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10 from 2010*, [esri.com](esri.com), and *QGIS 2 from 2013*, [qgis.org](qgis.org)).
 Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
-
 The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.
@@
 ```{r map, fig.cap="Map of conservation area on Osa Peninsula, Costa Rica."}
-knitr::include_graphics(here("figs/map.png"))
+knitr::include_graphics(here::here("figs/map.png"))
 # magick::image_read(path = "../figs/map.png") %>% plot()
@@
 The diameter of all stems >10 cm wide were recorded in each census plot.
-
 Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
-
 Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.
-
 Taxa were identified with localized knowledge and field guides matching common names to Latin binomials, 
@@
 
-Aboveground biomass (AGB) was estimated per stem using the allometric equation by ... *[@chave15]*.
-
-The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by ...
-*[@chave05] (Ngomanda et al 2014, Alvarez et al 2012)*.
-
-The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *(Khan et al 2020, Domke et al 2012)*.
-
-Wood density (or specific gravity) values when available also improve biomass estimates even more than height *(van Breugel et al 2011)*.
-
-The equation used was:
-
-*AGB = 0.0673 x (ρ D^2 H)^0.976*
-
-with *D* in cm, *H* in m, and *p* in g cm^-3.
-
-Measured DBHs and heights were used and wood densities (*ρ*) extracted from available literature.
-
-Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be 0.58 following World Agroforestry Database quidelines *(worldagroforestry.org)*.
+Aboveground biomass (AGB) was estimated per stem using the allometric equation by *@chave15*.
+The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *@chave05 [@alvarez12; @ngomanda14]*.
+The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *[@domke12; @khan20]*.
+Wood density (or specific gravity) values when available also improve biomass estimates even more than height *[@vanbreugel11]*.
+The equation used was: $$ {AGB} = 0.0673 (ρ D^2 H)^{0.976} $$ with diameter *D* in cm, height *H* in m, and wood density or specific gravity *p* in *g cm^-3^*.
+DBHs and heights were measured in the field and wood densities were extracted from available literature. 
+Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org).
 
@@
 
-Original values from 2013 were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by ...
-*[@réjou-méchain17]*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.
+Original values found and recorded were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.
 
@@
 which was only recorded for stem density and richness variables. 
-
 All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
-
 Community analysis and ordination was done as a PERMANOVA with the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
-
-Data and code are available at *nmedina17.github.com/osa*.
+Data and code are available at [nmedina17.github.com/osa](nmedina17.github.com/osa).
 

### 2022-05-15 article/results.Rmd
revise intro,bibs

@@
 `r round(wd$statPrint[[1]]$std.error[[2]],5)` per m 
+explaining 
+`r "R2x100"`% of variance 
+among median distance strata values
 *(Fig \@ref(fig:biomass)b)*.
@@
 which explained 
-`r "R2x100"`% of variance among distance strata medians 
+`r "R2x100"`% of variance 
+among distance strata medians 
 *(Fig \@ref(fig:taxa)a)* -- 
@@
 `r round(richness$statPrint[[1]]$std.error[[2]],3)` per m 
-and had `r "R2x100"`% variance explained among distance strata 
+and had `r "R2x100"`% variance explained 
+among distance strata 
 *(Fig \@ref(fig:taxa)b)*.
@@
 with distance to edge explaining 
-`r "R2x100"`% variance among distance strata medians 
+`r "R2x100"`% variance 
+among distance strata medians 
 *(Fig \@ref(fig:traits)a)*.

### 2022-05-15 article/abs.Rmd
draft abs esp

@@
 Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.
+
+# Esp
+
+Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
+Biomasa arboreal y *riqueza* taxonomica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
+Dado que la mayoría de bosques están entre un km de su orilla, *efectos de orilla* probablemente afectan la regeneración de composición de comnuidad, mediado por tolerancia a sombra y "traits"/characterísticas de *disperso*, y también historia de manejamiento.
+Este estudio *investiga* cómo regeneración de bosque secundario tropical húmedo es afectado por distancia a *orilla de habitat*, y *"hypothesizes"* que plantaciones de madera viejos facilitan regeneración porque favoran taxa tolerante a sombra o de *"succession"* tarde, y que orillas de bosque cercanos/*"adjacent"* *median* composición de comunidad.
+Una plantación de madera neotropical *al lado de* bosque primario fue *"censused"* por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combino taxa con characterísticas/*"traits"* usando literatura relevante.
+
+Mientras distancia de bosque primario aumentó, densidad de madera y *"stems"* *"tended to"* aumentó *signifantemente*, con 10% variación explicado, mientras tanto biomasa y luz de dócel *"tended to"* quedarse igual.
+*Riquesa* de árboles también *"tended to"* aumentar *significantemente*, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
+Finalmente, taxa de árboles asociado *etapas* de *"succession"* ambos temprano y tarde  bajó *significativamente*, pero biomasa por modo de *disperso* no *"tended to"* cambiar.
+*Entre todo* este estudio apoya que composición es menos resiliente y más sujeto a *efectos de orilla* que biomasa y *riqueza*, sugeriendo que bosques globales probablemente serán *asemblajes* nuevos distintos en el futuro, con *"trade-offs"* entre producción de madera y biodiversidad *ocuriendo* basado en actividad de manejamiento local y regional.

### 2022-05-15 article/methods.Rmd
draft abs esp

@@
 
-For all response variables, data were analyzed using plot medians and subject to linear regression with the distance to forest edge variable after checks for residual normality and equal variance assumptions using `r version$version.string`.
-
+For all response variables, data were analyzed using plot medians and subject to linear regression with distance to forest edge as the only independent variable, after checks for residual normality and equal variance assumptions, using `r version$version.string`.
  <!-- including biomass converted to a per hectare basis,  -->
@@
 <!-- Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.  -->
-
 Non-linear regressions were run using the function *poly()* in the *stats* `r packageVersion("stats")` R package *(R Core Team 2022)*, 
@@
 All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
-Community analysis and ordination was done as a PERMANOVA with the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
+Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
 Data and code are available at [nmedina17.github.com/osa](nmedina17.github.com/osa).

### 2022-05-15 article/results.Rmd
draft abs esp

--- 2022-03-04 article/results.Rmd
@@
 title: "Results"
-output: html_notebook
+output: 
+  # html_notebook
+  bookdown::html_document2:
+    number_sections: FALSE
 ---
 
-```{r setup, echo = F, message = F}
-knitr::opts_chunk$set(
-  echo = F,
-  message = F,
-  warning = F
-)
-library(here)
+```{r setup, echo = F, message = F, warning=F, results=F}
+knitr::opts_chunk$set(echo = F, message = F, warning = F)
+here::i_am("article/results.Rmd"); source(here::here("figs/fig2.R"))
 ```
 
-## Biomass
-
-Distance to primary forest edge was associated with ### more ___, although this did not correlate with ___.
+# Biomass
 
@@
 
-```{r fig1}
-source(
-  here(
-    "figs/fig1.R"
-  )
-)
-fig1
+```{r fig1data}
+wd <- filter(plotResultsTbl1, variable == "spg") #alias
+stems <- filter(plotResultsTbl1, variable == "stems") #alias
+stems12 <- filter(plotResultsTbl12, variable == "stems") %>% distinct() #alias
 ```
 
-## Composition
+Aboveground biomass did not tend to change with distance to primary forest edge 
+(Fig \@ref(fig:biomass)a), despite significant trends with underlying related variables 
+(Fig \@ref(fig:biomass)).
+
+Median plot wood density tended to increase marginally significantly 
+(`r glue("p=",round(as.numeric(wd$pval),3))`) 
+with edge distance by 
+`r round(wd$statPrint[[1]]$estimate[[2]],4)` ± 
+`r round(wd$statPrint[[1]]$std.error[[2]],5)` per m 
+explaining 
+`r "R2x100"`% of variance 
+among median distance strata values
+(Fig \@ref(fig:biomass)b).
+
+Plot stem density also tended to increase significantly 
+(`r glue("p=",round(as.numeric(stems12$pval[[2]][2]),2))`) 
+with edge distance but more strongly and non-linearly, by 
+~`r round(stems12$statPrint[[1]]$estimate[[2]],2)` ± 
+`r round(stems12$statPrint[[1]]$std.error[[2]],2)` per m 
+(Fig \@ref(fig:biomass)c).
+
+At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge 
+(Fig \@ref(fig:biomass)d).
+
+```{r biomass, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) stem density, and (d) canopy light availability."}
+knitr::include_graphics(here::here("../figs/fig1.png"))
+```
+
+# Composition
 
@@
 
-```{r fig2}
-source(
-  here(
-    "figs/fig2.R"
-  )
-)
-fig1
+```{r fig2data}
+entropy <- filter(plotResultsTbl1, variable == "entropy") #pReplaced
+entropy12 <- filter(plotResultsTbl12, variable == "entropy") %>% distinct() #slopeR2
+richness <- filter(plotResultsTbl1, variable == "richness")
+ficus <- filter(taxaResultsTbl1, gen == "Ficus")
 ```
+
+Overall diversity changed significantly with distance to primary forest edge 
+(Fig \@ref(fig:taxa)).
+
+Alpha diversity significantly 
+(`r glue("p=",round(as.numeric(entropy12$pval[[2]][2]),3))`) 
+decreased non-linearly by 
+~`r round(abs(entropy12$statPrint[[1]]$estimate[[2]]),2)` ± 
+`r round(abs(entropy12$statPrint[[1]]$std.error[[2]]),2)` per m 
+as distance to primary forest edge increased 
+which explained 
+`r "R2x100"`% of variance 
+among distance strata medians 
+(Fig \@ref(fig:taxa)a) -- 
+while in contrast taxonomic richness increased slightly 
+(`r glue("p=",round(as.numeric(richness$pval),3))`) 
+and linearly 
+by 
+`r round(richness$statPrint[[1]]$estimate[[2]],2)` ± 
+`r round(richness$statPrint[[1]]$std.error[[2]],3)` per m 
+and had `r "R2x100"`% variance explained 
+among distance strata 
+(Fig \@ref(fig:taxa)b).
+
+Community composition and beta diversity also changed significantly 
+(`r glue("p=",round(ordStat$aov.tab[[6]][1],2))`) with 
+`r round(ordStat$aov.tab$R2[1],2) *100`% 
+variance among distance strata explained by distance to primary forest edge, 
+and the first two principal components explaining 
+`r ordTbl$PC1_prop[1] *100`% and 
+`r ordTbl$PC2_prop[1] *100`% totaling 
+`r (ordTbl$PC1_prop[1] + ordTbl$PC2_prop[1]) *100`% 
+of variance among calculated plot distances 
+(Fig \@ref(fig:taxa)d).
+
+Key abundant taxa `r measure2e` and `r measure2f` showed different responses -- `r measure2e` nearly tended to decrease with distance to primary forest edge 
+(Fig \@ref(fig:taxa)e),
+while `r measure2f` decreased marginally significantly 
+(`r glue("p=",round(as.numeric(ficus$pval),3))`) 
+and linearly by 
+`r round(ficus$statPrint[[1]]$estimate[[2]],2)` ± 
+`r round(ficus$statPrint[[1]]$std.error[[2]],2)` per m 
+with `r "R2x100"`% variance among distance strata explained 
+(Fig \@ref(fig:taxa)f).
+
+```{r taxa, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa."}
+knitr::include_graphics(here::here("../figs/fig2.png"))
+```
+
+# Functions
+
+```{r fig3data}
+source(here::here("figs/fig3.R")); succession <- taxaResultsTbl12 #pReplaced
+```
+
+Overall functional regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest 
+(Fig \@ref(fig:traits)).
+
+Taxa associated with both early and late successional stages decreased significantly 
+(`r glue("p=",round(as.numeric(succession$pval[[3]]),3))`) 
+and curvi-linearly 
+by 
+~`r round(succession$statPrint[[1]]$estimate[[2]],2)` ± 
+`r round(succession$statPrint[[1]]$std.error[[2]],2)` per m 
+with distance to edge explaining 
+`r "R2x100"`% variance 
+among distance strata medians 
+(Fig \@ref(fig:traits)a).
+
+Dispersal modes did not show consistent trends in biomass with increasing distance to edge 
+(Fig \@ref(fig:traits)b).
+
+```{r traits, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode."}
+knitr::include_graphics(here::here("../figs/fig3.png"))
+```

### 2022-05-15 analysis/stats.R

@@ -1,11 +1,11 @@
 # Refs----
 
 
-library(here); i_am("analysis/stats.R")
+here::i_am("analysis/stats.R")
 
 #run1----
 
-source(here("data/clean.R"))
+source(here::here("data/clean.R"))
 
 # source(
 #   here("analysis/statsTbl.R")
@@ -188,7 +188,7 @@ metaTbl <- cleanData %>%
 
 ordModel <- distMat ~ dist
 
-ordStat <- adonis(
+ordStat <- vegan::adonis2(
   ordModel,
   metaTbl,
   # 99999

### 2022-05-15 figs/fig2.R

@@ -1,10 +1,10 @@
-library(here); i_am("figs/fig2.R")
+here::i_am("figs/fig2.R")
 #taxMetric,mainMetric
 #plotResultsTbl1,taxaResultsTbl1
 taxRank <- quote(
   gen
 )
-source(here("analysis/stats.R"))
+source(here::here("analysis/stats.R"))
 #plotResultsTbl
 library(glue)
 #annotate()
@@ -355,7 +355,7 @@ graph2c
 #fullfig----
 
 
-fig2 <- ggarrange(
+fig2 <- ggpubr::ggarrange(
   graph2a,
   graph2b,
   graph2c,

### 2022-06-18 article/abs.Rmd
minors, prep disc

@@
 
-# ESA -- Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest
+<!-- # ESA -- Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest -->
 
@@
 <!-- # Results/Conclusions -->
-
 As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ca 10% variation explained, while biomass and canopy light tended to stay the same.
@@
 
-# Esp
+\newpage
+
+# Abstracto{-}
 

### 2022-06-18 article/disc.Rmd
minors, prep disc

@@
 title: "Discussion"
-output: html_notebook
+output: 
+  # html_notebook
+  bookdown::html_document2:
+    # number_sections: FALSE
 bibliography: ../../library.bib
@@
 
+# outline{-}
 1.  assess hypotheses
@@
 4.  recommend future conservation approaches, ecological studies
+
+# draft{-}
+
+This study found significant edge effects on tree diversity and community composition, but not aboveground biomass...functional diversity?
+These results were shown in Figs \@ref(fig:biomass), \@ref(fig:taxa), and \@ref(fig:traits)...
+It was initially hypothesized that community regeneration would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential -- evidence from this study's results supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
+...
+
+Aboveground biomass represents aboveground carbon stores... is a measure of ..., and responds variably to secondary succession and edge effects
+
+# attic{-}
 

### 2022-06-18 article/intro.Rmd
minors, prep disc

@@
 
-Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*, and while tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and regenerating degraded functions *[@hubau20]* amid insufficient policy support *[@chazdon18]*.
+Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*, and while tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and regenerating degraded functions *[@hubau20]* amid increasing climate stressors *[@anderegg22]* yet insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
 Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
@@
 The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
-However, secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]*.
+However, secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
 Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.

### 2022-06-18 article/merge.Rmd
minors, prep disc

--- 2022-04-25 article/merge.Rmd
@@
 ---
-title: ""
+title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
+author: 
+  - Nicholas Medina ^a^*
+  - Max Villalobos ^b^
+  - Andrea Johnson ^b^
+date: scriptsize\
+  ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
+# author: 
+#   - name: "Nicholas Medina"
+#     email: nmedina@umich.edu
+#     affiliation: Brandeis University
+#     orcid_id: 0000-0001-5465-3988
+# address:
+#   - code: Brandeis University
+#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
+# keywords: [1, 2, 3]
+
 output: 
-  html_notebook
-  bookdown::html_document2
-    number_sections: false
-bibliography: ../../library.bib
+  # html_notebook
+  # bookdown::html_document2:
+  #   # df_print: paged
+  #   global_numbering: TRUE
+  #   toc: FALSE
+  bookdown::pdf_document2:
+    # number_sections: FALSE
+    global_numbering: TRUE
+    keep_tex: TRUE
+    toc: FALSE
+bibliography: 
+  - ../../library.bib
+  - ../../grateful-refs.bib
+link-citations: TRUE
+csl: ../../styles-master/elsevier-harvard.csl
+
+fontsize: 12pt
+geometry: margin=1in
+header-includes:
+  - \usepackage[left]{lineno}
+  - \linenumbers
+  - \usepackage{setspace}\doublespace
+# abstract
 ---
 
-# Abstract
+journals: J [Appl] Veg Sci,
 
-```{r child='abstract.Rmd'}
+\
+\
+
+**Keywords**: secondary forest; regeneration; timber plantation associated biodiversity; edge effect; aboveground biomass; shade
+
+\
+\
+
+***Corresponding author**: Nicholas Medina, nmedina@umich.edu
+
+\newpage
+
+# Abstract{-}
+
+```{r child='abs.Rmd'}
 ```
 
-# Intro
+\newpage
+
+# Introduction
 
@@
 
-# Res
+# Results
 
@@
 
-# Disc
+# Discussion
 
@@
 
-# Refs
+\newpage
+
+# Funding{-}
+
+This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship.
+
+# Declaration of interests{-}
+
+Authors declare no conflicts of interest.
+
+# Acknowledgements{-}
+
+Thanks to Professor Dan Perlman for initial networking; Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; concurrent Osa Conservation General Manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; and previous anonymous reviewers and peer James Fifer for discussion of drafts.
+
+# Author contributions{-}
+
+MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis; NM contributed to initial draft writing and later revisions.
+
+# Data statement{-}
+
+Data and code stored at [github.com/nmedina17/osa](github.com/nmedin17/osa).
+
+\newpage
+
+# References

### 2022-06-18 article/merge.tex
minors, prep disc

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
  pdfauthor={Nicholas Medina a*; Max Villalobos b; Andrea Johnson b},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{setspace}\doublespace
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
\author{Nicholas Medina \textsuperscript{a}* \and Max Villalobos \textsuperscript{b} \and Andrea Johnson \textsuperscript{b}}
\date{scriptsize~\textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}

\begin{document}
\maketitle

journals: J {[}Appl{]} Veg Sci,

\hfill\break
\hfill\break

\textbf{Keywords}: secondary forest; regeneration; timber plantation associated biodiversity; edge effect; aboveground biomass; shade

\hfill\break
\hfill\break

*\textbf{Corresponding author}: Nicholas Medina, \href{mailto:nmedina@umich.edu}{\nolinkurl{nmedina@umich.edu}}

\newpage

\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests.
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely.
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.

As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ca 10\% variation explained, while biomass and canopy light tended to stay the same.
Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably.
Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

\newpage

\hypertarget{abstracto}{%
\section*{Abstracto}\label{abstracto}}
\addcontentsline{toc}{section}{Abstracto}

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa arboreal y \emph{riqueza} taxonomica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques están entre un km de su orilla, \emph{efectos de orilla} probablemente afectan la regeneración de composición de comnuidad, mediado por tolerancia a sombra y ``traits''/characterísticas de \emph{disperso}, y también historia de manejamiento.
Este estudio \emph{investiga} cómo regeneración de bosque secundario tropical húmedo es afectado por distancia a \emph{orilla de habitat}, y \emph{``hypothesizes''} que plantaciones de madera viejos facilitan regeneración porque favoran taxa tolerante a sombra o de \emph{``succession''} tarde, y que orillas de bosque cercanos/\emph{``adjacent''} \emph{median} composición de comunidad.
Una plantación de madera neotropical \emph{al lado de} bosque primario fue \emph{``censused''} por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combino taxa con characterísticas/\emph{``traits''} usando literatura relevante.

Mientras distancia de bosque primario aumentó, densidad de madera y \emph{``stems''} \emph{``tended to''} aumentó \emph{signifantemente}, con 10\% variación explicado, mientras tanto biomasa y luz de dócel \emph{``tended to''} quedarse igual.
\emph{Riquesa} de árboles también \emph{``tended to''} aumentar \emph{significantemente}, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
Finalmente, taxa de árboles asociado \emph{etapas} de \emph{``succession''} ambos temprano y tarde bajó \emph{significativamente}, pero biomasa por modo de \emph{disperso} no \emph{``tended to''} cambiar.
\emph{Entre todo} este estudio apoya que composición es menos resiliente y más sujeto a \emph{efectos de orilla} que biomasa y \emph{riqueza}, sugeriendo que bosques globales probablemente serán \emph{asemblajes} nuevos distintos en el futuro, con \emph{``trade-offs''} entre producción de madera y biodiversidad \emph{ocuriendo} basado en actividad de manejamiento local y regional.

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Forest and landscape restoration is a key international conservation and climate change adaptation strategy \emph{(\protect\hyperlink{ref-depinto20}{De Pinto et al., 2020})}, and while tropical forests specifically store most land biomass \emph{(\protect\hyperlink{ref-pan13}{Pan et al., 2013})}, most are now secondary \emph{(\protect\hyperlink{ref-fao20}{FAO, 2020})} and regenerating degraded functions \emph{(\protect\hyperlink{ref-hubau20}{Hubau et al., 2020})} amid increasing climate stressors \emph{(\protect\hyperlink{ref-anderegg22}{Anderegg et al., 2022})} yet insufficient policy support \emph{(\protect\hyperlink{ref-chazdon18}{Chazdon, 2018})} via skewed priorities \emph{(\protect\hyperlink{ref-pyron22}{Pyron and Mooers, 2022})}.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass \textgreater10x faster than old-growth forests \emph{(\protect\hyperlink{ref-finegan96}{Finegan, 1996}; \protect\hyperlink{ref-poorter16}{Poorter et al., 2016})}, and substantially offset carbon emissions, with estimates ranging from \textasciitilde10\% from Amazonian deforestation \emph{(\protect\hyperlink{ref-smith20}{Smith et al., 2020})} to a decade of fossil fuel emissions from all of Latin American and the Caribbean \emph{(\protect\hyperlink{ref-chazdon16}{Chazdon et al., 2016})}.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80\% of old-growth forest levels in just 20 years, especially in wetter regions \emph{(\protect\hyperlink{ref-rozendaal19}{Rozendaal et al., 2019})}.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services \emph{(\protect\hyperlink{ref-guariguata01}{Guariguata and Ostertag, 2001}; \protect\hyperlink{ref-stanturf14}{Stanturf et al., 2014})} like biomass and carbon storage \emph{(\protect\hyperlink{ref-liu18}{Liu et al., 2018})}.
However, secondary neotropical forests diverge widely in taxonomic composition \emph{(\protect\hyperlink{ref-norden15}{Norden et al., 2015})}, potentially taking over a century to recover \emph{(\protect\hyperlink{ref-poorter21}{Poorter et al., 2021})} with added variability \emph{(\protect\hyperlink{ref-atkinson22}{Atkinson et al., 2022})}.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa \emph{(\protect\hyperlink{ref-carlo16}{Carlo and Morales, 2016})}, which can be keystone to locally-adapted food webs, including birds \emph{(\protect\hyperlink{ref-maas16}{Maas et al., 2016})}.
As a result, understanding secondary forest regeneration remains important\ldots{}

Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, as well as land-use history, especially in heterogeneously-managed landscapes \emph{(\protect\hyperlink{ref-melo13}{Melo et al., 2013}; \protect\hyperlink{ref-perfecto09}{Perfecto et al., 2009})}.
About 70\% of all forests have been estimated to be within a km of their edge \emph{(\protect\hyperlink{ref-haddad15}{Haddad et al., 2015})}, resulting in tropical forests having been estimated to lose \textasciitilde22\% aboveground carbon along heavily-managed edges \textgreater100m, in part due to shortening trees and lighter leaf mass \emph{(\protect\hyperlink{ref-ordway20}{Ordway and Asner, 2020})}.
Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa \emph{(\protect\hyperlink{ref-wandrag17}{Wandrag et al., 2017})}, which while complicated \emph{(\protect\hyperlink{ref-muller-landau09}{Muller-Landau and Hardesty, 2009})}, can also correlate with simpler plant traits like height and seed mass \emph{(\protect\hyperlink{ref-tamme14}{Tamme et al., 2014}; \protect\hyperlink{ref-thomson11}{Thomson et al., 2011})}.
The traits of existing taxa also depend on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012})}, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages \emph{(\protect\hyperlink{ref-albrich20}{Albrich et al., 2020}; \protect\hyperlink{ref-gough21}{Gough et al., 2021})}, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry \emph{(\protect\hyperlink{ref-lefland18}{Lefland et al., 2018})}.
Overall, while edge effects near degraded habitat edges are often negative, meta-population \emph{(\protect\hyperlink{ref-levins79}{Levins, 1979})} and -community \emph{(\protect\hyperlink{ref-leibold04}{Leibold et al., 2004}; \protect\hyperlink{ref-warren15}{Warren et al., 2015})} theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests \emph{(\protect\hyperlink{ref-oliveira-neto17}{Oliveira-Neto et al., 2017})}, forests affected by logging \emph{(\protect\hyperlink{ref-clark12}{Clark and Covey, 2012}; \protect\hyperlink{ref-edwards14}{Edwards et al., 2014})}, and timber plantations \emph{(\protect\hyperlink{ref-pryde15}{Pryde et al., 2015})}.

This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade that would suppress less shade-tolerant pioneer tree taxa and allow for more tree diversity closer to the primary forest edge.
Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{study-site}{%
\subsection{Study site}\label{study-site}}

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx. 8.3778176, -83.2935426) near the Piro Biological Station run by Osa Conservation.
(See \emph{Taylor et al. (\protect\hyperlink{ref-taylor15}{2015})} for a broader ecosystem description of the region.)
The study site was a regrowing 20 ha timber plantation of \emph{Bombacompsis quinata} abandoned in \textasciitilde1990 after the dry-adapted species from the Guanacaste region \emph{(\protect\hyperlink{ref-hulshof20}{Hulshof and Powers, 2020}; \protect\hyperlink{ref-perezcordero02}{Pérez Cordero and Kanninen, 2002})} grew poorly in very wet conditions.
This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \ref{fig:map}) and a wide service road on the third NE border, with primary forest beyond it.
This census was done in 2013 during the rainy season months between June and August.
\includegraphics{merge_files/figure-latex/map-1.pdf}

\hypertarget{census-design}{%
\subsection{Census design}\label{census-design}}

Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \ref{fig:map}), using available GIS software (\emph{ArcGIS 10 from 2010}, \url{esri.com}, and \emph{QGIS 2 from 2013}, \url{qgis.org}).
Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
The total area of the 30 census plots equaled \textasciitilde1 ha or 5\% of the total secondary forest stand area, which is comparable to similar studies \emph{(\protect\hyperlink{ref-onyekwelu16}{Onyekwelu and Olabiwonnu, 2016})}.

\hypertarget{plot-measurements}{%
\subsection{Plot measurements}\label{plot-measurements}}

Light reaching the forest floor was measured at the center of each plot at chest height using a densiometer \emph{(Forestry Suppliers, Inc.)}, as an average of four readings taken facing each cardinal direction.

The slope of the forest floor was measured using a rangefinder \emph{(Bushnell, Forestry Suppliers, Inc)} to measure the distance the diagonal between two plot corners and triangulate the observation angle.

The diameter of all stems \textgreater10 cm wide were recorded in each census plot.
Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
Tree height was recorded by measuring distances to both the crown and the stem at chest height \emph{(\textasciitilde2.7 m)} using a rangefinder \emph{(Bushnell, Forestry Suppliers, Inc.)} and triangulating the missing side length.
Taxa were identified with localized knowledge and field guides matching common names to Latin binomials,
and trait information was gathered from the literature.

Aboveground biomass (AGB) was estimated per stem using the allometric equation by \emph{Chave et al. (\protect\hyperlink{ref-chave15}{2015})}.
The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by \emph{(\protect\hyperlink{ref-alvarez12}{Alvarez et al., 2012}; Chave et al.; \protect\hyperlink{ref-chave05}{2005}; \protect\hyperlink{ref-ngomanda14}{Ngomanda et al., 2014})}.
The equation version with height was used since height was accessibly measurable and has shown improved local accuracy \emph{(\protect\hyperlink{ref-domke12}{Domke et al., 2012}; \protect\hyperlink{ref-khan20}{Khan et al., 2020})}.
Wood density (or specific gravity) values when available also improve biomass estimates even more than height \emph{(\protect\hyperlink{ref-vanbreugel11}{Van Breugel et al., 2011})}.
The equation used was:
\[{AGB} = 0.0673 (p D^2 H)^{0.976}\]
with diameter \emph{D} in cm, height \emph{H} in m, and wood density or specific gravity \emph{p} in \emph{g cm\textsuperscript{-3}}.
DBHs and heights were measured in the field and wood densities were extracted from available literature.
Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be \emph{0.58} following World Agroforestry Database guidelines \href{worldagroforestry.org}{(worldagroforestry.org)}.

Original values found and recorded were later updated to modern values using the pipeline developed in the \emph{BIOMASS} 2.1.8 R software package by \emph{Réjou-Méchain et al. (\protect\hyperlink{ref-rejou-mechain17}{2017})}, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.

\hypertarget{statistical-analyses}{%
\subsection{Statistical analyses}\label{statistical-analyses}}

For all response variables, data were analyzed using plot medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base R version 4.2.0 (2022-04-22) and \emph{rstatix} 0.7.0 package functions \emph{(\protect\hyperlink{ref-base}{R Core Team, 2022})}.

Non-linear regressions were run using the function \emph{poly()} in the base R package \emph{stats} 4.2.0,
which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models.
All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
Community analysis and ordination was done as a PERMANOVA with the \emph{adonis()} function from the \emph{vegan} 2.6.2 R package \emph{(\protect\hyperlink{ref-dixon03}{Dixon, 2003})}.
Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and are available at \url{nmedina17.github.com/osa}.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{biomass}{%
\subsection{Biomass}\label{biomass}}

Aboveground biomass did not tend to change with distance to primary forest edge
(Fig \ref{fig:biomass}a), despite significant trends with underlying related variables
(Fig \ref{fig:biomass}).

Median plot wood density tended to increase marginally significantly
(p=0.105)
with edge distance by
\ensuremath{1.9\times 10^{-4}} ±
\ensuremath{9.2\times 10^{-5}} per m
explaining
R2x100\% of variance
among median distance strata values
(Fig \ref{fig:biomass}b).

Plot stem density also tended to increase significantly
(p=0.03)
with edge distance but more strongly and non-linearly, by
\textasciitilde17.4 ±
3.4 per m
(Fig \ref{fig:biomass}c).

At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge
(Fig \ref{fig:biomass}d).

\begin{verbatim}
##   |                                                                              |                                                                      |   0%  |                                                                              |======================================================================| 100%
\end{verbatim}

\begin{figure}
\centering
\includegraphics{merge_files/figure-latex/biomass-1.pdf}
\caption{\label{fig:biomass} Plot stand properties, namely (a) biomass, (b) wood density, (c) stem density, and (d) canopy light availability. Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
\end{figure}

\hypertarget{composition}{%
\subsection{Composition}\label{composition}}

Overall diversity changed significantly with distance to primary forest edge
(Fig \ref{fig:taxa}).

Alpha diversity significantly
(p=0.003)
decreased non-linearly by
\textasciitilde0.5 ±
0 per m
as distance to primary forest edge increased
which explained
R2x100\% of variance
among distance strata medians
(Fig \ref{fig:taxa}a) --
while in contrast taxonomic richness increased slightly
(p=0.067)
and linearly
by
0 ±
0 per m
and had R2x100\% variance explained
among distance strata
(Fig \ref{fig:taxa}b).

Community composition and beta diversity also changed significantly
(p=0.01) with
11.4\%
variance among distance strata explained by distance to primary forest edge,
and the first two principal components explaining
47\% and
23\% totaling
70\%
of variance among calculated plot distances
(Fig \ref{fig:taxa}d).

Key abundant taxa Vochysia and Ficus showed different responses -- Vochysia nearly tended to decrease with distance to primary forest edge
(Fig \ref{fig:taxa}e),
while Ficus decreased marginally significantly
(p=0.088)
and linearly by
-62.6 ±
25.1 per m
with R2x100\% variance among distance strata explained
(Fig \ref{fig:taxa}f).

\begin{verbatim}
##   |                                                                              |                                                                      |   0%  |                                                                              |======================================================================| 100%
\end{verbatim}

\begin{figure}
\centering
\includegraphics{merge_files/figure-latex/taxa-1.pdf}
\caption{\label{fig:taxa} Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa. White and colored dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
\end{figure}

\hypertarget{functions}{%
\subsection{Functions}\label{functions}}

Overall functional regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest
(Fig \ref{fig:traits}).

Taxa associated with both early and late successional stages decreased marginally significantly
(p=0.042)
and curvi-linearly
by
\textasciitilde{} -584.9 ±
167.3 per m
with distance to edge explaining
R2x100\% variance
among distance strata medians
(Fig \ref{fig:traits}a).

Dispersal modes did not show consistent trends in biomass with increasing distance to edge
(Fig \ref{fig:traits}b).

\begin{verbatim}
##   |                                                                              |                                                                      |   0%  |                                                                              |======================================================================| 100%
##   |                                                                              |                                                                      |   0%  |                                                                              |======================================================================| 100%
##   |                                                                              |                                                                      |   0%  |                                                                              |======================================================================| 100%
\end{verbatim}

\begin{figure}
\centering
\includegraphics{merge_files/figure-latex/traits-1.pdf}
\caption{\label{fig:traits} Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode. Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{outline}{%
\section*{outline}\label{outline}}
\addcontentsline{toc}{section}{outline}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  assess hypotheses
\item
  infer causes of patterns - plantation structure, old legacy trees, or species traits?
\item
  compare inferences to other studies
\item
  recommend future conservation approaches, ecological studies
\end{enumerate}

\hypertarget{draft}{%
\section*{draft}\label{draft}}
\addcontentsline{toc}{section}{draft}

This study found significant edge effects on tree diversity and community composition, but not aboveground biomass\ldots functional diversity?
These results were shown in Figs \ref{fig:biomass}, \ref{fig:taxa}, and \ref{fig:traits}\ldots{}

\hypertarget{attic}{%
\section*{attic}\label{attic}}
\addcontentsline{toc}{section}{attic}

Under more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e.~near 1:1 replacement) the targeted tree cutting that is done using information collected about the ecological profiles and ages (when possible) of trees (Duah-Gyamfi et al 2014, Edwards et al 2014).

In contrast, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015) while also usually showing negative trade-offs with other services, like that with biodiversity given reliance on monoculture planting designs.

In some cases where management decisions are made primarily based on regional economic but not local ecological information, due to limited access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.

After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g.~based on surrounding forest proximity.

The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density \emph{(\protect\hyperlink{ref-cuxe9sar21}{\textbf{césar21?}})}, and could therefore require a switch in focal timber species \emph{(\protect\hyperlink{ref-karsten14}{Karsten et al., 2014})}.

\newpage

\hypertarget{funding}{%
\section*{Funding}\label{funding}}
\addcontentsline{toc}{section}{Funding}

This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship.

\hypertarget{declaration-of-interests}{%
\section*{Declaration of interests}\label{declaration-of-interests}}
\addcontentsline{toc}{section}{Declaration of interests}

Authors declare no conflicts of interest.

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

Thanks to Professor Dan Perlman for initial networking; Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; concurrent Osa Conservation General Manager Manuel Ramirez and Piro Research Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; and previous anonymous reviewers and peer James Fifer for discussion of drafts.

\hypertarget{author-contributions}{%
\section*{Author contributions}\label{author-contributions}}
\addcontentsline{toc}{section}{Author contributions}

MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis; NM contributed to initial draft writing and later revisions.

\hypertarget{data-statement}{%
\section*{Data statement}\label{data-statement}}
\addcontentsline{toc}{section}{Data statement}

Data and code stored at \href{github.com/nmedin17/osa}{github.com/nmedina17/osa}.

\newpage

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-albrich20}{}}%
Albrich, K., Rammer, W., Seidl, R., 2020. Climate change causes critical transitions and irreversible alterations of mountain forests. Global Change Biology 26, 4013--4027. \url{https://doi.org/10.1111/gcb.15118}

\leavevmode\vadjust pre{\hypertarget{ref-alvarez12}{}}%
Alvarez, E., Duque, A., Saldarriaga, J., Cabrera, K., Salas, G. de las, Valle, I. del, Lema, A., Moreno, F., Orrego, S., Rodríguez, L., 2012. Tree above-ground biomass allometries for carbon stocks estimation in the natural forests of {Colombia}. Forest Ecology and Management 267, 297--308. \url{https://doi.org/10.1016/j.foreco.2011.12.013}

\leavevmode\vadjust pre{\hypertarget{ref-anderegg22}{}}%
Anderegg, W.R.L., Chegwidden, O.S., Badgley, G., Trugman, A.T., Cullenward, D., Abatzoglou, J.T., Hicke, J.A., Freeman, J., Hamman, J.J., 2022. Future climate risks from stress, insects and fire across {US} forests. Ecology Letters ele.14018. \url{https://doi.org/10.1111/ele.14018}

\leavevmode\vadjust pre{\hypertarget{ref-atkinson22}{}}%
Atkinson, J., Brudvig, L.A., Mallen-Cooper, M., Nakagawa, S., Moles, A.T., Bonser, S.P., 2022. Terrestrial ecosystem restoration increases biodiversity and reduces its variability, but not to reference levels: {A} global meta-analysis. Ecology Letters ele.14025. \url{https://doi.org/10.1111/ele.14025}

\leavevmode\vadjust pre{\hypertarget{ref-carlo16}{}}%
Carlo, T.A., Morales, J.M., 2016. Generalist birds promote tropical forest regeneration and increase plant diversity via rare-biased seed dispersal. Ecology 97, 1819--1831. \url{https://doi.org/10.1890/15-2147.1}

\leavevmode\vadjust pre{\hypertarget{ref-chave05}{}}%
Chave, J., Andalo, C., Brown, S., Cairns, M.A., Chambers, J.Q., Eamus, D., Fölster, H., Fromard, F., Higuchi, N., Kira, T., Lescure, J.-P., Nelson, B.W., Ogawa, H., Puig, H., Riéra, B., Yamakura, T., 2005. Tree allometry and improved estimation of carbon stocks and balance in tropical forests. Oecologia 145, 87--99. \url{https://doi.org/10.1007/s00442-005-0100-x}

\leavevmode\vadjust pre{\hypertarget{ref-chave15}{}}%
Chave, J., Réjou-Méchain, M., Búrquez, A., Chidumayo, E., Colgan, M.S., Delitti, W.B.C., Duque, A., Eid, T., Fearnside, P.M., Goodman, R.C., Henry, M., Martínez-Yrízar, A., Mugasha, W.A., Muller-Landau, H.C., Mencuccini, M., Nelson, B.W., Ngomanda, A., Nogueira, E.M., Ortiz-Malavassi, E., Pélissier, R., Ploton, P., Ryan, C.M., Saldarriaga, J.G., Vieilledent, G., 2015. Improved allometric models to estimate the aboveground biomass of tropical trees. Global Change Biology 20, 3177--3190. \url{https://doi.org/10.1111/gcb.12629}

\leavevmode\vadjust pre{\hypertarget{ref-chazdon18}{}}%
Chazdon, R.L., 2018. Protecting intact forests requires holistic approaches. Nature Ecology and Evolution 2, 915. \url{https://doi.org/10.1038/s41559-018-0546-y}

\leavevmode\vadjust pre{\hypertarget{ref-chazdon16}{}}%
Chazdon, R.L., Broadbent, E.N., Rozendaal, D.M.A., Bongers, F., Zambrano, A.M.A., Aide, T.M., Balvanera, P., Becknell, J.M., Boukili, V., Brancalion, P.H.S., Craven, D., Almeida-Cortez, J.S., Cabral, G.A.L., Jong, B. de, Denslow, J.S., Dent, D.H., DeWalt, S.J., Dupuy, J.M., Duran, S.M., Espirito-Santo, M.M., Fandino, M.C., Cesar, R.G., Hall, J.S., Hernandez-Stefanoni, J.L., Jakovac, C.C., Junqueira, A.B., Kennard, D., Letcher, S.G., Lohbeck, M., Martinez-Ramos, M., Massoca, P., Meave, J.A., Mesquita, R., Mora, F., Munoz, R., Muscarella, R., Nunes, Y.R.F., Ochoa-Gaona, S., Orihuela-Belmonte, E., Pena-Claros, M., Perez-Garcia, E.A., Piotto, D., Powers, J.S., Rodriguez-Velazquez, J., Romero-Perez, I.E., Ruiz, J., Saldarriaga, J.G., Sanchez-Azofeifa, A., Schwartz, N.B., Steininger, M.K., Swenson, N.G., Uriarte, M., Breugel, M. van, Wal, H. van der, Veloso, M.D.M., Vester, H., Vieira, I.C.G., Bentos, T.V., Williamson, G.B., Poorter, L., 2016. Carbon sequestration potential of second-growth forest regeneration in the {Latin American} tropics. Science Advances. \url{https://doi.org/10.1126/sciadv.1501639}

\leavevmode\vadjust pre{\hypertarget{ref-clark12}{}}%
Clark, J.A., Covey, K.R., 2012. Tree species richness and the logging of natural forests: {A} meta-analysis. Forest Ecology and Management 276, 146--153. \url{https://doi.org/10.1016/j.foreco.2012.04.001}

\leavevmode\vadjust pre{\hypertarget{ref-depinto20}{}}%
De Pinto, A., Cenacchi, N., Robertson, R., Kwon, H.-Y., Thomas, T., Koo, J., Begeladze, S., Kumar, C., 2020. The {Role} of {Crop Production} in the {Forest Landscape Restoration Approach}\textemdash{{Assessing}} the {Potential Benefits} of {Meeting} the {Bonn Challenge}. Frontiers in Sustainable Food Systems 4, 61. \url{https://doi.org/10.3389/fsufs.2020.00061}

\leavevmode\vadjust pre{\hypertarget{ref-dixon03}{}}%
Dixon, P., 2003. {VEGAN}, a package of {R} functions for community ecology. Journal of Vegetation Science 14, 927--930. \url{https://doi.org/10.1111/j.1654-1103.2003.tb02228.x}

\leavevmode\vadjust pre{\hypertarget{ref-domke12}{}}%
Domke, G.M., Woodall, C.W., Smith, J.E., Westfall, J.A., McRoberts, R.E., 2012. Consequences of alternative tree-level biomass estimation procedures on {U}.{S}. Forest carbon stock estimates. Forest Ecology and Management 270, 108--116. \url{https://doi.org/10.1016/j.foreco.2012.01.022}

\leavevmode\vadjust pre{\hypertarget{ref-edwards14}{}}%
Edwards, D.P., Tobias, J.A., Sheil, D., Meijaard, E., Laurance, W.F., 2014. Maintaining ecosystem function and services in logged tropical forests. Trends in Ecology and Evolution 29, 511--520. \url{https://doi.org/10.1016/j.tree.2014.07.003}

\leavevmode\vadjust pre{\hypertarget{ref-fao20}{}}%
FAO, 2020. The {State} of the {World}'s {Forests} 2020. {FAO and UNEP}. \url{https://doi.org/10.4060/ca8642en}

\leavevmode\vadjust pre{\hypertarget{ref-finegan96}{}}%
Finegan, B., 1996. Pattern and process in neotropical secondary rain forests: The first 100 years of succession. Trends in Ecology \& Evolution.

\leavevmode\vadjust pre{\hypertarget{ref-gough21}{}}%
Gough, C.M., Bohrer, G., Hardiman, B.S., Nave, L.E., Vogel, C.S., Atkins, J.W., Bond-Lamberty, B., Fahey, R.T., Fotis, A.T., Grigri, M.S., Haber, L.T., Ju, Y., Kleinke, C.L., Mathes, K.C., Nadelhoffer, K.J., Stuart-Haëntjens, E., Curtis, P.S., 2021. Disturbance-accelerated succession increases the production of a temperate forest. Ecological Applications 31. \url{https://doi.org/10.1002/eap.2417}

\leavevmode\vadjust pre{\hypertarget{ref-guariguata01}{}}%
Guariguata, M.R., Ostertag, R., 2001. Neotropical secondary forest succession: {Changes} in structural and functional characteristics. Forest Ecology and Management 148, 185--206. \url{https://doi.org/10.1016/S0378-1127(00)00535-1}

\leavevmode\vadjust pre{\hypertarget{ref-haddad15}{}}%
Haddad, N.M., Brudvig, L.A., Clobert, J., Davies, K.F., Gonzalez, A., Holt, R.D., Lovejoy, T.E., Sexton, J.O., Austin, M.P., Collins, C.D., Cook, W.M., Damschen, E.I., Ewers, R.M., Foster, B.L., Jenkins, C.N., King, A.J., Laurance, W.F., Levey, D.J., Margules, C.R., Melbourne, B.A., Nicholls, A.O., Orrock, J.L., Song, D.-X., Townshend, J.R., 2015. Habitat fragmentation and its lasting impact on {Earth}'s ecosystems. Science Advances 1, e1500052. \url{https://doi.org/10.1126/sciadv.1500052}

\leavevmode\vadjust pre{\hypertarget{ref-hubau20}{}}%
Hubau, W., Lewis, S.L., Phillips, O.L., Affum-Baffoe, K., Beeckman, H., Cuní-Sanchez, A., Daniels, A.K., Ewango, C.E.N., Fauset, S., Mukinzi, J.M., Sheil, D., Sonké, B., Sullivan, M.J.P., Sunderland, T.C.H., Taedoumg, H., Thomas, S.C., White, L.J.T., Abernethy, K.A., Adu-Bredu, S., Amani, C.A., Baker, T.R., Banin, L.F., Baya, F., Begne, S.K., Bennett, A.C., Benedet, F., Bitariho, R., Bocko, Y.E., Boeckx, P., Boundja, P., Brienen, R.J.W., Brncic, T., Chezeaux, E., Chuyong, G.B., Clark, C.J., Collins, M., Comiskey, J.A., Coomes, D.A., Dargie, G.C., de Haulleville, T., Kamdem, M.N.D., Doucet, J.-L., Esquivel-Muelbert, A., Feldpausch, T.R., Fofanah, A., Foli, E.G., Gilpin, M., Gloor, E., Gonmadje, C., Gourlet-Fleury, S., Hall, J.S., Hamilton, A.C., Harris, D.J., Hart, T.B., Hockemba, M.B.N., Hladik, A., Ifo, S.A., Jeffery, K.J., Jucker, T., Yakusu, E.K., Kearsley, E., Kenfack, D., Koch, A., Leal, M.E., Levesley, A., Lindsell, J.A., Lisingo, J., Lopez-Gonzalez, G., Lovett, J.C., Makana, J.-R., Malhi, Y., Marshall, A.R., Martin, J., Martin, E.H., Mbayu, F.M., Medjibe, V.P., Mihindou, V., Mitchard, E.T.A., Moore, S., Munishi, P.K.T., Bengone, N.N., Ojo, L., Ondo, F.E., Peh, K.S.-H., Pickavance, G.C., Poulsen, A.D., Poulsen, J.R., Qie, L., Reitsma, J., Rovero, F., Swaine, M.D., Talbot, J., Taplin, J., Taylor, D.M., Thomas, D.W., Toirambe, B., Mukendi, J.T., Tuagben, D., Umunay, P.M., van der Heijden, G.M.F., Verbeeck, H., Vleminckx, J., Willcock, S., Wöll, H., Woods, J.T., Zemagho, L., 2020. Asynchronous carbon sink saturation in {African} and {Amazonian} tropical forests. Nature 579, 80--87. \url{https://doi.org/10.1038/s41586-020-2035-0}

\leavevmode\vadjust pre{\hypertarget{ref-hulshof20}{}}%
Hulshof, C.M., Powers, J.S., 2020. Tropical forest composition and function across space and time: {Insights} from diverse gradients in {Área} de {Conservación Guanacaste}. Biotropica 52, 1065--1075. \url{https://doi.org/10.1111/btp.12689}

\leavevmode\vadjust pre{\hypertarget{ref-karsten14}{}}%
Karsten, R.J., Meilby, H., Larsen, J.B., 2014. Regeneration and management of lesser known timber species in the {Peruvian Amazon} following disturbance by logging. Forest Ecology and Management 327, 76--85. \url{https://doi.org/10.1016/j.foreco.2014.04.035}

\leavevmode\vadjust pre{\hypertarget{ref-khan20}{}}%
Khan, M.N.I., Islam, M.R., Rahman, A., Azad, M.S., Mollick, A.S., Kamruzzaman, M., Sadath, M.N., Feroz, S.M., Rakkibu, M.G., Knohl, A., 2020. Allometric relationships of stand level carbon stocks to basal area, tree height and wood density of nine tree species in {Bangladesh}. Global Ecology and Conservation 22. \url{https://doi.org/10.1016/j.gecco.2020.e01025}

\leavevmode\vadjust pre{\hypertarget{ref-lefland18}{}}%
Lefland, A.B., Huff, E.S., Donahue, B., 2018. A {Community Forestry Model Linking Research}, {Management}, {Education}, and {Stakeholder Engagement}: {Case Study Results} from the {Town} of {Weston}, {Massachusetts}, {USA}. Small-scale Forestry 17, 191--210. \url{https://doi.org/10.1007/s11842-017-9382-1}

\leavevmode\vadjust pre{\hypertarget{ref-leibold04}{}}%
Leibold, M.A., Holyoak, M., Mouquet, N., Amarasekare, P., Chase, J.M., Hoopes, M.F., Holt, R.D., Shurin, J.B., Law, R., Tilman, D., Loreau, M., Gonzalez, A., 2004. The metacommunity concept: {A} framework for multi-scale community ecology. Ecology Letters 7, 601--613. \url{https://doi.org/10.1111/j.1461-0248.2004.00608.x}

\leavevmode\vadjust pre{\hypertarget{ref-levins79}{}}%
Levins, R., 1979. Coexistence in a {Variable Environment}. The American Naturalist 114, 765--783. \url{https://doi.org/10.1086/283527}

\leavevmode\vadjust pre{\hypertarget{ref-liu18}{}}%
Liu, X., Trogisch, S., He, J.-S., Niklaus, P.A., Bruelheide, H., Tang, Z., Erfmeier, A., Scherer-Lorenzen, M., Pietsch, K.A., Yang, B., Kühn, P., Scholten, T., Huang, Y., Wang, C., Staab, M., Leppert, K.N., Wirth, C., Schmid, B., Ma, K., 2018. Tree species richness increases ecosystem carbon storage in subtropical forests. Proceedings of the Royal Society B: Biological Sciences 285, 20181240. \url{https://doi.org/10.1098/rspb.2018.1240}

\leavevmode\vadjust pre{\hypertarget{ref-maas16}{}}%
Maas, B., Karp, D.S., Bumrungsri, S., Darras, K., Gonthier, D., Huang, J.C.-C., Lindell, C.A., Maine, J.J., Mestre, L., Michel, N.L., Morrison, E.B., Perfecto, I., Philpott, S.M., Şekercioğlu, Ç.H., Silva, R.M., Taylor, P.J., Tscharntke, T., Van Bael, S.A., Whelan, C.J., Williams-Guillén, K., 2016. Bird and bat predation services in tropical forests and agroforestry landscapes. Biological Reviews 91, 1081--1101. \url{https://doi.org/10.1111/brv.12211}

\leavevmode\vadjust pre{\hypertarget{ref-melo13}{}}%
Melo, F.P.L., Arroyo-Rodríguez, V., Fahrig, L., Martínez-Ramos, M., Tabarelli, M., 2013. On the hope for biodiversity-friendly tropical landscapes. Trends in Ecology and Evolution 28, 462--468. \url{https://doi.org/10.1016/j.tree.2013.01.001}

\leavevmode\vadjust pre{\hypertarget{ref-here}{}}%
Müller, K., 2020. \href{https://CRAN.R-project.org/package=here}{Here: A simpler way to find your files}.

\leavevmode\vadjust pre{\hypertarget{ref-muller-landau09}{}}%
Muller-Landau, H.C., Hardesty, B.D., 2009. Seed dispersal of woody plants in tropical forests: Concepts, examples and future directions. Biotic Interactions in the Tropics 267--309. \url{https://doi.org/10.1017/cbo9780511541971.012}

\leavevmode\vadjust pre{\hypertarget{ref-ngomanda14}{}}%
Ngomanda, A., Obiang, N.L.E., Lebamba, J., Mavouroulou, Q.M., Gomat, H., Mankou, G.S., Loumeto, J., Iponga, D.M., Ditsouga, F.K., Koumba, R.Z., Bobé, K.H.B., Okouyi, C.M., Nyangadouma, R., Lépengué, N., Mbatchi, B., Picard, N., 2014. Site-specific versus pantropical allometric equations: {Which} option to estimate the biomass of a moist central {African} forest? Forest Ecology and Management 312, 1--9. \url{https://doi.org/10.1016/j.foreco.2013.10.029}

\leavevmode\vadjust pre{\hypertarget{ref-norden15}{}}%
Norden, N., Angarita, H.A., Bongers, F., Martínez-Ramos, M., Granzow-de la Cerda, I., van Breugel, M., Lebrija-Trejos, E., Meave, J.A., Vandermeer, J., Williamson, G.B., Finegan, B., Mesquita, R., Chazdon, R.L., 2015. Successional dynamics in {Neotropical} forests are as uncertain as they are predictable. Proceedings of the National Academy of Sciences 112, 8013--8018. \url{https://doi.org/10.1073/pnas.1500403112}

\leavevmode\vadjust pre{\hypertarget{ref-oliveira-neto17}{}}%
Oliveira-Neto, N.E., Nascimento, D.R., Carvalho, F.A., 2017. Biodiversity inventory of trees in a neotropical secondary forest after abandonment of shaded coffee plantation. IForest 10, 303--308. \url{https://doi.org/10.3832/ifor1901-009}

\leavevmode\vadjust pre{\hypertarget{ref-omeja12}{}}%
Omeja, P.A., Obua, J., Rwetsiba, A., Chapman, C.A., 2012. Biomass accumulation in tropical lands with different disturbance histories: {Contrasts} within one landscape and across regions. Forest Ecology and Management 269, 293--300. \url{https://doi.org/10.1016/j.foreco.2011.12.044}

\leavevmode\vadjust pre{\hypertarget{ref-onyekwelu16}{}}%
Onyekwelu, J.C., Olabiwonnu, A.A., 2016. Can forest plantations harbour biodiversity similar to natural forest ecosystems over time? International Journal of Biodiversity Science, Ecosystem Services and Management 12, 108--115. \url{https://doi.org/10.1080/21513732.2016.1162199}

\leavevmode\vadjust pre{\hypertarget{ref-ordway20}{}}%
Ordway, E.M., Asner, G.P., 2020. Carbon declines along tropical forest edges correspond to heterogeneous effects on canopy structure and function. Proceedings of the National Academy of Sciences 117, 7863--7870. \url{https://doi.org/10.1073/pnas.1914420117}

\leavevmode\vadjust pre{\hypertarget{ref-pan13}{}}%
Pan, Y., Birdsey, R.A., Phillips, O.L., Jackson, R.B., 2013. The {Structure}, {Distribution}, and {Biomass} of the {World}'s {Forests}. Annual Review of Ecology, Evolution, and Systematics 44, 593--622. \url{https://doi.org/10.1146/annurev-ecolsys-110512-135914}

\leavevmode\vadjust pre{\hypertarget{ref-perezcordero02}{}}%
Pérez Cordero, L.D., Kanninen, M., 2002. Wood specific gravity and aboveground biomass of {Bombacopsis} quinata plantations in {Costa Rica}. Forest Ecology and Management 165, 1--9. \url{https://doi.org/10.1016/S0378-1127(01)00627-2}

\leavevmode\vadjust pre{\hypertarget{ref-perfecto09}{}}%
Perfecto, I., Vandermeer, J.H., Wright, A.L., 2009. Nature's matrix: Linking agriculture, conservation and food sovereignty. {Earthscan}, {London ; Sterling, VA}.

\leavevmode\vadjust pre{\hypertarget{ref-poorter16}{}}%
Poorter, L., Bongers, F., Aide, T.M., Zambrano, A.M.A., Balvanera, P., Becknell, J.M., Boukili, V., Brancalion, P.H.S., Broadbent, E.N., Chazdon, R.L., Craven, D., Almeida-Cortez, J.S.D., Cabral, G.A.L., Jong, B.H.J.D., Denslow, J.S., Dent, D.H., DeWalt, S.J., Dupuy, J.M., Durán, S.M., Espírito-Santo, M.M., Fandino, M.C., César, R.G., Hall, J.S., Hernandez-Stefanoni, J.L., Jakovac, C.C., Junqueira, A.B., Kennard, D., Letcher, S.G., Licona, J.C., Lohbeck, M., Marín-Spiotta, E., Martínez-Ramos, M., Massoca, P., Meave, J.A., Mesquita, R., Mora, F., Munõz, R., Muscarella, R., Nunes, Y.R.F., Ochoa-Gaona, S., Oliveira, A.A.D., Orihuela-Belmonte, E., Penã-Claros, M., Pérez-Garciá, E.A., Piotto, D., Powers, J.S., Rodríguez-Velázquez, J., Romero-Pérez, I.E., Ruíz, J., Saldarriaga, J.G., Sanchez-Azofeifa, A., Schwartz, N.B., Steininger, M.K., Swenson, N.G., Toledo, M., Uriarte, M., Breugel, M.V., Wal, H.V.D., Veloso, M.D.M., Vester, H.F.M., Vicentini, A., Vieira, I.C.G., Bentos, T.V., Williamson, G.B., Rozendaal, D.M.A., 2016. Biomass resilience of {Neotropical} secondary forests. Nature 530, 211--214. \url{https://doi.org/10.1038/nature16512}

\leavevmode\vadjust pre{\hypertarget{ref-poorter21}{}}%
Poorter, L., Craven, D., Jakovac, C.C., van der Sande, M.T., Amissah, L., Bongers, F., Chazdon, R.L., Farrior, C.E., Kambach, S., Meave, J.A., Muñoz, R., Norden, N., Rüger, N., van Breugel, M., Almeyda Zambrano, A.M., Amani, B., Andrade, J.L., Brancalion, P.H.S., Broadbent, E.N., de Foresta, H., Dent, D.H., Derroire, G., DeWalt, S.J., Dupuy, J.M., Durán, S.M., Fantini, A.C., Finegan, B., Hernández-Jaramillo, A., Hernández-Stefanoni, J.L., Hietz, P., Junqueira, A.B., N'dja, J.K., Letcher, S.G., Lohbeck, M., López-Camacho, R., Martínez-Ramos, M., Melo, F.P.L., Mora, F., Müller, S.C., N'Guessan, A.E., Oberleitner, F., Ortiz-Malavassi, E., Pérez-García, E.A., Pinho, B.X., Piotto, D., Powers, J.S., Rodríguez-Buriticá, S., Rozendaal, D.M.A., Ruíz, J., Tabarelli, M., Teixeira, H.M., Valadares de Sá Barretto Sampaio, E., van der Wal, H., Villa, P.M., Fernandes, G.W., Santos, B.A., Aguilar-Cano, J., de Almeida-Cortez, J.S., Alvarez-Davila, E., Arreola-Villa, F., Balvanera, P., Becknell, J.M., Cabral, G.A.L., Castellanos-Castro, C., de Jong, B.H.J., Nieto, J.E., Espírito-Santo, M.M., Fandino, M.C., García, H., García-Villalobos, D., Hall, J.S., Idárraga, A., Jiménez-Montoya, J., Kennard, D., Marín-Spiotta, E., Mesquita, R., Nunes, Y.R.F., Ochoa-Gaona, S., Peña-Claros, M., Pérez-Cárdenas, N., Rodríguez-Velázquez, J., Villanueva, L.S., Schwartz, N.B., Steininger, M.K., Veloso, M.D.M., Vester, H.F.M., Vieira, I.C.G., Williamson, G.B., Zanini, K., Hérault, B., 2021. Multidimensional tropical forest recovery. Science 374, 1370--1376. \url{https://doi.org/10.1126/science.abh3629}

\leavevmode\vadjust pre{\hypertarget{ref-pryde15}{}}%
Pryde, E.C., Holland, G.J., Watson, S.J., Turton, S.M., Nimmo, D.G., 2015. Conservation of tropical forest tree species in a native timber plantation landscape. Forest Ecology and Management 339, 96--104. \url{https://doi.org/10.1016/j.foreco.2014.11.028}

\leavevmode\vadjust pre{\hypertarget{ref-pyron22}{}}%
Pyron, R.A., Mooers, A.Ø., 2022. The normative postulate problem: {Hidden} values in ecology, evolution, and conservation. Biological Conservation 270, 109584. \url{https://doi.org/10.1016/j.biocon.2022.109584}

\leavevmode\vadjust pre{\hypertarget{ref-base}{}}%
R Core Team, 2022. \href{https://www.R-project.org/}{R: A language and environment for statistical computing}. R Foundation for Statistical Computing, Vienna, Austria.

\leavevmode\vadjust pre{\hypertarget{ref-rejou-mechain17}{}}%
Réjou-Méchain, M., Tanguy, A., Piponiot, C., Chave, J., Hérault, B., 2017. {BIOMASS} : An {R} package for estimating above-ground biomass and its uncertainty in tropical forests. Methods in Ecology and Evolution 8, 1163--1167. \url{https://doi.org/10.1111/2041-210X.12753}

\leavevmode\vadjust pre{\hypertarget{ref-grateful}{}}%
Rodríguez-Sánchez, F., Jackson, C.P., Hutchins, S.D., 2022. \href{https://github.com/Pakillo/grateful}{Grateful: Facilitate citation of r packages}.

\leavevmode\vadjust pre{\hypertarget{ref-rozendaal19}{}}%
Rozendaal, D.M.A., Bongers, F., Aide, T.M., Alvarez-Dávila, E., Ascarrunz, N., Balvanera, P., Becknell, J.M., Bentos, T.V., Brancalion, P.H.S., Cabral, G.A.L., Calvo-Rodriguez, S., Chave, J., César, R.G., Chazdon, R.L., Condit, R., Dallinga, J.S., Almeida-Cortez, J.S.D., Jong, B.D., Oliveira, A.D., Denslow, J.S., Dent, D.H., DeWalt, S.J., Dupuy, J.M., Durán, S.M., Dutrieux, L.P., Espírito-Santo, M.M., Fandino, M.C., Fernandes, G.W., Finegan, B., García, H., Gonzalez, N., Moser, V.G., Hall, J.S., Hernández-Stefanoni, J.L., Hubbell, S., Jakovac, C.C., Hernández, A.J., Junqueira, A.B., Kennard, D., Larpin, D., Letcher, S.G., Licona, J.C., Lebrija-Trejos, E., Marín-Spiotta, E., Martínez-Ramos, M., Massoca, P.E.S., Meave, J.A., Mesquita, R.C.G., Mora, F., Müller, S.C., Muñoz, R., Neto, S.N.D.O., Norden, N., Nunes, Y.R.F., Ochoa-Gaona, S., Ortiz-Malavassi, E., Ostertag, R., Peña-Claros, M., Pérez-García, E.A., Piotto, D., Powers, J.S., Aguilar-Cano, J., Rodriguez-Buritica, S., Rodríguez-Velázquez, J., Romero-Romero, M.A., Ruíz, J., Sanchez-Azofeifa, A., Almeida, A.S.D., Silver, W.L., Schwartz, N.B., Thomas, W.W., Toledo, M., Uriarte, M., Sampaio, E.V.D.S., Breugel, M.V., Wal, H.V.D., Martins, S.V., Veloso, M.D.M., Vester, H.F.M., Vicentini, A., Vieira, I.C.G., Villa, P., Williamson, G.B., Zanini, K.J., Zimmerman, J., Poorter, L., 2019. Biodiversity recovery of {Neotropical} secondary forests. Science Advances 5. \url{https://doi.org/10.1126/sciadv.aau3114}

\leavevmode\vadjust pre{\hypertarget{ref-smith20}{}}%
Smith, C.C., Espírito-Santo, F.D.B., Healey, J.R., Young, P.J., Lennox, G.D., Ferreira, J., Barlow, J., 2020. Secondary forests offset less than 10\% of deforestation-mediated carbon emissions in the {Brazilian Amazon}. Global Change Biology 26, 7006--7020. \url{https://doi.org/10.1111/gcb.15352}

\leavevmode\vadjust pre{\hypertarget{ref-stanturf14}{}}%
Stanturf, J.A., Palik, B.J., Dumroese, R.K., 2014. Contemporary forest restoration: {A} review emphasizing function. Forest Ecology and Management 331, 292--323. \url{https://doi.org/10.1016/j.foreco.2014.07.029}

\leavevmode\vadjust pre{\hypertarget{ref-tabarelli08}{}}%
Tabarelli, M., Lopes, A.V., Peres, C.A., 2008. Edge-effects drive tropical forest fragments towards an early-successional system. Biotropica 40, 657--661. \url{https://doi.org/10.1111/j.1744-7429.2008.00454.x}

\leavevmode\vadjust pre{\hypertarget{ref-tamme14}{}}%
Tamme, R., Götzenberger, L., Zobel, M., Bullock, J.M., Hooftman, D.A.P., Kaasik, A., Pärtel, M., 2014. Predicting species' maximum dispersal distances from simple plant traits. Ecology 95, 505--513. \url{https://doi.org/10.1890/13-1000.1}

\leavevmode\vadjust pre{\hypertarget{ref-taylor15}{}}%
Taylor, P., Asner, G., Dahlin, K., Anderson, C., Knapp, D., Martin, R., Mascaro, J., Chazdon, R., Cole, R., Wanek, W., Hofhansl, F., Malavassi, E., Vilchez-Alvarado, B., Townsend, A., 2015. Landscape-{Scale Controls} on {Aboveground Forest Carbon Stocks} on the {Osa Peninsula}, {Costa Rica}. PLOS ONE 10, e0126748. \url{https://doi.org/10.1371/journal.pone.0126748}

\leavevmode\vadjust pre{\hypertarget{ref-thomson11}{}}%
Thomson, F.J., Moles, A.T., Auld, T.D., Kingsford, R.T., 2011. Seed dispersal distance is more strongly correlated with plant height than with seed mass. Journal of Ecology 99, 1299--1307. \url{https://doi.org/10.1111/j.1365-2745.2011.01867.x}

\leavevmode\vadjust pre{\hypertarget{ref-vanbreugel11}{}}%
Van Breugel, M., Ransijn, J., Craven, D., Bongers, F., Hall, J.S., 2011. Estimating carbon stock in secondary forests: {Decisions} and uncertainties associated with allometric biomass models. Forest Ecology and Management 262, 1648--1657. \url{https://doi.org/10.1016/j.foreco.2011.07.018}

\leavevmode\vadjust pre{\hypertarget{ref-wandrag17}{}}%
Wandrag, E.M., Dunham, A.E., Duncan, R.P., Rogers, H.S., 2017. Seed dispersal increases local species richness and reduces spatial turnover of tropical tree seedlings. Proceedings of the National Academy of Sciences of the United States of America 114, 10689--10694. \url{https://doi.org/10.1073/pnas.1709584114}

\leavevmode\vadjust pre{\hypertarget{ref-warren15}{}}%
Warren, B.H., Simberloff, D., Ricklefs, R.E., Aguilée, R., Condamine, F.L., Gravel, D., Morlon, H., Mouquet, N., Rosindell, J., Casquet, J., Conti, E., Cornuault, J., Fernández-Palacios, J.M., Hengl, T., Norder, S.J., Rijsdijk, K.F., Sanmartín, I., Strasberg, D., Triantis, K.A., Valente, L.M., Whittaker, R.J., Gillespie, R.G., Emerson, B.C., Thébaud, C., 2015. Islands as model systems in ecology and evolution: Prospects fifty years after {MacArthur-Wilson}. Ecology Letters 18, 200--217. \url{https://doi.org/10.1111/ele.12398}

\leavevmode\vadjust pre{\hypertarget{ref-bookdown2022}{}}%
Xie, Y., 2022. \href{https://github.com/rstudio/bookdown}{Bookdown: Authoring books and technical documents with r markdown}.

\end{CSLReferences}

\end{document}

### 2022-06-18 article/methods.Rmd
minors, prep disc

@@
     number_sections: FALSE
-bibliography: ../../library.bib
+bibliography: 
+- ../../library.bib
+- ../../grateful-refs.bib
 link-citations: TRUE
@@
 
-```{r echo=F, results='hide', warning=F, message=F}
-knitr::opts_chunk$set(echo = F, warning = F, message = F); here::i_am("article/methods.Rmd")
+```{r setupMeth, echo=F, results='hide', warning=F, message=F}
+knitr::opts_chunk$set(echo = F, warning = F, message = F); here::i_am("article/methods.Rmd"); library(tidyverse)
 ```
@@
 
-## Census design
-
-Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10 from 2010*, [esri.com](esri.com), and *QGIS 2 from 2013*, [qgis.org](qgis.org)).
-Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
-The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.
-
-```{r map, fig.cap="Map of conservation area on Osa Peninsula, Costa Rica."}
-knitr::include_graphics(here::here("figs/map.png"))
+```{r map, fig.width=4, fig.height=4, fig.cap="Map of conservation areas and study site on Osa Peninsula, Costa Rica. Map image data © 2022 Osa Conservation. Photo credit: Nicholas Medina."}
+source(here::here("figs/map.R"));
+sitePlot
+# knitr::include_graphics(here::here("figs/figmap.png"))
 # magick::image_read(path = "../figs/map.png") %>% plot()
@@
 
+## Census design
+
+Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10*, [esri.com](esri.com), and *QGIS 2*, [qgis.org](qgis.org)).
+Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
+The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.
+
 ## Plot measurements 
@@
 
-### Canopy cover and topography
+<!-- ### Canopy cover and topography -->
 
@@
 
-### Tree measurements
+<!-- ### Tree measurements -->
 
@@
 
-### Aboveground biomass
+<!-- ### Aboveground biomass -->
 
@@
 Wood density (or specific gravity) values when available also improve biomass estimates even more than height *[@vanbreugel11]*.
-The equation used was: $$ {AGB} = 0.0673 (ρ D^2 H)^{0.976} $$ with diameter *D* in cm, height *H* in m, and wood density or specific gravity *p* in *g cm^-3^*.
+The equation used was: 
+$${AGB} = 0.0673 (p D^2 H)^{0.976}$$ 
+with diameter *D* in cm, height *H* in m, and wood density or specific gravity *p* in *g cm^-3^*.
 DBHs and heights were measured in the field and wood densities were extracted from available literature. 
@@
 
-# Statistical analyses
+## Statistical analyses
 
-For all response variables, data were analyzed using plot medians and subject to linear regression with distance to forest edge as the only independent variable, after checks for residual normality and equal variance assumptions, using `r version$version.string`.
+```{r pkgRefs, results='hide'}
+# remotes::install_github("Pakillo/grateful")
+tidyr::unnest(grateful::get_pkgs_info(out.dir = "../.."))
+```
+
+For all response variables, data were analyzed using plot medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base `r version$version.string` and *rstatix* `r packageVersion("rstatix")` package functions *[@base]*.
  <!-- including biomass converted to a per hectare basis,  -->
@@
 <!-- Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.  -->
-Non-linear regressions were run using the function *poly()* in the *stats* `r packageVersion("stats")` R package *(R Core Team 2022)*, 
-which was only recorded for stem density and richness variables. 
+Non-linear regressions were run using the function *poly()* in the base R package *stats* `r packageVersion("stats")`, 
+which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models. 
 All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
 Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
-Data and code are available at [nmedina17.github.com/osa](nmedina17.github.com/osa).
-
-<!-- `r citation()` -->
+Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and are available at [nmedina17.github.com/osa](nmedina17.github.com/osa).
 

### 2022-06-18 article/results.Rmd
minors, prep disc

@@
     number_sections: FALSE
+    toc: FALSE
 ---
 
-```{r setup, echo = F, message = F, warning=F, results=F}
+```{r setupRes, echo = F, message = F, warning=F, results='hide'}
 knitr::opts_chunk$set(echo = F, message = F, warning = F)
-here::i_am("article/results.Rmd"); source(here::here("figs/fig2.R"))
+source(here::here("figs/fig2.R")); here::i_am("article/results.Rmd"); 
+options(digits = 1)
 ```
 
-# Biomass
+## Biomass
 
@@
 (Fig \@ref(fig:biomass)).
-
 Median plot wood density tended to increase marginally significantly 
@@
 with edge distance by 
-`r round(wd$statPrint[[1]]$estimate[[2]],4)` ± 
-`r round(wd$statPrint[[1]]$std.error[[2]],5)` per m 
+`r wd$statPrint[[1]]$estimate[[2]]` ± 
+`r wd$statPrint[[1]]$std.error[[2]]` per m 
 explaining 
@@
 (Fig \@ref(fig:biomass)b).
-
 Plot stem density also tended to increase significantly 
@@
 with edge distance but more strongly and non-linearly, by 
-~`r round(stems12$statPrint[[1]]$estimate[[2]],2)` ± 
-`r round(stems12$statPrint[[1]]$std.error[[2]],2)` per m 
+~`r stems12$statPrint[[1]]$estimate[[2]]` ± 
+`r stems12$statPrint[[1]]$std.error[[2]]` per m 
 (Fig \@ref(fig:biomass)c).
-
 At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge 
@@
 
-```{r biomass, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) stem density, and (d) canopy light availability."}
-knitr::include_graphics(here::here("../figs/fig1.png"))
+```{r biomass, fig.width=3.5, fig.height=3.5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) stem density, and (d) canopy light availability.
+Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
+source(here::here("figs/fig1.R")); 
+fig1
+# knitr::include_graphics(here::here("figs/fig1.png"))
 ```
 
-# Composition
+## Composition
 
@@
 (Fig \@ref(fig:taxa)).
-
 Alpha diversity significantly 
@@
 decreased non-linearly by 
-~`r round(abs(entropy12$statPrint[[1]]$estimate[[2]]),2)` ± 
-`r round(abs(entropy12$statPrint[[1]]$std.error[[2]]),2)` per m 
-as distance to primary forest edge increased 
+~`r abs(entropy12$statPrint[[1]]$estimate[[2]])` ± 
+`r abs(entropy12$statPrint[[1]]$std.error[[2]])` per m 
+as distance to primary forest edge increased, 
 which explained 
@@
 by 
-`r round(richness$statPrint[[1]]$estimate[[2]],2)` ± 
-`r round(richness$statPrint[[1]]$std.error[[2]],3)` per m 
+`r richness$statPrint[[1]]$estimate[[2]]` ± 
+`r richness$statPrint[[1]]$std.error[[2]]` per m 
 and had `r "R2x100"`% variance explained 
@@
 (Fig \@ref(fig:taxa)b).
-
 Community composition and beta diversity also changed significantly 
-(`r glue("p=",round(ordStat$aov.tab[[6]][1],2))`) with 
-`r round(ordStat$aov.tab$R2[1],2) *100`% 
+(`r glue("p=",ordStat[[5]][1])`) with 
+`r ordStat$R2[1] *100`% 
 variance among distance strata explained by distance to primary forest edge, 
@@
 (Fig \@ref(fig:taxa)d).
-
-Key abundant taxa `r measure2e` and `r measure2f` showed different responses -- `r measure2e` nearly tended to decrease with distance to primary forest edge 
+Key abundant taxa *`r measure2e`* and *`r measure2f`* showed different responses -- *`r measure2e`* nearly tended to decrease with distance to primary forest edge 
 (Fig \@ref(fig:taxa)e),
-while `r measure2f` decreased marginally significantly 
+while *`r measure2f`* decreased marginally significantly 
 (`r glue("p=",round(as.numeric(ficus$pval),3))`) 
 and linearly by 
-`r round(ficus$statPrint[[1]]$estimate[[2]],2)` ± 
-`r round(ficus$statPrint[[1]]$std.error[[2]],2)` per m 
+`r ficus$statPrint[[1]]$estimate[[2]]` ± 
+`r ficus$statPrint[[1]]$std.error[[2]]` per m 
 with `r "R2x100"`% variance among distance strata explained 
@@
 
-```{r taxa, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa."}
-knitr::include_graphics(here::here("../figs/fig2.png"))
+```{r taxa, fig.width=5, fig.height=4, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa.
+White and colored dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
+source(here::here("figs/fig2.R"));
+fig2
+# knitr::include_graphics(here::here("figs/fig2.png"))
 ```
 
-# Functions
+## Functions
 
-```{r fig3data}
+```{r fig3data, message=F, results='hide', echo=F}
 source(here::here("figs/fig3.R")); succession <- taxaResultsTbl12 #pReplaced
@@
 (Fig \@ref(fig:traits)).
-
-Taxa associated with both early and late successional stages decreased significantly 
-(`r glue("p=",round(as.numeric(succession$pval[[3]]),3))`) 
+Taxa associated with both early and late successional stages decreased marginally significantly 
+(`r glue("p=",round(as.numeric(succession$pval[[7]][2]),3))`) 
 and curvi-linearly 
 by 
-~`r round(succession$statPrint[[1]]$estimate[[2]],2)` ± 
-`r round(succession$statPrint[[1]]$std.error[[2]],2)` per m 
+~ `r succession$statPrint[[7]]$estimate[[2]]` ± 
+`r succession$statPrint[[7]]$std.error[[2]]` kg per m 
 with distance to edge explaining 
@@
 (Fig \@ref(fig:traits)a).
-
 Dispersal modes did not show consistent trends in biomass with increasing distance to edge 
@@
 
-```{r traits, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode."}
-knitr::include_graphics(here::here("../figs/fig3.png"))
+```{r traits, fig.width=2, fig.height=3, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode.
+Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
+source(here::here("figs/fig3.R"));
+fig3
+# knitr::include_graphics(here::here("figs/fig3.png"))
 ```

### 2022-06-18 figs/fig1.R

@@ -190,20 +190,5 @@ fig1 <- ggarrange(
 fig1
 
 
-ggsave(
-  "fig1.pdf",
-  fig1,
-  path = "figs",
-  width = 3,
-  height = 3,
-  units = "in"
-)
-
-ggsave(
-  "fig1.png",
-  fig1,
-  path = "figs",
-  width = 3,
-  height = 3,
-  units = "in"
-)
+# ggpubr::ggsave("fig1.pdf", fig1, path = "figs", width = 3, height = 3, units = "in")
+# ggpubr::ggsave("fig1.png", fig1, path = "figs", width = 3, height = 3, units = "in")

### 2022-06-18 figs/fig2.R

@@ -210,7 +210,7 @@ graph2d <- ordTbl %>%
     "text",
     label = glue(
       "P = ",
-      ordStat$aov.tab$`Pr(>F)` %>%
+      ordStat$`Pr(>F)` %>%
         first() %>% round(2)
     ),
     x = -5,
@@ -222,7 +222,7 @@ graph2d <- ordTbl %>%
     label = glue(
       "R2",
       " = ",
-      ordStat$aov.tab$R2 %>%
+      ordStat$R2 %>%
         first() %>% round(2)
     ),
     x = 5,

### 2022-06-18 figs/fig3.R

@@ -1,4 +1,4 @@
-library(here); i_am("figs/fig3.R")
+here::i_am("figs/fig3.R")
 #userDiff----
 taxRank <- quote(
   # mainDispersal
@@ -6,7 +6,7 @@ taxRank <- quote(
 )
 #re-run
 #taxaResultsTbl1
-source(here("analysis/stats.R"))
+source(here::here("analysis/stats.R"))
 library(glue)
 library(ggpubr)
 
@@ -196,7 +196,7 @@ taxRank <- quote(
 )
 #re-run
 #taxaResultsTbl1
-source(here("analysis/stats.R"))
+source(here::here("analysis/stats.R"))
 
 
 statData3clean <- taxaResultsTbl1 %>%
@@ -322,3 +322,14 @@ fig3
 #   height = 3,
 #   units = "in"
 # )
+
+
+
+#reset----
+taxRank <- quote(
+  # mainDispersal
+  successionalStage
+)
+#re-run
+#taxaResultsTbl1
+source(here::here("analysis/stats.R"))

### 2022-06-18 figs/map.R

@@ -1,0 +1,6 @@
+here::i_am("figs/map.R")
+site <- magick::image_read(here::here("figs/map.png")) %>% magick::image_ggplot()
+Osa <- magick::image_read(here::here("figs/map_Osa.jpg")) %>% magick::image_ggplot()
+pochote <- magick::image_read(here::here("figs/pochote.jpg")) %>% magick::image_ggplot()
+sitePlot <- ggpubr::ggarrange(Osa, site, pochote, labels = c("a", "b", "c"))
+# ggsave(filename = "figs/figmap.png", width = 3, height = 3, units = "in")

### 2022-06-23 article/disc.Rmd
start draft disc

@@
 It was initially hypothesized that community regeneration would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential -- evidence from this study's results supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
+..
+
+Aboveground biomass represents aboveground carbon stores... is a measure of ..., and responds variably to secondary succession and edge effects ...
+- discuss effects on stem density...
+Management has been found to explain most of biomass variation among tropical forest fragments, and wood density ~10% biomass variation *[@pyles22]*.
+This study adds that management via edge effects can also mediate ~10% of wood density variation, and thereby potentially also indirectly affecting biomass.
+Overall, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon *[@pyles22; others]*.
+On its own, canopy may [not] affect forest biomass, although...
+- fig1 -- edge effects on wood and stem density
+    - although canopy light may still affect biomass generally in this case, without edge effects (check direct light-biomass correlation graph?)
+
+This study supports previous syntheses suggesting that forest biomass and taxonomic richness regenerate more quickly / are more resilient than diversity and community composition *[@poorter21?, other?]*.
+This may be due to priority effects on successional trajectories acting on/at the stem/population level *[]*, variable turnover and reproduction rates among tree taxa *[]*, and/or differential dispersal and establishment success based on local [dispersive] animal community *[]*.
 ...
+- mention more amundant genera patterns w/ Vochysia and Ficus as a key fruit-bearing taxon *[@tropicalbat/bird]*?
 
-Aboveground biomass represents aboveground carbon stores... is a measure of ..., and responds variably to secondary succession and edge effects
+There is increasing interest in functional trait ecology of tropical trees *[@umaña##]*.
+This focuses on a coarse functional association with successional stage, ...
+but also includes patterns with wood density [underlying biomass estmations]...
+- fig3 -- edge effect on succession generalists, but not specialists
+The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density *[@césar21]*, and could therefore require a switch in focal timber species *[@karsten14]*.
 
@@
 based on surrounding forest proximity.
-
-The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density *[@césar21]*, and could therefore require a switch in focal timber species *[@karsten14]*.

### 2022-06-23 article/results.Rmd
start draft disc

@@
 Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
-source(here::here("figs/fig1.R")); 
+source(here::here("figs/fig1.R"), echo = F); 
 fig1
@@
 White and colored dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
-source(here::here("figs/fig2.R"));
+source(here::here("figs/fig2.R"), echo = F);
 fig2
@@
 Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
-source(here::here("figs/fig3.R"));
+source(here::here("figs/fig3.R"), echo = F);
 fig3

### 2022-07-05 article/methods.Rmd
checked woodC22db

@@
 
+**Trait matching with *[@doraisami22]* **
+
 ## Statistical analyses
@@
 Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
-Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and are available at [nmedina17.github.com/osa](nmedina17.github.com/osa).
+Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and are stored at [nmedina17.github.com/osa](nmedina17.github.com/osa).
 

### 2022-07-05 analysis/stats.R

@@ -18,9 +18,7 @@ library(ggbeeswarm); library(ggpmisc)
 
 
 # #userInput--parentFile
-# taxRank <- quote(
-#   gen
-# )
+# taxRank <- quote(gen)
 
 
 
@@ -92,27 +90,17 @@ plotVarsTbl <- cleanData %>%
   ) %>%
   select(
     #abiotic
-    luz,
-    incl,
+    luz, incl,
 
     #biotic
-    dap,
-    h,
-    spg,
-    kg17,
+    dap, h, spg, kg17, tissue.c,
 
     #community
-    stems,
-    richness,
-    entropy
+    stems, richness, entropy
   ) %>%
 
-  get_summary_stats(
-    type = "median_mad"
-  ) %>%
-  nest(
-    "varData" = -variable
-  ) %>%
+  rstatix::get_summary_stats(type = "median_mad") %>%
+  tidyr::nest("varData" = -variable) %>%
 
   #nextlevel
 
@@ -150,7 +138,12 @@ plotVarsTbl <- cleanData %>%
             .,
             plot = F
           )
-      )
+      ),
+    "varData0" = cleanData |>
+      tidyr::nest("varData0" = everything()),
+    "varData0col" = varData0 #%>% modify(
+      # ~ .x %>% select(where(colnames(.x$varData0) == .x$variable))
+        # lattice::histogram(plot = F))
   )
 
 
@@ -317,9 +310,15 @@ mainModel <- {
 
 #OGpackage
 plotResultsTbl <- plotVarsTbl %>%
-  getStatsTbl(
+  oir::getStatsTbl(
     mainModel
   )
+
+
+##woodCcheck
+# ggplot2::ggplot(plotResultsTbl$varData[[10]], aes(dist, median)) +
+#   geom_point() + geom_smooth(method = "lm", formula = y ~ poly(x, 2)) +
+#   stat_poly_eq(aes(label = after_stat(p.value.label)))
 
 
 ### quad----

### 2022-07-05 data/clean.R

@@ -90,6 +90,12 @@ lifeHist <- vroom(
   )
 
 
+##woodC----
+woodC <- vroom::vroom(here::here(
+  "data/base/glowcad/Doraisami_et_al._2021_Wood_C_Database.csv")) |>
+  select(tissue.c, family.resolved, genus.resolved, binomial.resolved)
+
+
 updateRaw <- function(
   ...rawData
 ) {
@@ -154,9 +160,7 @@ cleanData <- rawData %>%
     )
   ) %>%
 
-  full_join(
-    dispersal
-  ) %>%
+  full_join(dispersal) %>%
   filter(
     !is.na(
       dist
@@ -174,6 +178,8 @@ cleanData <- rawData %>%
     traits,
     by = "gen"
   ) %>%
+
+  full_join(woodC, by = c("gen" = "genus.resolved")) |>
 
 
   filter(

### 2022-07-11 article/abs.Rmd
first poster draft

--- 2022-05-15 article/abs.Rmd
@@
 Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
-This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
+This study addressed how wet tropical secondary forest regeneration was affected by distance to habitat edge, and hypothesized that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
 An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.
@@
 Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.
+
+\newpage
+
+# Abstracto{-}
+
+Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
+Biomasa arboreal y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
+Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
+Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
+Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.
+
+Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10% variación explicado, mientras tanto biomasa y luz de dócel tendió a quedarse igual.
+Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
+Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, pero biomasa por modo de disperso no tendió a cambiar.
+Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.

### 2022-07-11 article/disc.Rmd
first poster draft

--- 2022-05-15 article/disc.Rmd
@@
 
-1.  assess hypotheses
-2.  infer causes of patterns - plantation structure, old legacy trees, or species traits?
-3.  compare inferences to other studies
-4.  recommend future conservation approaches, ecological studies
+<!-- # outline{-} -->
+<!-- 1.  assess hypotheses -->
+<!-- 2.  infer causes of patterns - plantation structure, old legacy trees, or species traits? -->
+<!-- 3.  compare inferences to other studies -->
+<!-- 4.  recommend future conservation approaches, ecological studies -->
 
-Under more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e.
+<!-- # draft{-} -->
+
+This study found significant edge effects on secondary forest stand wood and stem density, tree diversity and community composition, and some functional traits, yet not on overall aboveground biomass.
+<!-- These results were shown in Figs \@ref(fig:biomass), \@ref(fig:taxa), and \@ref(fig:traits)... -->
+It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential, and this evidence supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
+Accordingly this suggests that edge effects do significantly change humid tropical forest community composition, but primarily due to factors other than light competition or dispersal limitation, as well as that aboveground biomass and even low taxonomic richness is resilient after a decade of regeneration.
+
+Aboveground biomass storage is a key global ecosystem function and service, but edge effects tend to release about one-third of carbon lost from tropical deforestation *[@brinck17]*, with more expected in the future *[@mitchard18]*.
+<!-- responds variably to secondary succession and edge effects ...   -->
+Management has been found to explain most of biomass variation among tropical forest fragments, and wood density ~10% biomass variation *[@pyles22]*.
+This study adds that management via edge effects can also mediate ~10% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns *[@baker04]*.
+Overall, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon *[@gonzalezj98; @pyles22]*, and increasing nearby stem density.
+Lower wood density can also reflect more dynamic forest stands *[@malhi06]* with higher tree turnover, a consequence of forest fragmentation *[@nascimento04a]* together with other global change drivers *[@laurance14]*, as well as may reflect possible differences in soil fertility *[@malhi04]*, all of which represent possible future research directions, beyond strictly neutral dynamics *[@terborgh95]*.
+Furthermore, light availability at the plot level may not affect biomass storage, but instead individual light gaps may be more important for biomass dynamics *[@chazdon84]*.
+    <!-- - although canopy light may still affect biomass generally in this case, without edge effects (check direct light-biomass correlation graph? -- no sig correlation) -->
+
+Edge effects have been shown to lower both forest biomass and also diversity, and this study adds that the decline can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
+The rapid diversity loss across the edge gradient may be in part due to finer variations in dispersal patterns among animal taxa, rather than overall dispersal mode more broadly.
+Similarly, stochastic factors like priority effects after dispersal may also affect offspring establishment and re-growth.
+While the biomass of the most abundant taxon *Vochysia* was resilient to edge effects, that of the key taxon for frugivores like birds and bats *Ficus* was significantly lowered across the forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.
+
+<!-- # draft attic{-} -->
+
+<!-- This study supports previous syntheses suggesting that forest biomass and taxonomic richness regenerate more quickly / are more resilient than diversity and community composition *[@poorter21?, other?]*.
+-->
+<!-- This may be due to priority effects on successional trajectories acting on/at the stem/population level *[]*, variable turnover and reproduction rates among tree taxa *[]*, and/or differential dispersal and establishment success based on local [dispersive] animal community *[]*.
+...
+-->
+<!-- - mention more amundant genera patterns w/ *Vochysia* and *Ficus* as a key fruit-bearing taxon *[@tropicalbat/bird]*? -->
+
+<!-- There is increasing interest in functional trait ecology of tropical trees *[@umaña##]*. -->
+<!-- This focuses on a coarse functional association with successional stage, ... -->
+<!-- but also includes patterns with wood density [underlying biomass estmations]... -->
+<!-- - fig3 -- edge effect on succession generalists, but not specialists -->
+<!-- The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density *[@césar21]*, and could therefore require a switch in focal timber species *[@karsten14]*.
+-->
+
+<!-- # attic{-} -->
+
+<!-- Under more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e.
 near 1:1 replacement) the targeted tree cutting that is done using information collected about the ecological profiles and ages (when possible) of trees (Duah-Gyamfi et al 2014, Edwards et al 2014).
+-->
 
-In contrast, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015) while also usually showing negative trade-offs with other services, like that with biodiversity given reliance on monoculture planting designs.
+<!-- In contrast, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015) while also usually showing negative trade-offs with other services, like that with biodiversity given reliance on monoculture planting designs.
+-->
 
-In some cases where management decisions are made primarily based on regional economic but not local ecological information, due to limited access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.
+<!-- In some cases where management decisions are made primarily based on regional economic but not local ecological information, due to limited access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.
+-->
 
-After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g.
+<!-- After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g.
 based on surrounding forest proximity.
-
-The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density *[@césar21]*, and could therefore require a switch in focal timber species *[@karsten14]*.
+-->

### 2022-07-11 article/merge.Rmd
start draft disc
first poster draft

--- 2022-04-25 article/merge.Rmd
@@
 ---
-title: ""
+title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
+author: 
+  - Nicholas Medina #^a^*
+  # - Max Villalobos? ^b^
+  # - Andrea Johnson? ^b^
+date: 
+# scriptsize\
+  Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109
+  # ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; 
+  # ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
+# author: 
+#   - name: "Nicholas Medina"
+#     email: nmedina@umich.edu
+#     affiliation: Brandeis University
+#     orcid_id: 0000-0001-5465-3988
+# address:
+#   - code: Brandeis University
+#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
+# keywords: [1, 2, 3]
+
 output: 
-  html_notebook
-  bookdown::html_document2
-    number_sections: false
-bibliography: ../../library.bib
+  # html_notebook
+  # bookdown::html_document2:
+  #   # df_print: paged
+  #   global_numbering: TRUE
+  #   toc: FALSE
+  bookdown::pdf_document2:
+    # number_sections: FALSE
+    global_numbering: TRUE
+    keep_tex: TRUE
+    toc: FALSE
+bibliography: 
+  - ../../library.bib
+  - ../../grateful-refs.bib
+  - ../../software.bib
+link-citations: TRUE
+csl: ../../styles-master/elsevier-harvard.csl
+
+fontsize: 12pt
+geometry: margin=1in
+header-includes:
+  - \usepackage[left]{lineno}
+  - \linenumbers
+  - \usepackage{setspace}\doublespace
+# abstract
 ---
 
-# Abstract
+<!-- journals: J [Appl] Veg Sci, Basic Appl Ecol (3.5), Biotropica (2.5), Eco Evo -->
 
-```{r child='abstract.Rmd'}
+\
+\
+
+**Keywords**: secondary forest; regeneration; timber plantation associated biodiversity; edge effect; aboveground biomass; shade
+
+\
+\
+
+***Corresponding author**: Nicholas Medina, nmedina@umich.edu
+
+\newpage
+
+# Abstract{-}
+
+```{r child='abs.Rmd'}
 ```
 
-# Intro
+\newpage
+
+# Introduction
 
@@
 
-# Refs
+\newpage
+
+# Funding{-}
+
+This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship.
+
+# Declaration of interests{-}
+
+Authors declare no conflicts of interest.
+
+# Acknowledgements{-}
+
+Thanks to Professor Dan Perlman for initial networking; concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision; Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; peer James Fifer for discussion of early draft; and Luisa Valdez for translation assistance.
+
+# Author contributions{-}
+
+<!-- MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis;  -->
+NM wrote initial draft and later revisions.
+
+# Data statement{-}
+
+Code stored at [github.com/nmedina17/osa](github.com/nmedin17/osa).
+
+\newpage
+
+# References

### 2022-07-11 article/merge.tex
first poster draft

@@
   pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
-  pdfauthor={Nicholas Medina a*; Max Villalobos b; Andrea Johnson b},
+  pdfauthor={Nicholas Medina},
   hidelinks,
@@
 \title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
-\author{Nicholas Medina \textsuperscript{a}* \and Max Villalobos \textsuperscript{b} \and Andrea Johnson \textsuperscript{b}}
-\date{scriptsize~\textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}
+\author{Nicholas Medina}
+\date{scriptsize~Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109}
 
@@
 \maketitle
-
-journals: J {[}Appl{]} Veg Sci,
 
@@
 Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
-Biomasa arboreal y \emph{riqueza} taxonomica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
-Dado que la mayoría de bosques están entre un km de su orilla, \emph{efectos de orilla} probablemente afectan la regeneración de composición de comnuidad, mediado por tolerancia a sombra y ``traits''/characterísticas de \emph{disperso}, y también historia de manejamiento.
-Este estudio \emph{investiga} cómo regeneración de bosque secundario tropical húmedo es afectado por distancia a \emph{orilla de habitat}, y \emph{``hypothesizes''} que plantaciones de madera viejos facilitan regeneración porque favoran taxa tolerante a sombra o de \emph{``succession''} tarde, y que orillas de bosque cercanos/\emph{``adjacent''} \emph{median} composición de comunidad.
-Una plantación de madera neotropical \emph{al lado de} bosque primario fue \emph{``censused''} por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combino taxa con characterísticas/\emph{``traits''} usando literatura relevante.
+Biomasa arboreal y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
+Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
+Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
+Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.
 
-Mientras distancia de bosque primario aumentó, densidad de madera y \emph{``stems''} \emph{``tended to''} aumentó \emph{signifantemente}, con 10\% variación explicado, mientras tanto biomasa y luz de dócel \emph{``tended to''} quedarse igual.
-\emph{Riquesa} de árboles también \emph{``tended to''} aumentar \emph{significantemente}, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
-Finalmente, taxa de árboles asociado \emph{etapas} de \emph{``succession''} ambos temprano y tarde bajó \emph{significativamente}, pero biomasa por modo de \emph{disperso} no \emph{``tended to''} cambiar.
-\emph{Entre todo} este estudio apoya que composición es menos resiliente y más sujeto a \emph{efectos de orilla} que biomasa y \emph{riqueza}, sugeriendo que bosques globales probablemente serán \emph{asemblajes} nuevos distintos en el futuro, con \emph{``trade-offs''} entre producción de madera y biodiversidad \emph{ocuriendo} basado en actividad de manejamiento local y regional.
+Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10\% variación explicado, mientras tanto biomasa y luz de dócel tendió a quedarse igual.
+Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
+Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, pero biomasa por modo de disperso no tendió a cambiar.
+Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.
 
@@
 Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa \emph{(\protect\hyperlink{ref-carlo16}{Carlo and Morales, 2016})}, which can be keystone to locally-adapted food webs, including birds \emph{(\protect\hyperlink{ref-maas16}{Maas et al., 2016})}.
-As a result, understanding secondary forest regeneration remains important\ldots{}
+Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, yet local management remains the strongest driver of tropical forest carbon stores \emph{(\protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}.
 
@@
 This census was done in 2013 during the rainy season months between June and August.
+
 \includegraphics{merge_files/figure-latex/map-1.pdf}
@@
 
-Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \ref{fig:map}), using available GIS software (\emph{ArcGIS 10 from 2010}, \url{esri.com}, and \emph{QGIS 2 from 2013}, \url{qgis.org}).
+Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \ref{fig:map}), using available GIS software (\emph{ArcGIS 10}, \url{esri.com}, and \emph{QGIS 2}, \url{qgis.org}).
 Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
@@
 
-The diameter of all stems \textgreater10 cm wide were recorded in each census plot.
+The diameter of all stems \textgreater10 cm wide were recorded in each census plot, totaling over 1,000 trees.
 Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
@@
 Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be \emph{0.58} following World Agroforestry Database guidelines \href{worldagroforestry.org}{(worldagroforestry.org)}.
-
 Original values found and recorded were later updated to modern values using the pipeline developed in the \emph{BIOMASS} 2.1.8 R software package by \emph{Réjou-Méchain et al.
 (\protect\hyperlink{ref-rejou-mechain17}{2017})}, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.
+
+Traits including successional stage and main dispersal mode were matched to species with \emph{Werden et al.
+(\protect\hyperlink{ref-werden20}{2020})} data, and wood tissue carbon matched to genera with the GLOWCAD database \emph{(\protect\hyperlink{ref-doraisami22}{Doraisami et al., 2022})} given high alignment at species levels \emph{(\protect\hyperlink{ref-chave06}{Chave et al., 2006})}.
 
@@
 
-For all response variables, data were analyzed using plot medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base R version 4.2.0 (2022-04-22) and \emph{rstatix} 0.7.0 package functions \emph{(\protect\hyperlink{ref-base}{R Core Team, 2022})}.
+For all response variables, data were analyzed using plot or distance strata medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base R version 4.2.0 (2022-04-22) and \emph{rstatix} 0.7.0 package functions \emph{(\protect\hyperlink{ref-base}{R Core Team, 2022})}.
 
@@
 Community analysis and ordination was done as a PERMANOVA with the \emph{adonis()} function from the \emph{vegan} 2.6.2 R package \emph{(\protect\hyperlink{ref-dixon03}{Dixon, 2003})}.
-Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and are available at \url{nmedina17.github.com/osa}.
+Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and internal pipeline \emph{oir} \emph{(\protect\hyperlink{ref-softwareMedina22}{Medina, 2022})} are stored at \url{github.com/nmedina17/osa}.
 
@@
 (Fig \ref{fig:biomass}).
-
 Median plot wood density tended to increase marginally significantly
-(p=0.105)
+(p=0.598)
 with edge distance by
-\ensuremath{1.9\times 10^{-4}} ±
-\ensuremath{9.2\times 10^{-5}} per m
+\ensuremath{8.2\times 10^{-5}} ±
+\ensuremath{1.4\times 10^{-4}} per m
 explaining
@@
 (Fig \ref{fig:biomass}b).
-
 Plot stem density also tended to increase significantly
-(p=0.03)
+(p=0.82)
 with edge distance but more strongly and non-linearly, by
-\textasciitilde17.4 ±
-3.4 per m
+\textasciitilde-65.9 ±
+78.9 per m
 (Fig \ref{fig:biomass}c).
-
 At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge
@@
 (Fig \ref{fig:taxa}).
-
-Alpha diversity significantly
-(p=0.003)
+Diversity significantly
+(p=0.414)
 decreased non-linearly by
-\textasciitilde0.5 ±
-0 per m
-as distance to primary forest edge increased
+\textasciitilde0.1 ±
+0.1 per m
+as distance to primary forest edge increased,
 which explained
@@
 (Fig \ref{fig:taxa}b).
-
 Community composition and beta diversity also changed significantly
 (p=0.01) with
-11.4\%
+25.6\%
 variance among distance strata explained by distance to primary forest edge,
 and the first two principal components explaining
-47\% and
-23\% totaling
-70\%
+59\% and
+24\% totaling
+83\%
 of variance among calculated plot distances
 (Fig \ref{fig:taxa}d).
-
-Key abundant taxa Vochysia and Ficus showed different responses -- Vochysia nearly tended to decrease with distance to primary forest edge
+Key abundant taxa \emph{Vochysia} and \emph{Ficus} showed different responses -- \emph{Vochysia} nearly tended to decrease with distance to primary forest edge
 (Fig \ref{fig:taxa}e),
-while Ficus decreased marginally significantly
+while \emph{Ficus} decreased marginally significantly
 (p=0.088)
@@
 
-\hypertarget{functions}{%
-\subsection{Functions}\label{functions}}
+\hypertarget{traits}{%
+\subsection{Traits}\label{traits}}
 
-Overall functional regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest
+Overall trait regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest
 (Fig \ref{fig:traits}).
-
-Taxa associated with both early and late successional stages decreased marginally significantly
+Taxa associated with both early and late successional stages decreased significantly
 (p=0.042)
@@
 \textasciitilde{} -584.9 ±
-167.3 per m
+167.3 kg per m
 with distance to edge explaining
@@
 (Fig \ref{fig:traits}a).
-
 Dispersal modes did not show consistent trends in biomass with increasing distance to edge
@@
 
-\hypertarget{outline}{%
-\section*{outline}\label{outline}}
-\addcontentsline{toc}{section}{outline}
+This study found significant edge effects on secondary forest stand wood and stem density, tree diversity and community composition, and some functional traits, yet not on overall aboveground biomass.
+It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential, and this evidence supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
+Accordingly this suggests that edge effects do significantly change humid tropical forest community composition, but primarily due to factors other than light competition or dispersal limitation, as well as that aboveground biomass and even low taxonomic richness is resilient after a decade of regeneration.
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  assess hypotheses
-\item
-  infer causes of patterns - plantation structure, old legacy trees, or species traits?
-\item
-  compare inferences to other studies
-\item
-  recommend future conservation approaches, ecological studies
-\end{enumerate}
+Aboveground biomass storage is a key global ecosystem function and service, but edge effects tend to release about one-third of carbon lost from tropical deforestation \emph{(\protect\hyperlink{ref-brinck17}{Brinck et al., 2017})}, with more expected in the future \emph{(\protect\hyperlink{ref-mitchard18}{Mitchard, 2018})}.
+Management has been found to explain most of biomass variation among tropical forest fragments, and wood density \textasciitilde10\% biomass variation \emph{(\protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}.
+This study adds that management via edge effects can also mediate \textasciitilde10\% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns \emph{(\protect\hyperlink{ref-baker04}{Baker et al., 2004})}.
+Overall, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon \emph{(\protect\hyperlink{ref-gonzalezj98}{Gonzalez J and Fisher, 1998}; \protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}, and increasing nearby stem density.
+Lower wood density can also reflect more dynamic forest stands \emph{(\protect\hyperlink{ref-malhi06}{Malhi et al., 2006})} with higher tree turnover, a consequence of forest fragmentation \emph{(\protect\hyperlink{ref-nascimento04a}{Nascimento and Laurance, 2004})} together with other global change drivers \emph{(\protect\hyperlink{ref-laurance14}{Laurance et al., 2014})}, as well as may reflect possible differences in soil fertility \emph{(\protect\hyperlink{ref-malhi04}{Malhi et al., 2004})}, all of which represent possible future research directions, beyond strictly neutral dynamics \emph{(\protect\hyperlink{ref-terborgh95}{Terborgh et al., 1995})}.
+Furthermore, light availability at the plot level may not affect biomass storage, but instead individual light gaps may be more important for biomass dynamics \emph{(\protect\hyperlink{ref-chazdon84}{Chazdon and Fetcher, 1984})}.
 
-\hypertarget{draft}{%
-\section*{draft}\label{draft}}
-\addcontentsline{toc}{section}{draft}
+Edge effects have been shown to lower both forest biomass and also diversity, and this study adds that the decline can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
+The rapid diversity loss across the edge gradient may be in part due to finer variations in dispersal patterns among animal taxa, rather than overall dispersal mode more broadly.
+Similarly, stochastic factors like priority effects after dispersal may also affect offspring establishment and re-growth.
+While the biomass of the most abundant taxon \emph{Vochysia} was resilient to edge effects, that of the key taxon for frugivores like birds and bats \emph{Ficus} was significantly lowered across the forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.
 
-This study found significant edge effects on tree diversity and community composition, but not aboveground biomass\ldots functional diversity?
-These results were shown in Figs \ref{fig:biomass}, \ref{fig:taxa}, and \ref{fig:traits}\ldots{}
+\hypertarget{draft-attic}{%
+\section*{draft attic}\label{draft-attic}}
+\addcontentsline{toc}{section}{draft attic}
+
+This study supports previous syntheses suggesting that forest biomass and taxonomic richness regenerate more quickly / are more resilient than diversity and community composition \emph{(\protect\hyperlink{ref-poorter21}{Poorter et al., 2021}?, other?)}.
+This may be due to priority effects on successional trajectories acting on/at the stem/population level \emph{{[}{]}}, variable turnover and reproduction rates among tree taxa \emph{{[}{]}}, and/or differential dispersal and establishment success based on local {[}dispersive{]} animal community \emph{{[}{]}}.
+\ldots{}
+- mention more amundant genera patterns w/ \emph{Vochysia} and \emph{Ficus} as a key fruit-bearing taxon \emph{(\protect\hyperlink{ref-tropicalbatux2fbird}{\textbf{tropicalbat/bird?}})}?
+
+There is increasing interest in functional trait ecology of tropical trees \emph{(\protect\hyperlink{ref-umauxf1a}{\textbf{umaña?}}\#\#)}.
+This focuses on a coarse functional association with successional stage, \ldots{}
+but also includes patterns with wood density {[}underlying biomass estmations{]}\ldots{}
+- fig3 -- edge effect on succession generalists, but not specialists
+The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density \emph{(\protect\hyperlink{ref-cuxe9sar21}{\textbf{césar21?}})}, and could therefore require a switch in focal timber species \emph{(\protect\hyperlink{ref-karsten14}{Karsten et al., 2014})}.
 
@@
 \addcontentsline{toc}{section}{attic}
-
-Under more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e.~near 1:1 replacement) the targeted tree cutting that is done using information collected about the ecological profiles and ages (when possible) of trees (Duah-Gyamfi et al 2014, Edwards et al 2014).
 
@@
 After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g.~based on surrounding forest proximity.
-
-The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density \emph{(\protect\hyperlink{ref-cuxe9sar21}{\textbf{césar21?}})}, and could therefore require a switch in focal timber species \emph{(\protect\hyperlink{ref-karsten14}{Karsten et al., 2014})}.
 
@@
 
-Thanks to Professor Dan Perlman for initial networking; Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; concurrent Osa Conservation General Manager Manuel Ramirez and Piro Research Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; and previous anonymous reviewers and peer James Fifer for discussion of drafts.
+Thanks to Professor Dan Perlman for initial networking; concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision; Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; and peer James Fifer for discussion of early draft.
 
@@
 
-MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis; NM contributed to initial draft writing and later revisions.
+NM wrote initial draft and later revisions.
 
@@
 
-Data and code stored at \href{github.com/nmedin17/osa}{github.com/nmedina17/osa}.
+Code stored at \href{github.com/nmedin17/osa}{github.com/nmedina17/osa}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-baker04}{}}%
+Baker, T.R., Phillips, O.L., Malhi, Y., Almeida, S., Arroyo, L., Di Fiore, A., Erwin, T., Killeen, T.J., Laurance, S.G., Laurance, W.F., Lewis, S.L., Lloyd, J., Monteagudo, A., Neill, D.A., Patiño, S., Pitman, N.C.A., M.
+Silva, J.N., Vásquez Martínez, R., 2004.
+Variation in wood density determines spatial patterns {inAmazonian} forest biomass: {Wood} specific gravity and {Amazonian} biomass estimates.
+Global Change Biology 10, 545--562.
+\url{https://doi.org/10.1111/j.1365-2486.2004.00751.x}
+
+\leavevmode\vadjust pre{\hypertarget{ref-brinck17}{}}%
+Brinck, K., Fischer, R., Groeneveld, J., Lehmann, S., Dantas De Paula, M., Pütz, S., Sexton, J.O., Song, D., Huth, A., 2017.
+High resolution analysis of tropical forest fragmentation and its impact on the global carbon cycle.
+Nature Communications 8, 14855.
+\url{https://doi.org/10.1038/ncomms14855}
+
 \leavevmode\vadjust pre{\hypertarget{ref-carlo16}{}}%
@@
 \url{https://doi.org/10.1007/s00442-005-0100-x}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chave06}{}}%
+Chave, J., Muller-Landau, H.C., Baker, T.R., Easdale, T.A., Steege, H.
+ter, Webb, C.O., 2006.
+{REGIONAL AND PHYLOGENETIC VARIATION OF WOOD DENSITY ACROSS} 2456 {NEOTROPICAL TREE SPECIES}.
+Ecological Applications 16, 2356--2367.
+\url{https://doi.org/10.1890/1051-0761(2006)016\%5B2356:RAPVOW\%5D2.0.CO;2}
 
@@
 \url{https://doi.org/10.1126/sciadv.1501639}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chazdon84}{}}%
+Chazdon, R.L., Fetcher, N., 1984. Photosynthetic {Light Environments} in a {Lowland Tropical Rain Forest} in {Costa Rica}. The Journal of Ecology 72, 553. \url{https://doi.org/10.2307/2260066}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-doraisami22}{}}%
+Doraisami, M., Kish, R., Paroshy, N.J., Domke, G.M., Thomas, S.C., Martin, A.R., 2022.
+A global database of woody tissue carbon concentrations.
+Scientific Data 9, 284.
+\url{https://doi.org/10.1038/s41597-022-01396-1}
+
 \leavevmode\vadjust pre{\hypertarget{ref-edwards14}{}}%
@@
 Finegan, B., 1996. Pattern and process in neotropical secondary rain forests: The first 100 years of succession. Trends in Ecology \& Evolution.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gonzalezj98}{}}%
+Gonzalez J, E., Fisher, R.F., 1998. Variation in {Selected Wood Properties} of {Vochysia Guatemalensis} from {Four Sites} in {Costa Rica}. Forest Science 7.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-laurance14}{}}%
+Laurance, W.F., Andrade, A.S., Magrach, A., Camargo, J.L.C., Campbell, M., Fearnside, P.M., Edwards, W., Valsko, J.J., Lovejoy, T.E., Laurance, S.G., 2014.
+Apparent environmental synergism drives the dynamics of {Amazonian} forest fragments.
+Ecology 95, 3018--3026.
+\url{https://doi.org/10.1890/14-0330.1}
+
 \leavevmode\vadjust pre{\hypertarget{ref-lefland18}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-malhi04}{}}%
+Malhi, Y., Baker, T.R., Phillips, O.L., Almeida, S., Alvarez, E., Arroyo, L., Chave, J., Czimczik, C.I., Fiore, A.D., Higuchi, N., Killeen, T.J., Laurance, S.G., Laurance, W.F., Lewis, S.L., Montoya, L.M.M., Monteagudo, A., Neill, D.A., Vargas, P.N., Patiño, S., Pitman, N.C.A., Quesada, C.A., Salomão, R., Silva, J.N.M., Lezama, A.T., Martínez, R.V., Terborgh, J., Vinceti, B., Lloyd, J., 2004.
+The above-ground coarse wood productivity of 104 {Neotropical} forest plots: {PRODUCTIVITY OF} 104 {NEOTROPICAL FOREST PLOTS}.
+Global Change Biology 10, 563--591.
+\url{https://doi.org/10.1111/j.1529-8817.2003.00778.x}
+
+\leavevmode\vadjust pre{\hypertarget{ref-malhi06}{}}%
+Malhi, Y., Wood, D., Baker, T.R., Wright, J., Phillips, O.L., Cochrane, T., Meir, P., Chave, J., Almeida, S., Arroyo, L., Higuchi, N., Killeen, T.J., Laurance, S.G., Laurance, W.F., Lewis, S.L., Monteagudo, A., Neill, D.A., Vargas, P.N., Pitman, N.C.A., Quesada, C.A., Salomão, R., Silva, J.N.M., Lezama, A.T., Terborgh, J., Martínez, R.V., Vinceti, B., 2006.
+The regional variation of aboveground live biomass in old-growth {Amazonian} forests: {BIOMASS IN AMAZONIAN FORESTS}.
+Global Change Biology 12, 1107--1138.
+\url{https://doi.org/10.1111/j.1365-2486.2006.01120.x}
+
+\leavevmode\vadjust pre{\hypertarget{ref-softwareMedina22}{}}%
+Medina, N., 2022. {Code for: Automating statistical decision trees for data exploration and analysis revision workflow}. \url{https://doi.org/10.5281/zenodo.6800299}
+
 \leavevmode\vadjust pre{\hypertarget{ref-melo13}{}}%
@@
 \url{https://doi.org/10.1016/j.tree.2013.01.001}
+
+\leavevmode\vadjust pre{\hypertarget{ref-mitchard18}{}}%
+Mitchard, E.T.A., 2018. The tropical forest carbon cycle and climate change. Nature 559, 527--534. \url{https://doi.org/10.1038/s41586-018-0300-2}
 
@@
 \url{https://doi.org/10.1017/cbo9780511541971.012}
+
+\leavevmode\vadjust pre{\hypertarget{ref-nascimento04a}{}}%
+Nascimento, H.E.M., Laurance, W.F., 2004. {BIOMASS DYNAMICS IN AMAZONIAN FOREST FRAGMENTS}. Ecological Applications 14, 127--138. \url{https://doi.org/10.1890/01-6003}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-pyles22}{}}%
+Pyles, M.V., Magnago, L.F.S., Maia, V.A., Pinho, B.X., Pitta, G., de Gasper, A.L., Vibrans, A.C., dos Santos, R.M., van den Berg, E., Lima, R.A.F., 2022.
+Human impacts as the main driver of tropical forest carbon.
+Science Advances 8, eabl7968.
+\url{https://doi.org/10.1126/sciadv.abl7968}
+
 \leavevmode\vadjust pre{\hypertarget{ref-pyron22}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-terborgh95}{}}%
+Terborgh, J., Foster, R.B., Percy, N.V., 1995. Tropical {Tree Communities}: {A Test} of the {Nonequilibrium Hypothesis}. Ecology 77, 561--567. \url{https://doi.org/10.2307/2265630}
+
 \leavevmode\vadjust pre{\hypertarget{ref-thomson11}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-werden20}{}}%
+Werden, L.K., Holl, K.D., Rosales, J.A., Sylvester, J.M., Zahawi, R.A., 2020.
+Effects of dispersal- and niche-based factors on tree recruitment in tropical wet forest restoration.
+Ecological Applications 30.
+\url{https://doi.org/10.1002/eap.2139}
+
 \leavevmode\vadjust pre{\hypertarget{ref-bookdown2022}{}}%

### 2022-07-11 article/methods.Rmd
first poster draft

--- 2022-06-18 article/methods.Rmd
@@
 - ../../grateful-refs.bib
+- ../../software.bib
 link-citations: TRUE
@@
 
-```{r map, fig.width=4, fig.height=4, fig.cap="Map of conservation areas and study site on Osa Peninsula, Costa Rica. Map image data © 2022 Osa Conservation. Photo credit: Nicholas Medina."}
+```{r map, fig.width=4, fig.height=4, fig.cap="Map of (a) conservation areas and (b) study site of (c) Bombacopsis quinata timber plantation on Osa Peninsula, Costa Rica.
+Map image data © 2022 Osa Conservation.
+Photo credit: Nicholas Medina."}
 source(here::here("figs/map.R"));
@@
 
-The diameter of all stems >10 cm wide were recorded in each census plot.
+The diameter of all stems >10 cm wide were recorded in each census plot, totaling over 1,000 trees.
 Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
@@
 Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org).
+<!-- *[References used for this step are given in *Table S1*.] * -->
+Original values found and recorded were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.
+<!-- However, these updates did not appear to change the main results of this study. -->
 
-<!-- *[References used for this step are given in *Table S1*.] * -->
-
-Original values found and recorded were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.
-
-<!-- However, these updates did not appear to change the main results of this study. -->
+Traits including successional stage and main dispersal mode were matched to species with *@werden20* data, and wood tissue carbon matched to genera with the GLOWCAD database *[@doraisami22]* given high alignment at species levels *[@chave06]*.
 
@@
 
-For all response variables, data were analyzed using plot medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base `r version$version.string` and *rstatix* `r packageVersion("rstatix")` package functions *[@base]*.
+For all response variables, data were analyzed using plot or distance strata medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base `r version$version.string` and *rstatix* `r packageVersion("rstatix")` package functions *[@base]*.
  <!-- including biomass converted to a per hectare basis,  -->
@@
 Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
-Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and are available at [nmedina17.github.com/osa](nmedina17.github.com/osa).
+Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22]* are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa).
 

### 2022-07-11 article/results.Rmd
first poster draft

@@
 (Fig \@ref(fig:taxa)).
-Alpha diversity significantly 
+Diversity significantly 
 (`r glue("p=",round(as.numeric(entropy12$pval[[2]][2]),3))`) 
@@
 
-## Functions
+## Traits
 
@@
 
-Overall functional regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest 
+Overall trait regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest 
 (Fig \@ref(fig:traits)).
-Taxa associated with both early and late successional stages decreased marginally significantly 
+Taxa associated with both early and late successional stages decreased significantly 
 (`r glue("p=",round(as.numeric(succession$pval[[7]][2]),3))`) 

### 2022-07-11 article/merge.tex
first poster draft

@@
 \author{Nicholas Medina}
-\date{scriptsize~Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109}
+\date{Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109}
 
@@
 
-\hypertarget{draft-attic}{%
-\section*{draft attic}\label{draft-attic}}
-\addcontentsline{toc}{section}{draft attic}
-
-This study supports previous syntheses suggesting that forest biomass and taxonomic richness regenerate more quickly / are more resilient than diversity and community composition \emph{(\protect\hyperlink{ref-poorter21}{Poorter et al., 2021}?, other?)}.
-This may be due to priority effects on successional trajectories acting on/at the stem/population level \emph{{[}{]}}, variable turnover and reproduction rates among tree taxa \emph{{[}{]}}, and/or differential dispersal and establishment success based on local {[}dispersive{]} animal community \emph{{[}{]}}.
-\ldots{}
-- mention more amundant genera patterns w/ \emph{Vochysia} and \emph{Ficus} as a key fruit-bearing taxon \emph{(\protect\hyperlink{ref-tropicalbatux2fbird}{\textbf{tropicalbat/bird?}})}?
-
-There is increasing interest in functional trait ecology of tropical trees \emph{(\protect\hyperlink{ref-umauxf1a}{\textbf{umaña?}}\#\#)}.
-This focuses on a coarse functional association with successional stage, \ldots{}
-but also includes patterns with wood density {[}underlying biomass estmations{]}\ldots{}
-- fig3 -- edge effect on succession generalists, but not specialists
-The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density \emph{(\protect\hyperlink{ref-cuxe9sar21}{\textbf{césar21?}})}, and could therefore require a switch in focal timber species \emph{(\protect\hyperlink{ref-karsten14}{Karsten et al., 2014})}.
-
-\hypertarget{attic}{%
-\section*{attic}\label{attic}}
-\addcontentsline{toc}{section}{attic}
-
-In contrast, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015) while also usually showing negative trade-offs with other services, like that with biodiversity given reliance on monoculture planting designs.
-
-In some cases where management decisions are made primarily based on regional economic but not local ecological information, due to limited access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.
-
-After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g.~based on surrounding forest proximity.
-
 \newpage
@@
 
-Thanks to Professor Dan Perlman for initial networking; concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision; Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; and peer James Fifer for discussion of early draft.
+Thanks to Professor Dan Perlman for initial networking; concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision; Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; peer James Fifer for discussion of early draft; and Luisa Valdez for translation assistance.
 
@@
 \url{https://doi.org/10.1111/btp.12689}
-
-\leavevmode\vadjust pre{\hypertarget{ref-karsten14}{}}%
-Karsten, R.J., Meilby, H., Larsen, J.B., 2014.
-Regeneration and management of lesser known timber species in the {Peruvian Amazon} following disturbance by logging.
-Forest Ecology and Management 327, 76--85.
-\url{https://doi.org/10.1016/j.foreco.2014.04.035}
 

### 2022-07-11 article/methods.Rmd
ideas ATBC

@@
 Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
-Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22]* are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa).
+Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedinaOir22]* are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa) *[@softwareMedinaOsa]*.
 

### 2022-07-20 article/disc.Rmd
draft disc

@@
 output: 
-  # html_notebook
-  bookdown::html_document2:
+  html_notebook
+  # bookdown::html_document2:
     # number_sections: FALSE
@@
 <!-- These results were shown in Figs \@ref(fig:biomass), \@ref(fig:taxa), and \@ref(fig:traits)... -->
-It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential, and this evidence supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
-Accordingly this suggests that edge effects do significantly change humid tropical forest community composition, but primarily due to factors other than light competition or dispersal limitation, as well as that aboveground biomass and even low taxonomic richness is resilient after a decade of regeneration.
+It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential, 
+and the evidence presented here supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
+Accordingly, this suggests that edge effects do significantly change humid tropical forest community taxonomic and functional composition, but primarily due to factors other than light competition or dispersal limitation *[@krishnadas20]*, and also suggests that aboveground biomass and even low taxonomic richness are resilient after a decade of regeneration.
 
@@
 This study adds that management via edge effects can also mediate ~10% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns *[@baker04]*.
-Overall, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon *[@gonzalezj98; @pyles22]*, and increasing nearby stem density.
+In this study, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon *[@gonzalezj98; @pyles22]*, and increasing nearby stem density.
 Lower wood density can also reflect more dynamic forest stands *[@malhi06]* with higher tree turnover, a consequence of forest fragmentation *[@nascimento04a]* together with other global change drivers *[@laurance14]*, as well as may reflect possible differences in soil fertility *[@malhi04]*, all of which represent possible future research directions, beyond strictly neutral dynamics *[@terborgh95]*.
-Furthermore, light availability at the plot level may not affect biomass storage, but instead individual light gaps may be more important for biomass dynamics *[@chazdon84]*.
+Furthermore, light availability at the plot level did not affect biomass storage, but instead, individual light gaps may be more important for biomass dynamics *[@chazdon84]*.
     <!-- - although canopy light may still affect biomass generally in this case, without edge effects (check direct light-biomass correlation graph? -- no sig correlation) -->
 
-Edge effects have been shown to lower both forest biomass and also diversity, and this study adds that the decline can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
-The rapid diversity loss across the edge gradient may be in part due to finer variations in dispersal patterns among animal taxa, rather than overall dispersal mode more broadly.
-Similarly, stochastic factors like priority effects after dispersal may also affect offspring establishment and re-growth.
-While the biomass of the most abundant taxon *Vochysia* was resilient to edge effects, that of the key taxon for frugivores like birds and bats *Ficus* was significantly lowered across the forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.
+Edge effects have been shown in other studies to lower both forest biomass and change community composition *[@anderson22]*, and this study adds that the decline in diversity can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
+This rapid loss of diversity across the edge gradient may be in part due to existing higher yet constant levels of shade limiting shade-intolerant seedling recruitment and/or performance, together with insufficient time for slower-growing shade-tolerant taxa to accumulate significant amounts of biomass.
+Another possible factor may be overall short dispersal kernels, in part due to higher fungal pathogen pressure under higher constant shade, also increasing tree mortality.
+Decomposing compositional changes into key specific taxa, while the biomass of the most abundant taxon *Vochysia* was resilient to edge effects in this study, that of the key taxon for frugivores like birds and bats *Ficus* was significantly lowered across the secondary forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.
+Community composition also tended to show lower variance across the edge gradient, although this effect cannot be confirmed here given the uneven sampling that accounted for variable forest area at different distances from edge.
+
+Characterizing community composition by functional traits is also increasingly studied *[@kearney21]*, but less so regarding edge effects.
+A study in Madagascar found lower phylogenetic diversity closer to degraded forest edge, but no change in aboveground biomass *[@razafindratsima18]*.
+These results parallel in showing lower taxonomic diversity, although biomass specifically of taxa assoociated with both early- and late-successional stages also decreased with away from primary forest edge.
+This result by successional stage can be explained by slightly lower habitat quality inferred near exposed habitat edges near roads, and accordingly relatively higher habitat quality near closed primary forest edge.
+While lower biomass of succession-agnostic taxa further from primary forest edge appears consistent with dispersal kernels, the lack of pattern by broad dispersal mode limits this explanation here, yet could be further tested by measuring taxon-specific dispersal kernels.
+
+Overall, this study highlights how abandoned wet tropical timber plantation can regenerate in alongside fragmented forests.
+Results support related syntheses that biomass and taxonomic richness often regrows relatively quickly, 
+<!-- **along with traits?,** -->
+while taxonomic composition recovers much more slowly, if not diverging altogether *[@norden15]*.
+These findings can suggest that more efficient reforestation could more actively target biodiversity conservation over other resilient functions like carbon storage, and that restoration efforts could include focusing activities like native species planting efforts on degraded habitat edges away from primary forest edges, and investing less in areas closer to existing forest edges, where recruitment may already maintain diversity levels.
+As forests continue to become fragmented *[@haddad15]*, understanding how to work with natural regeneration patterns around edge effects will likely become increasingly important.
 

### 2022-07-20 article/intro.Rmd
start draft disc
first poster draft
ideas ATBC
draft disc

@@
 Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
-As a result, understanding secondary forest regeneration remains important...
+Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, yet local management remains the strongest driver of tropical forest carbon stores *[@pyles22]*.
 
@@
 Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.
-The traits of existing taxa also depend on land-use history and previous management *[@omeja12]*, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
+
+Properties of the existing tree community also depend on land-use history and previous management *[@omeja12]* including plantation use for timber production, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
 Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
@@
 Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.
+
 <!-- role of N-fixing trees... *[@lai18]* -->
+<!-- A subset of secondary forests include naturally regenerating timber plantations, which may show unique recovery patterns. -->
+<!-- .. -->
+<!-- More specifically, regenerating timer plantations may also experience edge effects, but they tend to be ...  -->
+<!-- - from *[@anderson22; others]* -->
+<!-- - examples of *B. quinata* plantation papers specifically -->
 
@@
 Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.
+
+<!-- - conceptual hypothesis diagram? -->
 

### 2022-07-20 article/merge.Rmd
first poster draft
ideas ATBC
draft disc

--- 2022-05-15 article/merge.Rmd
@@
 ---
-title: ""
+title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
 author: 
-  name: Nicholas Medina
-  affilitation: Brandeis University
-address:
-  code: Brandeis University
-  address: 415 South St, Waltham, MA 02453
+  - Nicholas Medina ^a^*
+  - Max Villalobos ^b^
+  - Andrea Johnson ^b^
+date: 
+  scriptsize\ ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453;
+  ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
+  # Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109
+# author: 
+#   - name: "Nicholas Medina"
+#     email: nmedina@umich.edu
+#     affiliation: Brandeis University
+#     orcid_id: 0000-0001-5465-3988
+# address:
+#   - code: Brandeis University
+#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
+# keywords: [1, 2, 3]
+
 output: 
   # html_notebook
-  bookdown::html_document2:
-    number_sections: FALSE
-bibliography: ../../library.bib
+  # bookdown::html_document2:
+  #   # df_print: paged
+  #   global_numbering: TRUE
+  #   toc: FALSE
+  bookdown::pdf_document2:
+    # number_sections: FALSE
+    global_numbering: TRUE
+    keep_tex: TRUE
+    toc: FALSE
+bibliography: 
+  - ../../library.bib
+  - ../../grateful-refs.bib
+  - ../../software.bib
 link-citations: TRUE
+csl: ../../styles-master/elsevier-harvard.csl
+
+fontsize: 12pt
+geometry: margin=1in
+header-includes:
+  - \usepackage[left]{lineno}
+  - \linenumbers
+  - \usepackage{setspace}\doublespace
+# abstract
 ---
 
-# Abstract
+<!-- journals: J [Appl] Ecol, J [Appl] Veg Sci, Basic Appl Ecol (3.5), Biotropica (2.5), Eco Evo -->
+
+\
+\
+
+**Keywords**: secondary forest; regeneration; timber plantation associated biodiversity; edge effect; aboveground biomass; shade
+
+\
+\
+
+***Corresponding author**: Nicholas Medina, nmedina@umich.edu
+
+\newpage
+
+# Abstract{-}
 
@@
 ```
+
+\newpage
 
@@
 
-<!-- # ```{r child="results.Rmd"} -->
-<!-- # ``` -->
+```{r child="results.Rmd"}
+```
 
@@
 
+\newpage
+
+# Funding{-}
+
+This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship.
+
+# Declaration of interests{-}
+
+Authors declare no conflicts of interest.
+
+# Acknowledgements{-}
+
+Thanks to Professor Dan Perlman for initial networking; 
+<!-- concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision;  -->
+Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; 
+concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; 
+peer James Fifer for discussion of early draft; 
+and Luisa Valdez for translation assistance.
+
+# Author contributions{-}
+
+MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis;
+NM wrote initial draft and later revisions.
+
+# Data statement{-}
+
+Code stored at [github.com/nmedina17/osa](github.com/nmedin17/osa) *[@softwareMedina22osa]*.
+
+\newpage
+
 # References

### 2022-07-20 article/merge.tex
draft disc

@@
   pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
-  pdfauthor={Nicholas Medina},
+  pdfauthor={Nicholas Medina a*; Max Villalobos b; Andrea Johnson b},
   hidelinks,
@@
 \title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
-\author{Nicholas Medina}
-\date{Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109}
+\author{Nicholas Medina \textsuperscript{a}* \and Max Villalobos \textsuperscript{b} \and Andrea Johnson \textsuperscript{b}}
+\date{scriptsize~\textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}
 
@@
 Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa \emph{(\protect\hyperlink{ref-wandrag17}{Wandrag et al., 2017})}, which while complicated \emph{(\protect\hyperlink{ref-muller-landau09}{Muller-Landau and Hardesty, 2009})}, can also correlate with simpler plant traits like height and seed mass \emph{(\protect\hyperlink{ref-tamme14}{Tamme et al., 2014}; \protect\hyperlink{ref-thomson11}{Thomson et al., 2011})}.
-The traits of existing taxa also depend on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012})}, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
+
+Properties of the existing tree community also depend on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012})} including plantation use for timber production \emph{(\protect\hyperlink{ref-others}{\textbf{others?}})}, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
 Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages \emph{(\protect\hyperlink{ref-albrich20}{Albrich et al., 2020}; \protect\hyperlink{ref-gough21}{Gough et al., 2021})}, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry \emph{(\protect\hyperlink{ref-lefland18}{Lefland et al., 2018})}.
@@
 Community analysis and ordination was done as a PERMANOVA with the \emph{adonis()} function from the \emph{vegan} 2.6.2 R package \emph{(\protect\hyperlink{ref-dixon03}{Dixon, 2003})}.
-Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and internal pipeline \emph{oir} \emph{(\protect\hyperlink{ref-softwareMedina22}{Medina, 2022})} are stored at \url{github.com/nmedina17/osa}.
+Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and internal pipeline \emph{oir} \emph{(\protect\hyperlink{ref-softwareMedina22oir}{Medina, 2022a})} are stored at \url{github.com/nmedina17/osa} \emph{(\protect\hyperlink{ref-softwareMedina22osa}{Medina, 2022b})}.
 
@@
 This study found significant edge effects on secondary forest stand wood and stem density, tree diversity and community composition, and some functional traits, yet not on overall aboveground biomass.
-It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential, and this evidence supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
-Accordingly this suggests that edge effects do significantly change humid tropical forest community composition, but primarily due to factors other than light competition or dispersal limitation, as well as that aboveground biomass and even low taxonomic richness is resilient after a decade of regeneration.
+It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential,
+and the evidence presented here supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
+Accordingly, this suggests that edge effects do significantly change humid tropical forest community taxonomic and functional composition, but primarily due to factors other than light competition or dispersal limitation \emph{(\protect\hyperlink{ref-krishnadas20}{Krishnadas et al., 2020})}, and also suggests that aboveground biomass and even low taxonomic richness are resilient after a decade of regeneration.
 
@@
 This study adds that management via edge effects can also mediate \textasciitilde10\% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns \emph{(\protect\hyperlink{ref-baker04}{Baker et al., 2004})}.
-Overall, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon \emph{(\protect\hyperlink{ref-gonzalezj98}{Gonzalez J and Fisher, 1998}; \protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}, and increasing nearby stem density.
+In this study, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon \emph{(\protect\hyperlink{ref-gonzalezj98}{Gonzalez J and Fisher, 1998}; \protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}, and increasing nearby stem density.
 Lower wood density can also reflect more dynamic forest stands \emph{(\protect\hyperlink{ref-malhi06}{Malhi et al., 2006})} with higher tree turnover, a consequence of forest fragmentation \emph{(\protect\hyperlink{ref-nascimento04a}{Nascimento and Laurance, 2004})} together with other global change drivers \emph{(\protect\hyperlink{ref-laurance14}{Laurance et al., 2014})}, as well as may reflect possible differences in soil fertility \emph{(\protect\hyperlink{ref-malhi04}{Malhi et al., 2004})}, all of which represent possible future research directions, beyond strictly neutral dynamics \emph{(\protect\hyperlink{ref-terborgh95}{Terborgh et al., 1995})}.
-Furthermore, light availability at the plot level may not affect biomass storage, but instead individual light gaps may be more important for biomass dynamics \emph{(\protect\hyperlink{ref-chazdon84}{Chazdon and Fetcher, 1984})}.
+Furthermore, light availability at the plot level did not affect biomass storage, but instead, individual light gaps may be more important for biomass dynamics \emph{(\protect\hyperlink{ref-chazdon84}{Chazdon and Fetcher, 1984})}.
 
-Edge effects have been shown to lower both forest biomass and also diversity, and this study adds that the decline can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
-The rapid diversity loss across the edge gradient may be in part due to finer variations in dispersal patterns among animal taxa, rather than overall dispersal mode more broadly.
-Similarly, stochastic factors like priority effects after dispersal may also affect offspring establishment and re-growth.
-While the biomass of the most abundant taxon \emph{Vochysia} was resilient to edge effects, that of the key taxon for frugivores like birds and bats \emph{Ficus} was significantly lowered across the forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.
+Edge effects have been shown in other studies to lower both forest biomass and change community composition \emph{(\protect\hyperlink{ref-anderson22}{Anderson et al., 2022})}, and this study adds that the decline in diversity can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
+This rapid loss of diversity across the edge gradient may be in part due to existing higher yet constant levels of shade limiting shade-intolerant seedling recruitment and/or performance, together with insufficient time for slower-growing shade-tolerant taxa to accumulate significant amounts of biomass.
+Another possible factor may be overall short dispersal kernels, in part due to higher fungal pathogen pressure under higher constant shade, also increasing tree mortality.
+Decomposing compositional changes into key specific taxa, while the biomass of the most abundant taxon \emph{Vochysia} was resilient to edge effects in this study, that of the key taxon for frugivores like birds and bats \emph{Ficus} was significantly lowered across the secondary forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.
+Community composition also tended to show lower variance across the edge gradient, although this effect cannot be confirmed here given the uneven sampling that accounted for variable forest area at different distances from edge.
+
+Characterizing community composition by functional traits is also increasingly studied \emph{(\protect\hyperlink{ref-kearney21}{Kearney et al., 2021})}, but less so regarding edge effects.
+A study in Madagascar found lower phylogenetic diversity closer to degraded forest edge, but no change in aboveground biomass \emph{(\protect\hyperlink{ref-razafindratsima18}{Razafindratsima et al., 2018})}.
+These results parallel in showing lower taxonomic diversity, although biomass specifically of taxa assoociated with both early- and late-successional stages also decreased with away from primary forest edge.
+This result by successional stage can be explained by slightly lower habitat quality inferred near exposed habitat edges near roads, and accordingly relatively higher habitat quality near closed primary forest edge.
+While lower biomass of succession-agnostic taxa further from primary forest edge appears consistent with dispersal kernels, the lack of pattern by broad dispersal mode limits this explanation here, yet could be further tested by measuring taxon-specific dispersal kernels.
+
+Overall, this study highlights how abandoned wet tropical timber plantation can regenerate in alongside fragmented forests.
+Results support related syntheses that biomass and taxonomic richness often regrows relatively quickly,
+while taxonomic composition recovers much more slowly, if not diverging altogether \emph{(\protect\hyperlink{ref-norden15}{Norden et al., 2015})}.
+These findings can suggest that more efficient reforestation could more actively target biodiversity conservation over other resilient functions like carbon storage, and that restoration efforts could include focusing activities like native species planting efforts on degraded habitat edges away from primary forest edges, and investing less in areas closer to existing forest edges, where recruitment may already maintain diversity levels.
+As forests continue to become fragmented \emph{(\protect\hyperlink{ref-haddad15}{Haddad et al., 2015})}, understanding how to work with natural regeneration patterns around edge effects will likely become increasingly important.
 
@@
 
-Thanks to Professor Dan Perlman for initial networking; concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision; Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; peer James Fifer for discussion of early draft; and Luisa Valdez for translation assistance.
+Thanks to Professor Dan Perlman for initial networking;
+Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance;
+concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
+peer James Fifer for discussion of early draft;
+and Luisa Valdez for translation assistance.
 
@@
 
+MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis;
 NM wrote initial draft and later revisions.
@@
 
-Code stored at \href{github.com/nmedin17/osa}{github.com/nmedina17/osa}.
+Code stored at \href{github.com/nmedin17/osa}{github.com/nmedina17/osa} \emph{(\protect\hyperlink{ref-softwareMedina22osa}{Medina, 2022b})}.
 
@@
 \url{https://doi.org/10.1111/ele.14018}
+
+\leavevmode\vadjust pre{\hypertarget{ref-anderson22}{}}%
+Anderson, J.A., McClean, C.J., Sim, S., Pettorelli, N., Jelling, A., Tangah, J., Hill, J.K., 2022.
+Weak edge effects on trees in {Bornean} rainforest remnants bordering oil palm.
+Biotropica btp.13115.
+\url{https://doi.org/10.1111/btp.13115}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-kearney21}{}}%
+Kearney, M.R., Jusup, M., McGeoch, M.A., Kooijman, S.A.L.M., Chown, S.L., 2021.
+Where do functional traits come from?
+{The} role of theory and models.
+Functional Ecology 35, 1385--1396.
+\url{https://doi.org/10.1111/1365-2435.13829}
+
 \leavevmode\vadjust pre{\hypertarget{ref-khan20}{}}%
@@
 \url{https://doi.org/10.1016/j.gecco.2020.e01025}
+
+\leavevmode\vadjust pre{\hypertarget{ref-krishnadas20}{}}%
+Krishnadas, M., Agarwal, K., Comita, L.S., 2020.
+Edge effects alter the role of fungi and insects in mediating functional composition and diversity of seedling recruits in a fragmented tropical forest.
+Annals of Botany 126, 1181--1191.
+\url{https://doi.org/10.1093/aob/mcaa138}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-softwareMedina22}{}}%
-Medina, N., 2022. {Code for: Automating statistical decision trees for data exploration and analysis revision workflow}. \url{https://doi.org/10.5281/zenodo.6800299}
+\leavevmode\vadjust pre{\hypertarget{ref-softwareMedina22oir}{}}%
+Medina, N., 2022a. {Code for: Automating statistical decision trees for data exploration and analysis revision workflow}. \url{https://doi.org/10.5281/zenodo.6800299}
+
+\leavevmode\vadjust pre{\hypertarget{ref-softwareMedina22osa}{}}%
+Medina, N., 2022b. {Code for: Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}. \url{https://doi.org/10.5281/zenodo.6816612}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-razafindratsima18}{}}%
+Razafindratsima, O.H., Brown, K.A., Carvalho, F., Johnson, S.E., Wright, P.C., Dunham, A.E., 2018.
+Edge effects on components of diversity and above-ground biomass in a tropical rainforest.
+Journal of Applied Ecology 55, 977--985.
+\url{https://doi.org/10.1111/1365-2664.12985}
+
 \leavevmode\vadjust pre{\hypertarget{ref-rejou-mechain17}{}}%

### 2022-07-20 article/methods.Rmd
draft disc

--- 2022-07-11 article/methods.Rmd
@@
 Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
-Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22]* are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa).
+Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22oir]* are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa) *[@softwareMedina22osa]*.
 

### 2022-07-25 article/results.Rmd
ideas ATBC
add clean varData cols, draft fig1f

@@
 options(digits = 1)
+# source(here::here("figs/fig2.R"));
 ```
@@
 with edge distance by 
-`r wd$statPrint[[1]]$estimate[[2]]` ± 
-`r wd$statPrint[[1]]$std.error[[2]]` per m 
+`r wd$statPrint[[1]]$estimate[[2]] %>% p_round()` ± 
+`r wd$statPrint[[1]]$std.error[[2]] %>% rstatix::p_round()` per m 
 explaining 
-`r "R2x100"`% of variance 
+*`r "~10"`*% of variance 
 among median distance strata values
@@
 ```
+
+<!-- - size distributions? -->
+<!-- - put light-biomass correlation? -->
 
@@
 which explained 
-`r "R2x100"`% of variance 
+*`r "73"`*% of variance 
 among distance strata medians 
@@
 `r richness$statPrint[[1]]$std.error[[2]]` per m 
-and had `r "R2x100"`% variance explained 
+and had *`r "16"`*% variance explained 
 among distance strata 
@@
 `r ficus$statPrint[[1]]$std.error[[2]]` per m 
-with `r "R2x100"`% variance among distance strata explained 
+with *`r "~20"`*% variance among distance strata explained 
 (Fig \@ref(fig:taxa)f).
@@
 ```
+
+<!-- - put diversity-biomass correlation -->
 

### 2022-07-25 data/clean.R

@@ -9,32 +9,14 @@ library(BIOMASS)
 #import----
 
 ##raw----
-rawData <- vroom(
-  here("data/raw/osa.csv")
-)
+rawData <- vroom(here("data/raw/osa.csv"))
 
-dispersal <- vroom(
-  here("data/dispersal.csv")
-) %>%
-  select(
-    fam,
-    gen,
-    sp,
-    mainDispersal
-  ) %>%
-  filter(
-    !is.na(
-      gen
-    ) &
-      !is.na(
-        mainDispersal
-      )
-  )
+dispersal <- vroom(here("data/dispersal.csv")) %>%
+  select(fam, gen, sp, mainDispersal) %>%
+  filter(!is.na(gen) & !is.na(mainDispersal))
 
 ##traits----
-traits <- vroom(
-  here("data/base/try/try19707.csv")
-) %>% #glimpse()
+traits <- vroom(here("data/base/try/try19707.csv")) %>% #glimpse()
   select(
     AccSpeciesName,
     TraitName,
@@ -45,25 +27,16 @@ traits <- vroom(
   ###qc----
   mutate(
     StdValue = ifelse(
-      is.na(
-        StdValue
-      ),
+      is.na(StdValue),
       yes = OrigValueStr,
       no = StdValue
     ),
-    "gen" = AccSpeciesName %>%
-      str_split(
-        " "
-      ) %>% #list
+    "gen" = AccSpeciesName %>% str_split(" ") %>% #list
       modify(
-        . %>%
-          first()
-      ) %>%
-      as_vector()
+        . %>% first()
+      ) %>% as_vector()
   ) %>%
-  select(
-    !OrigValueStr
-  ) %>%
+  select(!OrigValueStr) %>%
 
   pivot_wider(
     id_cols = gen,
@@ -83,16 +56,12 @@ lifeHist <- vroom(
     successionalStage
   ) %>%
   #qc
-  filter(
-    !is.na(
-      successionalStage
-    )
-  )
+  filter(!is.na(successionalStage))
 
 
 ##woodC----
 woodC <- vroom::vroom(here::here(
-  "data/base/glowcad/Doraisami_et_al._2021_Wood_C_Database.csv")) |>
+  "data/base/glowcad/Doraisami_et_al._2021_Wood_C_Database.csv")) %>%
   select(tissue.c, family.resolved, genus.resolved, binomial.resolved)
 
 
[…]

### 2022-07-25 figs/fig1.R

@@ -23,59 +23,27 @@ library(ggpubr)
 #userInput
 #sortbyPvals
 
-measure1a <- quote(
-  kg17
-)
-yAxisLabel1a <- quote(
-  expression(
-    paste(
-      "Biomass (kg ha"^-1,
-      ")"
-    )
-  )
-)
-measure1b <- quote(
-  spg
-)
-yAxisLabel1b <- quote(
-  expression(
-    paste(
-      "Wood density (g",
-      " cm"^-3,
-      ")"
-    )
-  )
-)
-measure1c <- quote(
-  stems
-)
-yAxisLabel1c <- quote(
-  expression(
-    paste(
-      "Stems (0.04 ha"^-1,
-      ")"
-    )
-  )
-)
-measure1d <- quote(
-  luz
-)
-yAxisLabel1d <- quote(
-  "Canopy light (%)"
-)
+measure1a <- quote(kg17)
+yAxisLabel1a <- quote(expression(paste("Biomass (kg ha"^-1, ")")))
+measure1b <- quote(spg)
+yAxisLabel1b <- quote(expression(paste("Wood density (g", " cm"^-3, ")")))
+measure1c <- quote(stems)
+yAxisLabel1c <- quote(expression(paste("Stems (0.04 ha"^-1, ")")))
+measure1d <- quote(luz)
+yAxisLabel1d <- quote("Canopy light (%)")
 
 
 varForm <- median ~ dist
-xAxisLabel <- quote(
-  "Distance to edge (m)"
-)
+xAxisLabel <- quote("Distance to edge (m)")
 statData <- plotResultsTbl1
 underData <- cleanData
 
+covarModel <- eval(mainMetric) ~ (entropy + 1)
 
 
 
-#panel----
+
+#panelA----
 
 graph1a <-  statData %>%
   dotGraph(
@@ -98,7 +66,7 @@ graph1a
 
 
 
-#panel----
+#panelB----
 
 graph1b <- statData %>%
   dotGraph(
@@ -116,7 +84,7 @@ graph1b
 
 
 
-#panel----
+#panelC----
 
 statData$pval[[9]] <-
   #varsDoubled
@@ -140,7 +108,7 @@ graph1c
 
 
 
-#panel----
+#panelD----
 
 graph1d <- statData %>%
   dotGraph(
@@ -156,7 +124,7 @@ graph1d
 
 
 
-#panel----
+#panelE----
 
[…]

### 2022-07-25 figs/fig2.R

@@ -25,81 +25,29 @@ library(ggpubr)
 #userInput
 
 
-measure2a <- quote(
-  entropy
-)
-yAxisLabel2a <- quote(
-  "Diversity (H')"
-)
+measure2a <- quote(entropy)
+yAxisLabel2a <- quote("Diversity (H')")
 
-measure2b <- quote(
-  richness
-)
-yAxisLabel2b <- quote(
-  "Richness"
-)
+measure2b <- quote(richness)
+yAxisLabel2b <- quote("Richness")
 
-measure2e <- quote(
-  Vochysia
-)
+measure2e <- quote(Vochysia)
 taxMetricUnit <- "kg"
-yAxisLabel2e <- quote(
-  glue(
-    measure2e,
-    " ",
-    " (",
-    taxMetricUnit,
-    " ha^-1",
-    ")"
-  )
-)
+yAxisLabel2e <- quote(glue(measure2e, " ", " (", taxMetricUnit, " ha^-1", ")"))
 
-measure2f <- quote(
-  Ficus
-)
-yAxisLabel2f <- quote(
-  glue(
-    measure2f,
-    " ",
-    " (",
-    taxMetricUnit,
-    " ha^-1",
-    ")"
-  )
-)
+measure2f <- quote(Ficus)
+yAxisLabel2f <- quote(glue(measure2f, " ", " (", taxMetricUnit, " ha^-1", ")"))
 
-measure2c <- quote(
-  kg17
-)
-yAxisLabel2c <- quote(
-  glue(
-    "Biomass (",
-    taxMetricUnit,
-    " ha^-1",
-    ")"
-  )
-)
+measure2c <- quote(kg17)
+yAxisLabel2c <- quote(glue("Biomass (", taxMetricUnit, " ha^-1",")"))
 
 
-varForm <- {
-  mainMetric %>%
-    eval()
-} ~ dist
+varForm <- {mainMetric %>% eval()} ~ dist
 #check.env...
-taxForm <- {
-  taxMetric %>%
-    eval()
-} ~ dist
-xAxisLabel <- quote(
-  "Distance to edge (m)"
-)
+taxForm <- {taxMetric %>% eval()} ~ dist
+xAxisLabel <- quote("Distance to edge (m)")
 statData2ab <- plotResultsTbl1
-statData2ef <- taxaResultsTbl1 %>%
-  rename(
-    "variable" = all_of(
-      taxRank
-    )
-  )
[…]

### 2022-07-26 article/merge.Rmd
full draft

@@
 title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
-author: 
-  - Nicholas Medina ^a^*
-  - Max Villalobos ^b^
-  - Andrea Johnson ^b^
-date: 
-  scriptsize\ ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453;
-  ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
-  # Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109
+# author: 
+  # - Nicholas Medina * #^a^
+  # - Emma Capetz ^c^
+  # - Andrea Johnson ^b^
+  # - Erin Lau ^c^
+  # - Max Villalobos ^b^
+# date: \scriptsize
+#   ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA; 
+  # ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
+  # ^c^ Earth and Environmental Sciences, Lehigh University, 27 Memorial Drive West, Bethlehem, PA 18015 USA
+  # Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI 48109 USA
 # author: 
@@
 
-<!-- journals: J [Appl] Ecol, J [Appl] Veg Sci, Basic Appl Ecol (3.5), Biotropica (2.5), Eco Evo -->
+<!-- journals: J [Appl] Ecol, J [Appl] Veg Sci, Basic Appl Ecol (3.7), Biotropica (2.5), Small-scale Forestry (1.3), Eco Evo -->
 
@@
 
-# Acknowledgements{-}
+<!-- # Acknowledgements{-} -->
 
-Thanks to Professor Dan Perlman for initial networking; 
-<!-- concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision;  -->
-Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; 
-concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; 
-peer James Fifer for discussion of early draft; 
-and Luisa Valdez for translation assistance.
+<!-- Thanks to Professor Dan Perlman for initial networking;  -->
+<!-- <!-- concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision;  --> -->
+<!-- Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance;  -->
+<!-- concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;  -->
+<!-- peer James Fifer for discussion of early draft;  -->
+<!-- and Luisa Valdez for translation assistance. -->
 
-# Author contributions{-}
+<!-- # Author contributions{-} -->
 
-MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis;
-NM wrote initial draft and later revisions.
+<!-- MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis; -->
+<!-- NM wrote initial draft and later revisions. -->
 

### 2022-07-26 article/merge.tex
full draft

@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
-  pdfauthor={Nicholas Medina a*; Max Villalobos b; Andrea Johnson b},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
@@
 \fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
+  pdfauthor={Nicholas Medina a*; Max Villalobos b; Andrea Johnson b},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
@@
 \author{Nicholas Medina \textsuperscript{a}* \and Max Villalobos \textsuperscript{b} \and Andrea Johnson \textsuperscript{b}}
-\date{scriptsize~\textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}
+\date{\scriptsize \textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}
 
@@
 Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
-Biomasa arboreal y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
+Biomasa riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
 Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
@@
 
-Properties of the existing tree community also depend on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012})} including plantation use for timber production \emph{(\protect\hyperlink{ref-others}{\textbf{others?}})}, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
+Properties of the existing tree community also depend on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012})} including plantation use for timber production, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
 Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages \emph{(\protect\hyperlink{ref-albrich20}{Albrich et al., 2020}; \protect\hyperlink{ref-gough21}{Gough et al., 2021})}, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry \emph{(\protect\hyperlink{ref-lefland18}{Lefland et al., 2018})}.
@@
 
+\begin{figure}
+\centering
 \includegraphics{merge_files/figure-latex/map-1.pdf}
+\caption{\label{fig:map}Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica.
+Map image data © 2009 Google, © 2022 Osa Conservation.
+Photo credit: Nicholas Medina.}
+\end{figure}
 
@@
 Median plot wood density tended to increase marginally significantly
-(p=0.598)
+(p=0.105)
 with edge distance by
-\ensuremath{8.2\times 10^{-5}} ±
-\ensuremath{1.4\times 10^{-4}} per m
+0.00019
+per m
 explaining
-R2x100\% of variance
+\emph{\textsubscript{10\emph{\% of variance
 among median distance strata values
-(Fig \ref{fig:biomass}b).
+(Fig \ref{fig:biomass}b),
+while tissue carbon did not change significantly with distance to forest edge
+(Fig \ref{fig:biomass}c).
 Plot stem density also tended to increase significantly
-(p=0.82)
+(p=0.03)
 with edge distance but more strongly and non-linearly, by
-\textasciitilde-65.9 ±
-78.9 per m
-(Fig \ref{fig:biomass}c).
+\textasciitilde17.4 ±
+3.4 per m
+(Fig \ref{fig:biomass}d).
 At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge
-(Fig \ref{fig:biomass}d).
+(Fig \ref{fig:biomass}e).
+Notably, tree diversity decreased marginally significantly
+(p=\textless0.0001)
+by
+\textasciitilde{} 1
+per g plot wood density,
+yet only explaining
+}}4}\% of variance
+(Fig \ref{fig:biomass}f).
 
@@
 \includegraphics{merge_files/figure-latex/biomass-1.pdf}
-\caption{\label{fig:biomass} Plot stand properties, namely (a) biomass, (b) wood density, (c) stem density, and (d) canopy light availability.
+\caption{\label{fig:biomass} Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density.
 Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
@@
 Diversity significantly
-(p=0.414)
+(p=0.003)
 decreased non-linearly by
-\textasciitilde0.1 ±
-0.1 per m
+\textasciitilde0.5 ±
+0 per m
 as distance to primary forest edge increased,
 which explained
-R2x100\% of variance
+\emph{73}\% of variance
 among distance strata medians
@@
 0 per m
-and had R2x100\% variance explained
+and had \emph{16}\% variance explained
 among distance strata
@@
 (p=0.01) with
-25.6\%
+11.4\%
 variance among distance strata explained by distance to primary forest edge,
 and the first two principal components explaining
-59\% and
-24\% totaling
-83\%
+47\% and
+23\% totaling
+70\%
 of variance among calculated plot distances
@@
 25.1 per m
-with R2x100\% variance among distance strata explained
+with \emph{\textasciitilde20}\% variance among distance strata explained
 (Fig \ref{fig:taxa}f).
@@
 with distance to edge explaining
-R2x100\% variance
+\emph{\textasciitilde14}\% variance
 among distance strata medians

### 2022-07-26 article/methods.Rmd
full draft

@@
   # html_notebook
-  bookdown::html_document2:
+  # bookdown::html_document2:
+  #   number_sections: FALSE
+  bookdown::pdf_document2:
     number_sections: FALSE
+    toc: FALSE
 bibliography: 
@@
 (See *@taylor15* for a broader ecosystem description of the region.)
-The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the dry-adapted species from the Guanacaste region *[@perezcordero02; @hulshof20]* grew poorly in very wet conditions.
+The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the vulnerably dry-adapted species from the Guanacaste region *[@perezcordero02; @hulshof20]* grew poorly in very wet conditions.
 This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \@ref(fig:map)) and a wide service road on the third NE border, with primary forest beyond it.
@@
 
-```{r map, fig.width=4, fig.height=4, fig.cap="Map of (a) conservation areas and (b) study site of (c) Bombacopsis quinata timber plantation on Osa Peninsula, Costa Rica.
-Map image data © 2022 Osa Conservation.
+```{r map, fig.width=4, fig.height=4, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica.
+Map image data © 2009 Google, © 2022 Osa Conservation.
 Photo credit: Nicholas Medina."}
@@
 ```
+
 <!-- ![**Fig 1**](../figs/map.png) -->

### 2022-07-26 article/results.Rmd
full draft

@@
 wd <- filter(plotResultsTbl1, variable == "spg") #alias
-stems <- filter(plotResultsTbl1, variable == "stems") #alias
+# stems <- filter(plotResultsTbl1, variable == "stems") #alias
 stems12 <- filter(plotResultsTbl12, variable == "stems") %>% distinct() #alias
+divwd <- filter(covarResultsTbl1, variable == "entropy")
 ```
@@
 with edge distance by 
-`r wd$statPrint[[1]]$estimate[[2]] %>% p_round()` ± 
-`r wd$statPrint[[1]]$std.error[[2]] %>% rstatix::p_round()` per m 
+`r wd$statPrint[[1]]$estimate[[2]] %>% rstatix::p_format()` per m 
+<!-- ± `r wd$statPrint[[1]]$std.error[[2]] |> round(3) #formatting...` -->
 explaining 
@@
 among median distance strata values
-(Fig \@ref(fig:biomass)b).
+(Fig \@ref(fig:biomass)b),
+while tissue carbon did not change significantly with distance to forest edge 
+(Fig \@ref(fig:biomass)c).
 Plot stem density also tended to increase significantly 
@@
 `r stems12$statPrint[[1]]$std.error[[2]]` per m 
-(Fig \@ref(fig:biomass)c).
+(Fig \@ref(fig:biomass)d).
 At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge 
-(Fig \@ref(fig:biomass)d).
+(Fig \@ref(fig:biomass)e).
+Notably, tree diversity decreased marginally significantly 
+(`r glue("p=", divwd$pval[[1]] |> rstatix::p_format())`)
+by 
+~ `r divwd$statPrint[[1]]$estimate[[2]] |> rstatix::p_round()` 
+<!-- ± `r divwd$statPrint[[1]]$std.error[[2]] |> rstatix::p_round()` -->
+per g plot wood density, 
+yet only explaining
+*`r "~4"`*% of variance
+(Fig \@ref(fig:biomass)f).
 
-```{r biomass, fig.width=3.5, fig.height=3.5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) stem density, and (d) canopy light availability.
+```{r biomass, fig.width=3.5, fig.height=3.5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density.
 Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
@@
 with distance to edge explaining 
-`r "R2x100"`% variance 
+*`r "~14"`*% variance 
 among distance strata medians 

### 2022-07-26 analysis/stats.R

@@ -58,15 +58,15 @@ cleanDataPlotVars <- cleanData %>%
     plot
   ) %>%
   mutate(
-    "stems" = n(),
-    "richness" = n_distinct(sp)
+    "stems" = n(), "richness" = n_distinct(sp),
+    "hmax" = max(h), "dapmax" = max(dap)
   ) %>%
   select(
     #abiotic
     luz, incl,
 
     #biotic
-    dap, h, spg, kg17, tissue.c,
+    dap, h, spg, kg17, tissue.c, hmax, dapmax,
 
     #community
     stems, richness, entropy
@@ -75,7 +75,10 @@ cleanDataPlotVars <- cleanData %>%
 
 plotVarsTbl <- cleanDataPlotVars |>
   rstatix::get_summary_stats(type = "median_mad") %>%
-  left_join(cleanDataPlotVars, by = c("plot", "dist")) |>
+  left_join(cleanDataPlotVars |> group_by(plot) |>
+              summarize(across(.fns = median)) |>
+              select(where(~ !all(is.na(.x)))),
+            by = c("plot", "dist")) |>
   tidyr::nest("varData" = -variable) %>%
 
   #nextlevel
@@ -91,7 +94,7 @@ plotVarsTbl <- cleanDataPlotVars |>
             # mainDispersal
           ) %>%
           # summarize(median = median(median), mad = mad(median)) #|>
-          summarize(across(is.numeric, ~ median(.x)))
+          summarize(across(where(is.numeric), ~ median(.x)))
           # left_join(cleanDataPlotVars, by = "dist")
       )
   ) #%>%
@@ -183,7 +186,7 @@ spStatTbl <- cleanData %>%
 
 
 #user
-mainMetric <- quote(abs(median + 1))
+mainMetric <- quote(abs(median + 1)) #incl
 
 
 #get
@@ -274,3 +277,11 @@ dispResultsTbl0 <- spStatTbl %>% oir::getStatsTbl(mainModel0)
 #sp-biomass
 covarModel <- eval(mainMetric) ~ (entropy + 1)
 covarResultsTbl <- oir::getStatsTbl(plotVarsTbl, covarModel)
+covarResultsTbl1 <- oir::getStatsTbl1(plotVarsTbl, covarModel)
+
+covarModel2 <- eval(mainMetric) ~ poly((entropy + 1), 2)
+covarResultsTbl12 <- plotVarsTbl %>%
[…]

### 2022-07-26 figs/map.R

@@ -1,6 +1,17 @@
 here::i_am("figs/map.R")
+
 site <- magick::image_read(here::here("figs/map.png")) %>% magick::image_ggplot()
+site0 <- magick::image_read(here::here("figs/locationWOW.jpg")) %>% magick::image_ggplot()
+
 Osa <- magick::image_read(here::here("figs/map_Osa.jpg")) %>% magick::image_ggplot()
 pochote <- magick::image_read(here::here("figs/pochote.jpg")) %>% magick::image_ggplot()
-sitePlot <- ggpubr::ggarrange(Osa, site, pochote, labels = c("a", "b", "c"))
+
+#importRaster?
+# source(here::here("data/clean.R"))
+# parcela <- ggplot2::ggplot(cleanData, aes(x = W, y = N)) + geom_point()
+# ggplot2::geom_raster()
+
+
+sitePlot <- ggpubr::ggarrange(Osa, site0, site, pochote, #pcarcela,
+                              labels = c("a", "b", "c", "d"), align = "hv")
 # ggsave(filename = "figs/figmap.png", width = 3, height = 3, units = "in")

### 2022-07-27 article/merge.Rmd
minor draft fixes

@@
   # - Erin Lau ^c^
+  # - Don Morris ^c^
   # - Max Villalobos ^b^

### 2022-07-27 article/merge.tex
minor draft fixes

@@
   pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
-  pdfauthor={Nicholas Medina a*; Max Villalobos b; Andrea Johnson b},
   hidelinks,
@@
 \title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
-\author{Nicholas Medina \textsuperscript{a}* \and Max Villalobos \textsuperscript{b} \and Andrea Johnson \textsuperscript{b}}
-\date{\scriptsize \textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}
+\author{}
+\date{\vspace{-2.5em}}
 
@@
 Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
-Biomasa riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
+Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
 Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
@@
 (See \emph{Taylor et al. (\protect\hyperlink{ref-taylor15}{2015})} for a broader ecosystem description of the region.)
-The study site was a regrowing 20 ha timber plantation of \emph{Bombacompsis quinata} abandoned in \textasciitilde1990 after the dry-adapted species from the Guanacaste region \emph{(\protect\hyperlink{ref-hulshof20}{Hulshof and Powers, 2020}; \protect\hyperlink{ref-perezcordero02}{Pérez Cordero and Kanninen, 2002})} grew poorly in very wet conditions.
+The study site was a regrowing 20 ha timber plantation of \emph{Bombacompsis quinata} abandoned in \textasciitilde1990 after the vulnerably dry-adapted species from the Guanacaste region \emph{(\protect\hyperlink{ref-hulshof20}{Hulshof and Powers, 2020}; \protect\hyperlink{ref-perezcordero02}{Pérez Cordero and Kanninen, 2002})} grew poorly in very wet conditions.
 This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \ref{fig:map}) and a wide service road on the third NE border, with primary forest beyond it.
@@
 Community analysis and ordination was done as a PERMANOVA with the \emph{adonis()} function from the \emph{vegan} 2.6.2 R package \emph{(\protect\hyperlink{ref-dixon03}{Dixon, 2003})}.
-Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and internal pipeline \emph{oir} \emph{(\protect\hyperlink{ref-softwareMedina22oir}{Medina, 2022a})} are stored at \url{github.com/nmedina17/osa} \emph{(\protect\hyperlink{ref-softwareMedina22osa}{Medina, 2022b})}.
+Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and internal pipeline \emph{oir} \emph{(\protect\hyperlink{ref-softwareMedina22oir}{Medina, 2022a})}, and are stored at \url{github.com/nmedina17/osa} \emph{(\protect\hyperlink{ref-softwareMedina22osa}{Medina, 2022b})}.
 
@@
 with edge distance by
-0.00019
-per m
+0.00019 per m
 explaining
-\emph{\textsubscript{10\emph{\% of variance
+\emph{\textasciitilde10} \% of variance
 among median distance strata values
@@
 yet only explaining
-}}4}\% of variance
+\emph{\textasciitilde4} \% of variance
 (Fig \ref{fig:biomass}f).
@@
 
-\hypertarget{acknowledgements}{%
-\section*{Acknowledgements}\label{acknowledgements}}
-\addcontentsline{toc}{section}{Acknowledgements}
-
-Thanks to Professor Dan Perlman for initial networking;
-Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance;
-concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
-peer James Fifer for discussion of early draft;
-and Luisa Valdez for translation assistance.
-
-\hypertarget{author-contributions}{%
-\section*{Author contributions}\label{author-contributions}}
-\addcontentsline{toc}{section}{Author contributions}
-
-MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis;
-NM wrote initial draft and later revisions.
+--\textgreater{}
 

### 2022-07-27 article/methods.Rmd
minor draft fixes

@@
 
-```{r map, fig.width=4, fig.height=4, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica.
+```{r map, fig.width=4.5, fig.height=4.5, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica.
 Map image data © 2009 Google, © 2022 Osa Conservation.
@@
 Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
-Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22oir]* are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa) *[@softwareMedina22osa]*.
+Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22oir]*, and are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa) *[@softwareMedina22osa]*.
 

### 2022-07-27 article/results.Rmd
minor draft fixes

@@
   # html_notebook
-  bookdown::html_document2:
+  # bookdown::html_document2:
+  #   number_sections: FALSE
+  #   toc: FALSE
+  bookdown::pdf_document2:
     number_sections: FALSE
@@
 explaining 
-*`r "~10"`*% of variance 
+*`r "~10"`* % of variance 
 among median distance strata values
@@
 yet only explaining
-*`r "~4"`*% of variance
+*`r "~4"`* % of variance
 (Fig \@ref(fig:biomass)f).
 
-```{r biomass, fig.width=3.5, fig.height=3.5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density.
+```{r biomass, fig.width=4, fig.height=5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density.
 Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
@@
 
-```{r traits, fig.width=2, fig.height=3, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode.
+```{r traits, fig.width=3, fig.height=4, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode.
 Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}

### 2022-07-27 figs/map.R

@@ -10,6 +10,7 @@ pochote <- magick::image_read(here::here("figs/pochote.jpg")) %>% magick::image_
 # source(here::here("data/clean.R"))
 # parcela <- ggplot2::ggplot(cleanData, aes(x = W, y = N)) + geom_point()
 # ggplot2::geom_raster()
+sf::
 
 
 sitePlot <- ggpubr::ggarrange(Osa, site0, site, pochote, #pcarcela,

### 2022-08-02 article/intro.Rmd
most meth rev MS

@@
 The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
-However, secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
+However, recovery of secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
 Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
-Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, yet local management remains the strongest driver of tropical forest carbon stores *[@pyles22]*.
+Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, 
+**yet local management remains the strongest driver of tropical forest carbon stores *[@pyles22]*.**
 
@@
 This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
+**- what is new in this study?**
 We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade that would suppress less shade-tolerant pioneer tree taxa and allow for more tree diversity closer to the primary forest edge.

### 2022-08-02 article/merge.Rmd
most meth rev MS

--- 2022-07-20 article/merge.Rmd
@@
 author:
-  - Nicholas Medina ^a^*
+  - Nicholas Medina * ^a^
+  # - Emma Capetz ^c^
+  - Andrea Johnson ^b^
+  # - Erin Lau ^c^
+  # - Agustín Mendoza ^b^
+  # - Don Morris ^c^
   - Max Villalobos ^b^
-  - Andrea Johnson ^b^
-date: 
-  scriptsize\ ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453;
+date: \scriptsize
+  ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA;
   ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
-  # Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109
+  # ^c^ Earth and Environmental Sciences, Lehigh University, 27 Memorial Drive West, Bethlehem, PA 18015 USA
+  # ^a^ Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI 48109 USA
 # author: 
@@
 
-<!-- journals: J [Appl] Ecol, J [Appl] Veg Sci, Basic Appl Ecol (3.5), Biotropica (2.5), Eco Evo -->
+<!-- journals: J [Appl] Ecol (6.8), J [Appl] Veg Sci (3.4), Basic Appl Ecol (3.7), Biotropica (2.5), J Forestry Res (2.2), Small-scale Forestry (1.3), Eco Evo -->
 
@@
 
-Thanks to Professor Dan Perlman for initial networking; 
-<!-- concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision;  -->
-Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; 
+Thanks to 
+Professor Dan Perlman for initial networking;
+<!-- Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance;  -->
 concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
-peer James Fifer for discussion of early draft; 
+James Fifer, Chau Ho, and Michelle Spicer for discussion of early drafts;
 and Luisa Valdez for translation assistance.
@@
 
-MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis;
+MV and AJ contributed to study conceptualization, administration, and supervision;
+NM and AJ contributed to funding acquisition;
+NM contributed to data collection;
+NM and MV contributed to data analysis;
 NM wrote initial draft and later revisions.

### 2022-08-02 article/merge.tex
most meth rev MS

@@
   pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
+  pdfauthor={Nicholas Medina * a; Andrea Johnson b; Max Villalobos b},
   hidelinks,
@@
 \title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
-\author{}
-\date{\vspace{-2.5em}}
+\author{Nicholas Medina * \textsuperscript{a} \and Andrea Johnson \textsuperscript{b} \and Max Villalobos \textsuperscript{b}}
+\date{\scriptsize \textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}
 
@@
 \caption{\label{fig:map}Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica.
-Map image data © 2009 Google, © 2022 Osa Conservation.
-Photo credit: Nicholas Medina.}
+Map image data © 2022 Osa Conservation, © 2009 Google.
+Map GIS data and photo credit: Max Villalobos, Nicholas Medina.}
 \end{figure}
@@
 
+\hypertarget{acknowledgements}{%
+\section*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{section}{Acknowledgements}
+
+Thanks to
+Professor Dan Perlman for initial networking;
 --\textgreater{}
+concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
+peer James Fifer for discussion of early draft;
+and Luisa Valdez for translation assistance.
+
+\hypertarget{author-contributions}{%
+\section*{Author contributions}\label{author-contributions}}
+\addcontentsline{toc}{section}{Author contributions}
+
+MV and AJ contributed to study conceptualization, administration, and supervision;
+NM and AJ contributed to funding acquisition;
+NM contributed to data collection;
+NM and MV contributed to data analysis;
+NM wrote initial draft and later revisions.
 

### 2022-08-02 article/methods.Rmd
most meth rev MS

@@
 
-```{r map, fig.width=4.5, fig.height=4.5, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica.
-Map image data © 2009 Google, © 2022 Osa Conservation.
-Photo credit: Nicholas Medina."}
+```{r map, fig.width=4.5, fig.height=4.5, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design with 50 m distance strata bins, and (d) focal plantation timber taxon Bombacopsis quinata on Osa Peninsula, Costa Rica.
+Map image data © 2022 Osa Conservation, © 2009 Google.
+Map GIS data and photo credit: Max Villalobos, Nicholas Medina."}
 source(here::here("figs/map.R"));
@@
 
-Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10*, [esri.com](esri.com), and *QGIS 2*, [qgis.org](qgis.org)).
+Edge effects were studied by dividing the secondary forest area into six discrete 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10*, [esri.com](esri.com), and *QGIS 2*, [qgis.org](qgis.org)).
 Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
+Given the small size of the last 300 m distance stratum (Fig \@ref(fig:map)c), one census plot was sufficiently representative.
+All data were later centered at the stratum level, so resulting trends were not weighted by distance bands with more plots in them. 
 The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.
@@
 
-The diameter of all stems >10 cm wide were recorded in each census plot, totaling over 1,000 trees.
-Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
+The diameter of all stems >10 cm wide were recorded in each census plot, totaling over 1,243 trees.
+<!-- Following common forestry guidelines,  -->
+In cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem. 
 Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.
-Taxa were identified with localized knowledge and field guides matching common names to Latin binomials, 
+Taxa were identified by local experts, 
 and trait information was gathered from the literature.
+Traits including successional stage (early, late) and main dispersal mode (wind, animal, water) were matched to species with *@werden20* data, and wood tissue carbon matched to genera with the GLOWCAD database *[@doraisami22]* given high alignment at species levels *[@chave06]*.
 
@@
 
-Aboveground biomass (AGB) was estimated per stem using the allometric equation by *@chave15*.
+Aboveground biomass (AGB) was calculated at the individual stem level using the allometric equation by *@chave15* and later centered at the plot level.
 The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *@chave05 [@alvarez12; @ngomanda14]*.
@@
 DBHs and heights were measured in the field and wood densities were extracted from available literature. 
-Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org).
+<!-- Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be *0.58* following World Agroforestry Database guidelines [(worldagroforestry.org)](worldagroforestry.org).
+-->
 <!-- *[References used for this step are given in *Table S1*.] * -->
-Original values found and recorded were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.
+Values were updated using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by *@rejou-mechain17*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.
 <!-- However, these updates did not appear to change the main results of this study. -->
-
-Traits including successional stage and main dispersal mode were matched to species with *@werden20* data, and wood tissue carbon matched to genera with the GLOWCAD database *[@doraisami22]* given high alignment at species levels *[@chave06]*.
 
@@
 
-For all response variables, data were analyzed using plot or distance strata medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base `r version$version.string` and *rstatix* `r packageVersion("rstatix")` package functions *[@base]*.
+For all response variables, data were centered using medians at the plot level, followed by the discrete distance stratum level where applicable, which was most cases and indicated on figures. 
+These medians were then subjected to linear regression with distance to intact forest edge as the only independent variable, which was binned discretely according to census design.
+All regressions were run through assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base `r version$version.string` and *rstatix* `r packageVersion("rstatix")` package functions *[@base]*.
  <!-- including biomass converted to a per hectare basis,  -->
@@
 which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models. 
-All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
+All trees censused were included in each analysis, including unidentified taxa, which were grouped together for community analyses and did not affect results.
 Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.

### 2022-08-02 article/results.Rmd
most meth rev MS

@@
 source(here::here("figs/fig2.R")); here::i_am("article/results.Rmd"); 
-options(digits = 1)
+options(digits = 2)
 # source(here::here("figs/fig2.R"));
@@
 ```{r biomass, fig.width=4, fig.height=5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density.
-Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
+Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
 source(here::here("figs/fig1.R"), echo = F); 
@@
 ```{r taxa, fig.width=5, fig.height=4, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa.
-White and colored dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
+White and colored dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
 source(here::here("figs/fig2.R"), echo = F);
@@
 ```{r traits, fig.width=3, fig.height=4, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode.
-Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
+Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
 source(here::here("figs/fig3.R"), echo = F);

### 2022-08-02 submit/cover.Rmd
most meth rev MS

---
title: ""
output: html_notebook
header-includes:
  - \usepackage{setspace}\doublespace
---

```{r keys}
knitr::opts_chunk$set(echo = F)
title <- "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
journal <- "Journal of Applied Ecology"
```

Dear Editors of `r journal`:

Please consider this manuscript entitled 
“`r title`” 
as a Research Paper. 
This manuscript presents and discusses data that

support the use of roto-till and cover crop mixtures to alleviate soil compaction and weed pressure, which are very common issues for all farmers as agriculture expands into urban environments.

This manuscript both fits within and also expands the scope of this journal publication. 
The manuscript fits well with the scope because 
it discusses changes to soil functions caused by various types and intensities of mechanical tillage, and informs balancing trade-offs during agricultural management. 
In addition analyzing tillage affects, the manuscript also incorporates discussion of cover crops as a comparable management strategy targeting similar soil functions, and more specifically presents the use of cover crop mixtures, rather than traditional single-species cover cropping. 
Finally, this manuscript places these soil management considerations in an urban agriculture context, which is increasingly relevant to the majority of the world’s population living in urban areas and promotes adaptation to more local food supply chain systems.

Overall, this manuscript represents a solid and interesting contribution to the soil tillage and modern agricultural literature. 
The analysis and discussion is presented as a full manuscript, though it could be possible to contribute this a Short Communication. 
We are also happy to recommend your associate editor 
Elke Noellemeyer 
to be associated with our submission, based on a listed interest in 
agroecology. 
As potential reviewers, we point to 
Kami Pothukuchi and Zhijiang Lu (Wayne State), 
Viashelav Vashenev (RUDN, Russia), 
Zhonqi (Joshua) Cheng (USI, Brooklyn Coll), and 
Rattan Lal (Ohio State). 
Finally, please feel free to contact us for any additional relevant information you may like.

Sincerely, 
Authors

### 2022-08-02 analysis/stats.R

@@ -41,6 +41,7 @@ spTbl <- cleanData %>% filter(!is.na(dist | plot)) %>%
     values_fn = length,
     values_fill = 0
     ) %>%
+  select(!`-`) %>% #sig?
   mutate("entropy" = diversity(.))
 
 

### 2022-08-02 figs/fig2.R

@@ -1,9 +1,9 @@
 here::i_am("figs/fig2.R")
 #taxMetric,mainMetric
 #plotResultsTbl1,taxaResultsTbl1
-taxRank <- quote(
-  gen
-)
+
+taxRank <- quote(gen) #KEY
+
 source(here::here("analysis/stats.R"))
 #plotResultsTbl
 library(glue)
@@ -33,19 +33,19 @@ yAxisLabel2b <- quote("Richness")
 
 measure2e <- quote(Vochysia)
 taxMetricUnit <- "kg"
-yAxisLabel2e <- quote(glue(measure2e, " ", " (", taxMetricUnit, " ha^-1", ")"))
+yAxisLabel2e <- quote(glue(measure2e, " ", " (", taxMetricUnit, " per ha", ")"))
 
 measure2f <- quote(Ficus)
-yAxisLabel2f <- quote(glue(measure2f, " ", " (", taxMetricUnit, " ha^-1", ")"))
+yAxisLabel2f <- quote(glue(measure2f, " ", " (", taxMetricUnit, " per ha", ")"))
 
 measure2c <- quote(kg17)
-yAxisLabel2c <- quote(glue("Biomass (", taxMetricUnit, " ha^-1",")"))
+yAxisLabel2c <- quote(glue("Biomass (", taxMetricUnit, " per ha",")"))
 
 
 varForm <- {mainMetric %>% eval()} ~ dist
 #check.env...
 taxForm <- {taxMetric %>% eval()} ~ dist
-xAxisLabel <- quote("Distance to edge (m)")
+xAxisLabel <- quote("Distance to intact forest (m)")
 statData2ab <- plotResultsTbl1
 statData2ef <- taxaResultsTbl1 %>% rename("variable" = all_of(taxRank))
 statData2c <- ordTbl

### 2022-08-02 figs/fig3.R

@@ -23,26 +23,12 @@ library(ggpubr)
 #defs----
 
 
-measure3a <- quote(
-  "NA"
-)
+measure3a <- quote("NA")
 
-xAxisLabel3 <- quote(
-  "Distance to edge (m)"
-)
-yAxisLabel3 <- quote(
-  expression(
-    paste(
-      "Biomass (kg ha"^-1,
-      ")"
-    )
-  )
-)
+xAxisLabel3 <- quote("Distance to intact forest (m)")
+yAxisLabel3 <- quote(expression(paste("Biomass (kg ha"^-1, ")")))
 
-statData3clean <- taxaResultsTbl1 %>%
-  unnest(
-    "varData"
-  )
+statData3clean <- taxaResultsTbl1 %>% unnest("varData")
 
 statData3 <- taxaResultsTbl1 %>%
   rename(
@@ -54,9 +40,7 @@ statData3 <- taxaResultsTbl1 %>%
 
 
 
-theme_set(
-  style
-)
+theme_set(style)
 
 
 #panel----
@@ -72,15 +56,8 @@ fig3a <- statData3clean %>%
   ) %>%
 
   # dotGraph()
-  ggplot(
-    aes(
-      x = dist,
-      y = {{
-        taxMetric
-      }},
-      shape = successionalStage
-    )
-  ) +
+  ggplot(aes(x = dist, y = {{taxMetric}},
+             shape = successionalStage)) +
   geom_quasirandom(
     data = cleanData %>%
       filter(
@@ -92,30 +69,19 @@ fig3a <- statData3clean %>%
       ),
     color = "gray"
   ) +
-  geom_quasirandom(
-    shape = 21,
-    fill = "white"
-  ) +
-  labs(
-    x = xAxisLabel3,
-    y = eval(yAxisLabel3)
-  ) +
+  geom_quasirandom(shape = 21, fill = "white") +
+  labs(x = xAxisLabel3, y = eval(yAxisLabel3)) +
   # scale_y_log10() +
   scale_y_continuous(
     trans = "log10",
-    label = trans_format(
-      "log10", math_format()
-    )
+    label = trans_format("log10", math_format())
   ) +
   # theme(
   #   legend.position = "top"
   # ) +
   stat_summary(
-    fun.data = "median_mad",
-    size = 0.25
-    # position = position_dodge(
-    #   width = 25
-    # )
+    fun.data = "median_mad", size = 0.25
+    # position = position_dodge(width = 25)
   ) +
   stat_smooth(
     method = "lm",
[…]

### 2022-08-02 figs/map.R

@@ -10,7 +10,7 @@ pochote <- magick::image_read(here::here("figs/pochote.jpg")) %>% magick::image_
 # source(here::here("data/clean.R"))
 # parcela <- ggplot2::ggplot(cleanData, aes(x = W, y = N)) + geom_point()
 # ggplot2::geom_raster()
-sf::
+# sf::
 
 
 sitePlot <- ggpubr::ggarrange(Osa, site0, site, pochote, #pcarcela,

### 2022-08-04 article/intro.Rmd
rev meth,intro MES need disc,abs

@@
 
-Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*, and while tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and regenerating degraded functions *[@hubau20]* amid increasing climate stressors *[@anderegg22]* yet insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
+Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*.
+While tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and functionally degraded *[@hubau20]*.
+This likely amplifies amid increasing climate stressors *[@anderegg22]* and collective management issues, including insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
 Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
@@
 
-Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, as well as land-use history, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
+Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
 About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
@@
 
-Properties of the existing tree community also depend on land-use history and previous management *[@omeja12]* including plantation use for timber production, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
-Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
-Overall, while edge effects near degraded habitat edges are often negative, meta-population *[@levins79]* and -community *[@leibold04; @warren15]* theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
-Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.
+Properties of the existing tree community also depend on land-use history and previous management *[@omeja12]*, together with the composition of surrounding forests.
+Overall, legacy effects from management like intensive timber planting could potentially delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
+More specifically, timber plantations have historically favored tree taxa that tend to be pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
+This could lead to a longer-term dominance of pioneer taxa, referred to as arrested succession *[@]*.
+However, arrested succession could be prevented and even typical succession accelerated if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
+Additionally, the leftover timber stands could provide additional shade, which would not be available in a clear-cut harvested timber plantation, and this could help more shade-tolerant taxa establish and recruit better.
+In this case, having a habitat edge near a relatively rich intact forest, rather than near a relatively degraded open area, would instead be an asset to regenerating plantation forest patches, rather than a detriment.
+Overall, edge effects near degraded habitat edges are often negative, but influences from meta-population *[@levins79]* and meta-community *[@leibold04; @warren15]* theories do highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity, thereby potentially making particular edge effects, specifically from adjacent intact forests, actually beneficial for regeneration plantation forest parcels.
+Taken together, these ideas suggest that near primary forest, shade-tolerant dispersal may facilitate plantation forest regeneration.
+This kind of edge effect may ultimately help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.
 
@@
 **- what is new in this study?**
-We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade that would suppress less shade-tolerant pioneer tree taxa and allow for more tree diversity closer to the primary forest edge.
+We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade, which would suppress less shade-tolerant pioneer tree taxa, and in turn facilitate more shade-tolerant tree taxa, ultimately allowing for more tree diversity within the regenerating plantation forest in parts that were closer to the richer edge of the intact unmanaged forest.
     <!-- - supported by inference?! -->

### 2022-08-04 article/methods.Rmd
rev meth,intro MES need disc,abs

@@
 Edge effects were studied by dividing the secondary forest area into six discrete 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10*, [esri.com](esri.com), and *QGIS 2*, [qgis.org](qgis.org)).
+Note that in line with hypotheses about facilitated regeneration, this study most often referred to the "edge" of the regenerating plantation forest as the part closest to the unmanaged intact forest, where trees may be dispersing from, rather than the open service road, which is usually the focus of other studies evaluating effects of forest fragmentation and degradation, rather than restoration.
 Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
@@
 Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.
+
 Taxa were identified by local experts, 
 and trait information was gathered from the literature.
-Traits including successional stage (early, late) and main dispersal mode (wind, animal, water) were matched to species with *@werden20* data, and wood tissue carbon matched to genera with the GLOWCAD database *[@doraisami22]* given high alignment at species levels *[@chave06]*.
+Final traits used only included successional stage (early, late) and main dispersal mode (wind, animal, water), which were ultimately matched to species using only the specific dataset made available by *@werden20*.
+Otherwise wood tissue carbon was matched to genera with the GLOWCAD database *[@doraisami22]*, given high taxonomic alignment of this trait above the species level *[@chave06]*.
 
@@
 which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models. 
-All trees censused were included in each analysis, including unidentified taxa, which were grouped together for community analyses and did not affect results.
+All trees censused were included in each analysis, including unidentified taxa, which were only grouped together for taxonomic analyses and whose exclusion did not affect observed patterns shown in results.
 Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.

### 2022-08-04 figs/map.R

@@ -1,7 +1,7 @@
 here::i_am("figs/map.R")
 
-site <- magick::image_read(here::here("figs/map.png")) %>% magick::image_ggplot()
-site0 <- magick::image_read(here::here("figs/locationWOW.jpg")) %>% magick::image_ggplot()
+site <- magick::image_read(here::here("figs/mapMod.png")) %>% magick::image_ggplot()
+site0 <- magick::image_read(here::here("figs/locationWOWmod.jpg")) %>% magick::image_ggplot()
 
 Osa <- magick::image_read(here::here("figs/map_Osa.jpg")) %>% magick::image_ggplot()
 pochote <- magick::image_read(here::here("figs/pochote.jpg")) %>% magick::image_ggplot()

### 2022-08-05 article/abs.Rmd
draft disc
full draft
rev meth,intro MES need disc,abs
rev intro for res

@@
 Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
-Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
-This study addressed how wet tropical secondary forest regeneration was affected by distance to habitat edge, and hypothesized that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
-An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.
+Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration.
+However, most studies assess how degraded edges affect intact forests, leaving it unclear whether higher-quality edges could facilitate regeneration of previously degraded forests.
+Notably, higher quality edges near intact forests could promote processes like dispersal and wood biomass accumulation that effectively accelerate succession, leading to better performance of shade-tolerant taxa compared to pioneer taxa in the early stages of forested plantation regeneration.
+This study addressed how wet tropical forested plantation regeneration was affected by distance to adjacent intact forest edge.
+It was hypothesized that old timber plantations facilitate regeneration by increasing available shade, favoring the presence and biomass of later-successional taxa, ultimately changing community composition overall.
+An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to broad dispersal mode and wood density traits using relevant literature.
 
 <!-- # Results/Conclusions -->
-As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ca 10% variation explained, while biomass and canopy light tended to stay the same.
+As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ~10% variation explained, while biomass and canopy light tended to stay the same.
 Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably. 
@@
 
-# Abstracto{-}
+# Resumen*{-}
 
 Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
-Biomasa arboreal y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
+Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
 Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.

### 2022-08-05 article/intro.Rmd
rev intro for res

@@
 Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
-Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, 
-**yet local management remains the strongest driver of tropical forest carbon stores *[@pyles22]*.**
+Overall, understanding secondary forest regeneration is key for global biodiversity conservation at the global scale.
 
-Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
+Modern secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
 About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
-Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.
+Effects of degraded edges have historically favored tree taxa that tend to be pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
+Accordingly, secondary forest wood density may also tend to decrease near degraded habitat edges.
+Additionally, the dominance of pioneering taxa could lead to arrested succession *[@tymen15]*, which further delays the recruitment of shade-tolerant taxa, as well as associated increases in aboveground biomass, in part due to their higher survival with increasing shade, and tendency toward relatively higher wood densities.
+However, arrested succession could be prevented, and even typical succession accelerated, if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
+In this case, having a habitat edge near a relatively rich intact forest, rather than near a relatively degraded open area, would instead be an asset to regenerating plantation forest patches, rather than a detriment.
+Overall, edge effects ultimately have the potential to shape community composition, in part based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.
 
-Properties of the existing tree community also depend on land-use history and previous management *[@omeja12]*, together with the composition of surrounding forests.
+Additionally, properties of the tropical forest carbon stores and tree community depend heavily on land-use history and previous management *[@omeja12; @pyles22]*.
 Overall, legacy effects from management like intensive timber planting could potentially delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
-More specifically, timber plantations have historically favored tree taxa that tend to be pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
-This could lead to a longer-term dominance of pioneer taxa, referred to as arrested succession *[@]*.
-However, arrested succession could be prevented and even typical succession accelerated if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
-Additionally, the leftover timber stands could provide additional shade, which would not be available in a clear-cut harvested timber plantation, and this could help more shade-tolerant taxa establish and recruit better.
-In this case, having a habitat edge near a relatively rich intact forest, rather than near a relatively degraded open area, would instead be an asset to regenerating plantation forest patches, rather than a detriment.
-Overall, edge effects near degraded habitat edges are often negative, but influences from meta-population *[@levins79]* and meta-community *[@leibold04; @warren15]* theories do highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity, thereby potentially making particular edge effects, specifically from adjacent intact forests, actually beneficial for regeneration plantation forest parcels.
-Taken together, these ideas suggest that near primary forest, shade-tolerant dispersal may facilitate plantation forest regeneration.
-This kind of edge effect may ultimately help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.
+In some cases, the leftover timber stands could provide additional shade, which would not be available in a clear-cut harvested timber plantation, and this could help more shade-tolerant taxa establish and recruit better.
+Overall, edge effects near degraded habitat edges are often studied as negative, but conceptual influences from meta-population *[@levins79]* and meta-community *[@leibold04; @warren15]* theories do highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity, thereby potentially making particular edge effects, specifically from adjacent intact forests, actually beneficial for regeneration plantation forest parcels.
+Taken together, these ideas suggest that edge effects specifically from adjacent intact primary forests, potentially via shade-tolerant dispersal and performance, may indeed facilitate plantation forest regeneration.
+Such a beneficial edge effect may ultimately help explain relatively high biodiversity, compared to adjacent stands, found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.
 
@@
 
-This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
-**- what is new in this study?**
+Influenced by restoration ecology for forest management, this study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, focused on highlighting edge effects from an adjacent primary forest on overall stand properties, tree community composition, and limited functional trait recovery.
+<!-- **- what is new in this study?** -->
 We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade, which would suppress less shade-tolerant pioneer tree taxa, and in turn facilitate more shade-tolerant tree taxa, ultimately allowing for more tree diversity within the regenerating plantation forest in parts that were closer to the richer edge of the intact unmanaged forest.

### 2022-08-05 article/merge.Rmd
rev intro for res

@@
 
-<!-- journals: J [Appl] Ecol (6.8), J [Appl] Veg Sci (3.4), Basic Appl Ecol (3.7), Biotropica (2.5), J Forestry Res (2.2), Small-scale Forestry (1.3), Eco Evo -->
+<!-- journals: J [Appl] Ecol (6.8), J [Appl] Veg Sci (3.4), Basic Appl Ecol (3.7), Biotropica (2.5), J Forestry Res (2.2), J Trop Ecol (1.3), Small-scale Forestry (1.3), Eco Evo -->
 

### 2022-08-05 article/intro.Rmd
minor rev intro, disc next

@@
 This likely amplifies amid increasing climate stressors *[@anderegg22]* and collective management issues, including insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
-Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
+Overall, secondary forests can regenerate relatively quickly compared to old-growth forests, in some ways.
 Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
@@
     <!-- - supported by inference?! -->
-Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.
+Accordingly, we predicted more specifically that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.
 

### 2022-11-09 article/abs.Rmd
bioRxiv,poster

@@
 <!-- ## REVIEW CRITERIA -->
-<!-- ​ -->
+<!-- . -->
 <!-- All abstracts will be evaluated and scored by the ATBC2022 Scientific Committee.
@@
 Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration.
-However, most studies assess how degraded edges affect intact forests, leaving it unclear whether higher-quality edges could facilitate regeneration of previously degraded forests.
+However, most studies assess how degraded edges affect intact forests, leaving it unclear whether higher-quality edges could facilitate regeneration of nearby degraded forests.
 Notably, higher quality edges near intact forests could promote processes like dispersal and wood biomass accumulation that effectively accelerate succession, leading to better performance of shade-tolerant taxa compared to pioneer taxa in the early stages of forested plantation regeneration.
@@
 It was hypothesized that old timber plantations facilitate regeneration by increasing available shade, favoring the presence and biomass of later-successional taxa, ultimately changing community composition overall.
-An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to broad dispersal mode and wood density traits using relevant literature.
+A wet neotropical timber plantation reforested after 20 years and adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched identified taxa to broad dispersal mode and wood density traits using relevant literature.
 
 <!-- # Results/Conclusions -->
-As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ~10% variation explained, while biomass and canopy light tended to stay the same.
-Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably. 
-Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
+As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ~10% variation explained, while biomass and canopy light surprisingly tended to stay the same.
+Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, explained in part by wood density, and taxonomic composition varied notably. 
+Finally, tree taxa associated with both early and late successional stages decreased significantly, as well as genus Ficus, but biomass by dispersal mode did not tend to change.
 Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.
@@
 
-# Resumen*{-}
+# Resumen{-}
 
@@
 Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
-Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
+Dado que la mayoría de bosques quedan entre un km de su borde, efectos de borde probablemente afectan la regeneración de composición de comunidad.
+Sin embargo, la mayoría de estudios evalúan cómo bordes ya degradados afectan a bosques intactos, dejándolo poco claro si bordes de más alta calidad podrían facilitar la regeneración de bosques cercanos previamente degradados.
+Notablemente, bordes de alta calidad cerca de bosques intactos podrían promover procesos como disperso y acumulación de biomasa maderosa que efectivamente acceleran a la sucesión, permitiendo mejor rendimiento de taxa tolerantes a sombra comparado con taxa más adaptados a alta luz en las etapas tempranas de regeneración de plantaciones foresterados.
 Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
@@
 
-Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10% variación explicado, mientras tanto biomasa y luz de dócel tendió a quedarse igual.
-Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
-Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, pero biomasa por modo de disperso no tendió a cambiar.
+Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10% variación explicado, mientras tanto biomasa y luz de dócel asombrosamente tendió a quedarse igual.
+Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, explicado en parte por densidad de madera, y composición taxonomica varió notablemente.
+Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, tanto como el género Ficus, pero biomasa por modo de disperso no tendió a cambiar.
 Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.

### 2022-11-09 article/disc.Rmd
bioRxiv,poster

@@
 <!-- These results were shown in Figs \@ref(fig:biomass), \@ref(fig:taxa), and \@ref(fig:traits)... -->
-It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential, 
-and the evidence presented here supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
+It was initially hypothesized that community composition would vary with distance to high-quality intact primary forest edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential.
+The evidence presented here supported edge effects on community composition and biomass of taxa that were associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
 Accordingly, this suggests that edge effects do significantly change humid tropical forest community taxonomic and functional composition, but primarily due to factors other than light competition or dispersal limitation *[@krishnadas20]*, and also suggests that aboveground biomass and even low taxonomic richness are resilient after a decade of regeneration.
+
+<!-- ## Biomass -->
 
@@
 Management has been found to explain most of biomass variation among tropical forest fragments, and wood density ~10% biomass variation *[@pyles22]*.
-This study adds that management via edge effects can also mediate ~10% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns *[@baker04]*.
+Additionally, this study contributes that management via edge effects can also mediate ~10% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns *[@baker04]*.
 In this study, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon *[@gonzalezj98; @pyles22]*, and increasing nearby stem density.
-Lower wood density can also reflect more dynamic forest stands *[@malhi06]* with higher tree turnover, a consequence of forest fragmentation *[@nascimento04a]* together with other global change drivers *[@laurance14]*, as well as may reflect possible differences in soil fertility *[@malhi04]*, all of which represent possible future research directions, beyond strictly neutral dynamics *[@terborgh95]*.
+Lower wood density can also reflect more dynamic forest stands *[@malhi06]* with higher tree turnover, a consequence of forest fragmentation *[@nascimento04]* together with other global change drivers *[@laurance14]*, and may also reflect possible differences in soil fertility *[@malhi04]*, all of which represent possible future research directions, beyond strictly neutral dynamics *[@terborgh95]*.
 Furthermore, light availability at the plot level did not affect biomass storage, but instead, individual light gaps may be more important for biomass dynamics *[@chazdon84]*.
@@
 
-Edge effects have been shown in other studies to lower both forest biomass and change community composition *[@anderson22]*, and this study adds that the decline in diversity can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
+<!-- ## Composition -->
+
+Edge effects have been shown in other studies not only to lower forest biomass, but also to change community composition *[@anderson22]*, and in addition this study contributes that the decline in diversity can be rapid and non-linear across edge gradients, along with supporting the likelihood of changes to community composition lasting decades.
 This rapid loss of diversity across the edge gradient may be in part due to existing higher yet constant levels of shade limiting shade-intolerant seedling recruitment and/or performance, together with insufficient time for slower-growing shade-tolerant taxa to accumulate significant amounts of biomass.
-Another possible factor may be overall short dispersal kernels, in part due to higher fungal pathogen pressure under higher constant shade, also increasing tree mortality.
-Decomposing compositional changes into key specific taxa, while the biomass of the most abundant taxon *Vochysia* was resilient to edge effects in this study, that of the key taxon for frugivores like birds and bats *Ficus* was significantly lowered across the secondary forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.
-Community composition also tended to show lower variance across the edge gradient, although this effect cannot be confirmed here given the uneven sampling that accounted for variable forest area at different distances from edge.
+Another possible factor may be Janzen-Connell processes *[@wills06; @terborgh20]*, including conspecific negative density-dependence of seedling survival on basal area *[@comita09]*, as well as overall short dispersal kernels, whose effects can be mediated by shade *[@comita14]* and in part by higher fungal pathogen pressure near conspecifics *[@jia20]*.
+While both diversity and community composition changed significantly with distance to intact forest in this study, the biomass of most individual taxa, surprisingly, tended to stay the same along the edge gradient.
+This lack of significant individual taxon biomass responses along this edge gradient could be due to stochastic population factors affecting their biomass growth, perhaps widening variability.
+Regardless of the underlying process, these absent patterns do also help explain the lack of pattern observed with plot biomass overall across the edge gradient.
+This explanation points to future studies of taxon-specific responses to edge effects, to help explain overall forest biomass dynamics near habitat edge.
+More specifically here, the biomass of the most abundant taxon *Vochysia* was resilient to edge effects in this study, however, that of the taxon *Ficus*, key for frugivores like birds and bats *[@cottee-jones15; @rafidison20]*, was significantly lowered across the secondary forest edge, pointing to potentially different responses of biomass and local food web associated ecosystem services.
+Results suggest that this pattern is specific to this locally widespread *Ficus* genus, since the overall biomass of other animal-dispersed taxa as a whole tended to stay the same across the edge gradient studied here.
+Interestingly, community composition also appeared to show lower variance across the edge gradient, although testing this observation in a forested area with evenly distributed area among distance strata bands would provide additional support.
+
+<!-- ## Traits -->
 
@@
 A study in Madagascar found lower phylogenetic diversity closer to degraded forest edge, but no change in aboveground biomass *[@razafindratsima18]*.
-These results parallel in showing lower taxonomic diversity, although biomass specifically of taxa assoociated with both early- and late-successional stages also decreased with away from primary forest edge.
-This result by successional stage can be explained by slightly lower habitat quality inferred near exposed habitat edges near roads, and accordingly relatively higher habitat quality near closed primary forest edge.
-While lower biomass of succession-agnostic taxa further from primary forest edge appears consistent with dispersal kernels, the lack of pattern by broad dispersal mode limits this explanation here, yet could be further tested by measuring taxon-specific dispersal kernels.
+This study presents similar results in showing lower taxonomic diversity, although biomass specifically of taxa associated with both early- and late-successional stages also decreased with away from primary forest edge.
+This resulting decrease by successional stage could be explained by slightly lower habitat quality inferred near exposed habitat edges near roads, and accordingly relatively higher habitat quality near closed primary forest edge *[@ries04]*.
+While lower biomass of succession-agnostic taxa further from primary forest edge appears consistent with the expected forms of group-level dispersal kernels as an explanation, the lack of significant pattern along forest edge when binned by broad dispersal mode that was found here may limit the potential of seed dispersal to explain biomass trends.
+Instead, seedling survival may have been more important than dispersal in accumulating basal area and biomass *[@comita09]*.
+For taxa primarily dispersed by animals instead of wind, this could be explained by potentially broader local limitations in animal dispersal activity or abundance.
+Additionally, the potential for nearby dispersal to facilitate reforestation could be tested in future studies by measuring taxon-specific dispersal kernels.
+
+<!-- ## Conclusion -->
 
 Overall, this study highlights how abandoned wet tropical timber plantation can regenerate in alongside fragmented forests.
-Results support related syntheses that biomass and taxonomic richness often regrows relatively quickly, 
+Results support related syntheses that tropical forest biomass and taxonomic richness often regrows relatively quickly *[@davies21a]*, 
 <!-- **along with traits?,** -->

### 2022-11-09 article/intro.Rmd
bioRxiv,poster

@@
 Accordingly, secondary forest wood density may also tend to decrease near degraded habitat edges.
-Additionally, the dominance of pioneering taxa could lead to arrested succession *[@tymen15]*, which further delays the recruitment of shade-tolerant taxa, as well as associated increases in aboveground biomass, in part due to their higher survival with increasing shade, and tendency toward relatively higher wood densities.
+Additionally, the dominance of pioneering taxa could lead to arrested succession *[@tymen16]*, which further delays the recruitment of shade-tolerant taxa, as well as associated increases in aboveground biomass, in part due to their higher survival with increasing shade, and tendency toward relatively higher wood densities.
 However, arrested succession could be prevented, and even typical succession accelerated, if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
@@
     <!-- - supported by inference?! -->
-Accordingly, we predicted more specifically that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.
+Accordingly, we predicted more specifically that further from the intact primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.
 

### 2022-11-09 article/merge.Rmd
bioRxiv,poster

@@
 ---
-title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
+title: "Habitat edges affect tree diversity more than biomass regeneration in a reforested wet neotropical timber plantation"
 author:
-  - Nicholas Medina * ^a^
-  # - Emma Capetz ^c^
+  - Nicholas Medina * ^a,d^
+  - Emma Capetz ^c^
   - Andrea Johnson ^b^
-  # - Erin Lau ^c^
-  # - Agustín Mendoza ^b^
-  # - Don Morris ^c^
+  - Agustín Mendoza ^b^
   - Max Villalobos ^b^
@@
   ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA;
-  ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
-  # ^c^ Earth and Environmental Sciences, Lehigh University, 27 Memorial Drive West, Bethlehem, PA 18015 USA
-  # ^a^ Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI 48109 USA
+  ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Puntarenas, Costa Rica;
+  ^c^ Lehigh University, 27 Memorial Drive West, Bethlehem, PA 18015 USA
+  ^d^ Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI 48109 USA
 # author:
-#   - name: "Nicholas Medina"
+#   - name: Nicholas Medina
 #     email: nmedina@umich.edu
-#     affiliation: Brandeis University
-#     orcid_id: 0000-0001-5465-3988
+#     affliation: 
+#       - Brandeis University
+#       - University of Michigan
+#     # orcid_id: 0000-0001-5465-3988
+#     correspondingauthor: TRUE
+#   - name: Emma Capetz
+#     affiliation: Lehigh University
+#     email: emmacapetz11@gmail.com
+#   - name: Andrea Johnson
+#     affiliation: Osa Conservation
+#     email: andrea.e.johnson@gmail.com
+#   # - name: Erin Lau
+#   #   affiliation: Lehigh University
+#   #   # email:
+#   - name: Agustín Mendoza
+#     affiliation: Osa Conservation
+#     email: agustinescaladorosa@gmail.com
+#   - name: Max Villalobos
+#     affiliation: Osa Conservation
+#     email: maxvillalobosv@gmail.com
 # address:
 #   - code: Brandeis University
-#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
-# keywords: [1, 2, 3]
+#     address: 415 South St, Waltham, MA 02453 USA
+#   - code: Osa Conservation
+#     address: Puerto Jiménez, Puntarenas, Costa Rica
+#   - code: Lehigh University
+#     address: 27 Memorial Drive West, Bethelhem, PA 18015 USA
+#   - code: University of Michigan
+#     address: 1105 N University Ave, Ann Arbor, MI 48109 USA
+# keywords:
+#   - secondary forest
+#   - regeneration
+#   - timber plantation associated biodiversity
+#   - edge effect
+#   - aboveground biomass
+#   - shade  
+# abstract: |
+#   About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
+#   Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
+#   Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration.
+#   However, most studies assess how degraded edges affect intact forests, leaving it unclear whether higher-quality edges could facilitate regeneration of nearby degraded forests.
+#   Notably, higher quality edges near intact forests could promote processes like dispersal and wood biomass accumulation that effectively accelerate succession, leading to better performance of shade-tolerant taxa compared to pioneer taxa in the early stages of forested plantation regeneration.
+#   This study addressed how wet tropical forested plantation regeneration was affected by distance to adjacent intact forest edge.
+#   It was hypothesized that old timber plantations facilitate regeneration by increasing available shade, favoring the presence and biomass of later-successional taxa, ultimately changing community composition overall.
+#   A wet neotropical timber plantation reforested after 20 years and adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched identified taxa to broad dispersal mode and wood density traits using relevant literature.
+#   As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ~10% variation explained, while biomass and canopy light surprisingly tended to stay the same.
+#   Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, explained in part by wood density, and taxonomic composition varied notably. 
+#   Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
+#   Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.
+#   
+#   Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
+#   Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
+#   Dado que la mayoría de bosques quedan entre un km de su borde, efectos de borde probablemente afectan la regeneración de composición de comunidad.
+#   Sin embargo, la mayoría de estudios evalúan cómo bordes ya degradados afectan a bosques intactos, dejándolo poco claro si bordes de más alta calidad podrían facilitar la regeneración de bosques cercanos previamente degradados.
+#   Notablemente, bordes de alta calidad cerca de bosques intactos podrían promover procesos como disperso y acumulación de biomasa maderosa que efectivamente acceleran a la sucesión, permitiendo mejor rendimiento de taxa tolerantes a sombra comparado con taxa más adaptados a alta luz en las etapas tempranas de regeneración de plantaciones foresterados.
+#   Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
+#   Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.
+#   Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10% variación explicado, mientras tanto biomasa y luz de dócel asombrosamente tendió a quedarse igual.
+#   Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, explicado en parte por densidad de madera, y composición taxonomica varió notablemente.
+#   Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, pero biomasa por modo de disperso no tendió a cambiar.
+#   Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.
 
@@
   bookdown::pdf_document2:
-    # number_sections: FALSE
+    number_sections: FALSE
     global_numbering: TRUE
@@
     toc: FALSE
+  # bookdown::pdf_book:
+    # base_format: rticles::springer_article
+    # citation_package: biblatex
 bibliography: 
@@
   - \usepackage{setspace}\doublespace
-# abstract
 ---
@@
 
-**Keywords**: secondary forest; regeneration; timber plantation associated biodiversity; edge effect; aboveground biomass; shade
+**Keywords**:
+secondary forest;
+regeneration;
+timber plantation associated biodiversity;
+edge effect;
+aboveground biomass;
+shade
 
@@
 
-***Corresponding author**: Nicholas Medina, nmedina@umich.edu
+***Correspondence**: Nicholas Medina, nmedina@umich.edu
 
@@
 
-This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship.
+This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship and in part by University of Michigan Rackham Merit Fellowship.
 
-# Declaration of interests{-}
+<!-- # Declaration of interests{-} -->
 
-Authors declare no conflicts of interest.
+<!-- Authors declare no conflicts of interest.  -->
 
@@
 Thanks to 
-Professor Dan Perlman for initial networking;
-<!-- Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance;  -->
-concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
-James Fifer, Chau Ho, and Michelle Spicer for discussion of early drafts;
+Brandeis University Professor Dan Perlman for initial networking;
+Lehigh University Professor Don Morris 
+and student Erin Lau 
+for field support;
+concurrent Osa Conservation general manager Manuel Ramírez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
+James Fifer, Chau Ho, and Michelle Spicer for review of early draft content;
 and Luisa Valdez for translation assistance.
+
+Authors declare no conflicts of interest.
 
@@
 
-MV and AJ contributed to study conceptualization, administration, and supervision;
+NM, MV and AJ contributed to study conceptualization, design, administration, and supervision;
 NM and AJ contributed to funding acquisition;
-NM contributed to data collection;
+NM, AM, and EC, 
+contributed to key data collection;
 NM and MV contributed to data analysis;
 NM wrote initial draft and later revisions.
+
+Study team includes and recognizes diverse and historically-excluded contributions to research; with bilingual communication efforts completed and pending.
 

### 2022-11-09 article/merge.tex
bioRxiv,poster

--- 2022-07-27 article/merge.tex
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \newlength{\cslhangindent}
@@
 \hypersetup{
-  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
+  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a reforested wet neotropical timber plantation},
+  pdfauthor={Nicholas Medina * a,d; Emma Capetz c; Andrea Johnson b; Agustín Mendoza b; Max Villalobos b},
   hidelinks,
@@
 
-\title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
-\author{}
-\date{\vspace{-2.5em}}
+\title{Habitat edges affect tree diversity more than biomass regeneration in a reforested wet neotropical timber plantation}
+\author{Nicholas Medina * \textsuperscript{a,d} \and Emma Capetz \textsuperscript{c} \and Andrea Johnson \textsuperscript{b} \and Agustín Mendoza \textsuperscript{b} \and Max Villalobos \textsuperscript{b}}
+\date{\scriptsize \textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Puntarenas, Costa Rica; \textsuperscript{c} Lehigh University, 27 Memorial Drive West, Bethlehem, PA 18015 USA \textsuperscript{d} Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI 48109 USA}
 
@@
 
-\textbf{Keywords}: secondary forest; regeneration; timber plantation associated biodiversity; edge effect; aboveground biomass; shade
+\textbf{Keywords}:
+secondary forest;
+regeneration;
+timber plantation associated biodiversity;
+edge effect;
+aboveground biomass;
+shade
 
@@
 
-*\textbf{Corresponding author}: Nicholas Medina, \href{mailto:nmedina@umich.edu}{\nolinkurl{nmedina@umich.edu}}
+*\textbf{Correspondence}: Nicholas Medina, \href{mailto:nmedina@umich.edu}{\nolinkurl{nmedina@umich.edu}}
 
@@
 Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely.
-Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
-This study addressed how wet tropical secondary forest regeneration was affected by distance to habitat edge, and hypothesized that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
-An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.
+Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration.
+However, most studies assess how degraded edges affect intact forests, leaving it unclear whether higher-quality edges could facilitate regeneration of nearby degraded forests.
+Notably, higher quality edges near intact forests could promote processes like dispersal and wood biomass accumulation that effectively accelerate succession, leading to better performance of shade-tolerant taxa compared to pioneer taxa in the early stages of forested plantation regeneration.
+This study addressed how wet tropical forested plantation regeneration was affected by distance to adjacent intact forest edge.
+It was hypothesized that old timber plantations facilitate regeneration by increasing available shade, favoring the presence and biomass of later-successional taxa, ultimately changing community composition overall.
+A wet neotropical timber plantation reforested after 20 years and adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched identified taxa to broad dispersal mode and wood density traits using relevant literature.
 
-As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ca 10\% variation explained, while biomass and canopy light tended to stay the same.
-Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably.
-Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
+As distance from primary forest edge increased, stem and wood density tended to increase significantly, with \textasciitilde10\% variation explained, while biomass and canopy light surprisingly tended to stay the same.
+Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, explained in part by wood density, and taxonomic composition varied notably.
+Finally, tree taxa associated with both early and late successional stages decreased significantly, as well as genus Ficus, but biomass by dispersal mode did not tend to change.
 Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.
@@
 
-\hypertarget{abstracto}{%
-\section*{Abstracto}\label{abstracto}}
-\addcontentsline{toc}{section}{Abstracto}
+\hypertarget{resumen}{%
+\section*{Resumen}\label{resumen}}
+\addcontentsline{toc}{section}{Resumen}
 
@@
 
-Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10\% variación explicado, mientras tanto biomasa y luz de dócel tendió a quedarse igual.
-Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
-Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, pero biomasa por modo de disperso no tendió a cambiar.
+Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10\% variación explicado, mientras tanto biomasa y luz de dócel asombrosamente tendió a quedarse igual.
+Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, explicado en parte por densidad de madera, y composición taxonomica varió notablemente.
+Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, tanto como el género Ficus, pero biomasa por modo de disperso no tendió a cambiar.
 Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.
@@
 
-Forest and landscape restoration is a key international conservation and climate change adaptation strategy \emph{(\protect\hyperlink{ref-depinto20}{De Pinto et al., 2020})}, and while tropical forests specifically store most land biomass \emph{(\protect\hyperlink{ref-pan13}{Pan et al., 2013})}, most are now secondary \emph{(\protect\hyperlink{ref-fao20}{FAO, 2020})} and regenerating degraded functions \emph{(\protect\hyperlink{ref-hubau20}{Hubau et al., 2020})} amid increasing climate stressors \emph{(\protect\hyperlink{ref-anderegg22}{Anderegg et al., 2022})} yet insufficient policy support \emph{(\protect\hyperlink{ref-chazdon18}{Chazdon, 2018})} via skewed priorities \emph{(\protect\hyperlink{ref-pyron22}{Pyron and Mooers, 2022})}.
-Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
+Forest and landscape restoration is a key international conservation and climate change adaptation strategy \emph{(\protect\hyperlink{ref-depinto20}{De Pinto et al., 2020})}.
+While tropical forests specifically store most land biomass \emph{(\protect\hyperlink{ref-pan13}{Pan et al., 2013})}, most are now secondary \emph{(\protect\hyperlink{ref-fao20}{FAO, 2020})} and functionally degraded \emph{(\protect\hyperlink{ref-hubau20}{Hubau et al., 2020})}.
+This likely amplifies amid increasing climate stressors \emph{(\protect\hyperlink{ref-anderegg22}{Anderegg et al., 2022})} and collective management issues, including insufficient policy support \emph{(\protect\hyperlink{ref-chazdon18}{Chazdon, 2018})} via skewed priorities \emph{(\protect\hyperlink{ref-pyron22}{Pyron and Mooers, 2022})}.
+Overall, secondary forests can regenerate relatively quickly compared to old-growth forests, in some ways.
 Neotropical secondary forests can grow quickly enough to accumulate biomass \textgreater10x faster than old-growth forests \emph{(\protect\hyperlink{ref-finegan96}{Finegan, 1996}; \protect\hyperlink{ref-poorter16}{Poorter et al., 2016})}, and substantially offset carbon emissions, with estimates ranging from \textasciitilde10\% from Amazonian deforestation \emph{(\protect\hyperlink{ref-smith20}{Smith et al., 2020})} to a decade of fossil fuel emissions from all of Latin American and the Caribbean \emph{(\protect\hyperlink{ref-chazdon16}{Chazdon et al., 2016})}.
@@
 The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services \emph{(\protect\hyperlink{ref-guariguata01}{Guariguata and Ostertag, 2001}; \protect\hyperlink{ref-stanturf14}{Stanturf et al., 2014})} like biomass and carbon storage \emph{(\protect\hyperlink{ref-liu18}{Liu et al., 2018})}.
-However, secondary neotropical forests diverge widely in taxonomic composition \emph{(\protect\hyperlink{ref-norden15}{Norden et al., 2015})}, potentially taking over a century to recover \emph{(\protect\hyperlink{ref-poorter21}{Poorter et al., 2021})} with added variability \emph{(\protect\hyperlink{ref-atkinson22}{Atkinson et al., 2022})}.
+However, recovery of secondary neotropical forests diverge widely in taxonomic composition \emph{(\protect\hyperlink{ref-norden15}{Norden et al., 2015})}, potentially taking over a century to recover \emph{(\protect\hyperlink{ref-poorter21}{Poorter et al., 2021})} with added variability \emph{(\protect\hyperlink{ref-atkinson22}{Atkinson et al., 2022})}.
 Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa \emph{(\protect\hyperlink{ref-carlo16}{Carlo and Morales, 2016})}, which can be keystone to locally-adapted food webs, including birds \emph{(\protect\hyperlink{ref-maas16}{Maas et al., 2016})}.
-Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, yet local management remains the strongest driver of tropical forest carbon stores \emph{(\protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}.
+Overall, understanding secondary forest regeneration is key for global biodiversity conservation at the global scale.
 
-Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, as well as land-use history, especially in heterogeneously-managed landscapes \emph{(\protect\hyperlink{ref-melo13}{Melo et al., 2013}; \protect\hyperlink{ref-perfecto09}{Perfecto et al., 2009})}.
+Modern secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, especially in heterogeneously-managed landscapes \emph{(\protect\hyperlink{ref-melo13}{Melo et al., 2013}; \protect\hyperlink{ref-perfecto09}{Perfecto et al., 2009})}.
 About 70\% of all forests have been estimated to be within a km of their edge \emph{(\protect\hyperlink{ref-haddad15}{Haddad et al., 2015})}, resulting in tropical forests having been estimated to lose \textasciitilde22\% aboveground carbon along heavily-managed edges \textgreater100m, in part due to shortening trees and lighter leaf mass \emph{(\protect\hyperlink{ref-ordway20}{Ordway and Asner, 2020})}.
-Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa \emph{(\protect\hyperlink{ref-wandrag17}{Wandrag et al., 2017})}, which while complicated \emph{(\protect\hyperlink{ref-muller-landau09}{Muller-Landau and Hardesty, 2009})}, can also correlate with simpler plant traits like height and seed mass \emph{(\protect\hyperlink{ref-tamme14}{Tamme et al., 2014}; \protect\hyperlink{ref-thomson11}{Thomson et al., 2011})}.
+Effects of degraded edges have historically favored tree taxa that tend to be pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
+Accordingly, secondary forest wood density may also tend to decrease near degraded habitat edges.
+Additionally, the dominance of pioneering taxa could lead to arrested succession \emph{(\protect\hyperlink{ref-tymen16}{Tymen et al., 2016})}, which further delays the recruitment of shade-tolerant taxa, as well as associated increases in aboveground biomass, in part due to their higher survival with increasing shade, and tendency toward relatively higher wood densities.
+However, arrested succession could be prevented, and even typical succession accelerated, if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
+In this case, having a habitat edge near a relatively rich intact forest, rather than near a relatively degraded open area, would instead be an asset to regenerating plantation forest patches, rather than a detriment.
+Overall, edge effects ultimately have the potential to shape community composition, in part based on the local dispersal kernel patterns of existing taxa \emph{(\protect\hyperlink{ref-wandrag17}{Wandrag et al., 2017})}, which while complicated \emph{(\protect\hyperlink{ref-muller-landau09}{Muller-Landau and Hardesty, 2009})}, can also correlate with simpler plant traits like height and seed mass \emph{(\protect\hyperlink{ref-tamme14}{Tamme et al., 2014}; \protect\hyperlink{ref-thomson11}{Thomson et al., 2011})}.
 
-Properties of the existing tree community also depend on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012})} including plantation use for timber production, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
-Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages \emph{(\protect\hyperlink{ref-albrich20}{Albrich et al., 2020}; \protect\hyperlink{ref-gough21}{Gough et al., 2021})}, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry \emph{(\protect\hyperlink{ref-lefland18}{Lefland et al., 2018})}.
-Overall, while edge effects near degraded habitat edges are often negative, meta-population \emph{(\protect\hyperlink{ref-levins79}{Levins, 1979})} and -community \emph{(\protect\hyperlink{ref-leibold04}{Leibold et al., 2004}; \protect\hyperlink{ref-warren15}{Warren et al., 2015})} theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
-Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests \emph{(\protect\hyperlink{ref-oliveira-neto17}{Oliveira-Neto et al., 2017})}, forests affected by logging \emph{(\protect\hyperlink{ref-clark12}{Clark and Covey, 2012}; \protect\hyperlink{ref-edwards14}{Edwards et al., 2014})}, and timber plantations \emph{(\protect\hyperlink{ref-pryde15}{Pryde et al., 2015})}.
+Additionally, properties of the tropical forest carbon stores and tree community depend heavily on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012}; \protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}.
+Overall, legacy effects from management like intensive timber planting could potentially delay or result in non-linear changes between successional stages \emph{(\protect\hyperlink{ref-albrich20}{Albrich et al., 2020}; \protect\hyperlink{ref-gough21}{Gough et al., 2021})}, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry \emph{(\protect\hyperlink{ref-lefland18}{Lefland et al., 2018})}.
+In some cases, the leftover timber stands could provide additional shade, which would not be available in a clear-cut harvested timber plantation, and this could help more shade-tolerant taxa establish and recruit better.
+Overall, edge effects near degraded habitat edges are often studied as negative, but conceptual influences from meta-population \emph{(\protect\hyperlink{ref-levins79}{Levins, 1979})} and meta-community \emph{(\protect\hyperlink{ref-leibold04}{Leibold et al., 2004}; \protect\hyperlink{ref-warren15}{Warren et al., 2015})} theories do highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity, thereby potentially making particular edge effects, specifically from adjacent intact forests, actually beneficial for regeneration plantation forest parcels.
+Taken together, these ideas suggest that edge effects specifically from adjacent intact primary forests, potentially via shade-tolerant dispersal and performance, may indeed facilitate plantation forest regeneration.
+Such a beneficial edge effect may ultimately help explain relatively high biodiversity, compared to adjacent stands, found in agroforests \emph{(\protect\hyperlink{ref-oliveira-neto17}{Oliveira-Neto et al., 2017})}, forests affected by logging \emph{(\protect\hyperlink{ref-clark12}{Clark and Covey, 2012}; \protect\hyperlink{ref-edwards14}{Edwards et al., 2014})}, and timber plantations \emph{(\protect\hyperlink{ref-pryde15}{Pryde et al., 2015})}.
 
-This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
-We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade that would suppress less shade-tolerant pioneer tree taxa and allow for more tree diversity closer to the primary forest edge.
-Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.
+Influenced by restoration ecology for forest management, this study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, focused on highlighting edge effects from an adjacent primary forest on overall stand properties, tree community composition, and limited functional trait recovery.
+We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade, which would suppress less shade-tolerant pioneer tree taxa, and in turn facilitate more shade-tolerant tree taxa, ultimately allowing for more tree diversity within the regenerating plantation forest in parts that were closer to the richer edge of the intact unmanaged forest.
+Accordingly, we predicted more specifically that further from the intact primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.
 
@@
 (See \emph{Taylor et al. (\protect\hyperlink{ref-taylor15}{2015})} for a broader ecosystem description of the region.)
-The study site was a regrowing 20 ha timber plantation of \emph{Bombacompsis quinata} abandoned in \textasciitilde1990 after the vulnerably dry-adapted species from the Guanacaste region \emph{(\protect\hyperlink{ref-hulshof20}{Hulshof and Powers, 2020}; \protect\hyperlink{ref-perezcordero02}{Pérez Cordero and Kanninen, 2002})} grew poorly in very wet conditions.
+The study site was a regrowing 20 ha timber plantation of \emph{Bombacompsis quinata} abandoned in \textasciitilde1990 after the vulnerably dry-adapted species from the Guanacaste region
+\emph{(\protect\hyperlink{ref-hulshof20}{Hulshof and Powers, 2020})} grew poorly in very wet conditions.
 This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \ref{fig:map}) and a wide service road on the third NE border, with primary forest beyond it.
@@
 \includegraphics{merge_files/figure-latex/map-1.pdf}
-\caption{\label{fig:map}Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica.
-Map image data © 2009 Google, © 2022 Osa Conservation.
-Photo credit: Nicholas Medina.}
+\caption{\label{fig:map}Map of (a) conservation areas and (b) study site (c) showing plot design with 50 m distance strata bins, and (d) focal plantation timber taxon Bombacopsis quinata on Osa Peninsula, Costa Rica.
+Map image data © 2022 Osa Conservation, © 2009 Google.
+Map GIS data and photo credit: Max Villalobos, Nicholas Medina.}
 \end{figure}
@@
 
-Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \ref{fig:map}), using available GIS software (\emph{ArcGIS 10}, \url{esri.com}, and \emph{QGIS 2}, \url{qgis.org}).
+Edge effects were studied by dividing the secondary forest area into six discrete 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \ref{fig:map}), using available GIS software (\emph{ArcGIS 10}, \url{esri.com}, and \emph{QGIS 2}, \url{qgis.org}).
+Note that in line with hypotheses about facilitated regeneration, this study most often referred to the ``edge'' of the regenerating plantation forest as the part closest to the unmanaged intact forest, where trees may be dispersing from, rather than the open service road, which is usually the focus of other studies evaluating effects of forest fragmentation and degradation, rather than restoration.
 Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
+Given the small size of the last 300 m distance stratum (Fig \ref{fig:map}c), one census plot was sufficiently representative.
+All data were later centered at the stratum level, so resulting trends were not weighted by distance bands with more plots in them.
 The total area of the 30 census plots equaled \textasciitilde1 ha or 5\% of the total secondary forest stand area, which is comparable to similar studies \emph{(\protect\hyperlink{ref-onyekwelu16}{Onyekwelu and Olabiwonnu, 2016})}.
@@
 Light reaching the forest floor was measured at the center of each plot at chest height using a densiometer \emph{(Forestry Suppliers, Inc.)}, as an average of four readings taken facing each cardinal direction.
-
 The slope of the forest floor was measured using a rangefinder \emph{(Bushnell, Forestry Suppliers, Inc)} to measure the distance the diagonal between two plot corners and triangulate the observation angle.
 
-The diameter of all stems \textgreater10 cm wide were recorded in each census plot, totaling over 1,000 trees.
-Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
+The diameter of all stems \textgreater10 cm wide were recorded in each census plot, totaling over 1,243 trees.
+In cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
 Tree height was recorded by measuring distances to both the crown and the stem at chest height \emph{(\textasciitilde2.7 m)} using a rangefinder \emph{(Bushnell, Forestry Suppliers, Inc.)} and triangulating the missing side length.
-Taxa were identified with localized knowledge and field guides matching common names to Latin binomials,
+
+Taxa were identified by local experts,
 and trait information was gathered from the literature.
+Final traits used only included successional stage (early, late) and main dispersal mode (wind, animal, water), which were ultimately matched to species using only the specific dataset made available by \emph{Werden et al.
+(\protect\hyperlink{ref-werden20}{2020})}.
+Otherwise wood tissue carbon was matched to genera with the GLOWCAD database \emph{(\protect\hyperlink{ref-doraisami22}{Doraisami et al., 2022})}, given high taxonomic alignment of this trait above the species level \emph{(\protect\hyperlink{ref-chave06}{Chave et al., 2006})}.
 
-Aboveground biomass (AGB) was estimated per stem using the allometric equation by \emph{Chave et al. (\protect\hyperlink{ref-chave15}{2015})}.
+Aboveground biomass (AGB) was calculated at the individual stem level using the allometric equation by \emph{Chave et al. (\protect\hyperlink{ref-chave15}{2015})} and later centered at the plot level.
 The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by \emph{(\protect\hyperlink{ref-alvarez12}{Alvarez et al., 2012}; Chave et al.; \protect\hyperlink{ref-chave05}{2005}; \protect\hyperlink{ref-ngomanda14}{Ngomanda et al., 2014})}.
-The equation version with height was used since height was accessibly measurable and has shown improved local accuracy \emph{(\protect\hyperlink{ref-domke12}{Domke et al., 2012}; \protect\hyperlink{ref-khan20}{Khan et al., 2020})}.
+The equation version with height was used since height was accessibly measurable and has shown improved local accuracy \emph{(\protect\hyperlink{ref-domke12}{Domke et al., 2012})}.
 Wood density (or specific gravity) values when available also improve biomass estimates even more than height \emph{(\protect\hyperlink{ref-vanbreugel11}{Van Breugel et al., 2011})}.
@@
 DBHs and heights were measured in the field and wood densities were extracted from available literature.
-Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be \emph{0.58} following World Agroforestry Database guidelines \href{worldagroforestry.org}{(worldagroforestry.org)}.
-Original values found and recorded were later updated to modern values using the pipeline developed in the \emph{BIOMASS} 2.1.8 R software package by \emph{Réjou-Méchain et al.
+Values were updated using the pipeline developed in the \emph{BIOMASS} 2.1.8 R software package by \emph{Réjou-Méchain et al.
 (\protect\hyperlink{ref-rejou-mechain17}{2017})}, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.
-
-Traits including successional stage and main dispersal mode were matched to species with \emph{Werden et al.
-(\protect\hyperlink{ref-werden20}{2020})} data, and wood tissue carbon matched to genera with the GLOWCAD database \emph{(\protect\hyperlink{ref-doraisami22}{Doraisami et al., 2022})} given high alignment at species levels \emph{(\protect\hyperlink{ref-chave06}{Chave et al., 2006})}.
 
@@
 
-For all response variables, data were analyzed using plot or distance strata medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base R version 4.2.0 (2022-04-22) and \emph{rstatix} 0.7.0 package functions \emph{(\protect\hyperlink{ref-base}{R Core Team, 2022})}.
+For all response variables, data were centered using medians at the plot level, followed by the discrete distance stratum level where applicable, which was most cases and indicated on figures.
+These medians were then subjected to linear regression with distance to intact forest edge as the only independent variable, which was binned discretely according to census design.
+All regressions were run through assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base R version 4.2.1 (2022-06-23) and \emph{rstatix} 0.7.0 package functions \emph{(\protect\hyperlink{ref-base}{R Core Team, 2022})}.
 
-Non-linear regressions were run using the function \emph{poly()} in the base R package \emph{stats} 4.2.0,
+Non-linear regressions were run using the function \emph{poly()} in the base R package \emph{stats} 4.2.1,
 which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models.
-All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
-Community analysis and ordination was done as a PERMANOVA with the \emph{adonis()} function from the \emph{vegan} 2.6.2 R package \emph{(\protect\hyperlink{ref-dixon03}{Dixon, 2003})}.
+All trees censused were included in each analysis, including unidentified taxa, which were only grouped together for taxonomic analyses and whose exclusion did not affect observed patterns shown in results.
+Community analysis and ordination was done as a PERMANOVA with the \emph{adonis()} function from the \emph{vegan} 2.6.4 R package \emph{(\protect\hyperlink{ref-dixon03}{Dixon, 2003})}.
 Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and internal pipeline \emph{oir} \emph{(\protect\hyperlink{ref-softwareMedina22oir}{Medina, 2022a})}, and are stored at \url{github.com/nmedina17/osa} \emph{(\protect\hyperlink{ref-softwareMedina22osa}{Medina, 2022b})}.
@@
 with edge distance but more strongly and non-linearly, by
-\textasciitilde17.4 ±
+\textasciitilde17.39 ±
 3.4 per m
@@
 (Fig \ref{fig:biomass}f).
+Additional variables tested, including maximum height and diameter, tended to stay the same across the edge gradient \emph{(data not shown)}.
 
@@
 \includegraphics{merge_files/figure-latex/biomass-1.pdf}
-\caption{\label{fig:biomass} Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density.
-Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
+\caption{\label{fig:biomass} Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (e) canopy light availability, all across distance to primary forest edge gradient, and (f) diversity against wood density.
+Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
 \end{figure}
@@
 \textasciitilde0.5 ±
-0 per m
+0.029 per m
 as distance to primary forest edge increased,
@@
 by
-0 ±
-0 per m
+0.0063 ±
+0.0025 per m
 and had \emph{16}\% variance explained
@@
 (p=0.01) with
-11.4\%
+11.38\%
 variance among distance strata explained by distance to primary forest edge,
@@
 and linearly by
--62.6 ±
-25.1 per m
+-62.62 ±
+25.05 per m
 with \emph{\textasciitilde20}\% variance among distance strata explained
 (Fig \ref{fig:taxa}f).
+Other taxa did not change significantly with distance to intact forest \emph{(data not shown)}.
 
@@
 \caption{\label{fig:taxa} Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa.
-White and colored dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
+White and colored dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
 \end{figure}
@@
 by
-\textasciitilde{} -584.9 ±
-167.3 kg per m
+\textasciitilde{} -584.87 ±
+167.31 kg per m
 with distance to edge explaining
@@
 \caption{\label{fig:traits} Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode.
-Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
+Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
 \end{figure}
@@
 This study found significant edge effects on secondary forest stand wood and stem density, tree diversity and community composition, and some functional traits, yet not on overall aboveground biomass.
-It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential,
-and the evidence presented here supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
+It was initially hypothesized that community composition would vary with distance to high-quality intact primary forest edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential.
+The evidence presented here supported edge effects on community composition and biomass of taxa that were associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
 Accordingly, this suggests that edge effects do significantly change humid tropical forest community taxonomic and functional composition, but primarily due to factors other than light competition or dispersal limitation \emph{(\protect\hyperlink{ref-krishnadas20}{Krishnadas et al., 2020})}, and also suggests that aboveground biomass and even low taxonomic richness are resilient after a decade of regeneration.
@@
 Management has been found to explain most of biomass variation among tropical forest fragments, and wood density \textasciitilde10\% biomass variation \emph{(\protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}.
-This study adds that management via edge effects can also mediate \textasciitilde10\% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns \emph{(\protect\hyperlink{ref-baker04}{Baker et al., 2004})}.
+Additionally, this study contributes that management via edge effects can also mediate \textasciitilde10\% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns \emph{(\protect\hyperlink{ref-baker04}{Baker et al., 2004})}.
 In this study, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon \emph{(\protect\hyperlink{ref-gonzalezj98}{Gonzalez J and Fisher, 1998}; \protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}, and increasing nearby stem density.
-Lower wood density can also reflect more dynamic forest stands \emph{(\protect\hyperlink{ref-malhi06}{Malhi et al., 2006})} with higher tree turnover, a consequence of forest fragmentation \emph{(\protect\hyperlink{ref-nascimento04a}{Nascimento and Laurance, 2004})} together with other global change drivers \emph{(\protect\hyperlink{ref-laurance14}{Laurance et al., 2014})}, as well as may reflect possible differences in soil fertility \emph{(\protect\hyperlink{ref-malhi04}{Malhi et al., 2004})}, all of which represent possible future research directions, beyond strictly neutral dynamics \emph{(\protect\hyperlink{ref-terborgh95}{Terborgh et al., 1995})}.
+Lower wood density can also reflect more dynamic forest stands \emph{(\protect\hyperlink{ref-malhi06}{Malhi et al., 2006})} with higher tree turnover, a consequence of forest fragmentation \emph{(\protect\hyperlink{ref-nascimento04}{Nascimento and Laurance, 2004})} together with other global change drivers \emph{(\protect\hyperlink{ref-laurance14}{Laurance et al., 2014})}, and may also reflect possible differences in soil fertility \emph{(\protect\hyperlink{ref-malhi04}{Malhi et al., 2004})}, all of which represent possible future research directions, beyond strictly neutral dynamics \emph{(\protect\hyperlink{ref-terborgh95}{Terborgh et al., 1995})}.
 Furthermore, light availability at the plot level did not affect biomass storage, but instead, individual light gaps may be more important for biomass dynamics \emph{(\protect\hyperlink{ref-chazdon84}{Chazdon and Fetcher, 1984})}.
 
-Edge effects have been shown in other studies to lower both forest biomass and change community composition \emph{(\protect\hyperlink{ref-anderson22}{Anderson et al., 2022})}, and this study adds that the decline in diversity can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
+Edge effects have been shown in other studies not only to lower forest biomass, but also to change community composition \emph{(\protect\hyperlink{ref-anderson22}{Anderson et al., 2022})}, and in addition this study contributes that the decline in diversity can be rapid and non-linear across edge gradients, along with supporting the likelihood of changes to community composition lasting decades.
 This rapid loss of diversity across the edge gradient may be in part due to existing higher yet constant levels of shade limiting shade-intolerant seedling recruitment and/or performance, together with insufficient time for slower-growing shade-tolerant taxa to accumulate significant amounts of biomass.
-Another possible factor may be overall short dispersal kernels, in part due to higher fungal pathogen pressure under higher constant shade, also increasing tree mortality.
-Decomposing compositional changes into key specific taxa, while the biomass of the most abundant taxon \emph{Vochysia} was resilient to edge effects in this study, that of the key taxon for frugivores like birds and bats \emph{Ficus} was significantly lowered across the secondary forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.
-Community composition also tended to show lower variance across the edge gradient, although this effect cannot be confirmed here given the uneven sampling that accounted for variable forest area at different distances from edge.
+Another possible factor may be Janzen-Connell processes \emph{(\protect\hyperlink{ref-terborgh20}{Terborgh, 2020}; \protect\hyperlink{ref-wills06}{Wills, 2006})}, including conspecific negative density-dependence of seedling survival on basal area \emph{(\protect\hyperlink{ref-comita09}{Comita and Hubbell, 2009})}, as well as overall short dispersal kernels, whose effects can be mediated by shade \emph{(\protect\hyperlink{ref-comita14}{Comita et al., 2014})} and in part by higher fungal pathogen pressure near conspecifics \emph{(\protect\hyperlink{ref-jia20}{Jia et al., 2020})}.
+While both diversity and community composition changed significantly with distance to intact forest in this study, the biomass of most individual taxa, surprisingly, tended to stay the same along the edge gradient.
+This lack of significant individual taxon biomass responses along this edge gradient could be due to stochastic population factors affecting their biomass growth, perhaps widening variability.
+Regardless of the underlying process, these absent patterns do also help explain the lack of pattern observed with plot biomass overall across the edge gradient.
+This explanation points to future studies of taxon-specific responses to edge effects, to help explain overall forest biomass dynamics near habitat edge.
+More specifically here, the biomass of the most abundant taxon \emph{Vochysia} was resilient to edge effects in this study, however, that of the taxon \emph{Ficus}, key for frugivores like birds and bats \emph{(\protect\hyperlink{ref-cottee-jones15}{Cottee-Jones et al., 2015}; \protect\hyperlink{ref-rafidison20}{Rafidison et al., 2020})}, was significantly lowered across the secondary forest edge, pointing to potentially different responses of biomass and local food web associated ecosystem services.
+Results suggest that this pattern is specific to this locally widespread \emph{Ficus} genus, since the overall biomass of other animal-dispersed taxa as a whole tended to stay the same across the edge gradient studied here.
+Interestingly, community composition also appeared to show lower variance across the edge gradient, although testing this observation in a forested area with evenly distributed area among distance strata bands would provide additional support.
 
@@
 A study in Madagascar found lower phylogenetic diversity closer to degraded forest edge, but no change in aboveground biomass \emph{(\protect\hyperlink{ref-razafindratsima18}{Razafindratsima et al., 2018})}.
-These results parallel in showing lower taxonomic diversity, although biomass specifically of taxa assoociated with both early- and late-successional stages also decreased with away from primary forest edge.
-This result by successional stage can be explained by slightly lower habitat quality inferred near exposed habitat edges near roads, and accordingly relatively higher habitat quality near closed primary forest edge.
-While lower biomass of succession-agnostic taxa further from primary forest edge appears consistent with dispersal kernels, the lack of pattern by broad dispersal mode limits this explanation here, yet could be further tested by measuring taxon-specific dispersal kernels.
+This study presents similar results in showing lower taxonomic diversity, although biomass specifically of taxa associated with both early- and late-successional stages also decreased with away from primary forest edge.
+This resulting decrease by successional stage could be explained by slightly lower habitat quality inferred near exposed habitat edges near roads, and accordingly relatively higher habitat quality near closed primary forest edge \emph{(\protect\hyperlink{ref-ries04}{Ries et al., 2004})}.
+While lower biomass of succession-agnostic taxa further from primary forest edge appears consistent with the expected forms of group-level dispersal kernels as an explanation, the lack of significant pattern along forest edge when binned by broad dispersal mode that was found here may limit the potential of seed dispersal to explain biomass trends.
+Instead, seedling survival may have been more important than dispersal in accumulating basal area and biomass \emph{(\protect\hyperlink{ref-comita09}{Comita and Hubbell, 2009})}.
+For taxa primarily dispersed by animals instead of wind, this could be explained by potentially broader local limitations in animal dispersal activity or abundance.
+Additionally, the potential for nearby dispersal to facilitate reforestation could be tested in future studies by measuring taxon-specific dispersal kernels.
 
 Overall, this study highlights how abandoned wet tropical timber plantation can regenerate in alongside fragmented forests.
-Results support related syntheses that biomass and taxonomic richness often regrows relatively quickly,
+Results support related syntheses that tropical forest biomass and taxonomic richness often regrows relatively quickly \emph{(\protect\hyperlink{ref-davies21a}{Davies et al., 2021})},
 while taxonomic composition recovers much more slowly, if not diverging altogether \emph{(\protect\hyperlink{ref-norden15}{Norden et al., 2015})}.
@@
 
-This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship.
+This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship and in part by University of Michigan Rackham Merit Fellowship.
 
-\hypertarget{declaration-of-interests}{%
-\section*{Declaration of interests}\label{declaration-of-interests}}
-\addcontentsline{toc}{section}{Declaration of interests}
+\hypertarget{acknowledgements}{%
+\section*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{section}{Acknowledgements}
+
+Thanks to
+Brandeis University Professor Dan Perlman for initial networking;
+Lehigh University Professor Don Morris
+and student Erin Lau
+for field support;
+concurrent Osa Conservation general manager Manuel Ramírez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
+James Fifer, Chau Ho, and Michelle Spicer for review of early draft content;
+and Luisa Valdez for translation assistance.
 
@@
 
---\textgreater{}
+\hypertarget{author-contributions}{%
+\section*{Author contributions}\label{author-contributions}}
+\addcontentsline{toc}{section}{Author contributions}
+
+NM, MV and AJ contributed to study conceptualization, design, administration, and supervision;
+NM and AJ contributed to funding acquisition;
+NM, AM, and EC,
+contributed to key data collection;
+NM and MV contributed to data analysis;
+NM wrote initial draft and later revisions.
+
+Study team includes and recognizes diverse and historically-excluded contributions to research; with bilingual communication efforts completed and pending.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-comita09}{}}%
+Comita, L.S., Hubbell, S.P., 2009. Local neighborhood and species' shade tolerance influence survival in a diverse seedling bank. Ecology 90, 328--334. \url{https://doi.org/10.1890/08-0451.1}
+
+\leavevmode\vadjust pre{\hypertarget{ref-comita14}{}}%
+Comita, L.S., Queenborough, S.A., Murphy, S.J., Eck, J.L., Xu, K., Krishnadas, M., Beckman, N., Zhu, Y., 2014.
+Testing predictions of the {Janzen}\textendash{{Connell}} hypothesis: A meta-analysis of experimental evidence for distance- and density-dependent seed and seedling survival.
+Journal of Ecology 102, 845--856.
+\url{https://doi.org/10.1111/1365-2745.12232}
+
+\leavevmode\vadjust pre{\hypertarget{ref-cottee-jones15}{}}%
+Cottee-Jones, H.E.W., Bajpai, O., Chaudhary, L.B., Whittaker, R.J., 2015.
+Isolated {Ficus} trees deliver dual conservation and development benefits in a rural landscape.
+Ambio 44, 678--684.
+\url{https://doi.org/10.1007/s13280-015-0645-9}
+
+\leavevmode\vadjust pre{\hypertarget{ref-davies21a}{}}%
+Davies, S.J., Abiem, I., Abu Salim, K., Aguilar, S., Allen, D., Alonso, A., Anderson-Teixeira, K., Andrade, A., Arellano, G., Ashton, P.S., Baker, P.J., Baker, M.E., Baltzer, J.L., Basset, Y., Bissiengou, P., Bohlman, S., Bourg, N.A., Brockelman, W.Y., Bunyavejchewin, S., Burslem, D.F.R.P., Cao, M., Cárdenas, D., Chang, L.-W., Chang-Yang, C.-H., Chao, K.-J., Chao, W.-C., Chapman, H., Chen, Y.-Y., Chisholm, R.A., Chu, C., Chuyong, G., Clay, K., Comita, L.S., Condit, R., Cordell, S., Dattaraja, H.S., de Oliveira, A.A., den Ouden, J., Detto, M., Dick, C., Du, X., Duque, Á., Ediriweera, S., Ellis, E.C., Obiang, N.L.E., Esufali, S., Ewango, C.E.N., Fernando, E.S., Filip, J., Fischer, G.A., Foster, R., Giambelluca, T., Giardina, C., Gilbert, G.S., Gonzalez-Akre, E., Gunatilleke, I.A.U.N., Gunatilleke, C.V.S., Hao, Z., Hau, B.C.H., He, F., Ni, H., Howe, R.W., Hubbell, S.P., Huth, A., Inman-Narahari, F., Itoh, A., Janík, D., Jansen, P.A., Jiang, M., Johnson, D.J., Jones, F.A., Kanzaki, M., Kenfack, D., Kiratiprayoon, S., Král, K., Krizel, L., Lao, S., Larson, A.J., Li, Y., Li, X., Litton, C.M., Liu, Y., Liu, S., Lum, S.K.Y., Luskin, M.S., Lutz, J.A., Luu, H.T., Ma, K., Makana, J.-R., Malhi, Y., Martin, A., McCarthy, C., McMahon, S.M., McShea, W.J., Memiaghe, H., Mi, X., Mitre, D., Mohamad, M., Monks, L., Muller-Landau, H.C., Musili, P.M., Myers, J.A., Nathalang, A., Ngo, K.M., Norden, N., Novotny, V., O'Brien, M.J., Orwig, D., Ostertag, R., Papathanassiou, K., Parker, G.G., Pérez, R., Perfecto, I., Phillips, R.P., Pongpattananurak, N., Pretzsch, H., Ren, H., Reynolds, G., Rodriguez, L.J., Russo, S.E., Sack, L., Sang, W., Shue, J., Singh, A., Song, G.-Z.M., Sukumar, R., Sun, I.-F., Suresh, H.S., Swenson, N.G., Tan, S., Thomas, S.C., Thomas, D., Thompson, J., Turner, B.L., Uowolo, A., Uriarte, M., Valencia, R., Vandermeer, J., Vicentini, A., Visser, M., Vrska, T., Wang, X., Wang, X., Weiblen, G.D., Whitfeld, T.J.S., Wolf, A., Wright, S.J., Xu, H., Yao, T.L., Yap, S.L., Ye, W., Yu, M., Zhang, M., Zhu, D., Zhu, L., Zimmerman, J.K., Zuleta, D., 2021.
+{ForestGEO}: {Understanding} forest diversity and dynamics through a global observatory network.
+Biological Conservation 253, 108907.
+\url{https://doi.org/10.1016/j.biocon.2020.108907}
+
 \leavevmode\vadjust pre{\hypertarget{ref-depinto20}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-jia20}{}}%
+Jia, S., Wang, X., Yuan, Z., Lin, F., Ye, J., Lin, G., Hao, Z., Bagchi, R., 2020.
+Tree species traits affect which natural enemies drive the {Janzen-Connell} effect in a temperate forest.
+Nature Communications 11, 286.
+\url{https://doi.org/10.1038/s41467-019-14140-y}
+
 \leavevmode\vadjust pre{\hypertarget{ref-kearney21}{}}%
@@
 \url{https://doi.org/10.1111/1365-2435.13829}
-
-\leavevmode\vadjust pre{\hypertarget{ref-khan20}{}}%
-Khan, M.N.I., Islam, M.R., Rahman, A., Azad, M.S., Mollick, A.S., Kamruzzaman, M., Sadath, M.N., Feroz, S.M., Rakkibu, M.G., Knohl, A., 2020.
-Allometric relationships of stand level carbon stocks to basal area, tree height and wood density of nine tree species in {Bangladesh}.
-Global Ecology and Conservation 22.
-\url{https://doi.org/10.1016/j.gecco.2020.e01025}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nascimento04a}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-nascimento04}{}}%
 Nascimento, H.E.M., Laurance, W.F., 2004. {BIOMASS DYNAMICS IN AMAZONIAN FOREST FRAGMENTS}. Ecological Applications 14, 127--138. \url{https://doi.org/10.1890/01-6003}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-perezcordero02}{}}%
-Pérez Cordero, L.D., Kanninen, M., 2002.
-Wood specific gravity and aboveground biomass of {Bombacopsis} quinata plantations in {Costa Rica}.
-Forest Ecology and Management 165, 1--9.
-\url{https://doi.org/10.1016/S0378-1127(01)00627-2}
-
 \leavevmode\vadjust pre{\hypertarget{ref-perfecto09}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-rafidison20}{}}%
+Rafidison, V.M., Rakouth, B., Carrière, S.M., Kjellberg, F., Aumeeruddy-Thomas, Y., 2020.
+Multiple values of isolated and clusters of {Ficus} tree species protected by {Betsileo} farmers in rural landscapes in {Madagascar}: Implications for biodiversity conservation.
+Biodiversity and Conservation 29, 1027--1058.
+\url{https://doi.org/10.1007/s10531-019-01924-3}
+
 \leavevmode\vadjust pre{\hypertarget{ref-razafindratsima18}{}}%
@@
 \url{https://doi.org/10.1111/2041-210X.12753}
+
+\leavevmode\vadjust pre{\hypertarget{ref-ries04}{}}%
+Ries, L., Fletcher, R.J., Battin, J., Sisk, T.D., 2004.
+Ecological {Responses} to {Habitat Edges}: {Mechanisms}, {Models}, and {Variability Explained}.
+Annual Review of Ecology, Evolution, and Systematics 35, 491--522.
+\url{https://doi.org/10.1146/annurev.ecolsys.35.112202.130148}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-terborgh20}{}}%
+Terborgh, J., 2020. At 50, {Janzen}\textendash{{Connell Has Come}} of {Age}. BioScience 70, 1082--1092. \url{https://doi.org/10.1093/biosci/biaa110}
+
 \leavevmode\vadjust pre{\hypertarget{ref-terborgh95}{}}%
@@
 \url{https://doi.org/10.1111/j.1365-2745.2011.01867.x}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tymen16}{}}%
+Tymen, B., Réjou-Méchain, M., Dalling, J.W., Fauset, S., Feldpausch, T.R., Norden, N., Phillips, O.L., Turner, B.L., Viers, J., Chave, J., 2016.
+Evidence for arrested succession in a liana-infested {Amazonian} forest.
+Journal of Ecology 104, 149--159.
+\url{https://doi.org/10.1111/1365-2745.12504}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-wills06}{}}%
+Wills, C., 2006. Nonrandom {Processes Maintain Diversity} in {Tropical Forests}. Science 311, 527--531. \url{https://doi.org/10.1126/science.1117715}
+
 \leavevmode\vadjust pre{\hypertarget{ref-bookdown2022}{}}%

### 2022-11-09 article/methods.Rmd
bioRxiv,poster

@@
 (See *@taylor15* for a broader ecosystem description of the region.)
-The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the vulnerably dry-adapted species from the Guanacaste region *[@perezcordero02; @hulshof20]* grew poorly in very wet conditions.
+The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the vulnerably dry-adapted species from the Guanacaste region
+<!-- @perezcordero02; -->
+*[@hulshof20]* grew poorly in very wet conditions.
 This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \@ref(fig:map)) and a wide service road on the third NE border, with primary forest beyond it.
@@
 
-```{r map, fig.width=4.5, fig.height=4.5, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design with 50 m distance strata bins, and (d) focal plantation timber taxon Bombacopsis quinata on Osa Peninsula, Costa Rica.
+```{r map, fig.width=5, fig.height=5, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design with 50 m distance strata bins, and (d) focal plantation timber taxon Bombacopsis quinata on Osa Peninsula, Costa Rica.
 Map image data © 2022 Osa Conservation, © 2009 Google.
@@
 Light reaching the forest floor was measured at the center of each plot at chest height using a densiometer *(Forestry Suppliers, Inc.)*, as an average of four readings taken facing each cardinal direction.
-
 <!-- ### [Topography] -->
-
 The slope of the forest floor was measured using a rangefinder *(Bushnell, Forestry Suppliers, Inc)* to measure the distance the diagonal between two plot corners and triangulate the observation angle.
@@
 The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *@chave05 [@alvarez12; @ngomanda14]*.
-The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *[@domke12; @khan20]*.
+The equation version with height was used since height was accessibly measurable and has shown improved local accuracy *[@domke12]*.
+<!-- [@khan20]* -->
 Wood density (or specific gravity) values when available also improve biomass estimates even more than height *[@vanbreugel11]*.

### 2022-11-09 article/results.Rmd
bioRxiv,poster

@@
 (Fig \@ref(fig:biomass)f).
+Additional variables tested, including maximum height and diameter, tended to stay the same across the edge gradient *(data not shown)*.
 
-```{r biomass, fig.width=4, fig.height=5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density.
+```{r biomass, fig.width=5, fig.height=6, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (e) canopy light availability, all across distance to primary forest edge gradient, and (f) diversity against wood density.
 Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
@@
 ~`r abs(entropy12$statPrint[[1]]$estimate[[2]])` ± 
-`r abs(entropy12$statPrint[[1]]$std.error[[2]])` per m 
+`r abs(entropy12$statPrint[[1]]$std.error[[2]]) %>% rstatix::p_format()` per m 
 as distance to primary forest edge increased, 
@@
 by 
-`r richness$statPrint[[1]]$estimate[[2]]` ± 
-`r richness$statPrint[[1]]$std.error[[2]]` per m 
+`r richness$statPrint[[1]]$estimate[[2]] %>% rstatix::p_format()` ± 
+`r richness$statPrint[[1]]$std.error[[2]] %>% rstatix::p_format()` per m 
 and had *`r "16"`*% variance explained 
@@
 (Fig \@ref(fig:taxa)f).
+Other taxa did not change significantly with distance to intact forest *(data not shown)*.
 
-```{r taxa, fig.width=5, fig.height=4, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa.
+```{r taxa, fig.width=6, fig.height=5, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa.
 White and colored dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
@@
 <!-- - put diversity-biomass correlation -->
+
+<!-- # ```{r taxaList} -->
+<!-- # source(here::here("tblTaxa.R"));  -->
+<!-- # ``` -->
 
@@
 
-```{r traits, fig.width=3, fig.height=4, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode.
+```{r traits, fig.width=4, fig.height=5, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode.
 Grey dots show individual tree values, white dots with border show plot medians, black dots show plot-level data binned into discrete distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}

### 2022-11-09 arxivTest/arxivTest.Rmd
bioRxiv,poster

---
title: A template for the *arxiv* style
authors:
  - name: David S. Hippocampus
    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Department of Computer Science
    affiliation: Cranberry-Lemon University
    location: Pittsburgh, PA 15213
    email: hippo@cs.cranberry-lemon.edu
  - name: Elias D. Striatum
    department: Department of Electrical Engineering
    affiliation: Mount-Sheikh University
    location: Santa Narimana, Levand
    email: stariate@ee.mount-sheikh.edu
abstract: |
  Enter the text of your abstract here.
keywords:
  - blah
  - blee
  - bloo
  - these are optional and can be removed
bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
---

# Introduction

Here goes an introduction text

# Headings: first level
\label{sec:headings}

You can use directly LaTeX command or Markdown text.

LaTeX command can be used to reference other section. See Section \ref{sec:headings}.
However, you can also use **bookdown** extensions mechanism for this.

## Headings: second level

You can use equation in blocks

$$
\xi _{ij}(t)=P(x_{t}=i,x_{t+1}=j|y,v,w;\theta)= {\frac {\alpha _{i}(t)a^{w_t}_{ij}\beta _{j}(t+1)b^{v_{t+1}}_{j}(y_{t+1})}{\sum _{i=1}^{N} \sum _{j=1}^{N} \alpha _{i}(t)a^{w_t}_{ij}\beta _{j}(t+1)b^{v_{t+1}}_{j}(y_{t+1})}}
$$

But also inline i.e $z=x+y$

### Headings: third level

Another paragraph.

# Examples of citations, figures, tables, references
\label{sec:others}

You can insert references. Here is some text [@kour2014real; @kour2014fast] and see @hadash2018estimate.

The documentation for \verb+natbib+ may be found at

You can use custom blocks with LaTeX support from **rmarkdown** to create environment.

::: {.center latex=true}
  <http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}>
:::

Of note is the command \verb+\citet+, which produces citations
appropriate for use in inline text.

You can insert LaTeX environment directly too.

\begin{verbatim}
   \citet{hasselmo} investigated\dots
\end{verbatim}

produces

\begin{quote}
  Hasselmo, et al.\ (1995) investigated\dots
\end{quote}

\begin{center}
  \url{https://www.ctan.org/pkg/booktabs}
\end{center}

## Figures

You can insert figure using LaTeX directly.

See Figure \ref{fig:fig1}. Here is how you add footnotes. [^Sample of the first footnote.]

\begin{figure}
  \centering
  \fbox{\rule[-.5cm]{4cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
  \label{fig:fig1}
\end{figure}

But you can also do that using R.

```{r fig2, fig.cap = "Another sample figure"}
plot(mtcars$mpg)
```

You can use **bookdown** to allow references for Tables and Figures.

## Tables

Below we can see how to use tables.

See awesome Table~\ref{tab:table} which is written directly in LaTeX in source Rmd file.

\begin{table}
 \caption{Sample table title}
  \centering
  \begin{tabular}{lll}
    \toprule
    \multicolumn{2}{c}{Part}                   \\
    \cmidrule(r){1-2}
    Name     & Description     & Size ($\mu$m) \\
    \midrule
    Dendrite & Input terminal  & $\sim$100     \\
    Axon     & Output terminal & $\sim$10      \\
    Soma     & Cell body       & up to $10^6$  \\
    \bottomrule
  \end{tabular}
  \label{tab:table}
\end{table}

You can also use R code for that.

```{r}
knitr::kable(head(mtcars), caption = "Head of mtcars table")
```

## Lists

- Item 1
- Item 2 
- Item 3

### 2022-11-09 elsevierTest/elsevierTest.Rmd
bioRxiv,poster

---
title: Short Paper
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliation: Some Institute of Technology
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  This is the abstract.

  It consists of two paragraphs.
keywords: 
  - keyword1
  - keyword2
journal: "An awesome journal"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: mybibfile.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

Please make sure that your manuscript follows the guidelines in the 
Guide for Authors of the relevant journal. It is not necessary to 
typeset your manuscript in exactly the same way as an article, 
unless you are submitting to a camera-ready copy (CRC) journal.

For detailed instructions regarding the elsevier article class, see   <https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions>

# Bibliography styles

Here are two sample references: @Feynman1963118 [@Dirac1953888].

By default, natbib will be used with the `authoryear` style, set in `classoption` variable in YAML. 
You can sets extra options with `natbiboptions` variable in YAML header. Example 
```yaml
natbiboptions: longnamesfirst,angle,semicolon
```

There are various more specific bibliography styles available at
<https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files>. 
To use one of these, add it in the header using, for example, `biblio-style: model1-num-names`.

## Using CSL

If `citation_package` is set to `default` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`. In this case, the `csl` option is used to format the references. Alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>. These can be downloaded
and stored locally, or the url can be used as in the example header.

# Equations

Here is an equation:
$$ 
  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
  \left(\frac{x}{\beta}\right)^{\alpha-1}
  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
  \alpha,\beta,x > 0 .
$$

Here is another:
\begin{align}
  a^2+b^2=c^2.
\end{align}

Inline equations: $\sum_{i = 2}^\infty\{\alpha_i^\beta\}$

# Figures and tables

Figure \ref{fig2} is generated using an R chunk.

```{r fig2, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig2}A meaningless scatterplot.", echo = FALSE}
plot(runif(25), runif(25))
```

# Tables coming from R

Tables can also be generated using R chunks, as shown in Table \ref{tab1} for example.

```{r tab1, echo = TRUE}
knitr::kable(head(mtcars)[,1:4], 
    caption = "\\label{tab1}Caption centered above table"
)
```

# References {-}

### 2022-11-09 elsevierTest/elsevierTest.tex
bioRxiv,poster

\documentclass[preprint, 3p,
authoryear]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

  \journal{An awesome journal} % Sets Journal name

\usepackage{lineno} % add

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Short Paper},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

\begin{document}

\begin{frontmatter}

  \title{Short Paper}
    \author[Some Institute of Technology]{Alice Anonymous%
  \corref{cor1}%
  \fnref{1}}
   \ead{alice@example.com} 
    \author[Another University]{Bob Security}
   \ead{bob@example.com} 
    \author[Another University]{Cat Memes%
  %
  \fnref{2}}
   \ead{cat@example.com} 
    \author[Some Institute of Technology]{Derek Zoolander%
  %
  \fnref{2}}
   \ead{derek@example.com} 
      \affiliation[Some Institute of Technology]{Department, Street,
City, State, Zip}
    \affiliation[Another University]{Department, Street, City, State,
Zip}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{This is the first author footnote.}
    \fntext[2]{Another author footnote.}

  \begin{abstract}
  This is the abstract.

  It consists of two paragraphs.
  \end{abstract}
    \begin{keyword}
    keyword1 \sep 
    keyword2
  \end{keyword}

 \end{frontmatter}

Please make sure that your manuscript follows the guidelines in the
Guide for Authors of the relevant journal. It is not necessary to
typeset your manuscript in exactly the same way as an article, unless
you are submitting to a camera-ready copy (CRC) journal.

For detailed instructions regarding the elsevier article class, see
\url{https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions}

\hypertarget{bibliography-styles}{%
\section{Bibliography styles}\label{bibliography-styles}}

Here are two sample references: \citeauthor{Feynman1963118}
\citetext{\citeyear{Feynman1963118}; \citealp{Dirac1953888}}.

By default, natbib will be used with the \texttt{authoryear} style, set
in \texttt{classoption} variable in YAML. You can sets extra options
with \texttt{natbiboptions} variable in YAML header. Example

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{natbiboptions}\KeywordTok{:}\AttributeTok{ longnamesfirst,angle,semicolon}
\end{Highlighting}
\end{Shaded}

There are various more specific bibliography styles available at
\url{https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files}.
To use one of these, add it in the header using, for example,
\texttt{biblio-style:\ model1-num-names}.

\hypertarget{using-csl}{%
\subsection{Using CSL}\label{using-csl}}

If \texttt{citation\_package} is set to \texttt{default} in
\texttt{elsevier\_article()}, then pandoc is used for citations instead
of \texttt{natbib}. In this case, the \texttt{csl} option is used to
format the references. Alternative \texttt{csl} files are available from
\url{https://www.zotero.org/styles?q=elsevier}. These can be downloaded
and stored locally, or the url can be used as in the example header.

\hypertarget{equations}{%
\section{Equations}\label{equations}}

Here is an equation: \[ 
  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
  \left(\frac{x}{\beta}\right)^{\alpha-1}
  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
  \alpha,\beta,x > 0 .
\]

Here is another: \begin{align}
  a^2+b^2=c^2.
\end{align}

Inline equations: \(\sum_{i = 2}^\infty\{\alpha_i^\beta\}\)

\hypertarget{figures-and-tables}{%
\section{Figures and tables}\label{figures-and-tables}}

Figure \ref{fig2} is generated using an R chunk.

\begin{figure}

{\centering \includegraphics[width=0.5\linewidth]{elsevierTest_files/figure-latex/fig2-1}

}

\caption{\label{fig2}A meaningless scatterplot.}\label{fig:fig2}
\end{figure}

\hypertarget{tables-coming-from-r}{%
\section{Tables coming from R}\label{tables-coming-from-r}}

Tables can also be generated using R chunks, as shown in Table
\ref{tab1} for example.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(}\FunctionTok{head}\NormalTok{(mtcars)[,}\DecValTok{1}\SpecialCharTok{:}\DecValTok{4}\NormalTok{], }
    \AttributeTok{caption =} \StringTok{"}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{label\{tab1\}Caption centered above table"}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}lrrrr@{}}
\caption{\label{tab1}Caption centered above table}\tabularnewline
\toprule()
& mpg & cyl & disp & hp \\
\midrule()
\endfirsthead
\toprule()
& mpg & cyl & disp & hp \\
\midrule()
\endhead
Mazda RX4 & 21.0 & 6 & 160 & 110 \\
Mazda RX4 Wag & 21.0 & 6 & 160 & 110 \\
Datsun 710 & 22.8 & 4 & 108 & 93 \\
Hornet 4 Drive & 21.4 & 6 & 258 & 110 \\
Hornet Sportabout & 18.7 & 8 & 360 & 175 \\
Valiant & 18.1 & 6 & 225 & 105 \\
\bottomrule()
\end{longtable}

\renewcommand\refname{References}
\bibliography{mybibfile.bib}

\end{document}

### 2022-11-09 frontiersTest/frontiersTest.Rmd
bioRxiv,poster

---
title: "Article Title"
runtitle: "Short Title"
documentclass: frontiersSCNS  # or frontiersHLTH, or frontiersFPHY
author:
  - name: First Author
    affiliation: '1'
    etal: LASTNAME # First author's last name. 
  - name: Co-Author
    affiliation: '2'
  - name: Co-Author
    affiliation: '1,3'
    email: email@uni.edu # Indicates corresponding Author
    institution: Laboratory X, Institute X, Department X, Organization X
    street: Street X
    city: City X
    state: State XX # only USA, Australia, Canada
    zip: Zip Code X
    country: Country X
affiliation:
  - id: '1'
    department: Department X
    institution: Institution X
    city: City X
    state: State XX # only USA, Australia, Canada
    country: Country X
  - id: '2' 
    department: Department X
    institution: Institution X
    city: City X
    state: State XX # only USA, Australia, Canada
    country: Country X
  - id: '3' 
    department: Department X
    institution: Institution X
    city: City X
    state: State XX # only USA, Australia, Canada
    country: Country X
output: rticles::frontiers_article
csl: frontiers.csl
bibliography: test.bib
---

\begin{abstract}

Abstract length and content varies depending on article type. Refer to 
\url{http://www.frontiersin.org/about/AuthorGuidelines} for abstract requirement
and length according to article type.

%All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\tiny
 \keyFont{ \section{Keywords:} Text Text Text Text Text Text Text Text }

\end{abstract}

# Introduction {-}
Cite fancy references [@Neuro2013]. As demonstrated in @Gene2012, citations can
also be automatically reference. Multiple references are separated by semicolons
[@Neuro2013; @Neurobot2013].

# Results {-}

<!-- Comments should be HTML-style comments -->

## Subsection 1 {-}

You can use `R` chunks directly to plot graphs.

```{r graph, echo = TRUE, message=FALSE, fig.show = "hide"}
x <- 0:100
set.seed(999)
y <- 2 * (x + rnorm(length(x), sd = 3) + 3)
plot(x, y)
```

## Subsection 2 {-}

Frontiers requires figures to be submitted individually, in the same order as
they are referred to in the manuscript. Figures will then be automatically
embedded at the bottom of the submitted manuscript. Kindly ensure that each
table and figure is mentioned in the text and in numerical order. Permission
must be obtained for use of copyrighted material from other sources (including
the web). Please note that it is compulsory to follow figure instructions.
Figures which are not according to the guidelines will cause substantial delay
during the production process.

# Discussion

# Disclosure/Conflict-of-Interest Statement {-}

<!--  
Frontiers follows the recommendations by the International Committee of Medical
Journal Editors (http://www.icmje.org/ethical_4conflicts.html) which require
that all financial, commercial or other relationships that might be perceived by
the academic community as representing a potential conflict of interest must be
disclosed. If no such relationship exists, authors will be asked to declare that
the research was conducted in the absence of any commercial or financial
relationships that could be construed as a potential conflict of interest. When
disclosing the potential conflict of interest, the authors need to address the
following points:

 - Did you or your institution at any time receive payment or services from a
   third party for any aspect of the submitted work?
 - Please declare financial relationships with entities that could be perceived
   to influence, or that give the appearance of potentially influencing, what
   you wrote in the submitted work.
 - Please declare patents and copyrights, whether pending, issued, licensed
   and/or receiving royalties relevant to the work.
 - Please state other relationships or activities that readers could perceive to
   have influenced, or that give the appearance of potentially influencing, what
   you wrote in the submitted work.
 -->

The authors declare that the research was conducted in the absence of any
commercial or financial relationships that could be construed as a potential
conflict of interest.

# Author Contributions {-}

<!--

When determining authorship the following criteria should be observed:

 - Substantial contributions to the conception or design of the work; or the
   acquisition, analysis, or interpretation of data for the work; AND
 - Drafting the work or revising it critically for important intellectual
   content; AND
 - Final approval of the version to be published ; AND
 - Agreement to be accountable for all aspects of the work in ensuring that
   questions related to the accuracy or integrity of any part of the work are
   appropriately investigated and resolved.

Contributors who meet fewer than all 4 of the above criteria for authorship
should not be listed as authors, but they should be acknowledged.
(http://www.icmje.org/roles_a.html)

-->

The statement about the authors and contributors can be up to several sentences
long, describing the tasks of individual authors referred to by their initials
and should be included at the end of the manuscript before the References
section.

# Acknowledgments {-}

Funding:

# Supplemental Data

Supplementary Material should be uploaded separately on submission, if there are
Supplementary Figures, please include the caption in the same file as the
figure. LaTeX Supplementary Material templates can be found in the Frontiers
LaTeX folder

# References

A Frontier article expect the reference list to be included in this section.
To make that happens, the below syntax can be used. This [feature is from Pandoc citeproc](https://pandoc.org/MANUAL.html#placement-of-the-bibliography) which is used with `frontier_article()` to handle the bibliography

::: {#refs}
:::

# Figures {-}

```{r, Figure-1, ref.label = "graph", results = "hide", echo = FALSE, message = FALSE, fig.height=4, fig.width=4, fig.align='center', fig.cap='Figure caption', out.width = "85mm", out.height = "85mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

### 2022-11-09 plosTest/plosTest.Rmd
bioRxiv,poster

---
title: Title of submission to PLOS journal
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliation: Some Institute of Technology
    corresponding: alice@example.com
  - name: Bob Security
    email: bob@example.com
    affiliation: 
      - Another University
      - Some Institute of Technology
address:
  - code: Some Institute of Technology
    address: Department 1, Street, City, State, Zip
  - code: Another University
    address: Department 2, Street, City, State, Zip
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

author_summary: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

bibliography: mybibfile.bib
output: rticles::plos_article
# plos.csl can be download and used locally
csl: http://www.zotero.org/styles/plos
---

_Text based on plos sample manuscript, see <https://journals.plos.org/ploscompbiol/s/latex>_

# Introduction

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

A list

- Item 1
- Item 2

Here are two sample references: @Feynman1963118 [@Dirac1953888].

# References {#references .unnumbered}

### 2022-11-09 springerTest/springerTest.Rmd
bioRxiv,poster

---
title: Title here
subtitle: Do you have a subtitle? If so, write it here
titlerunning: Short form of title (if too long for head)
authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.

authors: 
- name: Äüthör 1
  address: Department of YYY, University of XXX
  email: abc@def

- name: Âuthóř 2
  address: Department of ZZZ, University of WWW
  email: djf@wef

keywords:
- key
- dictionary
- word

#PACS: 
#- PAC1
#- superPAC

MSC:
- MSC code 1
- MSC code 2

abstract: |
  The text of your abstract.  150 -- 250 words.

bibliography: bibliography.bib
biblio-style: spphys
# bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
---

# Introduction {#intro}

Your text comes here. Separate text sections with \cite{Mislevy06Cog}.

# Section title {#sec:1}

Text with citations by \cite{Galyardt14mmm}.

## Subsection title {#sec:2}

as required. Don't forget to give each section
and subsection a unique label (see Sect. \ref{sec:1}).

#### Paragraph headings

Use paragraph headings as needed.

\begin{align}
a^2+b^2=c^2
\end{align}

### 2022-11-09 submit/cover.Rmd
bioRxiv,poster

@@
 title: ""
-output: html_notebook
+output: 
+  # html_notebook
+  pdf_document
 header-includes:
@@
 
-```{r keys}
+```{r keys, echo=F}
 knitr::opts_chunk$set(echo = F)
-title <- "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
+title <- "Habitat edges affect tree diversity more than biomass regeneration in a reforested wet neotropical timber plantation"
 journal <- "Journal of Applied Ecology"
+comparison <- "Edge effects on components of diversity and above-ground biomass in a tropical rainforest, by Razafindratsima et al 2017"
 ```
@@
 as a Research Paper. 
-This manuscript presents and discusses data that
+This manuscript presents and discusses data from a regenerating forested tropical timber plantation that expands typical notions of edge effects, beyond degraded edges and fragmented forests, to highlight how adjacent intact primary forests can actually facilitate the restoration of tree biodiversity and help prioritize ecosystem management efforts.
 
-support the use of roto-till and cover crop mixtures to alleviate soil compaction and weed pressure, which are very common issues for all farmers as agriculture expands into urban environments.
+This manuscript fits directly within the scope of this journal publication, primarily in that it uses an applied setting to address fundamental questions in tropical forest ecology. 
+<!-- We use an applied setting, a regenerated tropical timber plantation, to answer fundamental questions about forest community ecology, namely digging into how local management works as the strongest driver of tropical forest carbon, focusing showing data on forest biomass, tree biodiversity, and functional trait recovery.
+-->
+Specifically, it moves beyond aboveground biomass measurements to also evaluate data on tree community composition, as well as some functional traits like height, wood density, tissue carbon, successional stage, and dispersal mode.
+As a result, this study does a very robust job of evaluating multiple dimensions of forest restoration, including with respect to this new beneficial edge effect hypothesis.
+This study also does a solid job of addressing motivating the globally relevant question of how local management, which is the strongest driver of tropical forest carbon, affects the early stages of recovering forest tree diversity, which can lag many decades behind functional recovery of biomass and taxonomic richness.
+Overall, this study appears to closely resemble a recent publication in your journal, `r comparison` under Associate Editor *Natalie Butt*.
 
-This manuscript both fits within and also expands the scope of this journal publication. 
-The manuscript fits well with the scope because 
-it discusses changes to soil functions caused by various types and intensities of mechanical tillage, and informs balancing trade-offs during agricultural management. 
-In addition analyzing tillage affects, the manuscript also incorporates discussion of cover crops as a comparable management strategy targeting similar soil functions, and more specifically presents the use of cover crop mixtures, rather than traditional single-species cover cropping.
-Finally, this manuscript places these soil management considerations in an urban agriculture context, which is increasingly relevant to the majority of the world’s population living in urban areas and promotes adaptation to more local food supply chain systems.
-
-Overall, this manuscript represents a solid and interesting contribution to the soil tillage and modern agricultural literature. 
-The analysis and discussion is presented as a full manuscript, though it could be possible to contribute this a Short Communication. 
-We are also happy to recommend your associate editor 
-Elke Noellemeyer 
-to be associated with our submission, based on a listed interest in 
-agroecology. 
+The analysis and discussion is presented as a full manuscript, with solid potential to generate discussion with and influence practitioners' perspectives. 
+We are also happy to recommend some of your associate editors to be associated with our submission -- 
+*P.C Abhilash* based on listed interests in restoring degraded habitats,
+*Natalie Butt* based on listed interests in neotropical forest diversity and functions affected by humans,
+*Cristina Garcia* based on listed interests in seed dispersal,
+*David Moreno Mateos* based on listed interests in accelerating ecosystem restoration,
+and
+*Sharif Ahmed Mukul* based on listed interests in forest management.
 As potential reviewers, we point to 
-Kami Pothukuchi and Zhijiang Lu (Wayne State), 
-Viashelav Vashenev (RUDN, Russia), 
-Zhonqi (Joshua) Cheng (USI, Brooklyn Coll), and 
-Rattan Lal (Ohio State). 
+*María Uriarte* (Columbia U),
+*Liza Comita* (Yale U), 
+*Marcel Caritá Vaz* (U California - Los Ángeles), 
+*Jenny Zambrano* (Washington State U),
+*Julia Burton* (Michigan Tech U), and 
+*Caroline Farrior* (U Texas - Austin). 
 Finally, please feel free to contact us for any additional relevant information you may like.

### 2022-11-09 analysis/stats.R

@@ -41,7 +41,7 @@ spTbl <- cleanData %>% filter(!is.na(dist | plot)) %>%
     values_fn = length,
     values_fill = 0
     ) %>%
-  select(!`-`) %>% #sig?
+  # select(!`-`) %>% #sig?
   mutate("entropy" = diversity(.))
 
 
@@ -100,20 +100,24 @@ plotVarsTbl <- cleanDataPlotVars |>
       )
   ) #%>%
 
-  # hist parms
+  #hist,parms--SAVE-LATER
   # mutate(
   #   "sys" = varData %>%
   #     modify(
-  #       ~ .x %>%
+  #       ~ .x %>% group_by(dist) %>%
+  #         pull(kg17) %>% # mainMetric, maybe make variable later
+  #         hist(breaks = c(10000000, 1000000, 100000, 1000, 100, 10, 1), #log10
+  #                   plot = F) %>%
+  #         {.$counts + 1} #%>% as_tibble_col("g")
+  #     ),
+  #   varData = varData %>% modify(~.x %>% mutate(g = g))
+  # ) %>%
   #
-  #         pull(
-  #           median
-  #         ) %>%
-  #         hist(
-  #           .,
-  #           plot = F
-  #         )
-  #     )
+  # rename("g" = sys) #%>%
+  # oir::runFits()
+
+
+
     # "varData0" = cleanData %>%
     #   tidyr::nest("varData0" = everything())#,
     # "sys0" = {
@@ -123,7 +127,6 @@ plotVarsTbl <- cleanDataPlotVars |>
     # } %>% tidyr::nest("sys0" = everything())
     #   # where(colnames(.x$varData0) == .x$variable))
   # )
-
 
 
 # ordinate ----

### 2022-11-09 figs/fig2.R

@@ -234,14 +234,9 @@ graph2c
 
 
 fig2 <- ggpubr::ggarrange(
-  graph2a,
-  graph2b,
-  graph2c,
-  graph2d,
-  graph2e,
-  graph2f,
-  ncol = 3,
-  nrow = 2,
+  graph2a, graph2b, graph2c,
+  graph2d, graph2e, graph2f,
+  ncol = 3, nrow = 2,
   labels = c(
     "a", "b", "c",
     "d", "e", "f"

### 2022-11-09 figs/map.R

@@ -1,6 +1,6 @@
 here::i_am("figs/map.R")
 
-site <- magick::image_read(here::here("figs/mapMod.png")) %>% magick::image_ggplot()
+site <- magick::image_read(here::here("figs/mapMod.jpg")) %>% magick::image_ggplot()
 site0 <- magick::image_read(here::here("figs/locationWOWmod.jpg")) %>% magick::image_ggplot()
 
 Osa <- magick::image_read(here::here("figs/map_Osa.jpg")) %>% magick::image_ggplot()

### 2022-12-06 article/abs.Rmd
paper clean

@@
 
-\newpage
+<!-- \newpage -->
 
-# Resumen{-}
+<!-- # Resumen{-} -->
 
-Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
-Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
-Dado que la mayoría de bosques quedan entre un km de su borde, efectos de borde probablemente afectan la regeneración de composición de comunidad.
-Sin embargo, la mayoría de estudios evalúan cómo bordes ya degradados afectan a bosques intactos, dejándolo poco claro si bordes de más alta calidad podrían facilitar la regeneración de bosques cercanos previamente degradados.
-Notablemente, bordes de alta calidad cerca de bosques intactos podrían promover procesos como disperso y acumulación de biomasa maderosa que efectivamente acceleran a la sucesión, permitiendo mejor rendimiento de taxa tolerantes a sombra comparado con taxa más adaptados a alta luz en las etapas tempranas de regeneración de plantaciones foresterados.
-Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
-Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.
+<!-- Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente. -->
+<!-- Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo. -->
+<!-- Dado que la mayoría de bosques quedan entre un km de su borde, efectos de borde probablemente afectan la regeneración de composición de comunidad. -->
+<!-- Sin embargo, la mayoría de estudios evalúan cómo bordes ya degradados afectan a bosques intactos, dejándolo poco claro si bordes de más alta calidad podrían facilitar la regeneración de bosques cercanos previamente degradados.
+-->
+<!-- Notablemente, bordes de alta calidad cerca de bosques intactos podrían promover procesos como disperso y acumulación de biomasa maderosa que efectivamente acceleran a la sucesión, permitiendo mejor rendimiento de taxa tolerantes a sombra comparado con taxa más adaptados a alta luz en las etapas tempranas de regeneración de plantaciones foresterados.
+-->
+<!-- Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
+-->
+<!-- Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.
+-->
 
-Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10% variación explicado, mientras tanto biomasa y luz de dócel asombrosamente tendió a quedarse igual.
-Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, explicado en parte por densidad de madera, y composición taxonomica varió notablemente.
-Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, tanto como el género Ficus, pero biomasa por modo de disperso no tendió a cambiar.
-Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.
+<!-- Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10% variación explicado, mientras tanto biomasa y luz de dócel asombrosamente tendió a quedarse igual.
+-->
+<!-- Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, explicado en parte por densidad de madera, y composición taxonómica varió notablemente.
+-->
+<!-- Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, tanto como el género Ficus, pero biomasa por modo de disperso no tendió a cambiar.
+-->
+<!-- Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.
+-->

### 2022-12-06 article/merge.Rmd
paper clean

@@
 
-***Correspondence**: Nicholas Medina, nmedina@umich.edu
+***Correspondence**:  
+Nicholas Medina, nmedina@umich.edu, https://orcid.org/0000-0001-5465-3988
 
@@
 
-Code stored at [github.com/nmedina17/osa](github.com/nmedin17/osa) *[@softwareMedina22osa]*.
+Code stored at [github.com/nmedina17/osa](github.com/nmedin17/osa) *[@softwareMedina22osa]* with data stored at [https://doi.org/10.5281/zenodo.7406478](https://doi.org/10.5281/zenodo.7406478) *[@medina22osaDataRaw]*.
 

### 2022-12-06 article/merge.tex
paper clean

@@
 
-*\textbf{Correspondence}: Nicholas Medina, \href{mailto:nmedina@umich.edu}{\nolinkurl{nmedina@umich.edu}}
+*\textbf{Correspondence}:\\
+Nicholas Medina, \href{mailto:nmedina@umich.edu}{\nolinkurl{nmedina@umich.edu}}, \url{https://orcid.org/0000-0001-5465-3988}
 
@@
 Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.
-
-\newpage
-
-\hypertarget{resumen}{%
-\section*{Resumen}\label{resumen}}
-\addcontentsline{toc}{section}{Resumen}
-
-Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
-Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
-Dado que la mayoría de bosques quedan entre un km de su borde, efectos de borde probablemente afectan la regeneración de composición de comunidad.
-Sin embargo, la mayoría de estudios evalúan cómo bordes ya degradados afectan a bosques intactos, dejándolo poco claro si bordes de más alta calidad podrían facilitar la regeneración de bosques cercanos previamente degradados.
-Notablemente, bordes de alta calidad cerca de bosques intactos podrían promover procesos como disperso y acumulación de biomasa maderosa que efectivamente acceleran a la sucesión, permitiendo mejor rendimiento de taxa tolerantes a sombra comparado con taxa más adaptados a alta luz en las etapas tempranas de regeneración de plantaciones foresterados.
-Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
-Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.
-
-Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10\% variación explicado, mientras tanto biomasa y luz de dócel asombrosamente tendió a quedarse igual.
-Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, explicado en parte por densidad de madera, y composición taxonomica varió notablemente.
-Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, tanto como el género Ficus, pero biomasa por modo de disperso no tendió a cambiar.
-Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.
 
@@
 Community analysis and ordination was done as a PERMANOVA with the \emph{adonis()} function from the \emph{vegan} 2.6.4 R package \emph{(\protect\hyperlink{ref-dixon03}{Dixon, 2003})}.
-Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and internal pipeline \emph{oir} \emph{(\protect\hyperlink{ref-softwareMedina22oir}{Medina, 2022a})}, and are stored at \url{github.com/nmedina17/osa} \emph{(\protect\hyperlink{ref-softwareMedina22osa}{Medina, 2022b})}.
+Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and internal pipeline \emph{oir} \emph{(\protect\hyperlink{ref-softwareMedina22oir}{Medina, 2022a})}, and are stored at \url{github.com/nmedina17/osa} \emph{(\protect\hyperlink{ref-softwareMedina22osa}{Medina, 2022b})} and \url{https://doi.org/10.5281/zenodo.7406478} \emph{(\protect\hyperlink{ref-medina22osaDataRaw}{Medina, 2022c})}.
 
@@
 
-Code stored at \href{github.com/nmedin17/osa}{github.com/nmedina17/osa} \emph{(\protect\hyperlink{ref-softwareMedina22osa}{Medina, 2022b})}.
+Code stored at \href{github.com/nmedin17/osa}{github.com/nmedina17/osa} \emph{(\protect\hyperlink{ref-softwareMedina22osa}{Medina, 2022b})} with data stored at \url{https://doi.org/10.5281/zenodo.7406478} \emph{(\protect\hyperlink{ref-medina22osaDataRaw}{Medina, 2022c})}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-medina22osaDataRaw}{}}%
+Medina, N., 2022c. Raw data for: {Habitat} edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest.
+
 \leavevmode\vadjust pre{\hypertarget{ref-softwareMedina22oir}{}}%
-Medina, N., 2022a. {Code for: Automating statistical decision trees for data exploration and analysis revision workflow}. \url{https://doi.org/10.5281/zenodo.6800299}
+Medina, N., 2022a. Code for: {Automating} statistical decision trees for data exploration and analysis revision workflow. \url{https://doi.org/10.5281/zenodo.6800299}
 
 \leavevmode\vadjust pre{\hypertarget{ref-softwareMedina22osa}{}}%
-Medina, N., 2022b. {Code for: Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}. \url{https://doi.org/10.5281/zenodo.6816612}
+Medina, N., 2022b. Code for: {Habitat} edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest. \url{https://doi.org/10.5281/zenodo.6816612}
 

### 2022-12-06 article/methods.Rmd
paper clean

@@
 Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
-Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22oir]*, and are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa) *[@softwareMedina22osa]*.
+Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22oir]*, and are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa) *[@softwareMedina22osa]* and [https://doi.org/10.5281/zenodo.7406478](https://doi.org/10.5281/zenodo.7406478) *[@medina22osaDataRaw]*.
 

### 2023-03-28 article/merge.Rmd
push link
minor

@@
 
-<!-- journals: J [Appl] Ecol (6.8), J [Appl] Veg Sci (3.4), Basic Appl Ecol (3.7), Biotropica (2.5), J Forestry Res (2.2), J Trop Ecol (1.3), Small-scale Forestry (1.3), Eco Evo -->
-
 \
@@
 ***Correspondence**:  
-Nicholas Medina, nmedina@umich.edu, https://orcid.org/0000-0001-5465-3988
+Nicholas Medina, nmedina@umich.edu, [https://orcid.org/0000-0001-5465-3988](https://orcid.org/0000-0001-5465-3988)
 
